---
title: "Initial quality control of single-cell RNA-sequencing"
author: "Anna Aulicino, Kévin Rue-Albrecht"
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Import QC metric collected by _MultiQC_ after read preprocessing.
vignette: >
  %\VignetteIndexEntry{Bioconductor style for HTML documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# Requirements

This vignette requires the following packages to be installed:

```{r pkgs, message=FALSE}
library(xlsx)
library(edgeR)
library(scater)
library(ggplot2)
library(RColorBrewer)
library(ComplexHeatmap)
library(pander)
library(dplyr)
library(ggrepel)
library(scran)
library(rtracklayer)
```

# Pre-processed data

## Overview

Prior to this vignette, paired-end reads (2x75 bp) were:

* Quality controlled using
  [FastQC](http://www.bioinformatics.babraham.ac.uk/projects/fastqc)
  (version `0.11.4`),
* Aligned using [HISAT2](https://ccb.jhu.edu/software/hisat2)
  (version `2.0.3b`),
* Assigned to annotated features using
  [featureCounts](http://bioinf.wehi.edu.au/featureCounts)
  (version `1.5.0-p2`).

**Note:**
Adapter and quality trimming using
[Trim Galore!](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore)
(version `0.4.1`; using _cutadapt_ version `1.10`) was tested,
but the marginal improvement did not justify the extra computational
time and disk space required to process and store the additional data.

## Composite genome (_Ensembl_)

Composite genome Fasta and GTF files were formed by concatenating the
primary assembly of the _Homo sapiens_ genome build _GRCh38_
(excluding haplotypes and patches) and the ERCC RNA Spike-In Mix:

```
cat $genomeFasta $erccFasta > $compositeFasta
cat $genomeGTF $erccGTF > $compositeGTF
```

The Fasta and GTF files for the human genome were obtained from the
_Ensembl FTP site_ [release 86](ftp://ftp.ensembl.org/pub//release-86).
Namely, the downloaded files are:

* `Homo_sapiens.GRCh38.dna.primary_assembly.fa.gz`
* `Homo_sapiens.GRCh38.86.chr.gtf.gz`

The ERCC Fasta and GTF files were obtained from the _ThermoFisher_
[website](https://www.thermofisher.com/order/catalog/product/4456739).

## Genome index (_HISAT2_)

The composite genome was indexed as follows:

```
hisat2-build $compositeFasta $hisatIdx
```

## Read quality control (_FastQC_)

Raw sequenced reads were quality-controlled using
[FastQC](#http://www.bioinformatics.babraham.ac.uk/projects/fastqc/)
as follows:

```
fastqc \
  --outdir $fastqcFolder \
  --nogroup \
  $fastqFile
```

An example _FastQC_ report is embedded below:

<iframe height="500" width="100%"
src="fastqc/WTCHG_305265_201231_2_fastqc.html"></iframe>

## Alignment (_HISAT2_)

Raw (_i.e._, not trimmed.) read pairs were aligned as follows:

```
hisat2 \
    -x ${hisatIdx} \
    -1 ${fastqFolder}/${cell}_1.fastq.gz \
    -2 ${fastqFolder}/${cell}_2.fastq.gz \
    -S $hisatDir/${cell}.sam \
    1> $logDir/${cell}.o 2> $logDir/${cell}.e
```

The resulting SAM files were compressed to BAM format as follows:

```
samtools view -hb -o $hisatDir/${cell}.bam $hisatDir/${cell}.sam
```

Finally, SAM files were then removed as follows:

```
rm -f $hisatDir/${cell}.sam
```

## Read assignment (_featureCounts_) {#featureCounts}

Uniquely aligned read pairs were assigned to annotated genomic feaures and
ERCC RNA spike-in features, and counted in an unstranded manner as follows:

```
featureCounts \
    --primary \
    --ignoreDup \
    -p \
    -a $compositeGTF \
    -o $countDir/$cell \
    $bamDir/${cell}.bam \
    1> $logDir/$cell.out 2> $logDir/$cell.err
```

## _MultiQC_

Reports of the various preprocessing steps above were collated using [MultiQC](http://multiqc.info). The overall report in embedded below:

<iframe height="500" width="100%" src="MultiQC/final_pipeline.html"></iframe>

# Import data

## Phenotype information

Let us import experimental information and reformat some of the experimental
factors:

```{r importPheno}
pheno <- read.xlsx("expdata/samples.xlsx", sheetName="pheno", row.names=1)
pheno$sample <- as.character(pheno$sample)
pheno$Lane <- as.factor(pheno$Lane)
pheno$Infection <- factor(pheno$Infection, c("Mock","D23580","LT2","Blank"))
pheno$Status <- factor(
  pheno$Status, c("uninfected","exposed","infected","Blank","BULK")
)
```

## Counts

To import count data that is stored in a separate file for each sample,
let us use the `r Biocpkg("edgeR")` `readDGE` method:

```{r readDGE}
RG <- readDGE(
  rownames(pheno),
  "counts",
  c(1, 3),
  paste(pheno$Time, pheno$Infection, pheno$Status, sep = "_"),
  pheno$sample
)
```

## `SCESet`

Let us assemble the phenotype information and count data above into
`r Biocpkg("scater")` `SCESet` objects:

* one that contains all samples together
* one for the blank samples
* one for the bulk samples
* one for the single cells

**Note:**
This distinction will later be important to compute QC metrics within each
data set.

First, let us match the sample identifier associated with the phenotype
information to that associated with the count data:

```{r renamePheno}
rownames(pheno) <- rownames(RG$samples)
```

We may then create the four data sets:

```{r sce.all}
pd.all <- new("AnnotatedDataFrame", data = pheno)
sce.all <- newSCESet(countData = RG$counts, phenoData = pd.all)
dim(sce.all)
```

```{r sce.blank}
idx.blank <- pheno$Status == "Blank"
pd.blank <- new("AnnotatedDataFrame", data = droplevels(pheno[idx.blank,]))
sce.blank <- newSCESet(countData = RG$counts[,idx.blank], phenoData = pd.blank)
dim(sce.blank)
```
```{r sce.bulk}
idx.bulk<- pheno$Status == "BULK"
pd.bulk <- new("AnnotatedDataFrame", data = droplevels(pheno[idx.bulk,]))
sce.bulk <- newSCESet(countData = RG$counts[,idx.bulk], phenoData = pd.bulk)
dim(sce.bulk)
```

```{r sce.sc}
idx.sc <- pheno$Status %in% c("exposed","infected","uninfected")
pd.sc <- new("AnnotatedDataFrame", data = droplevels(pheno[idx.sc,]))
sce.sc <- newSCESet(countData = RG$counts[,idx.sc], phenoData = pd.sc)
dim(sce.sc)
```

We may then free some memory by removing unused objects:

```{r cleanup_SCESet}
rm(pd.all, pd.blank, pd.bulk, pd.sc, pheno, RG)
```

**Note:**
In particular, `SCESet` objects have a much smaller memory footprint than
the equivalent data stored in a `DGEList`.

# QC metrics
## _MultiQC_
### Import QC metrics

Importantly, _MultiQC_ collated metrics for each sample at each step of
preprocessing---from raw reads to assignmed counts---; those QC metrics
already provide an idea of technical issues in certain samples.

Let us import the QC metrics produced by _MultiQC_:

```{r importMultiQC}
multiqc <- read.delim("MultiQC/final_pipeline_data/multiqc_general_stats.txt")
multiqc$Sample <- gsub("WTCHG_[[:digit:]]+_", "", multiqc$Sample)
```

Let us annotated those metrics with phenotype information:

```{r addPhenoMultiQC}
multiqc <- merge(
  pData(sce.all), multiqc, by.x = "sample", by.y = "Sample",
  sort = FALSE
)
```

Let us confirm that the samples are ordered identically to the `sce.all`
object:

```{r checkOrder}
all(multiqc$sample == rownames(pData(sce.all)))
```

### Assigned read pairs / Library size (_featureCounts_) {#featureCounts}

Excessively small library size may indicate technial issues during
library preparation or low quality sample.
Moreover, library size excessively different from the
majority of samples may introduce bias due to under- or over-sampling of
RNA fragments.

Let us identify as outliers single cells
for which the number of assigned read pairs (_i.e._ library size)
is further than `3` MADs (median absolute deviations) from the median
of _single cell_ samples:

```{r multiQC_outliers_assigned}
metric <- log10(multiqc$featureCounts_Assigned[idx.sc])
multiqc.assigned.M <- median(metric)
multiqc.assigned.MAD <- mad(metric)
multiqc.assigned.outliers <-
  abs(metric - multiqc.assigned.M) > 3 * multiqc.assigned.MAD
table(multiqc.assigned.outliers)
```

In particular, the lower cut-off value is:

```{r assigned_lowerCutOff}
10^(multiqc.assigned.M - 3 * multiqc.assigned.MAD)
```

Let us visualise those outliers, along with blank and bulk samples
for reference:

```{r multiQC_ggplot_assigned}
ggplot(multiqc, aes(sample, featureCounts_Assigned)) +
  geom_point(aes(colour = Status, shape = Infection)) +
  ggrepel::geom_text_repel(
    aes(label = sample), subset(multiqc[idx.sc,], multiqc.assigned.outliers),
    alpha = 0.4
  ) +
  geom_hline(
    yintercept = 10^(multiqc.assigned.M + 3*c(1,-1)*multiqc.assigned.MAD),
    colour = "red"
  ) +
  geom_hline(yintercept = 10^(multiqc.assigned.M)) +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  )
```

### GC content (_FastQC_)

GC content is species-specific; it can be estimated from the annotated transcriptome or the sequenced data. Therefore, GC content excessively
different from the majority of samples can indicate issues during
library preparation or contamination.

Let us identify as outliers samples for which GC content is
further than 3 MADs (median absolute deviations) from the median
of _single cell_ samples:

```{r multiQC_outliers_GC}
metric <- multiqc$FastQC_percent_gc[idx.sc]
multiqc.GC.M <- median(metric)
multiqc.GC.MAD <- mad(metric)
multiqc.GC.outliers <-
  abs(metric - multiqc.GC.M) > 3 * multiqc.GC.MAD
table(multiqc.GC.outliers)
```

Let us visualise the outlier, along with blank and bulk samples
for reference:

```{r multiQC_ggplot_GC}
ggplot(multiqc, aes(sample, FastQC_percent_gc)) +
  geom_point(aes(colour = Status, shape = Infection)) +
  ggrepel::geom_text_repel(
    aes(label = sample), subset(multiqc[idx.sc,], multiqc.GC.outliers),
    alpha = 0.4
  ) +
  geom_hline(
    yintercept = multiqc.GC.M + 3*c(1,-1)*multiqc.GC.MAD,
    colour = "red"
  ) +
  geom_hline(yintercept = multiqc.GC.M) +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  )
```

### Alignment rate (_HISAT2_ / _bowtie2_)

Technical issues introduced during library preparation may also result
in excessively _low_ alignment rate.
Nevertheless, libraries with slightly lower alignment rates may still yield
a library size comparable to the majority of samples.
For that reason, let us apply a slightly more relaxed cut-off of `5` MADs
to identify as outliers _single cells_ with excessively _low_ alignment rate:

```{r multiQC_outliers_align}
metric <- multiqc$Bowtie.2_overall_alignment_rate[idx.sc]
multiqc.aligned.M <- median(metric)
multiqc.aligned.MAD <- mad(metric)
multiqc.aligned.outliers <-
  metric < multiqc.aligned.M - 5 * multiqc.aligned.MAD
table(multiqc.aligned.outliers)
```

Let us visualise the outlier, along with blank and bulk samples
for reference:

```{r multiQC_ggplot_align}
ggplot(multiqc, aes(sample, Bowtie.2_overall_alignment_rate)) +
  geom_point(aes(colour = Status, shape = Infection)) +
  ggrepel::geom_text_repel(
    aes(label = sample), subset(multiqc[idx.sc,], multiqc.aligned.outliers),
    alpha = 0.4
  ) +
  geom_hline(
    yintercept = multiqc.aligned.M + 5*c(1,-1)*multiqc.aligned.MAD,
    colour = "red",
    linetype = 2:1
  ) +
  geom_hline(yintercept = multiqc.aligned.M) +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  ) +
  ylab("HISAT.2_overall_alignment_rate")
```

### Percentage assigned (_HISAT2_ / _bowtie2_)

After alignment, unusually _low_ assignment rate of reads to annotated genomic
features may be indicative of technical issue; for instance during
poly-A selection.

Let us identify as outliers _single cells_ with excessively _low_
assignment rate:

```{r multiQC_outliers_assign}
metric <- multiqc$featureCounts_percent_assigned[idx.sc]
multiqc.assignedPerc.M <- median(metric)
multiqc.assignedPerc.MAD <- mad(metric)
multiqc.assignedPerc.outliers <-
  metric < multiqc.assignedPerc.M - 3 * multiqc.assignedPerc.MAD
table(multiqc.assigned.outliers)
```

Let us visualise the outlier, along with blank and bulk samples
for reference:

```{r multiQC_ggplot_assign}
ggplot(multiqc, aes(sample, featureCounts_percent_assigned)) +
  geom_point(aes(colour = Status, shape = Infection)) +
  ggrepel::geom_text_repel(
    aes(label = sample), subset(multiqc[idx.sc,], multiqc.assignedPerc.outliers),
    alpha = 0.4
  ) +
  geom_hline(
    yintercept = multiqc.assignedPerc.M + 3*c(1,-1)*multiqc.assignedPerc.MAD,
    colour = "red",
    linetype = 2:1
  ) +
  geom_hline(yintercept = multiqc.assignedPerc.M) +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  )
```

### _MultiQC_ summary

```{r multiQC_summary}
outlier.vars <- ls(pattern = "outliers")
colSums(sapply(outlier.vars, function(x){get(x)}))
table(rowSums(sapply(outlier.vars, function(x){get(x)})))
```

## Blanks
### Highest expressed features

Blank samples are designed to detected technical issues during library
preparation; in particular contamination, as only ERCC spike-in features are
expected in those samples.

First, let us obtain the identifier of ERCC spike-in features:

```{r ERCC_ids}
ERCCs <- grep("^ERCC-[[:digit:]]+$", featureNames(sce.blank), value = TRUE)
```

The `r Biocpkg("scater")` `calculateQCMetrics` may then be used to calculate
the relevant QC metrics:

```{r blank_QC}
sce.blank <- calculateQCMetrics(sce.blank, list(ERCC = ERCCs))
```

In the `SCESet` object augmnented by QC metrics, let us visualise the
most abundant features which displays the expected over-representation of
ERCC spike-in features:

```{r blank_highExprs}
plotQC(sce.blank, type = "highest-expression")
```

## _scater_
### Calculation of QC metrics

The following sections require the estimation of several QC metrics that
leverage control features and control samples, as means to identify
outliers among single cells.
For this purpose, let us first import the `gene` records from the
same GTF file that was supplied to _featureCounts_, to ensure coherent
annotations are used throughout this analysis:

```{r}
gtfData <- import.gff("expdata/genes_with_ERCC.gtf", feature.type = "gene")
```

Let us extract from those annotations the identifier of mitochondrial gene
features; those will be used in a [later section](#MTcontent):

```{r MT_ids}
MTs <- mcols(gtfData)[seqnames(gtfData) == "MT", "gene_id"]
```

Let us apply the `r Biocpkg("scater")` `calculateQCMetrics` to calculate
the relevant QC metrics in the _single cell_ data set:

```{r sc_QC}
sce.sc <- calculateQCMetrics(
  sce.sc,
  feature_controls = list(ERCC = ERCCs, MT = MTs)
)
```

For reference purpose only, let us also calculate
the relevant QC metrics in the complete data set:

```{r all_QC}
sce.all <- calculateQCMetrics(
  sce.all,
  feature_controls = list(ERCC = ERCCs, MT = MTs),
  cell_controls = list(
    Bulk = which(sce.all$Status == "BULK"),
    Blank = which(sce.all$Status == "Blank")
  )
)
```

### Library size

The _library size_ refers to the total count of reads assigned to 
annotated features in each individua sample.
As an internal validation, let us verify that identifying outliers
using the counts imported into the `SCESet` produce the same results as
the [earlier](#featureCounts) use of _MultiQC_ reports of read pairs
assignment by _featureCounts_.

First let us identify as outliers single cells
for which the library size
is further than `3` MADs (median absolute deviations) from the median
of _single cell_ samples:

```{r scater_outliers_totalCounts}
metric <- sce.sc$log10_total_counts
scater.totalCounts.M <- median(metric)
scater.totalCounts.MAD <- mad(metric)
scater.totalCounts.outliers <-
  abs(metric - scater.totalCounts.M) > 3 * scater.totalCounts.MAD
table(scater.totalCounts.outliers)
```

In particular, the lower cut-off value is:

```{r totalCounts_lowerCutOff}
10^(scater.totalCounts.M - 3 * scater.totalCounts.MAD)
```

Let us visualise those outliers, along with blank and bulk samples
for reference:

```{r scater_ggplot_totalCounts}
ggplot(pData(sce.all), aes(sample, log10_total_counts)) +
  geom_point(aes(colour = Status, shape = Infection)) +
  geom_text_repel(
    aes(label = sample), subset(pData(sce.sc), scater.totalCounts.outliers),
    alpha = 0.4
  ) +
  geom_hline(
    yintercept = scater.totalCounts.M + 3*c(1,-1)*scater.totalCounts.MAD,
    colour = "red"
  ) +
  geom_hline(yintercept = scater.totalCounts.M) +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  )
```

### Library complexity: bulks, single cells, blanks {#libraryComplexity}

_Library complexity_ refers to the proportion of assigned reads accounted for
by the _N_ most abundant features within each individual sample, with values
for _N_ usually ranging from tens to hundreds.
It is expected to observe a lower library complexity in single cells relative
to bulk samples, due to higher dropout events in single cells, and the union of
multiple phenotypes within bulk samples.
In contrast, blank samples are expected to have extremely low library
complexity, as only `92` ERCC spike-ins sequences are expected in those
libraries, in contrast to thousands of genomic features in biological samples.

In the `SCESet` object augmnented by QC metrics, let us visualise the
library complexity profile against experimental phenotype information:

```{r all_libraryComplexity}
plot(
  sce.all, block1 = "Status", block2 = "Infection",
  colour_by = "Time", nfeatures = 300, exprs_values = "counts"
)
```

First, from a QC perspective:

* Blanks samples show a very low library complexity, as expected.
* Bulk samples show a generally high library complexity, as expected.
* Single cells generally show a complexity profile comparable to that of bulk
  samples.
* A small subset of single cells (_e.g._, `exposed:D23580:6h`,`exposed:LT2:4h`, 
  `infected:D23580:4h`, `infected:D23580:6h`, `infected:LT2:6h`) show
  suspiciously low library complexity.
  Those single cells must therefore be treated as outliers, and be removed from
  the data set prior to feature selection, normalisation, and subsequent
  analyses.

Now, from a biological perspective, an interesting observation may be made:

* Library complexity seems to decrease with `Time` in `exposed` and `infected`
  samples in cells exposed to or infected by bacteria, in contrast to
  control uninfected cells.
  This _may_ indicate a polarisation of single cells toward
  specialised phenotype(s) after infection or exposure.
  + `Mock` (_i.e._, `uninfected`) single cells tend to show a consistent,
    relatively high library complexity
  + Both `LT2` and `D23580` infections show a gradual decrease of library
    complexity over time. This observation _might_ be more pronounced
    for `D23580`.

### Proportion of library assigned to the most abundant features

Although abundantly expressed features are an expected aspect of cell biology
(_e.g._, myosin in muscle cells, clusters of differentiation in immune cells)
which may contribute to the identification and classification of cell types, overly abundant feature in RNA-Seq libraries
may indicate issues during library preparation; for instance, cDNA
amplification bias.
Moreover, when the most abundant features account for an excessive proportion
of the library, expression levels for the majority of genes may not be
estimated with sufficient accuracy.

Let us identify as outliers single cells
for which the `50`, `100`, and `200` most abundant features account for an
excessively _high_ proportion of the library size:

```{r scater_outliers_top50}
metric <- sce.sc$pct_counts_top_50_features
scater.top50.M <- median(metric)
scater.top50.MAD <- mad(metric)
scater.top50.outliers <-
  metric > scater.top50.M + 3 * scater.top50.MAD
table(scater.top50.outliers)
```

```{r scater_outliers_top100}
metric <- sce.sc$pct_counts_top_100_features
scater.top100.M <- median(metric)
scater.top100.MAD <- mad(metric)
scater.top100.outliers <-
  metric > scater.top100.M + 3 * scater.top100.MAD
table(scater.top100.outliers)
```

```{r scater_outliers_top200}
metric <- sce.sc$pct_counts_top_200_features
scater.top200.M <- median(metric)
scater.top200.MAD <- mad(metric)
scater.top200.outliers <-
  metric > scater.top200.M + 3 * scater.top200.MAD
table(scater.top200.outliers)
```

Let us visualise those outliers, along with blank and bulk samples
for reference:

```{r scater_ggplot_top50}
ggplot(pData(sce.all), aes(sample, pct_counts_top_50_features)) +
  geom_point(aes(colour = Status, shape = Infection)) +
  geom_text_repel(
    aes(label = sample), subset(pData(sce.sc), scater.top50.outliers),
    alpha = 0.4
  ) +
  geom_hline(
    yintercept = scater.top50.M + 3 * c(-1,1) * scater.top50.MAD,
    colour = "red", linetype = 2:1
  ) +
  geom_hline(yintercept = scater.top50.M) +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  )
```

```{r scater_ggplot_top100}
ggplot(pData(sce.all), aes(sample, pct_counts_top_100_features)) +
  geom_point(aes(colour = Status, shape = Infection)) +
  geom_text_repel(
    aes(label = sample), subset(pData(sce.sc), scater.top100.outliers),
    alpha = 0.4
  ) +
  geom_hline(
    yintercept = scater.top100.M + 3 * c(-1,1) * scater.top100.MAD,
    colour = "red", linetype = 2:1
  ) +
  geom_hline(yintercept = scater.top100.M) +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  )
```

```{r scater_ggplot_top200}
ggplot(pData(sce.all), aes(sample, pct_counts_top_200_features)) +
  geom_point(aes(colour = Status, shape = Infection)) +
  geom_text_repel(
    aes(label = sample), subset(pData(sce.sc), scater.top200.outliers),
    alpha = 0.4
  ) +
  geom_hline(
    yintercept = scater.top200.M + 3 * c(-1,1) * scater.top200.MAD,
    colour = "red", linetype = 2:1
  ) +
  geom_hline(yintercept = scater.top200.M) +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  )
```

In this last figure (_i.e._, `200` most abundant features),
note how two `infected:D23580` samples appear near the cut-off threshold.
This observation can be related to the overall library complexity figure
shown [earlier](#libraryComplexity).

### Count of detected features

A notion closely related to library complexity is the count of features
detected within each individual sample. Although the definition of "detection
level" may be debated, we will accept here any feature with non-zero count.

Abnormal counts of detected features may be indicative of technical issue
during RNA capture and library preparation; for instance,
excessively _low_ counts of detected features may indicate RNA degradation,
while excessively _high_ counts of detected features may indicate capture of
multiple cells.

Due to the considerable amount of variation that results from the relaxed
definition of detection level above, let us apply a somewhat stringent cut-off
of `2` MADs to identify as outliers _single cells_ with excessively _low_
counts of detected features:

```{r scater_outliers_features}
metric <- sce.sc$total_features
scater.features.M <- median(metric)
scater.features.MAD <- mad(metric)
scater.features.outliers <-
  metric < scater.features.M - 2 * scater.features.MAD
table(scater.features.outliers)
```

Let us visualise those outliers, along with blank and bulk samples
for reference:

```{r scater_ggplot_features}
ggplot(pData(sce.all), aes(sample, total_features)) +
  geom_point(aes(colour = Status, shape = Infection)) +
  geom_text_repel(
    aes(label = sample), subset(pData(sce.sc), scater.features.outliers),
    alpha = 0.4
  ) +
  geom_hline(
    yintercept = scater.features.M + 2 * c(-1,1) * scater.features.MAD,
    colour = "red", linetype = 1:2
  ) +
  geom_hline(yintercept = scater.features.M) +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  )
```

### ERCC content {#ERCCoutliers}

An identical quantity of ERCC spike-in features is added to each sample
prior to library preparation. As a result, technical variation and issues
during library preparation may be estimated from the proportion of reads
assigned to those spike-in features; an excessive proportion of reads
assigned to ERCC spike-in features may indicate eukaryotic RNA degradation,
while excessive variation may indicate cDNA amplication bias.

Due to the limited amount of variation that results from the very low levels
of ERCC spike-in features generally detected,
let us apply a somewhat relaxed cut-off of `5` MADs
to identify as outliers _single cells_ with excessively _high_ ERCC content:

```{r scater_outliers_ercc}
metric <- sce.sc$pct_counts_feature_controls_ERCC
scater.ERCC.M <- median(metric)
scater.ERCC.MAD <- mad(metric)
scater.ERCC.outliers <- metric > scater.ERCC.M + 5 * scater.ERCC.MAD
table(scater.ERCC.outliers)
```

Let us visualise those outliers, along with blank and bulk samples
for reference:

```{r scater_ggplot_ERCC}
ggplot(pData(sce.all), aes(sample, pct_counts_feature_controls_ERCC)) +
  geom_point(aes(colour = Status, shape = Infection)) +
  geom_text_repel(
    aes(label = sample), subset(pData(sce.sc), scater.ERCC.outliers),
    alpha = 0.6
  ) +
  geom_hline(
    yintercept = scater.ERCC.M + 5 * c(-1,1) * scater.ERCC.MAD,
    colour = "red", linetype = 2:1
  ) +
  geom_hline(yintercept = scater.ERCC.M) +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  )
```

### Mitochondrial content {#MTcontent}

Similarly to the ERCC spike-in features, mitochondrial RNA may be used
as an endogenous source of control features.
Albeit mitochondrial gene expression levels may be altered by physiological
processes, the limited count of genes encoded on the mitochondrial chromosome
implies a similarly limited proportion of reads assigned to those gene
features.
Consequently, abnormal proportions of reads assigned to mitochondrial gene
features may be indicative of technical issue during RNA capture and
library preparation, or abnormal count of mitochondria within the single cells.

Let us identify as outliers single cells for which the proportion of reads
assigned to mitochondrial gene features is is further than `3` MADs
from the median of _single cell_ samples:

```{r scater_outliers_MT}
metric <- sce.sc$pct_counts_feature_controls_MT
scater.MT.M <- median(metric)
scater.MT.MAD <- mad(metric)
scater.MT.outliers <-
  metric > scater.MT.M + 5 * scater.MT.MAD
table(scater.MT.outliers)
```

Let us visualise those outliers, along with blank and bulk samples
for reference:

```{r scater_ggplot_MT}
ggplot(pData(sce.all), aes(sample, pct_counts_feature_controls_MT)) +
  geom_point(aes(colour = Status, shape = Infection)) +
  geom_text_repel(
    aes(label = sample), subset(pData(sce.sc), scater.MT.outliers),
    alpha = 0.4
  ) +
  geom_hline(
    yintercept = scater.MT.M + 5 * c(-1,1) * scater.MT.MAD,
    colour = "red", linetype = 2:1
  ) +
  geom_hline(yintercept = scater.MT.M) +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  )
```

## Global summary of sample-level QC metrics
### Collection of QC filters

First, let us collect the QC filters described in the previous sections as
into a single summary `data.frame`:

```{r QC_collectOutliers}
outlier.vars <- ls(pattern = "\\.outliers$")
outliers.df <- data.frame(lapply(
  outlier.vars,
  function(x){ifelse(get(x), "Fail", "Pass")}
))
colnames(outliers.df) <- gsub("\\.outliers$", "", outlier.vars)
rownames(outliers.df) <- sampleNames(sce.sc)
```

Let us also define an order of sample that groups single cells by
experimental phenotypes:

```{r sampleSupervisedOrder}
sample.order.supervised <- order(sce.sc$Time, sce.sc$Status, sce.sc$Infection)
```

### Counts of outliers across filters

First, let us examine the count of _single cells_ that fail each of the QC
metrics described in the previous sections:

```{r QC_countByMetric}
colSums(outliers.df == "Fail")
```

Secondly, we may also examine the count of _single cells_, according to the
number of QC filters that they have failed:

```{r QC_countByFailed}
table(rowSums(outliers.df == "Fail"))
```

Let us define the single cells that did not fail any of the QC filter:

```{r QC_finalOutliers}
idx.pass <- rowSums(outliers.df == "Fail") == 0
table(idx.pass)
```

### Heat map summary view

Let us visually summarise _single cells_ that pass each filter as a heat map.

First, let us define colours used to indicate phenotype information:

```{r palette9}
col9 <- brewer.pal(10, "Set3")[c(1:8, 10)]
```

We may then display the QC metrics failed by each single cell as red cells
in an heat map:

```{r QC_heatmap}
hm_row <- rowAnnotation(
  df = pData(sce.sc)[sample.order.supervised,c("Time", "Status", "Infection")],
  col = list(
    Time = c("2h" = col9[1], "4h" = col9[2], "6h" = col9[3]),
    Status=c("uninfected"=col9[4],"exposed"=col9[5],"infected"=col9[6]),
    Infection = c("Mock" = col9[7], "LT2" = col9[8], "D23580" = col9[9])
  )
)
hm <- Heatmap(
  as.matrix(outliers.df)[sample.order.supervised,],
  name = "Filter",
  col = c("Pass" = "lightblue", "Fail" = "red"),
  cluster_rows = FALSE, show_column_dend = FALSE, show_row_names = FALSE
)
draw(hm + hm_row)
```

### Phenotype information of single cell outliers

We may summarise experimental phenotype information for the outliers samples:

```{r QC_summary}
varLabels(sce.sc)[1:10]
summary(pData(sce.sc)[
  !idx.pass,c("Infection","Status","Time","Lane","Plate","Well")
])
```

Or alternatively display the full information in a table:

```{r QC_tableOutliers, results='asis'}
pander::pandoc.table(
  arrange(
    cbind(
      pData(sce.sc)[
        !idx.pass,c("Infection","Status","Time","Lane","Plate","Well")
      ],
      Filters = rowSums(outliers.df == "Fail")[!idx.pass]
    ),
    desc(Filters)
  ),
  "Experimental phenotype information for single cell outliers."
)
```

# Single cells selection
## Subsetting

Let us select the single cells that passed all QC metrics into a new `SCESet`:

```{r subset_goodCells}
sce.pass <- sce.sc[,idx.pass]
```

```{r save_sce.pass, include=FALSE}
saveRDS(sce.pass, file.path("rds", "sce.pass.rds"))
```

## Experimental phenotype information

Let us examine the count of cells remaining in each experimental group:

```{r goodCells_pheno}
table(pData(sce.pass)[,c("Infection","Status","Time")])
```

# Feature selection
## Overview of feature detection levels

Having selected single cells that pass all quality control criteria in the
previous sections, let us identify features that are robustly detected across
single cells in the data set.
Indeed, features detected in a small count of cells, or expressed at very
low levels carry limited information for differential expression analyses;
moreover, those features negatively affect the performance of normalisation
algorithms, due to the large number of dropout events (zero count).

**cite paper 67 from McCarthy et al; Brennecke et al., 2013**

For this purpose, let us first calculate updated QC metrics in the subsetted
data set using once more the `r Biocpkg("scater")` `calculateQCMetrics` method:

```{r pass_calculateQCMetrics}
sce.pass <- calculateQCMetrics(
  sce.pass, feature_controls = list(ERCC = ERCCs, MT = MTs)
)
```

We may then examine for each feature the proportion of cells in which it is
detected (non-zero expression level) against its average expression level
across all cells:

```{r plotExprsFreqVsMean}
plotExprsFreqVsMean(
  sce.pass,
  feature_controls = fData(sce.pass)$is_feature_control_ERCC
)
```

## Definition of detected feature

Let us set define low-abundance features as those with an average count below a
filter threshold of `1`:

```{r ave.counts}
ave.counts <- rowMeans(counts(sce.pass))
keep <- ave.counts >= 1
sum(keep)
```

To check whether the chosen threshold is suitable, let us examine the
distribution of log-means across all genes:

```{r detectVsMean_hist}
hist(
  log10(ave.counts), breaks = 100,
  main = "", xlab = expression(Log[10]~"average count", col="grey80")
)
abline(v = log10(1), col = "blue", lwd = 2, lty = 2)
```

We may compare the above figure with the a smoother representation of the
count of cells in which each feature is detected against the average
log~10~-transformed count of reads assigned to that same feature, while
marking the same cut-off threshold of an average `1` count across all cells:

```{r smoothScatter}
numcells <- nexprs(sce.pass, byrow = TRUE)
is.ercc <- fData(sce.pass)$is_feature_control_ERCC
smoothScatter(
  log10(ave.counts), numcells, xlab=expression(Log[10]~"average count"),
  ylab="Number of expressing cells"
)
points(
  log10(ave.counts[is.ercc]), numcells[is.ercc],
  col = "red", pch = 16, cex = 0.5
)
abline(v = log10(1), col = "blue", lwd = 2, lty = 2)
```

## Selection of detected features

Let us now select the features defined as detected in the above sections
into a new `SCESet`:

```{r subset_features}
sce.filtered <- sce.pass[keep,]
dim(sce.filtered)
```

# Normalisation
## Sum factors

Having identified the cells of good quality and the features robustly
detected across those cells, we may now normalise expression levels between
cells using an approach suitable for single cell data sets.

Considering the multifactorial experimental design of this data set, and the
expected differential expression in `exposed` and `infected` cells,
let us cluster similar cells together prior to normalising the cells in each
cluster using the deconvolution method implemented in the `r Biocpkg("scran")`
package:

```{r quickCluster}
clusters <- quickCluster(sce.filtered)
```

In this case, none of the cells were assigned to any cluster, which suggests
that despite the expected differences between experimental groups, the data set
is sufficiently homogenous for the normalisation procedure to be applied to
all cells equally:

```{r computeSumFactors}
sce.filtered <- computeSumFactors(sce.filtered, clusters = clusters)
summary(sizeFactors(sce.filtered))
```

In this case, the size factors are somewhat correlated with library size,
while significant scatter is visible around the trend;
augmenting the figure with experimental phenotype information emphasises
how the effect of library size on the size factor of individual cells is
subject to factors related to the experimental design of the experiment.

In particular, the figure below emphasises how the effect of library size is
more important in `uninfected` cells, and at the earliest time point
(*i.e.*, `2h`), when the biological effect of infection may not have fully
affect the transcriptome of most cells yet.

In contrast, at later time points (*i.e.*, `4h` and `6h`), trends indicate a
markedly weaker effect of library size on size factor.
This suggest that other sources of systematic differences between cells
arose, and that differential expression is to be expected
between experimental groups at those time points:

```{r sizeFactor_librarySize}
ggplot(
  pData(sce.filtered), aes(total_counts / 1E6, size_factor)) +
  geom_point(aes(colour = Time, shape = Status)) +
  labs(x = "Library size (millions)", y = "Size factor") +
  stat_smooth(
    aes(colour = Time, linetype = Status),
    method = "lm", se = TRUE, size = 0.5, alpha = 0.1
  )
```

Let us examine further the effect of experimental covariates on the size
factors:

```{r sizeFactor_expFactors}
ggplot(
  pData(sce.filtered),
  aes(interaction(Infection, Status, Time, sep = " - "), size_factor)) +
  geom_boxplot(aes(alpha = Time, fill = Infection, linetype = Status)) +
  coord_flip() +
  labs(y = "Size factor", x = "")
```

In a different perspective, the above figure emphasises again how
size factors are affected by experimental factors in this data set;
in particular at the later time points, size factors for cells infected by
or exposed to bacteria display markedly distinct size factors,
irrespective of library size.
This suggests that `Infection`, and `Status` drive more important
systematic differences between cells than library size at those time points,
having likely reshaped the transcriptional profile of stimulated cells.

## Spike factors

To ensure normalization is performed correctly, let us compute a separate set
of size factors for the spike-in set. This assumes
that none of the spike-in transcripts are differentially expressed.

```{r computeSpikeFactors}
setSpike(sce.filtered) <- "ERCC"
sce.filtered <- computeSpikeFactors(
  sce.filtered, type = "ERCC", general.use = FALSE
)
```

We may then examine the spike factors calculated:

```{r spikeFactors_ggplot}
ggplot(pData(sce.filtered), aes(total_counts / 1E6, size_factor_ERCC)) +
  geom_point(aes(colour = Time, shape = Status)) +
  stat_smooth(method = "lm") +
  labs(x = "Library size", y = "Spike factor")
```

In contrast to size factors calculated using endogenous features, spike factors
calculated using ERCC spike-in display a much weaker correlation with
library size, and no trend with experimental factors.
This reinforces the absence of extreme outliers after the removal
of cells with abnormal ERCC content in an [earlier section](#ERCCoutliers);
moreover, it suggests that counts assigned to individual ERCC spike-in features
are similar across all the cells retained after quality control.

## Apply size factors to normalise

The count data are used to compute normalized log-expression values for use in
downstream analyses.
Each value is defined as the log-ratio of each count to the size factor for the
corresponding cell, after adding a prior count of `1` to avoid undefined values
at zero counts.
Division by the size factor ensures that any cell-specific biases are removed.
Spike-in-specific size factors present are applied to normalize the spike-in
transcripts separately from the endogenous features:

```{r normalise}
sce.filtered <- normalize(sce.filtered)
```

# Relative importance of experimental factors

Having normalised the expression data for detected features in cells that
passed quality control, the relative importance of experimental factors---both
technical and biological---provides insight into uninteresting technical
factors that may need to be accounted for when explaining variance in the
expression data.

Let us examine the proportion of variance explained by known experimental
factors, as well as the total spike-in count in each cell:

```{r plotExplanatoryVariables}
plotExplanatoryVariables(
  sce.filtered,
  exprs_values = "norm_exprs",
  variables = c(
    "counts_feature_controls_ERCC",
    "log10_counts_feature_controls_ERCC",
    "Plate", "Lane", "Time", "Status", "Infection"
    )
)
```

The above figure reinforces the effect of `Time` as the primary source of
variance in this expression data set, with substantial proportion of variance
explained for a subset of features. Next are `Status` and `Infection`
factors, while other technical factors share explanation of lower amount
of the residual variance in the data set.

# Clustering

Having normalised the expression data for detected features
in single cells that passed quality control, let us examine the unsupervised
clustering of those cells using different approaches.

## Principal component analysis (PCA)

Let us initially focus our attention on the two principal components that
explain the largest amount of variance in the endogenous features.

**Note 1:** Some important default values are:

* `ntop=500`: only the `500` most variable features are used
* `scale_features=TRUE`: standardises expression values so that each feature
  has unit variance; normalises the relative importance of features expressed
  at high and low levels in the calculation of principal components.

**Note 2:** Some important non-default values are:

* `exprs_values`: the normalised log~2~-transformed CPM are used.
* `feature_set`: only endogenous features are used.

```{r plotPCASCESet_PC1and2}
plotPCASCESet(
  sce.filtered, ncomponents = 2, exprs_values = "norm_exprs",
  colour_by = "Status", shape_by = "Time",
  feature_set = !isSpike(sce.filtered)
)
```

The above figure emphasises the importance of the `Time` factor, significantly
associated with `PC1`; in the figure, cells progress from the earliest time
points at the lowest coordinates on `PC1` to the latest time points at the
highest coordinates of `PC1`.

In addition, `PC2` contributes to the increasingly marked separation of
stimulated cells (*i.e.*, `infected` and `stimulated`) from `uninfected` cells;
`uninfected` cells generally appear at lower coordinates than their stimulated
counterparts on both `PC2` and `PC1`, with this trend becoming increasingly
apparent over time.

Overall, the observations above suggest different transcriptional *paths* taken
by cells in their experimental stimuli.

The analysis may be extended to more principal components; however, by
definition those principal will report decreasing amount of variance in the
data set:

```{r plotPCASCESet_PC1to3}
plotPCASCESet(
  sce.filtered, ncomponents = 3, exprs_values = "norm_exprs",
  colour_by = "Status", shape_by = "Time",
  feature_set = !isSpike(sce.filtered)
)
```

We may also use the `shape` aesthetic to examine the clustering of samples
according to their `Infection` phenotype:

```{r plotPCASCESet_infection}
plotPCASCESet(
  sce.filtered, ncomponents = 3, exprs_values = "norm_exprs",
  colour_by = "Infection", shape_by = "Time",
  feature_set = !isSpike(sce.filtered)
)
```

## t-distributed stochastic neighbour embedding (t-SNE)

Let us apply the t-distributed stochastic neighbour embedding (t-SNE) dimensionality reduction technique on the normalised expression of the
`500` most variable endogenous features to check for substructure in the data set:

```{r plotTSNE_status}
sce.filtered <- plotTSNE(
  sce.filtered, ncomponents = 2, exprs_values = "norm_exprs",
  colour_by = "Status", shape_by = "Time", return_SCESet = TRUE,
  feature_set = !isSpike(sce.filtered),
  rand_seed = 1794
)
```

While the `Status` factor is investigated in the above figure,
we may also examine the `Infection` factor below:

```{r plotTSNE_infection}
plotReducedDim.SCESet(
  sce.filtered, ncomponents = 2, colour_by = "Infection", shape_by = "Time"
)
```


Alternatively, we may overlay the expression level of a particular gene onto
the data points.

```{r plotTSNE_TNF}
plotGeneId <- subset(
    mcols(gtfData), gene_name == "TNF", select = gene_id, drop = TRUE
  )
plotReducedDim.SCESet(
  sce.filtered, ncomponents = 2,
  colour_by = plotGeneId, shape_by = "Time"
  ) +
  guides(colour = guide_colourbar(title = "TNF"))
```

## Diffusion map

Finally, let us also produce a diffusion map plot of two components as a
dimensionality reduction of the endogenous features:

```{r plotDiffusionMapSCESet}
plotDiffusionMapSCESet(
  sce.filtered, exprs_values = "norm_exprs",
  colour_by = "Status", shape_by = "Time",
  feature_set = !isSpike(sce.filtered),
  rand_seed = 1794
)
```

# Saved session

Let us save the current session state for later use:

```{r saveSession}
save.image(file = "01_Preprocessing_QC.RData")
```

# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
[^1]: References needed.
