---
title: "Dimensionality reduction of mini-bulks"
author: "Kevin Rue-Albrecht"
date: "06/01/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
stopifnot(
  require(DESeq2),
  require(ggrepel)
)
sce.filtered <- readRDS("rds/sce.minibulks.filtered_excludeD3.rds")
```

# Normalise using DESeq2

To avoid a warning raised by `DESeq2`, let us 'clean' phenotype levels
to avoid the use of unsafe characters
(*i.e.* characters that may be interpreted by R as having a special meaning):

```{r}
sce.filtered$Group_Safe <- factor(
  gsub(
    "Violet \\+", "Violet",
    gsub(
      "STM-", "",
      sce.filtered$Group
    )
  ),
  gsub(
    "Violet \\+", "Violet",
    gsub(
      "STM-", "",
      levels(sce.filtered$Group)
    )
  )
)
```

Make a `DESeqDataSet` and estimate the size factors:

```{r dds}
dds <- DESeqDataSetFromMatrix(
  countData = assay(sce.filtered, "counts"),
  colData = colData(sce.filtered)[,c("Group_Safe","Donor")],
  design= ~ Donor + Group_Safe
)
dds <- estimateSizeFactors(dds)
```

Extract transformed values:

```{r}
se.vsd <- vst(dds, blind=FALSE)
```

# Apply PCA

Compute the principal components:

```{r}
pca <- prcomp(t(assay(se.vsd)), scale. = TRUE)
```

Join with phenotype information:

```{r}
stopifnot(
  all(colnames(sce.filtered) == rownames(pca$x))
)
pca.data <- cbind(
  data.frame(pca$x),
  data.frame(colData(sce.filtered))
)
```

```{r pca_plot}
ggplot(pca.data, aes(PC1, PC2)) +
  geom_point(
    aes(colour = Infection, shape = Status),
    size = 2
  ) +
  geom_text_repel(
    aes(label = paste(Time, Donor, sep = "_")),
    alpha = 0.4,
    min.segment.length = unit(0, "lines")
  ) +
  theme_minimal()
```

Removing donor `D3` (here, simply hiding the data points; not recomputing the PCA!)

```{r}
ggplot(
  subset(pca.data, Donor != "D3"),
  aes(PC1, PC2)
) +
  geom_point(
    aes(colour = Infection, shape = Status),
    size = 2
  ) +
  geom_text_repel(
    aes(label = paste(Time, Donor, sep = "_")),
    alpha = 0.4,
    min.segment.length = unit(0, "lines")
  ) +
  theme_minimal()
```

# PCA without donor 3

Compute the principal components:

```{r}
pca.excludeD3 <- prcomp(t(assay(se.vsd[,se.vsd$Donor != "D3"])), scale. = TRUE)
```

Join with phenotype information:

```{r}
stopifnot(
  all(colnames(sce.filtered) == rownames(pca$x))
)
pca.excludeD3.data <- cbind(
  data.frame(pca.excludeD3$x),
  data.frame(colData(sce.filtered[,sce.filtered$Donor != "D3"]))
)
```
