---
title: "Differential expression between unsupervised clusters using *scde*"
bibliography:
  bibtex.bib
---

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
stopifnot(
  require(scater),
  require(scran),
  require(ggplot2),
  require(ComplexHeatmap),
  require(dplyr),
  require(ggrepel),
  require(RColorBrewer),
  require(goseq)
)
sce.norm <- readRDS("rds/sce.norm.tSNE.rds")
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,2,4)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.norm$Time)[1:3]
names(col.infection) <- levels(sce.norm$Infection)
names(col.status) <- levels(sce.norm$Status)
sce.endo <- sce.norm[!isSpike(sce.norm),]
outdir <- "05_cluster_out"; dir.create(outdir, showWarnings = FALSE)
# Subset by time point
sce.2h <- sce.endo[,sce.endo$Time == '2h']
sce.6h <- sce.endo[,sce.endo$Time == '6h']
sce.4h <- sce.endo[,sce.endo$Time == '4h']
# Filter detected genes
filterCounts <- function(m, counts = 10, cells = 10){
  apply(m, 1, function(e){
    return(sum(e >= counts) >= cells)
  })
}
sg.2h <- droplevels(sce.2h$Group)
keep.2h <- filterCounts(counts(sce.2h))
cd.2h <- counts(sce.2h)[keep.2h,]; storage.mode(cd.2h) <- 'integer'
sg.4h <- droplevels(sce.4h$Group)
keep.4h <- filterCounts(counts(sce.4h))
cd.4h <- counts(sce.4h)[keep.4h,]; storage.mode(cd.4h) <- 'integer'
sg.6h <- droplevels(sce.6h$Group)
keep.6h <- filterCounts(counts(sce.6h))
cd.6h <- counts(sce.6h)[keep.6h,]; storage.mode(cd.6h) <- 'integer'
```

<!-- Cleaner code based on 08_SCDE_v8
Remove ERCC spike-ins
!! Split cells by time
!! Retain genes detected at each time (10 counts, 5 cells, 1 group)
Identify cell clusters @ time
<!-- Error model: {cells @ time; genes @ time} <- same as 08_SCDE
Prior @ time
DE: {group on all cells with NA; no batch correction; P-value < 0.01}
-->

# Prepare data

**Note:**
for this analysis, the data is prepared identically to the
[differential expression between experimental groups using *scde*](05_SCDE.html):

* ERCC spike-in features are removed,
* Cells in the data set is subsetted by time point,
* Genes in the data set are subsetted for those detected at sufficient levels
  within each time point,

# Identify clusters {#identifyClusters .tabset}

Let us carry out an unsupervised identification of clusters within
each time point.

## 2h

```{r clusters.2h}
clusters <- quickCluster(
  sce.2h, min.size=min(table(sce.2h$Treatment)),
  assay.type="counts")
sce.2h$quickCluster <- clusters
qc.2h <- rep(NA, ncol(sce.endo)); names(qc.2h) <- colnames(sce.endo)
qc.2h[colnames(sce.2h)] <- as.numeric(levels(clusters)[clusters])
sce.endo$quickCluster.2h <- as.factor(qc.2h)
```

## 4h

```{r clusters.4h}
clusters <- quickCluster(
  sce.4h, min.size=min(table(sce.4h$Treatment)),
  assay.type="counts")
sce.4h$quickCluster <- clusters
qc.4h <- rep(NA, ncol(sce.endo)); names(qc.4h) <- colnames(sce.endo)
qc.4h[colnames(sce.4h)] <- as.numeric(levels(clusters)[clusters])
sce.endo$quickCluster.4h <- as.factor(qc.4h)
```

## 6h

```{r clusters.6h}
clusters <- quickCluster(
  sce.6h, min.size=min(table(sce.6h$Treatment)),
  assay.type="counts")
sce.6h$quickCluster <- clusters
qc.6h <- rep(NA, ncol(sce.endo)); names(qc.6h) <- colnames(sce.endo)
qc.6h[colnames(sce.6h)] <- as.numeric(levels(clusters)[clusters])
sce.endo$quickCluster.6h <- as.factor(qc.6h)
```

```{r, echo=FALSE, eval=FALSE}
saveRDS(sce.endo, "rds/sce.endo_clusters.rds")
```

# tSNE {.tabset}

Let us first define:

* a set of colours used to indicate *cluster membership* consistently
  across the upcoming sections:

```{r col.sets}
col.cluster <- brewer.pal(12, "Set3")[c(1,12,3,4,9)]
names(col.cluster) <- c(1:3,0,"NA")
```

## 2h

```{r plotReducedDim.2h, echo=FALSE}
tmpGG <- data.frame(
  reducedDim(sce.endo, "TSNE"),
  Cluster = sce.endo$quickCluster.2h,
  Treatment = gsub("_", " ", sce.endo$Treatment)
)
ggplot(tmpGG, aes(X1, X2, shape = Treatment, colour = Cluster)) +
  geom_point() +
  theme_bw() +
  labs(x="Dimension 1",y="Dimension 2",title="2h") +
  scale_color_manual(values = col.cluster, na.value = col.cluster["NA"])
```

```{r reducedDimension_pdf.2h, include=FALSE}
ggsave(file.path(outdir, "tSNE_clusters_2h.pdf"), width = 6, height = 4)
```

Cross-tabulated counts of cells by cluster and experimental treatment:

```{r table.2h, results='asis', echo=FALSE}
pander::pandoc.table(
  with(colData(sce.endo), table(Treatment, quickCluster.2h))
)
```

Visualised as a bar plot:

```{r clusterBarplot.2h, echo=FALSE}
tmpGG <- ggplot(data.frame(colData(sce.2h))) +
  stat_count(aes(x = gsub("_", " ", Treatment), fill = quickCluster)) +
  facet_grid(~ ifelse(
    quickCluster == 0, " No cluster" , gsub("^", "cluster ", quickCluster)
  )) +
  scale_fill_manual(values = col.cluster, na.value = col.cluster["NA"]) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle=90,hjust=1,vjust=0.5,face="bold")
  ) +
  labs(x = NULL, y = "Cells", fill = "Cluster")
tmpGG
ggsave(file.path(outdir, "cluster_count_2h.pdf"), width = 9, height = 4)
ggsave(
  file.path(outdir, "cluster_count_2h_wrap.pdf"),
  ggplot(data.frame(colData(sce.2h))) +
    stat_count(aes(
      x = interaction(Infection, Status, sep = "\n"),
      fill = quickCluster
    )) +
    facet_wrap(~ ifelse(
      quickCluster == 0, " No cluster" , gsub("^", "cluster ", quickCluster)
    )) +
    scale_fill_manual(values = col.cluster, na.value = col.cluster["NA"]) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle=90,hjust=1,vjust=0.5,face="bold")
    ) +
    labs(x = NULL, y = "Cells", fill = "Cluster"),
  width = 9, height = 8
)
```

## 4h

```{r plotReducedDim.4h, echo=FALSE}
tmpGG <- data.frame(
  reducedDim(sce.endo, "TSNE"),
  Cluster = sce.endo$quickCluster.4h,
  Treatment = gsub("_", " ", sce.endo$Treatment),
  label = ifelse(
    sce.endo$Treatment == "LT2_infected" & sce.endo$quickCluster.4h == 2, 2, NA
  )
)
ggplot(tmpGG, aes(X1, X2, shape = Treatment, colour = Cluster)) +
  geom_point() +
  theme_bw() +
  # theme(legend.position = "bottom", legend.box = "vertical") +
  labs(x="Dimension 1",y="Dimension 2",title="4h") +
  scale_color_manual(values = col.cluster, na.value = col.cluster["NA"]) #+
  #geom_point(aes(colour = "red", size = as.numeric(label)), colour = "red", alpha = 0.25) # highlight
```

```{r reducedDimension_pdf.4h, include=FALSE}
ggsave(file.path(outdir, "tSNE_clusters_4h.pdf"), width = 6, height = 4)
```

Cross-tabulated counts of cells by cluster and experimental treatment:

```{r table.4h, results='asis', echo=FALSE}
pander::pandoc.table(
  with(colData(sce.endo), table(Treatment, quickCluster.4h))
)
```

Visualised as a bar plot:

```{r clusterBarplot.4h, echo=FALSE}
tmpGG <- ggplot(data.frame(colData(sce.4h))) +
  stat_count(aes(gsub("_", " ", Treatment), fill = quickCluster)) +
  facet_grid(~ gsub("^", "Cluster ", quickCluster)) +
  scale_fill_manual(values = col.cluster, na.value = col.cluster["NA"]) +
  theme_minimal() + theme(
    axis.text.x = element_text(angle=90,hjust=1,vjust=0.5,face="bold")
  ) + labs(y = "Cells", fill = "Cluster", x = NULL)
tmpGG
ggsave(file.path(outdir, "cluster_count_4h.pdf"), width = 9, height = 4)
ggsave(
  file.path(outdir, "cluster_count_4h_wrap.pdf"),
  ggplot(data.frame(colData(sce.4h))) +
    stat_count(aes(
      x = interaction(Infection, Status, sep = "\n"),
      fill = quickCluster
    )) +
    facet_wrap(~ ifelse(
      quickCluster == 0, " No cluster" , gsub("^", "cluster ", quickCluster)
    ), nrow = 1) +
    scale_fill_manual(values = col.cluster, na.value = col.cluster["NA"]) +
    theme_bw() +
    theme(
      axis.text.x = element_text(
        angle=90,hjust=1,vjust=0.5,face="bold", size = rel(1.25))
    ) +
    labs(x = NULL, y = "Cells", fill = "Cluster"),
  width = 9, height = 5
)
ggsave(
  file.path(outdir, "cluster_count_4h_vertical.pdf"),
  ggplot(data.frame(colData(sce.4h))) +
    stat_count(aes(
      x = interaction(Infection, Status, sep = "\n"),
      fill = quickCluster
    )) +
    facet_wrap(~ ifelse(
      quickCluster == 0, " No cluster" , gsub("^", "cluster ", quickCluster)
    ), ncol = 1) +
    scale_fill_manual(values = col.cluster, na.value = col.cluster["NA"]) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle=90,hjust=1,vjust=0.5,face="bold")
    ) +
    labs(x = NULL, y = "Cells", fill = "Cluster"),
  width = 9, height = 18
)
```

## 6h

```{r plotReducedDim.6h, echo=FALSE}
tmpGG <- data.frame(
  reducedDim(sce.endo, "TSNE"),
  Cluster = sce.endo$quickCluster.6h,
  Treatment = gsub("_", " ", sce.endo$Treatment)
)
ggplot(tmpGG, aes(X1, X2, shape = Treatment, colour = Cluster)) +
  geom_point() + theme_bw() + labs(x="Dimension 1",y="Dimension 2",title="6h") +
  scale_color_manual(values = col.cluster, na.value = col.cluster["NA"])
```

```{r reducedDimension_pdf.6h, include=FALSE}
ggsave(file.path(outdir, "tSNE_clusters_6h.pdf"), width = 6, height = 4)
```

Cross-tabulated counts of cells by cluster and experimental treatment:

```{r table.6h, results='asis', echo=FALSE}
pander::pandoc.table(
  with(colData(sce.endo), table(Treatment, quickCluster.6h))
)
```

Visualised as a bar plot:

```{r clusterBarplot.6h, echo=FALSE}
tmpGG <- ggplot(data.frame(colData(sce.6h))) +
  stat_count(aes(x = Treatment, fill = quickCluster)) +
  facet_grid(~ gsub("^", "cluster ", quickCluster)) +
  scale_fill_manual(values = col.cluster, na.value = col.cluster["NA"]) +
  theme_minimal() + theme(
    axis.text.x = element_text(angle=90,hjust=1,vjust=0.5,face="bold")
  ) + labs(y = "Cells", fill = "Cluster")
tmpGG
ggsave(file.path(outdir, "cluster_count_6h.pdf"), width = 9, height = 4)
ggsave(
  file.path(outdir, "cluster_count_6h_wrap.pdf"),
  ggplot(data.frame(colData(sce.6h))) +
    stat_count(aes(
      x = interaction(Infection, Status, sep = "\n"),
      fill = quickCluster
    )) +
    facet_wrap(~ ifelse(
      quickCluster == 0, " No cluster" , gsub("^", "cluster ", quickCluster)
    ), nrow = 1) +
    scale_fill_manual(values = col.cluster, na.value = col.cluster["NA"]) +
    theme_bw() +
    theme(
      axis.text.x = element_text(
        angle=90,hjust=1,vjust=0.5,face="bold", size = rel(1.25))
    ) +
    labs(x = NULL, y = "Cells", fill = "Cluster"),
  width = 9, height = 5
)
```
