---
title: "Markers of clusters using *scde*"
bibliography:
  bibtex.bib
editor_options: 
  chunk_output_type: console
---

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
stopifnot(
  require(scater),
  require(scran),
  require(ggplot2),
  require(RColorBrewer)
)
sce.endo <- readRDS("rds/sce.endo_clusters.rds")
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,4,2)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.endo$Time)[1:3]
names(col.infection) <- levels(sce.endo$Infection)
names(col.status) <- levels(sce.endo$Status)
outdir <- "06_density_out"; dir.create(outdir, showWarnings = FALSE)
```

# Helper values and functions

Let us first define some values useful to design colour scales:

```{r colours}
range.exprs <- range(assay(sce.endo, "logcounts"))
col.gene <- RColorBrewer::brewer.pal(11, "RdYlBu")
col.treatment <- RColorBrewer::brewer.pal(9, "Paired")[c(9,3,1,4,2)]
names(col.treatment) <- levels(sce.endo$Treatment)
```

Let us then define a few convenience functions:

* to fetch the gene ID associated with a gene name

```{r geneNameToID}
geneNameToID <- function(x){
  geneId <- with(rowData(sce.endo), gene_id[match(x, gene_name)])
  if (length(geneId) > 1){
    warning("Multiple IDs found. Use ID instead")
    return(data.frame(
      gene_name = x,
      gene_id = geneId
    ))
  } else if (length(geneId) == 0){
    stop("gene_name not found")
  }
  return(geneId)
}
```

* to fetch the necessary expression and phenotype data

```{r geneDataByName}
geneDataByName <- function(x){
  geneId <- geneNameToID(x)
  ggData <- data.frame(
    gene = assay(sce.endo, "logcounts")[geneId,],
    colData(sce.endo)[,c("Time","Infection","Status","Treatment")]
  )
  return(ggData)
}
```

* to draw a histogram

```{r histByName}
histByName <- function(x){
  geneId <- geneNameToID(x)
  ggData <- geneDataByName(x)
  # return(ggData)
  ggplot(ggData, aes(gene)) +
    facet_grid(gsub("_", "\n", Treatment)~Time, scales = "free_y") +
    geom_histogram(aes(colour=Treatment,fill=Treatment)) +
    scale_color_manual(values = col.treatment) +
    scale_fill_manual(values = col.treatment) +
    scale_y_continuous(breaks = seq(0, 30, 5)) +
    labs(
      title = sprintf("%s", x),
      subtitle = sprintf("%s", geneId),
      x = "log-counts",
      colour = "Treatment") +
    theme_bw()
}
```

* to draw a density plot

```{r densityByName}
densityByName <- function(x){
  geneId <- geneNameToID(x)
  ggData <- geneDataByName(x)
  # return(ggData)
  ggplot(ggData, aes(gene)) +
    facet_grid(gsub("_", "\n", Treatment) ~ Time, scales = "free_y") +
    geom_density(aes(colour = Treatment)) +
    geom_rug() +
    scale_color_manual(values = col.treatment) +
    labs(
      title = sprintf("%s", x),
      subtitle = sprintf("%s", geneId),
      x = "log-counts",
      colour = "Treatment") +
    theme_bw()
}
```


```{r plotByName, echo=FALSE}
plotByNameTimeRow <- function(x){
  geneId <- geneNameToID(x)
  ggData <- geneDataByName(x)
  # return(ggData)
  ggplot(ggData, aes(gene)) +
    facet_grid(Time~gsub("_", "\n", Treatment), scales = "free_y") +
    geom_density(aes(colour = Treatment)) +
    scale_color_manual(values = col.treatment) +
    scale_y_continuous(breaks = seq(0, 30, 5)) +
    labs(
      title = sprintf("%s", x),
      subtitle = sprintf("%s", geneId),
      x = "log-counts",
      colour = "Treatment") +
    theme_bw()
}
```

# Genes (alphabetical order) {.tabset}

The above convenience function make it straightforward to produce a figure
for any gene of interest:

## AIM2 {.tabset}

### Histogram

```{r AIM2_hist, echo=FALSE, message=FALSE}
geneName <- "AIM2"
histByName(geneName)
```

### Density

```{r AIM2_density, echo=FALSE, message=FALSE}
geneName <- "AIM2"
densityByName(geneName)
```

## APOC1 {.tabset}

### Histogram

```{r APOC1_hist, echo=FALSE, message=FALSE}
geneName <- "APOC1"
histByName(geneName)
```

### Density

```{r APOC1_density, echo=FALSE, message=FALSE}
geneName <- "APOC1"
densityByName(geneName)
```

## APOE {.tabset}

### Histogram

```{r APOE_hist, echo=FALSE, message=FALSE}
geneName <- "APOE"
histByName(geneName)
```

### Density

```{r APOE_density, echo=FALSE, message=FALSE}
geneName <- "APOE"
densityByName(geneName)
```

## CD83 {.tabset}

### Histogram

```{r CD83_hist, echo=FALSE, message=FALSE}
geneName <- "CD83"
histByName(geneName)
```

### Density

```{r CD83_density, echo=FALSE, message=FALSE}
geneName <- "CD83"
densityByName(geneName)
```

## CALM1 {.tabset}

### Histogram

```{r CALM1_hist, echo=FALSE, message=FALSE}
geneName <- "CALM1"
histByName(geneName)
```

### Density

```{r CALM1_density, echo=FALSE, message=FALSE}
geneName <- "CALM1"
densityByName(geneName)
```

## CTSL {.tabset}

### Histogram

```{r CTSL_hist, echo=FALSE, message=FALSE}
geneName <- "CTSL"
histByName(geneName)
```

### Density

```{r CTSL_density, echo=FALSE, message=FALSE}
geneName <- "CTSL"
densityByName(geneName)
```

## DEFB1 {.tabset}

### Histogram

```{r DEFB1_hist, echo=FALSE, message=FALSE}
geneName <- "DEFB1"
histByName(geneName)
```

### Density

```{r DEFB1_density, echo=FALSE, message=FALSE}
geneName <- "DEFB1"
densityByName(geneName)
```

## DUSP4 {.tabset}

### Histogram

```{r DUSP4_hist, echo=FALSE, message=FALSE}
geneName <- "DUSP4"
histByName(geneName)
```

### Density

```{r DUSP4_density, echo=FALSE, message=FALSE}
geneName <- "DUSP4"
densityByName(geneName)
```

## IFIT1 {.tabset}

### Histogram

```{r IFIT1_hist, echo=FALSE, message=FALSE}
geneName <- "IFIT1"
histByName(geneName)
```

### Density

```{r IFIT1_density, echo=FALSE, message=FALSE}
geneName <- "IFIT1"
densityByName(geneName)
```

## IFIT2 {.tabset}

### Histogram

```{r IFIT2_hist, echo=FALSE, message=FALSE}
geneName <- "IFIT2"
histByName(geneName)
```

### Density

```{r IFIT2_density, echo=FALSE, message=FALSE}
geneName <- "IFIT2"
densityByName(geneName)
```

## IFIT3 {.tabset}

### Histogram

```{r IFIT3_hist, echo=FALSE, message=FALSE}
geneName <- "IFIT3"
histByName(geneName)
```

### Density

```{r IFIT3_density, echo=FALSE, message=FALSE}
geneName <- "IFIT3"
densityByName(geneName)
```

## IFITM3 {.tabset}

### Histogram

```{r IFITM3_hist, echo=FALSE, message=FALSE}
geneName <- "IFITM3"
histByName(geneName)
```

### Density

```{r IFITM3_density, echo=FALSE, message=FALSE}
geneName <- "IFITM3"
densityByName(geneName)
```

## IL1B {.tabset}

### Histogram

```{r IL1B_hist, echo=FALSE, message=FALSE}
geneName <- "IL1B"
histByName(geneName)
```

### Density

```{r IL1B_density, echo=FALSE, message=FALSE}
geneName <- "IL1B"
densityByName(geneName)
```

## IL12B {.tabset}

### Histogram

```{r IL12B_hist, echo=FALSE, message=FALSE}
geneName <- "IL12B"
histByName(geneName)
```

### Density

```{r IL12B_density, echo=FALSE, message=FALSE}
geneName <- "IL12B"
densityByName(geneName)
```

## OASL {.tabset}

### Histogram

```{r OASL_hist, echo=FALSE, message=FALSE}
geneName <- "OASL"
histByName(geneName)
```

### Density

```{r OASL_density, echo=FALSE, message=FALSE}
geneName <- "OASL"
densityByName(geneName)
```

## RSAD2 {.tabset}

### Histogram

```{r RSAD2_hist, echo=FALSE, message=FALSE}
geneName <- "RSAD2"
histByName(geneName)
```

### Density

```{r RSAD2_density, echo=FALSE, message=FALSE}
geneName <- "RSAD2"
densityByName(geneName)
```

## SIGLEC1 {.tabset}

### Histogram

```{r SIGLEC1_hist, echo=FALSE, message=FALSE}
geneName <- "SIGLEC1"
histByName(geneName)
```

### Density

```{r SIGLEC1_density, echo=FALSE, message=FALSE}
geneName <- "SIGLEC1"
densityByName(geneName)
```

## SLC2A9 {.tabset}

### Histogram

```{r SLC2A9_hist, echo=FALSE, message=FALSE}
geneName <- "SLC2A9"
histByName(geneName)
```

### Density

```{r SLC2A9_density, echo=FALSE, message=FALSE}
geneName <- "SLC2A9"
densityByName(geneName)
```

## TBL3 {.tabset}

### Histogram

```{r TBL3_hist, echo=FALSE, message=FALSE}
geneName <- "TBL3"
histByName(geneName)
```

### Density

```{r TBL3_density, echo=FALSE, message=FALSE}
geneName <- "TBL3"
densityByName(geneName)
```

## TLR8 {.tabset}

### Histogram

```{r TLR8_hist, echo=FALSE, message=FALSE}
geneName <- "TLR8"
histByName(geneName)
```

### Density

```{r TLR8_density, echo=FALSE, message=FALSE}
geneName <- "TLR8"
densityByName(geneName)
```

## TNFAIP2 {.tabset}

### Histogram

```{r TNFAIP2_hist, echo=FALSE, message=FALSE}
geneName <- "TNFAIP2"
histByName(geneName)
```

### Density

```{r TNFAIP2_density, echo=FALSE, message=FALSE}
geneName <- "TNFAIP2"
densityByName(geneName)
```

## WARS {.tabset}

### Histogram

```{r WARS_hist, echo=FALSE, message=FALSE}
geneName <- "WARS"
histByName(geneName)
```

### Density

```{r WARS_density, echo=FALSE, message=FALSE}
geneName <- "WARS"
densityByName(geneName)
```
