---
title: "Gene expression levels overlaid onto tSNE view"
---

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
stopifnot(
  require(scater),
  require(scran),
  require(ggplot2),
  require(SummarizedExperiment),
  require(RColorBrewer),
  require(circlize),
  require(ComplexHeatmap)
)
sce.endo <- readRDS("rds/sce.endo_clusters.rds")
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,2,4)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.endo$Time)[1:3]
names(col.infection) <- levels(sce.endo$Infection)
names(col.status) <- levels(sce.endo$Status)
outdir <- "06_gene_TSNE_out"; dir.create(outdir, showWarnings = FALSE)
pdfWidth <- 7; pdfHeight <- 5
pdfName <- "%s_gene_TSNE.pdf"
```

# Helper values and functions

Let us first define some values useful to design colour scales:

```{r colours}
range.exprs <- range(assay(sce.endo, "logcounts"))
col.gene <- RColorBrewer::brewer.pal(11, "RdYlBu")
col.treatment <- RColorBrewer::brewer.pal(9, "Paired")[c(9,3,1,4,2)]
names(col.treatment) <- levels(sce.endo$Treatment)
```

Let us then define a few convenience functions:

* to fetch the gene ID associated with a gene name

```{r geneNameToID}
geneNameToID <- function(x){
  geneId <- with(rowData(sce.endo), gene_id[match(x, gene_name)])
  if (length(geneId) > 1){
    warning("Multiple IDs found. Use ID instead")
    return(data.frame(
      gene_name = x,
      gene_id = geneId
    ))
  } else if (length(geneId) == 0){
    stop("gene_name not found")
  }
  return(geneId)
}
```

* to fetch the necessary expression and phenotype data

```{r geneDataByName}
geneDataByName <- function(x){
  geneId <- geneNameToID(x)
  ggData <- data.frame(
    gene = assay(sce.endo, "logcounts")[geneId,],
    dimension1 = reducedDim(sce.endo, "TSNE")[,1],
    dimension2 = reducedDim(sce.endo, "TSNE")[,2],
    colData(sce.endo)[,c("Time","Infection","Status","Treatment")]
  )
  return(ggData)
}
```

* to draw the figure

```{r}
plotByName <- function(x){
  geneId <- geneNameToID(x)
  ggData <- geneDataByName(x)
  # return(ggData)
  ggplot(ggData, aes(dimension1, dimension2, shape = paste(Infection, Status))) +
    geom_point(aes(colour = gene), alpha = 0.75, size = 1.5) +
    scale_color_gradient2(
      low = col.gene[11], mid = col.gene[5], high = col.gene[1],
      midpoint = mean(range.exprs),
      limits = range.exprs
    ) +
    scale_fill_manual(values = col.treatment) +
    labs(
      title = sprintf("%s", x),
      subtitle = sprintf("%s", geneId),
      x = "Dimension 1",
      y = "Dimension 2",
      colour = "logcounts",
      fill = "Treatment",
      shape = "Treatment") +
    theme_bw()
}
```

# Genes (alphabetical order) {.tabset}

The above convenience function make it straightforward to produce a figure
for any gene of interest:

## IL12B

```{r IL12B, echo=FALSE}
geneName <- "IL12B"
plotByName(geneName)
```

```{r IL12B_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName)),
  height = pdfHeight, width = pdfWidth)
```

## IL1B

```{r IL1B, echo=FALSE}
geneName <- "IL1B"
plotByName(geneName)
```

```{r IL1B_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName)),
  height = pdfHeight, width = pdfWidth)
```
