---
title: "Overview of normalised expression data"
---

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
stopifnot(
  require(scater),
  require(scran),
  require(ggplot2),
  require(SummarizedExperiment),
  require(RColorBrewer),
  require(circlize),
  require(ComplexHeatmap)
)
sce.norm <- readRDS("rds/sce.norm.rds")
sce.all <- readRDS("rds/sce.all.SCE.rds")
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,4,2)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.norm$Time)[1:3]
names(col.infection) <- levels(sce.norm$Infection)
names(col.status) <- levels(sce.norm$Status)
outdir <- "06_reduceDim_out"; dir.create(outdir, showWarnings = FALSE)
```

# Plotting themes

Let us first define theme elements used throughout various figures
in the following sections:

```{r PCAtheme}
PCAtheme <- theme_bw() + theme(
    legend.position = "bottom", legend.box = "vertical"
  )
```


# Relative importance of experimental factors {#explanatoryVariables .tabset}

Having normalised the expression data for detected features in cells that
passed quality control, the relative importance of experimental factors---both
technical and biological---provides insight into uninteresting technical
factors that may need to be accounted for when explaining variance in the
expression data.

## Explanatory variables

Let us examine the proportion of variance explained by known experimental
factors, as well as the total spike-in count in each cell:

```{r plotExplanatoryVariables, echo=FALSE}
plotExplanatoryVariables(
  sce.norm,
  xprs_values = "logcounts",
  variables = c(
    "counts_feature_controls_ERCC",
    "log10_counts_feature_controls_ERCC",
    "Plate", "Lane", "Time", "Status", "Infection"
    )
)
```

The above figure reinforces the effect of `Time` as the primary source of
variance in gene expression, with substantial proportion of variance
explained for a subset of features. Next are `Status` and `Infection`
factors, while other technical factors share explanation of lower amount
of the residual variance in the data set.

**Note:**
Due to technical constraints, the the representation of time points is uneven
across the four plates used for library preparation.
In summary, while 75% of each plate is balanced, the remaining 25% are:

* a block of single cells collected at 2h
* a block of single cells collected at 4h
* a block of single cells collected at 6h
* an even mixture of cells from 2h, 4h, and 6h

As a result, some variance is expected to associate with the `Plate` factor,
as it is partially confounded with the powerful `Time` factor.

For reference, the plate design is shown in the second tab.

## Plate design (full)

```{r plateDesign_full, echo=FALSE}
plateDesign <-
  dplyr::mutate(
    as.data.frame(colData(sce.all)[,c("Plate","Well","Time","Infection","Status")]),
    Row = as.factor(gsub("([[:alpha:]])[[:digit:]]*", "\\1", Well)),
    Column = as.factor(as.numeric(gsub("[[:alpha:]]([[:digit:]]*)", "\\1", Well))),
    Plate = gsub("P", "Plate ", Plate)
  )
ggplot(plateDesign, aes(Column, Row, fill = Time, colour = Infection, alpha = Status)) +
  geom_tile(size = 1) + facet_wrap(~ Plate) + labs(x="",y="") +
  scale_y_discrete(limits = rev(levels(plateDesign$Row))) +
  theme_minimal() + theme(panel.grid = element_blank()) +
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Set1") +
  guides(
    fill = guide_legend(order = 1),
    colour = guide_legend(order = 2),
    shape = guide_legend(order = 3)
  )
```

## Plate design (Cells passed QC)

```{r plateDesign_pass, echo=FALSE}
plateDesign <-
  dplyr::mutate(
    as.data.frame(colData(sce.norm)[,c("Plate","Well","Time","Infection","Status")]),
    Row = as.factor(gsub("([[:alpha:]])[[:digit:]]*", "\\1", Well)),
    Column = as.factor(as.numeric(gsub("[[:alpha:]]([[:digit:]]*)", "\\1", Well))),
    Plate = gsub("P", "Plate ", Plate)
  )
ggplot(plateDesign, aes(Column, Row, fill = Time, colour = Infection, alpha = Status)) +
  geom_tile(size = 1) + facet_wrap(~ Plate) + labs(x="",y="") +
  scale_y_discrete(limits = rev(levels(plateDesign$Row))) +
  theme_minimal() + theme(panel.grid = element_blank()) +
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Set1") +
  scale_alpha_discrete(range = c(0.1, 3/5)) +
  guides(
    fill = guide_legend(order = 1),
    colour = guide_legend(order = 2),
    shape = guide_legend(order = 3)
  )
```

# Dimensionality reduction methods

Let us examine the qualitatite distribution of cells using different
dimensionality reduction approaches.

## Principal component analysis (PCA) {#PCA}

### All cells

Let us initially focus our attention on the two principal components that
explain the largest amount of variance in the endogenous features.

**Note 1:** Some important *default* values are:

* `ntop=500`: only the `500` most variable features are used
* `scale_features=TRUE`: standardises expression values so that each feature
  has unit variance; this procedure normalises the relative importance of
  features expressed at high and low levels in the calculation of principal
  components.

**Note 2:** Some important *non-default* values are:

* `exprs_values="norm_exprs"`:
  the normalised log~2~-transformed CPM are used.
* `feature_set=!isSpike(sce.norm)`:
  ERCC spike-in features are not considered
  (see this [earlier section](05_normalisation.html#setSpike) where the
  set of spike-in features is defined).

```{r plotPCASCESet}
sce.norm <- runPCA(sce.norm, ncomponents = 3, exprs_values = "logcounts",
  feature_set = NULL, scale_features = TRUE, pca_data_input = "logcounts")
```

#### Principal components 1 and 2 {.tabset}

##### Treatment

```{r plotPCASCESet_Group, fig.height=6, echo=FALSE}
ggPCA <- data.frame(
  reducedDim(sce.norm, "PCA")[,paste0("PC", 1:2)],
  colData(sce.norm)[,c("Group","Time","Infection","Status","Treatment","Plate")]
)
ggplot(ggPCA, aes(PC1, PC2)) +
  geom_point(aes(colour = Treatment, shape = Time)) + geom_rug() +
  PCAtheme + theme(legend.text = element_text(size = 6))
```

The above figure emphasises the importance of the `Time` factor, significantly
associated with `PC1`; in the figure, cells progress from the earliest time
points at the lowest coordinates on `PC1` to the latest time points at the
highest coordinates of `PC1`.

In addition, `PC2` contributes to the increasingly marked separation of
stimulated cells (*i.e.*, `infected` and `stimulated`,
in orange, green, red, and blue) from `uninfected` cells (in blue);
`uninfected` cells generally appear at lower coordinates than their `infected`
and `exposed` counterparts on both `PC2` and `PC1`, with this trend becoming
increasingly apparent over time.

Overall, the observations above suggest different transcriptional trajectories
taken by cells in response to experimental stimuli.

##### Time

```{r plotPCASCESet_Time, fig.height=6, echo=FALSE}
ggPCA <- data.frame(
  reducedDim(sce.norm, "PCA")[,paste0("PC", 1:2)],
  colData(sce.norm)[,c("Group","Time","Infection","Status","Treatment","Plate")]
)
ggplot(ggPCA, aes(PC1, PC2)) +
  geom_point(aes(colour = Time, shape = Infection)) +
  scale_color_manual(values = col.time) + geom_rug() +
  PCAtheme + theme(legend.text = element_text(size = 6))
```

##### Infection

```{r plotPCASCESet_Infection, fig.height=6, echo=FALSE}
ggPCA <- data.frame(
  reducedDim(sce.norm, "PCA")[,paste0("PC", 1:2)],
  colData(sce.norm)[,c("Group","Time","Infection","Status","Treatment","Plate")]
)
ggplot(ggPCA, aes(PC1, PC2)) +
  geom_point(aes(colour = Infection, shape = Time)) +
  scale_color_manual(values = col.infection) + geom_rug() +
  PCAtheme + theme(legend.text = element_text(size = 6))
```

##### Status

```{r plotPCASCESet_Status, fig.height=6, echo=FALSE}
ggPCA <- data.frame(
  reducedDim(sce.norm, "PCA")[,paste0("PC", 1:2)],
  colData(sce.norm)[,c("Group","Time","Infection","Status","Treatment","Plate")]
)
ggplot(ggPCA, aes(PC1, PC2)) +
  geom_point(aes(colour = Status, shape = Time)) +
  scale_color_manual(values = col.status) + geom_rug() +
  PCAtheme + theme(legend.text = element_text(size = 6))
```

```{r plotPCASCESet_Status_pdf, include=FALSE}
ggPCA <- data.frame(
  reducedDim(sce.norm, "PCA")[,paste0("PC", 1:2)],
  colData(sce.norm)[,c("Group","Time","Infection","Status","Treatment","Plate")]
)
ggplot(ggPCA, aes(PC1, PC2)) +
  geom_point(aes(colour = Status, shape = Time)) +
  scale_color_manual(values = col.status) + geom_rug() +
  PCAtheme + theme(
    legend.text = element_text(size = 6),
    legend.position = "right")
ggsave(file.path(outdir, "PCA_status.pdf"), height = 5, width = 7)
```

#### Additional principal components {.tabset}

The analysis may be extended to more principal components; however, by
definition those principal components will report decreasing amount of variance
in the data set:

##### Percentage variance explained

Percentage of variance explained by prinicipal components:

```{r PercentVariance_PCA, echo=FALSE}
ggplot(
  data.frame(
    PercentVariance = head(attr(reducedDim(sce.norm, "PCA"), "percentVar"), 10),
    PC = factor(1:10)
  )
) + labs(x = "Principal component", y = "Percentage variance explained") +
  geom_col(aes(PC, PercentVariance)) + theme_minimal()
```

```{r PercentVariance_pdf, include=FALSE}
ggsave(file.path(outdir, "PCA_variance_explained.pdf"), height = 5, width = 7)
```


##### Treatment

```{r plotPCASCESet_PC1to3_Treatment, echo=FALSE}
plotReducedDim(
  sce.norm, use_dimred = "PCA", ncomponents = 3, exprs_values = "logcounts",
  colour_by = "Treatment", shape_by = "Time") +
  PCAtheme + theme(legend.text = element_text(size = 6))
```

##### Time

```{r plotPCASCESet_PC1to3_Time, echo=FALSE, warning=FALSE}
plotReducedDim(
  sce.norm, use_dimred = "PCA", ncomponents = 3,
  colour_by = "Time", shape_by = "Plate") +
  PCAtheme + scale_color_manual(values = col.time)
```

##### Status

```{r plotPCASCESet_PC1to3_Status, echo=FALSE, warning=FALSE}
plotReducedDim(
  sce.norm, use_dimred = "PCA", ncomponents = 3,
  colour_by = "Status", shape_by = "Time") +
  PCAtheme + scale_color_manual(values = col.status)
```

##### Infection

```{r plotPCASCESet_PC1to3_Infection, echo=FALSE, warning=FALSE}
plotReducedDim(
  sce.norm, use_dimred = "PCA", ncomponents = 3,
  colour_by = "Infection", shape_by = "Time") +
  PCAtheme + scale_color_manual(values = col.infection)
```

### PCA analyses within time points {.tabset}

In the above [section](#PCA), the overwhelming influence of the `Time` factor
likely shadows the relatively weaker effect of other biologically relevant
factors `Infection` and `Status`.

Let us perform PCA at each time point:

```{r time.sce}
time.sce <- list()
for (time in levels(sce.norm$Time)){
  tmp.sce <- sce.norm[,sce.norm$Time == time]
  keep_feature <- (rowMeans(counts(tmp.sce)) >= 1)
  tmp.sce <- tmp.sce[keep_feature,]
  tmp.sce <- runPCA(
    tmp.sce, ncomponents = 3, exprs_values = "logcounts",
    feature_set = !isSpike(tmp.sce, "ERCC")
  )
  time.sce[[time]] <- tmp.sce
  rm(tmp.sce)
}
rm(time)
```

**Note:**
Below, the tab name indicates the experimental factor used to *colour*
single cells (*i.e.*, `Infection` or `Status`); the other factor is then
used to *shape* each data point.

#### 2h {.tabset}

##### Infection

```{r 2h-Infection, fig.height=6, echo=FALSE, warning=FALSE}
plotReducedDim(
  time.sce[["2h"]], use_dimred = "PCA", ncomponents = 3,
  colour_by = "Infection", shape_by = "Status") +
  PCAtheme + scale_color_manual(values = col.infection) +
  guides(colour = guide_legend(title = "Infection"))
```

##### Status

```{r 2h-Status, fig.height=6, echo=FALSE, warning=FALSE}
plotReducedDim(
  time.sce[["2h"]], use_dimred = "PCA", ncomponents = 3,
  colour_by = "Status", shape_by = "Infection") +
  PCAtheme + scale_color_manual(values = col.status) +
  guides(colour = guide_legend(title = "Status"))
```

##### Treatment

```{r 2h-Treatment, fig.height=6, echo=FALSE}
plotReducedDim(
  time.sce[["2h"]], use_dimred = "PCA", ncomponents = 3,
  colour_by = "Treatment") +
  PCAtheme
```

#### 4h {.tabset}

##### Infection

```{r 4h-Infection, fig.height=6, echo=FALSE, warning=FALSE}
plotReducedDim(
  time.sce[["2h"]], use_dimred = "PCA", ncomponents = 3,
  colour_by = "Infection", shape_by = "Status") +
  PCAtheme + scale_color_manual(values = col.infection) +
  guides(colour = guide_legend(title = "Infection"))
```

##### Status

```{r 4h-Status, fig.height=6, echo=FALSE, warning=FALSE}
plotReducedDim(
  time.sce[["2h"]], use_dimred = "PCA", ncomponents = 3,
  colour_by = "Status", shape_by = "Infection") +
  PCAtheme + scale_color_manual(values = col.status) +
  guides(colour = guide_legend(title = "Status"))
```

##### Treatment

```{r 4h-Treatment, fig.height=6, echo=FALSE}
plotReducedDim(
  time.sce[["2h"]], use_dimred = "PCA", ncomponents = 3,
  colour_by = "Treatment") + PCAtheme
```

#### 6h {.tabset}

##### Infection

```{r 6h-Infection, fig.height=6, echo=FALSE, warning=FALSE}
plotReducedDim(
  time.sce[["2h"]], use_dimred = "PCA", ncomponents = 3,
  colour_by = "Infection", shape_by = "Status") +
  PCAtheme + scale_color_manual(values = col.infection) +
  guides(colour = guide_legend(title = "Infection"))
```

##### Status

```{r 6h-Status, fig.height=6, echo=FALSE, warning=FALSE}
plotReducedDim(
  time.sce[["2h"]], use_dimred = "PCA", ncomponents = 3,
  colour_by = "Status", shape_by = "Infection") +
  PCAtheme + scale_color_manual(values = col.status) +
  guides(colour = guide_legend(title = "Status"))
```

##### Treatment

```{r 6h-Treatment, fig.height=6, echo=FALSE}
plotReducedDim(
  time.sce[["2h"]], use_dimred = "PCA", ncomponents = 3,
  colour_by = "Treatment") + PCAtheme
```

## t-distributed stochastic neighbour embedding (t-SNE)

### Experimental factors {.tabset}

Let us apply the t-distributed stochastic neighbour embedding (t-SNE) dimensionality reduction technique on the normalised expression of the
**500** most variable endogenous features to check for substructure in the data set.
Importantly, let us set the `perplexity` parameter of the `Rtsne` function to
half the average size of an experimental group of single cells
(*i.e.*, `r round(mean(table(sce.norm$Group)) / 2)`; see this
[earlier section](04_feature_QC.html#selectFeatures) for the count of cells
per group).

**Note**: "In t-SNE, the perplexity may be viewed as a knob that sets the number
of effective nearest neighbors. It is comparable with the number of nearest
neighbors $k$ that is employed in many manifold learners
(see the GitHub website of
[Laurens van der Maaten](https://lvdmaaten.github.io/tsne/)).
As a consequence, an appropriate value of perplexity is necessary to
allow data points (here, single cells) to form clusters with a reasonable
number of nearest neighbours;
instead, excessively large values of perplexity usually result
in large 'balls' of uniformly distributed points, as all data points attempt to be equidistant from each other:

```{r plotTSNE}
sce.norm <- plotTSNE(
  sce.norm, ncomponents = 2, exprs_values = "logcounts",
  colour_by = "Status", shape_by = "Time",
  feature_set = !isSpike(sce.norm),
  perplexity = round(mean(table(sce.norm$Group)) / 2),
  rand_seed = 1794,
  return_SCE = TRUE, draw_plot = TRUE
)
```

```{r saveRDS, include=FALSE}
saveRDS(sce.norm, 'rds/sce.norm.tSNE.rds')
```

```{r postertSNE, include=FALSE}
posterPCA <- reducedDim(sce.norm, "TSNE")[,1:2]
colnames(posterPCA) <- paste("Dimension", 1:2)
posterSamples <- rownames(posterPCA)
posterPCA <- cbind(
  posterPCA,
  data.frame(colData(sce.norm)[posterSamples, c("Status","Infection","Time")])
  )
lc <- brewer.pal(3, "Set1")[c(1,3,2)] # RGB
posterLabels <- data.frame(
  x = c(-20, -2.5, -5, 10, 15),
  y = c(-12.5, -5, 10, -2.5, 10),
  label = c("2h", "4h", "6h", "4h", "6h"),
  col = c("black", lc[1], lc[1], "black", "black")
)
posterPCA <- ggplot(
  posterPCA,
  aes_string("`Dimension 1`","`Dimension 2`")) +
  geom_point(aes(colour = Status, shape = Infection)) + 
  scale_color_manual(values = col.status) +
  geom_label(aes(x, y, label = label), posterLabels, inherit.aes = FALSE) +
  theme_light()
posterPCA
ggsave(
  file.path(outdir, "poster_tSNE_right.pdf"),
  posterPCA,
  width = 7, height = 5)
ggsave(
  file.path(outdir, "poster_tSNE_right_blankAxes.pdf"),
  posterPCA + theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) + labs(x=NULL, y=NULL),
  width = 7, height = 5)
ggsave(
  file.path(outdir, "poster_tSNE_bottom.pdf"),
  posterPCA + theme(
    legend.position = "bottom", legend.box = "vertical"
  ),
  width = 6, height = 6)
```

#### Status {#tSNEstatus}

```{r plotTSNE_status, echo=FALSE}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2,
  colour_by="Status",shape_by="Time") + PCAtheme
```

#### Infection

```{r plotTSNE_infection, echo=FALSE}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2,
  colour_by="Infection",shape_by="Time") + PCAtheme
```

#### Treatment

```{r plotTSNE_Treatment, echo=FALSE}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2,
  colour_by="Treatment",shape_by="Time") + PCAtheme
```

#### Plate (batch)

We may overlay the plate identifier onto the tSNE coordinates,
to demonstrate the absence of batch effect associated with the preparation
of sequencing libraries:

```{r plotTSNE_Plate, echo=FALSE}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2,
  colour_by="Plate",shape_by="Time") + PCAtheme
```

#### quickCluster (normalisation)

We may also overlay the clusters identify during
[normalisation](05_normalisation.html) onto the t-SNE:

```{r plotTSNE_quickCluster, echo=FALSE}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2,
  colour_by="Cluster",shape_by="Time") + PCAtheme
```

### Genes {.tabset}

Alternatively, we may also overlay the expression level of a particular gene
onto the data points as a colour gradient.

For this purpose, let us first set a common colour scale that ranges from
the lowest to the highest expression values observed in the data set:

```{r exprScale}
exprScale <- range(assay(sce.norm, "logcounts"))
colourScale <- rev(brewer.pal(9, "YlGnBu"))
```

#### TNF

```{r plotTSNE_TNF, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="TNF", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="TNF", limits=exprScale, colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### IFNB1

```{r plotTSNE_IFNB1, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="IFNB1", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="IFNB1",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### IL12B

```{r plotTSNE_IL12B, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="IL12B", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="IL12B",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### IL36G

```{r plotTSNE_IL36G, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="IL36G", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="IL36G",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### IL1B

```{r plotTSNE_IL1B, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="IL1B", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="IL1B",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD1A

```{r plotTSNE_CD1A, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="CD1A", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="CD1A",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD1B

```{r plotTSNE_CD1B, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="CD1B", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="CD1B",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD1C

```{r plotTSNE_CD1C, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="CD1C", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="CD1C",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### RAMP1

```{r plotTSNE_RAMP1, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="RAMP1", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="RAMP1",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD40

```{r plotTSNE_CD40, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="CD40", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="CD40",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD80

```{r plotTSNE_CD80, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="CD80", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="CD80",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD83

```{r plotTSNE_CD83, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="CD83", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="CD83",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### ALOX15

```{r plotTSNE_ALOX15, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="ALOX15", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="ALOX15",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD209

```{r plotTSNE_CD209, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="CD209", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="CD209",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CTSB

```{r plotTSNE_CTSB, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="CTSB", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="CTSB",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD86

```{r tSNE_CD86, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="CD86", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="CD86",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

```{r tSNE_CD86_pdf, include=FALSE}
ggsave(
  file.path(outdir, "20170525_lab-meeting_tSNE_CD86.pdf"),
  width = 5, height = 5)
```

#### MARCH1

```{r tSNE_MARCH1, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="MARCH1", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="MARCH1",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

```{r tSNE_MARCH1_pdf, include=FALSE}
ggsave(
  file.path(outdir, "20170525_lab-meeting_tSNE_MARCH1.pdf"),
  width = 5, height = 5)
```

#### IL10

```{r tSNE_IL10, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim(
  sce.norm, use_dimred = "TSNE", ncomponents = 2, shape_by = "Time",
  colour_by = subset(rowData(sce.norm), gene_name=="IL10", "gene_id", drop=TRUE)) +
  PCAtheme +
  scale_colour_gradientn(name="IL10",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1), shape = guide_legend(order = 2))
```

```{r tSNE_IL10_pdf, include=FALSE}
ggsave(
  file.path(outdir, "20170525_lab-meeting_tSNE_IL10.pdf"), width = 5, height = 5)
```

## Diffusion map

Finally, let us also produce a diffusion map plot of two components as a
dimensionality reduction of the endogenous features:

```{r plotDiffusionMapSCESet}
plotDiffusionMap(
  sce.norm,
  colour_by = "Status", shape_by = "Time",
  feature_set = !isSpike(sce.norm, "ERCC"),
  rand_seed = 1794) + PCAtheme
```

# Pairwise distance beween samples in tSNE

The high-dimensional nature of gene expression data sets presents significant
challenges for many analytical and visualisation methods.
In particular, the set of low-dimensional coordinates produced by the tSNE
technique can greatly facilitate the calculation of distance matrices, and
the identification of robust clusters of cells in the data set.

## Subset time points {.tabset}

Considering the overwhelming effect of the `Time` factor, let us
examine pairwise distance and hierarchical clustering of cells within each
time-point:

### 2h

```{r sce.2h}
sce.2h <- sce.norm[,sce.norm$Time == '2h']
dim(sce.2h)
```

### 4h

```{r sce.4h}
sce.4h <- sce.norm[,sce.norm$Time == '4h']
dim(sce.4h)
```

### 6h

```{r sce.6h}
sce.6h <- sce.norm[,sce.norm$Time == '6h']
dim(sce.6h)
```

## Estimate pairwise Euclidian distance in tSNE {.tabset}

### 2h

```{r d.e_2h}
d.e.2h <- dist(reducedDim(sce.2h, "TSNE"), diag = TRUE, upper = TRUE)
mat.e.2h <- as.matrix(d.e.2h)
h.e.2h <- hclust(d.e.2h)
ord.mat.2h <- mat.e.2h[h.e.2h$order, h.e.2h$order]
rm(d.e.2h)
```

### 4h

```{r d.e_4h}
d.e.4h <- dist(reducedDim(sce.4h, "TSNE"), diag = TRUE, upper = TRUE)
mat.e.4h <- as.matrix(d.e.4h)
h.e.4h <- hclust(d.e.4h)
ord.mat.4h <- mat.e.4h[h.e.4h$order, h.e.4h$order]
rm(d.e.4h)
```

### 6h

```{r d.e_6h}
d.e.6h <- dist(reducedDim(sce.6h, "TSNE"), diag = TRUE, upper = TRUE)
mat.e.6h <- as.matrix(d.e.6h)
h.e.6h <- hclust(d.e.6h)
ord.mat.6h <- mat.e.6h[h.e.6h$order, h.e.6h$order]
rm(d.e.6h)
```

## Prepare heatmaps {.tabset}

Let us first define a colour gradient common to all three time-points:

```{r distRange}
distRange <- range(c(mat.e.2h, mat.e.4h, mat.e.6h))
colorMap.eucl <- colorRamp2(
  c(min(distRange), median(distRange), max(distRange)),
  c("blue","white","red"))
```

### 2h

```{r ht_column_2h}
ht_column <- HeatmapAnnotation(
  df = data.frame(colData(sce.2h)[,c("Time", "Infection", "Status")]),
  col = list(Time = col.time, Status = col.status, Infection = col.infection)
)
ht.2h <- Heatmap(
  mat.e.2h, colorMap.eucl, name = "2h", column_title = "2h",
  top_annotation = ht_column,
  row_dend_reorder=rev(nrow(mat.e.2h):1),
  column_dend_reorder=rev(ncol(mat.e.2h):1),
  show_row_dend = FALSE, show_row_names = FALSE, show_column_names = FALSE
)
```

### 4h

```{r ht_column_4h}
ht_column <- HeatmapAnnotation(
  df = data.frame(colData(sce.2h)[,c("Time", "Infection", "Status")]),
  col = list(Time = col.time, Status = col.status, Infection = col.infection)
)
ht.4h <- Heatmap(
  mat.e.4h, colorMap.eucl, name = "4h", column_title = "4h",
  top_annotation = ht_column,
  row_dend_reorder=rev(nrow(mat.e.4h):1),
  column_dend_reorder=rev(ncol(mat.e.4h):1),
  show_row_dend = FALSE, show_row_names = FALSE, show_column_names = FALSE
)
```

### 6h

```{r ht_column_6h}
ht_column <- HeatmapAnnotation(
  df = data.frame(colData(sce.6h)[,c("Time", "Infection", "Status")]),
  col = list(Time = col.time, Status = col.status, Infection = col.infection)
)
ht.6h <- Heatmap(
  mat.e.6h, colorMap.eucl, name = "6h", column_title = "6h",
  top_annotation = ht_column,
  row_dend_reorder=rev(nrow(mat.e.6h):1),
  column_dend_reorder=rev(ncol(mat.e.6h):1),
  show_row_dend = FALSE, show_row_names = FALSE, show_column_names = FALSE
)
```

## Heat maps {.tabset}

### 2h

```{r heatmap.2h, echo=FALSE}
draw(ht.2h)
```

```{r heatmap.2h_pdf, include=FALSE}
pdf(file.path(outdir, "heatmap_pairwise_2h.pdf"), height = 5, width = 5)
draw(ht.2h, show_heatmap_legend = FALSE)
dev.off()
```

### 4h

```{r heatmap.4h, echo=FALSE}
draw(ht.4h)
```

```{r heatmap.4h_pdf, include=FALSE}
pdf(file.path(outdir, "heatmap_pairwise_4h.pdf"), height = 5, width = 5)
draw(ht.4h, show_heatmap_legend = FALSE)
dev.off()
```

### 6h

```{r heatmap.6h, echo=FALSE}
draw(ht.6h)
```

```{r heatmap.6h_pdf, include=FALSE}
pdf(file.path(outdir, "heatmap_pairwise_6h.pdf"), height = 5, width = 5)
draw(ht.6h, show_heatmap_legend = FALSE)
dev.off()
```

# Heat maps of selected marker genes

It is also possible to use the above hierarchical clustering of cells
(based on tSNE coordinates) to preserve the order of cells in heat maps
used to display the expression levels of selected genes, thereby
emphasising particular gene expression differences between clusters of cells
defined by their broader transcriptome.

Let us first define:

* relevant gene identifiers:

```{r geneIds.ht}
geneNames.ht <- list(
  "1. Trans. reprog." = c("DUSP1","DUSP2","SOCS3","IER3","NFKBIZ"),
  "2. Interm." = c("IL1B","IL1A","IFNB1","CCL7","TNF"),
  "3. Late" = c("IL6","IL10","IL12B","IL27")
)
geneIds.ht <- with(
  rowData(sce.norm),
  gene_id[match(unlist(geneNames.ht), gene_name)]
)
```

* a colour scale of expression levels common to the individual heat maps:

```{r exprsRange}
exprsRange <- range(assay(sce.norm, "logcounts"))
colorMap.exprs <- colorRamp2(
  c(min(exprsRange), median(exprsRange), max(exprsRange)),
  c("blue","white","red"))
```

## Prepare heat maps {.tabset}

### 2h

```{r heatmap_gene_2h}
ht_column <- HeatmapAnnotation(
  df = data.frame(colData(sce.2h)[,c("Time", "Infection", "Status")]),
  col = list(Time = col.time, Status = col.status, Infection = col.infection)
)
exprs.2h <- assay(sce.2h, "logcounts")[geneIds.ht,]
rownames(exprs.2h) <- unlist(geneNames.ht)
ht_gene_2h <- Heatmap(
  exprs.2h, colorMap.exprs, name = "2h", column_title = "2h",
  top_annotation = ht_column, row_names_side = "left",
  cluster_rows = FALSE, cluster_columns = h.e.2h,
  show_row_dend = FALSE, show_column_names = FALSE,
  split = rep(names(geneNames.ht), times = lengths(geneNames.ht))
)
```

### 4h

```{r heatmap_gene_4h}
ht_column <- HeatmapAnnotation(
  df = data.frame(colData(sce.4h)[,c("Time", "Infection", "Status")]),
  col = list(Time = col.time, Status = col.status, Infection = col.infection)
)
exprs.4h <- assay(sce.4h, "logcounts")[geneIds.ht,]
rownames(exprs.4h) <- unlist(geneNames.ht)
ht_gene_4h <- Heatmap(
  exprs.4h, colorMap.exprs, name = "4h", column_title = "4h",
  top_annotation = ht_column, row_names_side = "left",
  cluster_rows = FALSE, cluster_columns = h.e.4h,
  show_row_dend = FALSE, show_column_names = FALSE,
  split = rep(names(geneNames.ht), times = lengths(geneNames.ht))
)
```

### 6h

```{r heatmap_gene_6h}
ht_column <- HeatmapAnnotation(
  df = data.frame(colData(sce.6h)[,c("Time", "Infection", "Status")]),
  col = list(Time = col.time, Status = col.status, Infection = col.infection)
)
exprs.6h <- assay(sce.6h, "logcounts")[geneIds.ht,]
rownames(exprs.6h) <- unlist(geneNames.ht)
ht_gene_6h <- Heatmap(
  exprs.6h, colorMap.exprs, name = "6h", column_title = "6h",
  top_annotation = ht_column, row_names_side = "left",
  cluster_rows = FALSE, cluster_columns = h.e.6h,
  show_row_dend = FALSE, show_column_names = FALSE,
  split = rep(names(geneNames.ht), times = lengths(geneNames.ht))
)
```

## Heat maps {.tabset}

### all

```{r ht_gene_list, echo=FALSE}
ht_gene_list <-
  Heatmap(
    exprs.2h, colorMap.exprs, name = "n.e.", column_title = "2h",
    top_annotation = HeatmapAnnotation(
      df = data.frame(colData(sce.2h)[,c("Time", "Status", "Infection")]),
        col = list(
          Time = col.time,
          Status = col.status,
          Infection = col.infection
        ),
      show_legend = TRUE # FALSE
    ),
    row_names_side = "left", row_names_gp = gpar(fontsize = 8),
    row_title_gp=gpar(fontsize=10), cluster_rows=FALSE,
    cluster_columns=h.e.2h, show_column_names=FALSE,
    show_row_dend = FALSE, show_row_names = TRUE, show_heatmap_legend=FALSE,
    split = rep(names(geneNames.ht), times = lengths(geneNames.ht))
  ) +
  Heatmap(
    exprs.4h, colorMap.exprs, name = "4h", column_title = "4h",
    top_annotation = HeatmapAnnotation(
      df = data.frame(colData(sce.4h)[,c("Time", "Status", "Infection")]),
        col = list(
          Time = col.time,
          Status = col.status,
          Infection = col.infection
        ),
      show_legend = FALSE
    ),
    cluster_rows=FALSE, cluster_columns=h.e.4h, show_column_names = FALSE,
    show_row_dend = FALSE, show_row_names = FALSE, show_heatmap_legend = FALSE
  ) +
  Heatmap(
    exprs.6h, colorMap.exprs, name = "6h", column_title = "6h",
    top_annotation = HeatmapAnnotation(
      df = data.frame(colData(sce.6h)[,c("Time", "Status", "Infection")]),
        col = list(
          Time = col.time,
          Status = col.status,
          Infection = col.infection
        ),
      show_legend = FALSE
    ),
    cluster_rows=FALSE, cluster_columns=h.e.6h, show_column_names = FALSE,
    show_row_dend = FALSE, show_row_names = FALSE, show_heatmap_legend = FALSE
  )
draw(ht_gene_list)
```

```{r ht_gene_list_pdf, include=FALSE}
pdf(file.path(outdir, "heatmap_genes_tSNE.pdf"), height = 4, width = 8)
draw(ht_gene_list)
dev.off()
```

```{r ht_gene_list_noLegend_pdf, include=FALSE}
pdf(file.path(outdir, "heatmap_genes_tSNE_noLegend.pdf"), height = 4, width = 8)
draw(ht_gene_list, show_annotation_legend = FALSE)
dev.off()
```


### 2h

```{r ht_gene_2h, echo=FALSE}
draw(ht_gene_2h)
```

### 4h

```{r ht_gene_4h, echo=FALSE}
draw(ht_gene_4h)
```

### 6h

```{r ht_gene_6h, echo=FALSE}
draw(ht_gene_6h)
```

