---
title: "Gene expression levels overlaid onto tSNE view"
editor_options: 
  chunk_output_type: console
---

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
stopifnot(
  require(scater),
  require(scran),
  require(ggplot2),
  require(SummarizedExperiment),
  require(RColorBrewer),
  require(circlize),
  require(ComplexHeatmap)
)
sce.norm <- readRDS("rds/sce.norm.rds")
sce.endo <- sce.norm[!isSpike(sce.norm, "ERCC"),]
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,4,2)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.endo$Time)[1:3]
names(col.infection) <- levels(sce.endo$Infection)
names(col.status) <- levels(sce.endo$Status)
outdir <- "13_violins_out"; dir.create(outdir, showWarnings = FALSE)
pdfWidth <- 7; pdfHeight <- 4
cathepsinWidth <- 6; cathepsinHeight <- 3
pdfName <- "%s_%s.pdf"
sce.endo$TreatmentLabel <- with(colData(sce.endo), factor(
  gsub("_", "\n", Treatment),
  gsub("_", "\n", levels(Treatment))
))
```

# Helper values and functions

Let us first define some values useful to design colour scales:

```{r colours}
range.exprs <- range(assay(sce.endo, "logcounts"))
col.treatment <- RColorBrewer::brewer.pal(9, "Paired")[c(9,3,4,1,2)]
names(col.treatment) <- gsub("_","\n",levels(sce.endo$TreatmentLabel))
```

Let us then define a few convenience functions:

* to fetch the gene ID associated with a gene name

```{r geneNameToID}
geneNameToID <- function(x){
  geneId <- subset(rowData(sce.endo), gene_name == x, "gene_id", drop = TRUE)
  if (length(geneId) > 1){
    warning("Multiple IDs found. Use ID instead")
    return(data.frame(
      gene_name = x,
      gene_id = geneId
    ))
  } else if (length(geneId) == 0){
    stop("gene_name not found")
  }
  return(geneId)
}
```

* to fetch the necessary expression and phenotype data

```{r geneDataByName}
geneDataByName <- function(x){
  geneId <- geneNameToID(x)
  ggData <- data.frame(
    gene = assay(sce.endo, "logcounts")[geneId,],
    colData(sce.endo)[,c("Time","Infection","Status","TreatmentLabel")]
  )
  return(ggData)
}
```

* to draw the figure, faceted by time

```{r plotByNameFacetTime}
plotByNameFacetTime <- function(x){
  geneId <- geneNameToID(x)
  ggData <- geneDataByName(x)
  # return(ggData)
  ggplot(ggData, aes(TreatmentLabel, gene)) +
    facet_grid(Time~.) +
    geom_violin(
      aes(colour = TreatmentLabel, fill = TreatmentLabel),
      alpha = 0.2) +
    geom_jitter(
      aes(colour = TreatmentLabel), height = 0, width = 0.2,
      alpha = 0.5, size = 0.4) +
    scale_color_manual(values = col.treatment) +
    scale_fill_manual(values = col.treatment) +
    labs(
      title = sprintf("%s", x),
      subtitle = sprintf("%s", geneId),
      x = "Treatment",
      y = "log-counts",
      colour = "Treatment",
      fill = "Treatment") +
    scale_y_continuous(limits = range.exprs) +
    theme_bw() +
    theme(
      legend.key.height = unit(1.75, "lines")
    )
}
```

* to draw the figure, for a single time point

```{r plotByName}
plotByNameTime <- function(x, time){
  geneId <- geneNameToID(x)
  ggData <- geneDataByName(x)
  ggData <- subset(ggData, Time == time)
  # return(ggData)
  ggplot(ggData, aes(TreatmentLabel, gene)) +
    geom_violin(
      aes(colour = TreatmentLabel, fill = TreatmentLabel), scale = "width",
      alpha = 0.4) +
    geom_jitter(
      aes(colour = TreatmentLabel), height = 0, width = 0.2,
      alpha = 0.5, size = 0.4) +
    scale_color_manual(values = col.treatment) +
    scale_fill_manual(values = col.treatment) +
    labs(
      title = sprintf("%s (%s)", x, time),
      subtitle = sprintf("%s", geneId),
      x = "Treatment",
      y = "log-counts",
      colour = "Treatment",
      fill = "Treatment") +
    scale_y_continuous(limits = range.exprs) +
    theme_bw() +
    theme(
      legend.key.height = unit(1.75, "lines"),
      axis.text.y = element_text(size = rel(1.5))
    )
}
```

# Genes (alphabetical order) {.tabset}

The above convenience function make it straightforward to produce a figure
for any gene of interest:

## CCL24

```{r CCL24, echo=FALSE, fig.height=7}
geneName <- "CCL24"
plotByNameFacetTime(geneName)
```

```{r CCL24_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "2h")),
  plotByNameTime(geneName, "2h"),
  height = pdfHeight, width = pdfWidth)
```

## CYBB

```{r CYBB, echo=FALSE, fig.height=7}
geneName <- "CYBB"
plotByNameFacetTime(geneName)
```

```{r CYBB_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "2h")),
  plotByNameTime(geneName, "2h"),
  height = pdfHeight, width = pdfWidth)
```

## EBI3

```{r EBI3, echo=FALSE, fig.height=7}
geneName <- "EBI3"
plotByNameFacetTime(geneName)
```

```{r EBI3_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "all")),
  plotByNameTime(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
```

## IL12B

```{r IL12B, echo=FALSE, fig.height=7}
geneName <- "IL12B"
plotByNameFacetTime(geneName)
```

```{r IL12B_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "all")),
  height = 10, width = pdfWidth)
```

## IL1B

```{r IL1B, echo=FALSE, fig.height=7}
geneName <- "IL1B"
plotByNameFacetTime(geneName)
```

```{r IL1B_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "all")),
  height = 10, width = pdfWidth)
```

## CTSD

```{r CTSD, echo=FALSE, fig.height=7}
geneName <- "CTSD"
plotByNameFacetTime(geneName)
```

```{r CTSD_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameTime(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameTime(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## CTSL

```{r CTSL, echo=FALSE, fig.height=7}
geneName <- "CTSL"
plotByNameFacetTime(geneName)
```

```{r CTSL_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameTime(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameTime(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## CTSS

```{r CTSS, echo=FALSE, fig.height=7}
geneName <- "CTSS"
plotByNameFacetTime(geneName)
```

```{r CTSS_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameTime(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameTime(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## LGMN

```{r LGMN, echo=FALSE, fig.height=7}
geneName <- "LGMN"
plotByNameFacetTime(geneName)
```

```{r LGMN_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "2h")),
  plotByNameTime(geneName, "2h"),
  height = pdfHeight, width = pdfWidth)
```

## MS4A4A

```{r MS4A4A, echo=FALSE, fig.height=7}
geneName <- "MS4A4A"
plotByNameFacetTime(geneName)
```

```{r MS4A4A_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameTime(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
```

## PLAT

```{r PLAT, echo=FALSE, fig.height=7}
geneName <- "PLAT"
plotByNameFacetTime(geneName)
```

```{r PLAT_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameTime(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
```

## MARCH1

```{r MARCH1, echo=FALSE, fig.height=7}
geneName <- "MARCH1"
plotByNameFacetTime(geneName)
```

```{r MARCH1_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameTime(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

