---
title: "qPCR validation"
author: "Kevin Rue-Albrecht"
date: "21/11/2017"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
stopifnot(
  require(readxl),
  require(SummarizedExperiment),
  require(ggplot2),
  require(ggrepel)
)
sce.endo <- readRDS("rds/sce.endo_clusters.rds")
scde_groups <- readRDS("rds/scde_group.rds")
addGeneName <- function(x){
  x <- cbind(
    gene_name = with(rowData(sce.endo), gene_name[match(rownames(x), gene_id)]),
    x
  )
  return(x)
}
dirOut <- "_002_out"; dir.create(dirOut)
```

```{r}
qPCR_uninfected <- data.frame(read_excel("../expdata/qPCR_FC.xlsx", sheet = "uninfected"))
qPCR_direct <- data.frame(read_excel("../expdata/qPCR_FC.xlsx", sheet = "direct"))
```

# Uninfected

```{r}
replacements <- data.frame(
  old = c(
    'D23580 exposed',
    'D23580 infected',
    'LT2 exposed',
    'LT2 infected',
    'D23 Violet Vs Exposed',
    'LT2 Violet Vs Exposed',
    'D23 Violet Vs LT2 violet'
  ),
  new = c(
    "%s_STM-D23580_Exposed-%s_Mock_Uninfected",
    '%s_STM-D23580_Violet +-%s_Mock_Uninfected',
    '%s_STM-LT2_Exposed-%s_Mock_Uninfected',
    '%s_STM-LT2_Violet +-%s_Mock_Uninfected',
    '%s_STM-D23580_Violet +-%s_STM-D23580_Exposed',
    '%s_STM-LT2_Violet +-%s_STM-LT2_Exposed',
    '%s_STM-D23580_Violet +-%s_STM-LT2 violet'
  )
)
qPCR_uninfected$Contrast <- with(
  replacements,
  new[match(qPCR_uninfected$contrast, old)]
)
qPCR_uninfected$Contrast <- with(
  qPCR_uninfected,
  sprintf(as.character(Contrast), time, time)
)
stopifnot(
  all(qPCR_uninfected$Contrast %in% names(scde_groups))
)
```

First two columns of the final table: `gene` and `Contrast`

```{r}
finalTable <- unique(qPCR_uninfected[,c("gene", "Contrast", "time")])
```

Mean log2FC of qPCR for each contrast:

```{r}
addTable <- aggregate(logFC_qPCR ~ gene + Contrast + time, qPCR_uninfected, "mean")
colnames(addTable) <- gsub("^logFC_qPCR$", "meanLog2FC", colnames(addTable))
finalTable <- merge(finalTable, addTable)
```

standard deviation of log2FC in qPCR for each contrast:

```{r}
addTable <- aggregate(logFC_qPCR ~ gene + Contrast + time, qPCR_uninfected, "sd")
colnames(addTable) <- gsub("^logFC_qPCR$", "sdLog2FC", colnames(addTable))
finalTable <- merge(finalTable, addTable)
```

number of replicates for qPCR for each contrast:

```{r}
addTable <- aggregate(logFC_qPCR ~ gene + Contrast + time, qPCR_uninfected, "length")
colnames(addTable) <- gsub("^logFC_qPCR$", "replicates", colnames(addTable))
finalTable <- merge(finalTable, addTable)
```

Compute the SEM

```{r}
finalTable <- dplyr::mutate(finalTable, SEMlog2FC = sdLog2FC / sqrt(replicates))
```

Add the mle, lb and ub from the corresponding scde computations:

```{r}
finalTable[,c("mle","lb","ub")] <- NA_real_
for (iRow in seq_len(nrow(finalTable))){
  tmp_Contrast <- finalTable[iRow, "Contrast"]
  tmp_gene <- finalTable[iRow, "gene"]
  tmp_scde <- addGeneName(scde_groups[[tmp_Contrast]])
  tmp_scde <- subset(tmp_scde, gene_name == tmp_gene, c("lb","mle","ub"))
  stopifnot(dim(tmp_scde) == c(1,3))
  finalTable[iRow, c("lb","mle","ub")] <- tmp_scde
}
stopifnot(sum(is.na(finalTable)) == 0)
```

Contrast without time

```{r}
finalTable$ContrastNoTime <- gsub(
  "_", " ",
  gsub(
    "-Mock", " vs. \nMock",
    gsub("[[:digit:]]h_", "", finalTable$Contrast)
  )
)
```


```{r}
rangeFC <- max(abs(with(finalTable, range(c(meanLog2FC, mle)))))
ggplot(
  subset(finalTable, gene != "NIT1"),
  aes(mle, meanLog2FC)
  ) +
  facet_grid(time ~ ContrastNoTime) +
  # geom_smooth(method = "glm", alpha = 0.1, size = 0.2) +
  geom_point(colour = "blue") +
  geom_abline(slope = 1, intercept = 0, colour = "red", linetype = "dashed") +
  geom_hline(yintercept = 0, alpha = 0.25, size = 2) +
  geom_vline(xintercept = 0, alpha = 0.25, size = 2) +
  labs(
    x = "mle of log2 fold-change (single-cell; scde)",
    y = "mean log2 fold-change (qPCR)"
  ) +
  geom_text_repel(
    aes(label = gene),
    min.segment.length = unit(0, "lines"), size = 3) +
  coord_fixed() +
  scale_x_continuous(breaks = seq(-10, 10, 5), limits = c(-rangeFC, rangeFC)) +
  scale_y_continuous(breaks = seq(-10, 10, 5), limits = c(-rangeFC, rangeFC)) +
  theme_bw()
ggsave(file.path(dirOut, "uninfected_plot.pdf"), width = 10, height = 5)
```

```{r}
with(
  subset(finalTable, gene != "NIT1"),
  cor.test(mle, meanLog2FC)
)
```

# Direct

```{r}
replacements <- data.frame(
  old = c(
    "D23 Violet Vs Exposed",
    "LT2 Violet Vs Exposed",
    "D23 Violet Vs LT2 violet"
  ),
  new = c(
    "%s_STM-D23580_Violet +-%s_STM-D23580_Exposed",
    "%s_STM-LT2_Violet +-%s_STM-LT2_Exposed",
    "%s_STM-D23580_Violet +-%s_STM-LT2_Violet +"
  )
)
qPCR_direct$Contrast <- with(
  replacements,
  new[match(qPCR_direct$contrast, old)]
)
qPCR_direct$Contrast <- with(
  qPCR_direct,
  sprintf(as.character(Contrast), time, time)
)
stopifnot(
  all(qPCR_direct$Contrast %in% names(scde_groups))
)
```

First two columns of the final table: `gene` and `Contrast`

```{r}
finalTable <- unique(qPCR_direct[,c("gene", "Contrast", "time")])
```

Mean log2FC of qPCR for each contrast:

```{r}
addTable <- aggregate(logFC_qPCR ~ gene + Contrast + time, qPCR_direct, "mean")
colnames(addTable) <- gsub("^logFC_qPCR$", "meanLog2FC", colnames(addTable))
finalTable <- merge(finalTable, addTable)
```

standard deviation of log2FC in qPCR for each contrast:

```{r}
addTable <- aggregate(logFC_qPCR ~ gene + Contrast + time, qPCR_direct, "sd")
colnames(addTable) <- gsub("^logFC_qPCR$", "sdLog2FC", colnames(addTable))
finalTable <- merge(finalTable, addTable)
```

number of replicates for qPCR for each contrast:

```{r}
addTable <- aggregate(logFC_qPCR ~ gene + Contrast + time, qPCR_direct, "length")
colnames(addTable) <- gsub("^logFC_qPCR$", "replicates", colnames(addTable))
finalTable <- merge(finalTable, addTable)
```

Compute the SEM

```{r}
finalTable <- dplyr::mutate(finalTable, SEMlog2FC = sdLog2FC / sqrt(replicates))
```

Add the mle, lb and ub from the corresponding scde computations:

```{r}
finalTable[,c("mle","lb","ub")] <- NA_real_
for (iRow in seq_len(nrow(finalTable))){
  tmp_Contrast <- finalTable[iRow, "Contrast"]
  tmp_gene <- finalTable[iRow, "gene"]
  tmp_scde <- addGeneName(scde_groups[[tmp_Contrast]])
  tmp_scde <- subset(tmp_scde, gene_name == tmp_gene, c("lb","mle","ub"))
  stopifnot(dim(tmp_scde) == c(1,3))
  finalTable[iRow, c("lb","mle","ub")] <- tmp_scde
}
stopifnot(sum(is.na(finalTable)) == 0)
```

Contrast without time

```{r}
finalTable$ContrastNoTime <- gsub(
  "_", " ",
  gsub(
    "-STM", " vs. \nSTM",
    gsub("[[:digit:]]h_", "", finalTable$Contrast)
  )
)
```


```{r}
rangeFC <- max(abs(with(finalTable, range(c(meanLog2FC, mle)))))
ggplot(
  subset(finalTable, gene != "NIT1"),
  aes(mle, meanLog2FC)
  ) +
  facet_grid(time ~ ContrastNoTime) +
  # geom_smooth(method = "glm", alpha = 0.1, size = 0.2) +
  geom_point(colour = "blue") +
  geom_abline(slope = 1, intercept = 0, colour = "red", linetype = "dashed") +
  geom_hline(yintercept = 0, alpha = 0.25, size = 2) +
  geom_vline(xintercept = 0, alpha = 0.25, size = 2) +
  labs(
    x = "mle of log2 fold-change (single-cell; scde)",
    y = "mean log2 fold-change (qPCR)"
  ) +
  geom_text_repel(
    aes(label = gene),
    min.segment.length = unit(0, "lines"), size = 3) +
  coord_fixed() +
  scale_x_continuous(breaks = seq(-10, 10, 5), limits = c(-rangeFC, rangeFC)) +
  scale_y_continuous(breaks = seq(-10, 10, 5), limits = c(-rangeFC, rangeFC)) +
  theme_bw()
ggsave(file.path(dirOut, "direct_plot.pdf"), width = 10, height = 5)
```

```{r}
with(
  subset(finalTable, gene != "NIT1"),
  cor.test(mle, meanLog2FC)
)
```

