---
title: "_004_HVG_heatmap_kmeans2"
author: "Kevin Rue-Albrecht"
date: "17/12/2017"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
stopifnot(
  require(ComplexHeatmap),
  require(RColorBrewer),
  requireNamespace("circlize")
)
sce.norm <- readRDS("rds/sce.norm.tSNE.rds")
# Table of HVGs
hvg.out <- read.csv("05_HVG_out/hvg.out.csv", row.names = 1)
# Prepare colours
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,2,4)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.norm$Time)
names(col.infection) <- levels(sce.norm$Infection)
names(col.status) <- levels(sce.norm$Status)
dirOut <- "_004_out"; dir.create(dirOut)
```

# Set heat map color scale

```{r}
exprsRange <- range(assay(sce.norm, "logcounts"))
colorMap <- circlize::colorRamp2(
  c(min(exprsRange), median(exprsRange)*0.9, max(exprsRange)*0.9),
  c("black","purple","yellow")
)
```

# Order by decreasing biological variance

```{r}
hvg.out <- hvg.out[order(hvg.out$bio, decreasing = TRUE),]
sce.hvg <- sce.norm[rownames(hvg.out),]
sce.hvg
```

# Ensure there are no more spike ins

```{r}
sce.endo <- sce.hvg[!isSpike(sce.hvg),]
sce.endo
```

# Take top N features (by decreasing biological variance)

```{r}
sce.top <- sce.endo[1:500,]
sce.top
```

# Make heat map

## Cluster HVGs across all cells

```{r}
hvg.endo.d <- dist(assay(sce.top, "logcounts"))
hvg.endo.clust <- hclust(hvg.endo.d)
rm(hvg.endo.d)
```

## Cluster samples within each experimental group

```{r}
sample.order <- c()
expGroups <- unique(colData(sce.top)[,c("Time","Infection","Status")])
expGroups <- dplyr::arrange(data.frame(expGroups), Time, Status, Infection)
for (groupIndex in seq_len(nrow(expGroups))){
  time <- expGroups$Time[groupIndex]
  infection <- expGroups$Infection[groupIndex]
  status <- expGroups$Status[groupIndex]
  sample.index <- with(colData(sce.top), which(
    Time == time & Infection == infection & Status == status
  ))
  hvg.sample.d <- dist(t(assay(sce.top[,sample.index], "logcounts")))
  local.order <- sample.index[hclust(hvg.sample.d)$order]
  sample.order <- c(sample.order, local.order)
  rm(sample.index, local.order, time, infection, status)
}
rm(expGroups, groupIndex)
sce.reorder <- sce.top[,sample.order]
```

## Split samples by time point

```{r}
sce.2h <- sce.reorder[,sce.reorder$Time == "2h"]
sce.4h <- sce.reorder[,sce.reorder$Time == "4h"]
sce.6h <- sce.reorder[,sce.reorder$Time == "6h"]
```

## Draw heat map annotations

```{r}
ha_2h <- HeatmapAnnotation(
  df = colData(sce.2h)[,c("Time", "Status", "Infection")],
  col = list(
    Time = col.time,
    Status = col.status,
    Infection = col.infection
  )
)
ha_4h <- HeatmapAnnotation(
  df = colData(sce.4h)[,c("Time", "Status", "Infection")],
  col = list(
    Time = col.time,
    Status = col.status,
    Infection = col.infection
  ), show_legend = FALSE
)
ha_6h <- HeatmapAnnotation(
  df = colData(sce.6h)[,c("Time", "Status", "Infection")],
  col = list(
    Time = col.time,
    Status = col.status,
    Infection = col.infection
  ), show_legend = FALSE
)
```

## Draw main heat maps

```{r}
ht_2h <- Heatmap(
  assay(sce.2h, "logcounts"), col = colorMap,
  name = "logcounts", column_title = "2h",
  top_annotation = ha_2h,
  row_order = hvg.endo.clust$order, column_order = NULL,
  cluster_rows = hvg.endo.clust, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE, width = unit(4, "inches")
)
ht_4h <- Heatmap(
  assay(sce.4h, "logcounts"), col = colorMap,
  name = "logcounts.4h", column_title = "4h",
  top_annotation = ha_4h,
  row_order = hvg.endo.clust$order, column_order = NULL,
  cluster_rows = hvg.endo.clust, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE,
  show_heatmap_legend = FALSE, width = unit(4, "inches")
)
ht_6h <- Heatmap(
  assay(sce.6h, "logcounts"), col = colorMap,
  name = "logcounts.6h", column_title = "6h",
  top_annotation = ha_6h,
  row_order = hvg.endo.clust$order, column_order = NULL,
  cluster_rows = hvg.endo.clust, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE,
  show_heatmap_legend = FALSE , width = unit(4, "inches")
)
```

```{r}
pdf(file.path(dirOut, "heatmap_panel_time.pdf"), height = 6, width = 12)
draw(ht_2h + ht_4h + ht_6h)
dev.off()
```

# Augment heat map with HVG grouping

## Cut the HVG tree in blocks {.tabset}

### Set height (100)

Show what an arbitrary cut at height 100 looks like:

```{r}
pdf(file.path(dirOut, "dendrogram_genes_height.pdf"), height = 6, width = 10)
plot(hvg.endo.clust, labels = FALSE)
abline(h = 100, col = "red")
dev.off()
```

```{r}
h_100 <- cutree(hvg.endo.clust, h = 100)
table(h_100) # 33 clusters at height 100
```

`r length(unique(h_100))` clusters at height 100.

### Set number of clusters (10)

Cut tree in 10 clusters:

```{r}
k_10 <- cutree(hvg.endo.clust, k = 10)
table(k_10)
```

Find a height that would correspond to a cut into 10 clusters:

```{r}
cut10.h <- mean(sort(hvg.endo.clust$height, decreasing = TRUE)[9:10])
```

```{r}
pdf(file.path(dirOut, "dendrogram_genes_number_10.pdf"), height = 6, width = 10)
plot(hvg.endo.clust, labels = FALSE)
abline(h = cut10.h, col = "red")
dev.off()
```

## Draw heat map of HVG grouping

```{r}
ht_tree <- Heatmap(
  sprintf("%2d", k_10), col = brewer.pal(10, "Paired"),
  name = "k", column_title = "k",
  row_order = hvg.endo.clust$order, column_order = NULL,
  cluster_rows = hvg.endo.clust, cluster_columns = FALSE, show_row_dend = FALSE,
  show_row_names = FALSE, show_column_names = FALSE,
  width = unit(1/2, "inches"),
  split = 10
)
```

```{r}
pdf(file.path(dirOut, "heatmap_panel_time_k_split.pdf"), height = 10, width = 15)
draw(
  object = ht_2h + ht_4h + ht_6h + ht_tree,
  main_heatmap = 4,
  gap = unit(1, "mm")
)
dev.off()
```

# Select representative genes by cluster

```{r, include=FALSE}
gene_k <- data.frame(
  rowData(sce.6h)[
    hvg.endo.clust$order,c("gene_id","gene_name")],
  k_10 = k_10[hvg.endo.clust$order]
)
write.csv(gene_k, file.path(dirOut, "gene_k_10.csv"), row.names = FALSE)
```

```{r}
anno.xlsx <- file.path("_004_in/gene_k_10.xlsx")
anno.data <- xlsx::read.xlsx(anno.xlsx, sheetName = "gene_k_10")
anno.subset <- subset(anno.data, Heatmap == "x", select=c("gene_name","gene_id"))
gene.idx = match(anno.subset$gene_id, rownames(sce.6h))
ra <- rowAnnotation(
  link = row_anno_link(
    at = gene.idx, labels = as.character(anno.subset$gene_name),
    labels_gp = gpar(cex=0.3), link_width = unit(0.5, "inches")
    ), width = max_text_width(labels)
)
```

```{r}
pdf(file.path(dirOut, "heatmap_panel_time_k_split_annoLink.pdf"), height = 10, width = 17)
draw(
  object = ht_2h + ht_4h + ht_6h + ht_tree + ra,
  main_heatmap = 4,
  gap = unit(1, "mm")
)
dev.off()
```

# Representative genes by cluster (no split)

```{r}
ht_tree_noSplit <- Heatmap(
  sprintf("%2d", k_10), col = brewer.pal(10, "Paired"),
  name = "k", column_title = "k",
  row_order = hvg.endo.clust$order, column_order = NULL,
  cluster_rows = hvg.endo.clust, cluster_columns = FALSE, show_row_dend = FALSE,
  show_row_names = FALSE, show_column_names = FALSE,
  width = unit(1/2, "inches")
)
```

```{r}
pdf(file.path(dirOut, "heatmap_panel_time_k_annoLink.pdf"), height = 10, width = 15)
draw(
  object = ht_2h + ht_4h + ht_6h + ht_tree_noSplit + ra,
  main_heatmap = 4,
  gap = unit(1, "mm")
)
dev.off()
```

