---
title: "Quality control of mini-bulk samples"
bibliography:
  bibtex.bib
editor_options: 
  chunk_output_type: console
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
stopifnot(
  require(SingleCellExperiment),
  require(ggplot2),
  require(RColorBrewer),
  require(scater),
  require(ggrepel),
  require(ComplexHeatmap)
)
sce.mini <- readRDS("rds/sce.minibulks.rds")
outdir <- "002_out"; dir.create(outdir, showWarnings = FALSE)
sce.mini$GroupLabel <- with(colData(sce.mini), factor(
  gsub("_", " ", Group),
  gsub("_", " ", levels(Group))
))
```

# Plotting themes

First of all, let us define:

* theme elements used throughout various figures in the following sections:

```{r sampleQCtheme}
sampleQCtheme <-
  theme_minimal() +
  theme(
    axis.ticks.x = element_blank(), axis.text.x = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical")
```

```{r groupQCtheme}
groupQCtheme <-
  theme_bw() +
  theme(
    axis.text.x = element_text(angle=90, hjust = 1, vjust = 0.5, size = rel(0.6)),
    legend.position = "bottom", legend.box = "vertical",
    plot.title = element_text(hjust = 0.5)
  )
```

* a set of colours used to indicate phenotype information consistently
  across the subsequent figures:

```{r colours}
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,4,2)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.mini$Time)[1:3]
names(col.infection) <- levels(sce.mini$Infection)
names(col.status) <- levels(sce.mini$Status)
```

* let us also initialise a `data.frame` that will be used to store outliers
  identified in each of the quality control filters defined below:

```{r outliers}
outliers <- data.frame(row.names = colnames(sce.mini))
```

# FastQC / MultiQC

## GC content (*FastQC*)

GC content is species-specific; it can be estimated from the annotated transcriptome or the sequenced data. Therefore, GC content excessively
different from the majority of samples can indicate issues during
library preparation or contamination.

Let us identify as outliers single cells that display a GC content
excessively *different* from the median of *single cell* samples,
here beyond **3** MADs:

```{r GC}
outliers$GC <- isOutlier(sce.mini$FastQC_percent_gc, 3)
table(outliers$GC)
```

Let us visualise outliers, along with blank and bulk samples for reference:

```{r plot.GC, echo=FALSE, fig.height=5}
GC.M <- median(sce.mini$FastQC_percent_gc)
GC.MAD <- mad(sce.mini$FastQC_percent_gc)
plot.GC <-
  ggplot(data.frame(colData(sce.mini)), aes(Sample, FastQC_percent_gc)) +
  # geom_rug(sides = "l", alpha=0.25) +
  geom_hline(
    yintercept = GC.M + 3 * c(1,-1) * GC.MAD, colour = "red", alpha = 0.5) +
  geom_hline(yintercept = GC.M, alpha=0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Time), data.frame(colData(sce.mini)[outliers$GC,]), alpha = 0.6,
    min.segment.length = unit(0, "lines")) +
  labs(y = "% GC (FastQC)", x = "Samples") +
  scale_y_continuous(limits = c(0,100)) +
  scale_colour_manual(values = col.infection) +
  sampleQCtheme
# plot.GC
ggsave(
  file.path(outdir, "GC_sample.pdf"),
  plot.GC + guides(colour = "none", shape = "none"), width = 6, height = 4
)
plot.GC <-
  ggplot(
    data.frame(colData(sce.mini)),
    aes(GroupLabel, FastQC_percent_gc)) +
  geom_violin(aes(colour = Infection), alpha = 0.25) +
  geom_hline(
    yintercept = GC.M + 3 * c(1,-1) * GC.MAD, colour = "red",alpha = 0.5) +
  geom_hline(yintercept = GC.M, alpha=0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$GC,]),
    alpha = 0.6, min.segment.length = unit(0, "lines"), size = rel(2)) +
  labs(title = "% GC (FastQC)", x = NULL, y = NULL) +
  scale_y_continuous(limits = c(0,100)) +
  scale_colour_manual(values = col.infection) +
  groupQCtheme
plot.GC
ggsave(
  file.path(outdir, "GC_group.pdf"),
  plot.GC + guides(colour = "none", shape = "none"), width = 6, height = 4
)
```

For reference:

* single cells
    - median: `r sprintf("%.0f%%", GC.M)`
* cut-offs
    - upper bound: `r sprintf("%.0f%%", GC.M+GC.MAD*3)`
    - lower bound: `r sprintf("%.0f%%", GC.M-GC.MAD*3)`

## Alignment rate (*HISAT2*)

Technical issues introduced during library preparation may result
in an excessively *low* alignment rate.
Nevertheless, libraries with slightly lower alignment rates may still yield
a library size comparable to the majority of samples.

For those reasons reasons,
let us identify as outliers single cells with an alignment rate
excessively *lower* than the median of *single cell* samples,
here beyond **3** MADs:

```{r AlignedPercent}
outliers$AlignedPercent <- isOutlier(sce.mini$Bowtie.2_overall_alignment_rate, 3, "lower")
table(outliers$AlignedPercent)
```

Let us visualise outliers, along with blank and bulk samples for reference:

```{r plot.aligned, echo=FALSE, fig.height=5}
align.M <- median(sce.mini$Bowtie.2_overall_alignment_rate)
align.MAD <- mad(sce.mini$Bowtie.2_overall_alignment_rate)
plot.aligned <-
  ggplot(
    data.frame(colData(sce.mini)),
    aes(Sample, Bowtie.2_overall_alignment_rate)) +
  geom_hline(yintercept = align.M, alpha = 0.5) +
  geom_hline(
    yintercept = align.M + 3*c(1,-1)*align.MAD,
    colour = "red", linetype = 2:1, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Time),
    data.frame(colData(sce.mini))[outliers$AlignedPercent,],
    alpha = 0.6, min.segment.length = unit(0, "lines")) +
  scale_y_continuous(limits = c(0,100)) +
  scale_colour_manual(values = col.infection) +
  labs(y = "Overall alignment rate (HISAT2)", x = "Samples") + sampleQCtheme
# plot.aligned
ggsave(
  file.path(outdir, "alignment_rate_sample.pdf"),
  plot.aligned + guides(colour = "none", shape = "none"), width = 6, height = 4
)
plot.aligned <-
  ggplot(
    data.frame(colData(sce.mini)),
    aes(GroupLabel, Bowtie.2_overall_alignment_rate)) +
  geom_violin(aes(colour = Infection), alpha = 0.25) +
  geom_hline(
    yintercept = align.M + 3 * c(1,-1) * align.MAD,
    colour = "red", linetype = 2:1, alpha = 0.5) +
  geom_hline(yintercept = align.M, alpha=0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$AlignedPercent,]),
    alpha = 0.6, min.segment.length = unit(0, "lines"), size = rel(2)) +
  labs(title = "Overall alignment rate (HISAT2)", y=NULL, x=NULL) +
  scale_y_continuous(limits = c(0,100)) +
  scale_colour_manual(values = col.infection) +
  groupQCtheme
plot.aligned
ggsave(
  file.path(outdir, "alignment_rate_group.pdf"),
  plot.aligned + guides(colour = "none", shape = "none"), width = 6, height = 4
)
```

For reference:

* single cells
    - median: `r sprintf("%.0f%%", align.M)`
* cut-off
    - lower bound: `r sprintf("%.0f%%", align.M-align.MAD*3)`

```{r posterAligned, include=FALSE}
posterAligned <-
  ggplot(data.frame(colData(sce.mini)), aes(Sample, Bowtie.2_overall_alignment_rate)) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_hline(
    yintercept = align.M + 3*c(1,-1)*align.MAD,
    colour ="red", linetype = 2:1) +
  geom_hline(yintercept = align.M) +
  geom_text_repel(
    aes(label = Time),
    data.frame(colData(sce.mini))[outliers$AlignedPercent,],
    alpha = 0.4, min.segment.length = unit(0, "lines")) +
  scale_y_continuous(limits = c(0,100)) +
  ylab("Overall alignment rate (HISAT2)") +
  scale_color_manual(values = col.infection) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
ggsave(
  file.path(outdir, "poster_alignedQC.pdf"), posterAligned,
  width = 9, height = 6)
```

## Assigned reads / Library size (*featureCounts*) {#assigned}

An excessively small count of aligned reads assigned to 
annotated genomics features (*i.e.*, library size) may indicate
technical issues during library preparation or low quality samples.
Moreover, library size excessively different from the
majority of samples may introduce
bias due to under- or over-sampling of RNA fragments, and negatively affect
the performance of between-sample normalisation algorithms.

Let us identify as outliers single cells
with a count of assigned read pairs (*i.e.* library size)
excessively *different* from the median of *single cell* samples,
here beyond **3** MADs:

```{r AssignedCount}
outliers$AssignedCount <- isOutlier(sce.mini$featureCounts_Assigned, 3, log = TRUE)
table(outliers$AssignedCount)
```

Let us visualise outliers, along with blank and bulk samples, for reference:

```{r plot.cAssigned, echo=FALSE, fig.height=5}
cAssigned.M <- median(log10(sce.mini$featureCounts_Assigned))
cAssigned.MAD <- mad(log10(sce.mini$featureCounts_Assigned))
plot.cAssigned <-
  ggplot(data.frame(colData(sce.mini)), aes(Sample, featureCounts_Assigned)) +
  geom_hline(
    yintercept = 10^(cAssigned.M + 3*c(1,-1)*cAssigned.MAD),
    colour = "red", alpha = 0.5) +
  geom_hline(yintercept = 10^(cAssigned.M), alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Time), data.frame(colData(sce.mini))[outliers$AssignedCount,],
    alpha = 0.6, min.segment.length = unit(0, "lines")
  ) +
  scale_color_manual(values = col.infection) +
  labs(y = "Count assigned (featureCounts)", x = "Samples") +
  sampleQCtheme
# plot.cAssigned
ggsave(
  file.path(outdir, "assigned_count_sample.pdf"),
  plot.cAssigned + guides(colour = "none", shape = "none"), width = 6, height = 4
)
plot.cAssigned <-
  ggplot(
    data.frame(colData(sce.mini)),
    aes(GroupLabel, featureCounts_Assigned)) +
  geom_violin(aes(colour = Infection), alpha = 0.25) +
  geom_hline(
    yintercept = 10^(cAssigned.M + 3 * c(1,-1) * cAssigned.MAD),
    colour = "red", alpha = 0.5) +
  geom_hline(yintercept = 10^cAssigned.M, alpha=0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$AssignedCount,]),
    alpha = 0.6, min.segment.length = unit(0, "lines"), size = rel(2)) +
  labs(title = "Count assigned (featureCounts)", y = NULL, x = NULL) +
  scale_colour_manual(values = col.infection) +
  groupQCtheme
plot.cAssigned
ggsave(
  file.path(outdir, "assigned_count_group.pdf"),
  plot.cAssigned + guides(colour = "none", shape = "none"), width = 6, height = 4
)
```

For reference:

* single cells
    - median: `r sprintf("%.1fe+06", 10^cAssigned.M/1E6)`
* cut-offs
    - lower bound:
      `r sprintf("%.1fe+06", 10^(cAssigned.M-cAssigned.MAD*3)/1E6)`
    - upper bound:
      `r sprintf("%.1fe+06", 10^(cAssigned.M+cAssigned.MAD*3)/1E6)`

## Percentage assigned (*featureCounts*)

Similarly, unusually *low* assignment rate of reads to annotated genomic
features may indicate technical issue; for instance during
poly-A selection.

Let us identify as outliers single cells
with a rate of assigned reads
excessively *lower* than the median of *single cell* samples,
here beyond **3** MADs:

```{r AssignedPercent}
outliers$AssignedPercent <- isOutlier(sce.mini$featureCounts_percent_assigned, 3)
table(outliers$AssignedPercent)
```

Let us visualise outliers, along with blank and bulk samples, for reference:

```{r plot.pAssigned, echo=FALSE, fig.height=5}
pAssigned.M <- median(sce.mini$featureCounts_percent_assigned)
pAssigned.MAD <- mad(sce.mini$featureCounts_percent_assigned)
plot.pAssigned <-
  ggplot(data.frame(colData(sce.mini)), aes(Sample, featureCounts_percent_assigned)) +
  geom_hline(
    yintercept = (pAssigned.M + 3*c(1,-1)*pAssigned.MAD), colour = "red",
    linetype = 2:1, alpha = 0.5) +
  geom_hline(yintercept = pAssigned.M, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Time), data.frame(colData(sce.mini))[outliers$AssignedPercent,],
    alpha = 0.6, min.segment.length = unit(0, "lines")
  ) +
  scale_y_continuous(limits = c(0,100)) +
  scale_colour_manual(values = col.infection) +
  labs(y = "% assigned (featureCounts)", x = "Samples") +
  sampleQCtheme
# plot.pAssigned
ggsave(
  file.path(outdir, "assigned_rate_sample.pdf"),
  plot.pAssigned + guides(colour = "none", shape = "none"), width = 6, height = 4
)
plot.pAssigned <-
  ggplot(
    data.frame(colData(sce.mini)),
    aes(GroupLabel, featureCounts_percent_assigned)) +
  geom_violin(aes(colour = Infection), alpha = 0.25) +
  geom_hline(
    yintercept = pAssigned.M + 3 * c(1,-1) * pAssigned.MAD,
    colour = "red", linetype = 2:1, alpha = 0.5) +
  geom_hline(yintercept = pAssigned.M, alpha=0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$AssignedPercent,]),
    alpha = 0.6, min.segment.length = unit(0, "lines"), size = rel(2)) +
  labs(title = "% assigned (featureCounts)", x=NULL, y=NULL) +
  scale_y_continuous(limits = c(0,100)) +
  scale_colour_manual(values = col.infection) +
  groupQCtheme
plot.pAssigned
ggsave(
  file.path(outdir, "alignment_rate_group.pdf"),
  plot.pAssigned + guides(colour = "none", shape = "none"), width = 6, height = 4
)
```

For reference:

* single cells
    - median: `r sprintf("%.0f%%", pAssigned.M)`
* cut-off
    - lower bound:
      `r sprintf("%.0f%%", pAssigned.M-pAssigned.MAD*3)`

## Summary

Let us display the count of cells identified as outliers for each of the
filters so far:

```{r fastqc.fr}
fastqc.fr <- FilterRules(colnames(outliers))
fastqc.fr
```

```{r fastqc.fr_table, results='asis', echo=FALSE}
pander::pandoc.table(
  data.frame(
    Fail = summary(evalSeparately(fastqc.fr, outliers)),
    Pass = summary(!evalSeparately(fastqc.fr, outliers))
))
```

**Note:**
In the above table, the first row simply indicates the total count
of cells before any filtering;
conversely, the last row indicates the count of cells that failed or
passed, respectively, _all_ filters.

# *scater*

## Calculation of QC metrics {#calculateQCMetrics}

The following sections require the estimation of several QC metrics that
leverage control features, as means to identify outliers among single cells
(there are no control samples available in this data set).

Let us apply the [scater](http://bioconductor.org/packages/scater)
`calculateQCMetrics` to calculate the relevant QC metrics in the
data set, using set of spike-in control features:

```{r sc_QC}
sce.mini <- calculateQCMetrics(
  sce.mini,
  feature_controls = list(
    ERCC = isSpike(sce.mini, "ERCC"),
    MT = as.logical(seqnames(sce.mini) == "MT")
  )
)
```

## Library size

The library size refers to the total count of reads assigned to 
annotated features in each individual sample.

Here, let us simply show the direct relation between
`log10_total_counts` computed by `calculateQCMetrics`
and the `featureCounts_Assigned` counts
detected by [MultiQC](http://multiqc.info) in
an earlier section [above](#assigned):

```{r scater_outliers_totalCounts}
with(
  data.frame(colData(sce.mini)),
  all(abs(log10_total_counts == log10(featureCounts_Assigned+1)))
)
```

As a result, this metric is identical to the *featureCounts_Assigned*
metric described [above](#assigned); and thereby would
unnecessarily identify the same outliers.

## Library complexity {#libraryComplexity}

*Library complexity* refers to the proportion of assigned reads accounted for
by the $N$ most abundant features within each individual sample, with values
for $N$ usually ranging from tens to hundreds.
It is expected to observe a lower library complexity in single cells relative
to bulk samples, due to higher dropout events in single cells, and the union of
multiple phenotypes within bulk samples.
In contrast, blank samples are expected to have extremely low library
complexity, as only up to **92** ERCC spike-ins sequences are expected in those
libraries, in contrast to thousands of genomic features in biological samples.

In the `SingleCellExperiment` object augmnented by QC metrics,
let us visualise the
library complexity profile against experimental phenotype information:

```{r all_libraryComplexity}
plotScater(
  sce.mini, block1 = "Status", block2 = "Infection",
  colour_by = "Time", nfeatures = 300, exprs_values = "counts",
  line_width = 0.5
)
```

First, from a QC perspective:

* Blank samples show a very low library complexity, as expected.
* Bulk samples show a generally high library complexity, as expected.
* Single cells generally show a complexity profile comparable to that of bulk
  samples, indicative of healthy dendritic cells and unbiased sampling of
  diverse transcripts.
* A small subset of single cells (*e.g.*, `exposed:D23580:6h`,`exposed:LT2:4h`, 
  `infected:D23580:4h`, `infected:D23580:6h`, `infected:LT2:6h`) show
  suspiciously low library complexity; suggestive of technical issues during
  library prepration (*e.g.*, amplification bias, dropout events).
  Those single cells must therefore be treated as outliers, and be removed from
  the data set prior to feature selection, normalisation, and subsequent
  analyses.

Now, from a biological perspective, interesting observations may be made:

* Library complexity seems to decrease with `Time` in `exposed` and `infected`
  samples in cells exposed to or infected by bacteria, in contrast to
  control uninfected cells.
  This may indicate a polarisation of single cells toward
  specialised phenotype(s) after infection or exposure
  (*e.g.*, cytokine production, antigen presentation).
  + `Mock` (*i.e.*, `uninfected`) single cells tend to show a consistent,
    relatively high library complexity
  + Both `LT2` and `D23580` infections show a gradual decrease of library
    complexity over time. This observation *might* be more pronounced
    for `D23580`.

## Most abundant features

Although abundantly expressed features are an expected aspect of cell biology
(*e.g.*, myosin in muscle cells, clusters of differentiation in immune cells)
which may contribute to the identification and classification of cell types,
overly abundant features in RNA-Seq libraries
may indicate issues during library preparation;
in particular, cDNA amplification bias.
Moreover, when a small number of highly abundant features
account for an excessive proportion of sequenced reads,
expression levels for the vast majority of other genes may not be
estimated with sufficient accuracy.

### Identification {.tabset}

Let us identify as outliers single cells
in which the `50`, `100`, and `200` most abundant features account for a
proportion of library size
excessively *higher* than the median of *single cell* samples,
here beyond **3** MADs:

#### Top 50

```{r scater_outliers_top50}
outliers$top50 <- isOutlier(sce.mini$pct_counts_top_50_features, 3)
table(outliers$top50)
```

#### Top 100

```{r scater_outliers_top100}
outliers$top100 <- isOutlier(sce.mini$pct_counts_top_100_features, 3)
table(outliers$top100)
```

#### Top 200

```{r scater_outliers_top200}
outliers$top200 <- isOutlier(sce.mini$pct_counts_top_200_features, 3)
table(outliers$top200)
```

### Visualisation {.tabset}

Let us visualise outliers, along with blank and bulk samples, for reference:

#### 50

```{r plot.top50, echo=FALSE, fig.height=5}
top50.M <- median(sce.mini$pct_counts_top_50_features)
top50.MAD <- mad(sce.mini$pct_counts_top_50_features)
plot.top50 <-
  ggplot(data.frame(colData(sce.mini)), aes(Sample, pct_counts_top_50_features)) +
  geom_hline(
    yintercept = top50.M + 3*c(-1,1)*top50.MAD,
    colour = "red", linetype = 2:1, alpha = 0.5) +
  geom_hline(yintercept = top50.M, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$top50,]),
    alpha = 0.6, min.segment.length = unit(0, "lines")) +
  labs(y = "% counts (top 50 features)", x = "Samples") +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = col.infection) +
  sampleQCtheme
# plot.top50
ggsave(
  file.path(outdir, "complexity_50_sample.pdf"),
  plot.top50 + guides(colour = "none", shape = "none"), width = 6, height = 4
)
plot.top50 <-
  ggplot(
    data.frame(colData(sce.mini)),
    aes(GroupLabel, pct_counts_top_50_features)) +
  geom_violin(aes(colour = Infection), alpha = 0.25) +
  geom_hline(
    yintercept = top50.M + 3*c(-1,1)*top50.MAD,
    colour = "red", linetype = 2:1, alpha = 0.5) +
  geom_hline(yintercept = top50.M, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$top50,]),
    alpha = 0.6, min.segment.length = unit(0, "lines"), size = rel(2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = col.infection) +
  labs(title = "% counts (top 50 features)", x=NULL, y=NULL) +
  groupQCtheme
plot.top50
ggsave(
  file.path(outdir, "complexity_50_group.pdf"),
  plot.top50 + guides(colour = "none", shape = "none"), width = 6, height = 4
)
```

For reference:

* single cells
    - median: `r sprintf("%.0f%%", top50.M)`
* cut-off
    - upper bound: `r sprintf("%.0f%%", top50.M+top50.MAD*3)`

#### 100

```{r plot.top100, echo=FALSE, fig.height=5}
top100.M <- median(sce.mini$pct_counts_top_100_features)
top100.MAD <- mad(sce.mini$pct_counts_top_100_features)
plot.top100 <-
  ggplot(data.frame(colData(sce.mini)), aes(Sample, pct_counts_top_100_features)) +
  geom_hline(
    yintercept = top100.M + 3*c(-1,1)*top100.MAD,
    colour = "red", linetype = 2:1, alpha=0.4) +
  geom_hline(yintercept = top100.M, alpha=0.4) +
  geom_point(aes(colour = Infection, shape = Status), alpha=0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$top100,]),
    alpha = 0.6, min.segment.length = unit(0, "lines")) +
  scale_y_continuous(limits = c(0,100)) +
  scale_colour_manual(values = col.infection) +
  labs(y = "% counts (top 100 features)", x = "Samples") +
  sampleQCtheme
# plot.top100
ggsave(
  file.path(outdir, "complexity_100_sample.pdf"),
  plot.top100 + guides(colour = "none", shape = "none"), width = 6, height = 4
)
plot.top100 <-
  ggplot(
    data.frame(colData(sce.mini)),
    aes(GroupLabel, pct_counts_top_100_features)) +
  geom_violin(aes(colour = Infection), alpha = 0.25) +
  geom_hline(
    yintercept = top100.M + 3*c(-1,1)*top100.MAD,
    colour = "red", linetype = 2:1, alpha = 0.5) +
  geom_hline(yintercept = top100.M, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$top100,]),
    alpha = 0.6, min.segment.length = unit(0, "lines"), size = rel(2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = col.infection) +
  labs(title = "% counts (top 100 features)", x=NULL, y=NULL) +
  groupQCtheme
plot.top100
ggsave(
  file.path(outdir, "complexity_100_group.pdf"),
  plot.top100 + guides(colour = "none", shape = "none"), width = 6, height = 4
)
```

For reference:

* single cells
    - median: `r sprintf("%.0f%%", top100.M)`
* cut-off
    - upper bound: `r sprintf("%.0f%%", top100.M+top100.MAD*3)`

**Note**:
Two `infected:D23580` samples appear near (below) the cut-off
threshold.
This observation can be related to the overall library complexity figure
shown [earlier](#libraryComplexity).

#### 200

```{r scater_ggplot_top200, echo=FALSE, fig.height=5}
top200.M <- median(sce.mini$pct_counts_top_200_features)
top200.MAD <- mad(sce.mini$pct_counts_top_200_features)
plot.top200 <-
  ggplot(data.frame(colData(sce.mini)), aes(Sample, pct_counts_top_200_features)) +
  geom_hline(
    yintercept = top200.M + 3*c(-1,1)*top200.MAD,
    colour = "red", linetype = 2:1, alpha=0.4) +
  geom_hline(yintercept = top200.M, alpha=0.4) +
  geom_point(aes(colour = Infection, shape = Status), alpha=0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$top200,]),
    alpha = 0.6, min.segment.length = unit(0, "lines")) +
  scale_y_continuous(limits = c(0,100)) +
  scale_colour_manual(values = col.infection) +
  labs(y = "% counts (top 100 features)", x = "Samples") +
  sampleQCtheme
# plot.top200
ggsave(
  file.path(outdir, "complexity_200_sample.pdf"),
  plot.top200 + guides(colour = "none", shape = "none"), width = 6, height = 4
)
plot.top200 <-
  ggplot(
    data.frame(colData(sce.mini)),
    aes(GroupLabel, pct_counts_top_200_features)) +
  geom_violin(aes(colour = Infection), alpha = 0.25) +
  geom_hline(
    yintercept = top200.M + 3*c(-1,1)*top200.MAD,
    colour = "red", linetype = 2:1, alpha = 0.5) +
  geom_hline(yintercept = top200.M, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$top200,]),
    alpha = 0.6, min.segment.length = unit(0, "lines"), size = rel(2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = col.infection) +
  labs(title = "% counts (top 200 features)", x=NULL, y=NULL) +
  groupQCtheme
plot.top200
ggsave(
  file.path(outdir, "complexity_200_group.pdf"),
  plot.top100 + guides(colour = "none", shape = "none"), width = 6, height = 4
)
```

For reference:

* single cells
    - median: `r sprintf("%.0f%%", top200.M)`
* cut-off
    - upper bound: `r sprintf("%.0f%%", top200.M+top200.MAD*3)`

**Note**:
Two `infected:D23580` samples appear near (below) the cut-off threshold.
This observation can be related to the overall library complexity figure
shown [earlier](#libraryComplexity).

## Count of detected features

A notion closely related to library complexity is the count of features
detected within each individual sample. Although the definition of "detection
level" may be debated, we will consider here any feature with non-zero count.

Abnormal counts of detected features may be indicative of technical issue
during RNA capture and library preparation; for instance,
excessively *low* counts of detected features may indicate RNA degradation,
while excessively *high* counts of detected features may indicate capture of
multiple cells.

Owing to the considerable amount of variation that results from the relaxed
definition of detection level above, let us apply a somewhat stringent cut-off
of **2** MADs to identify as outliers single cells
with a count of detected features
excessively *lower* than the median of *single cell* samples,
here beyond **2** MADs::

```{r scater_outliers_features}
outliers$FeaturesCount <- isOutlier(sce.mini$total_features, 2, "lower")
table(outliers$FeaturesCount)
```

Let us visualise outliers, along with blank and bulk samples, for reference:

```{r plot.features, echo=FALSE, fig.height=5}
features.M <- median(sce.mini$total_features)
features.MAD <- mad(sce.mini$total_features)
plot.features <-
  ggplot(data.frame(colData(sce.mini)), aes(Sample, total_features)) +
  geom_hline(yintercept = features.M, alpha = 0.5) +
  geom_hline(
    yintercept = features.M + 2 * c(-1,1) * features.MAD, colour = "red",
    linetype = 1:2, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$FeaturesCount,]),
    alpha = 0.6, min.segment.length = unit(0, "lines")) +
  scale_colour_manual(values = col.infection) +
  scale_y_continuous(limits=c(0,max(sce.mini$total_features))*1.1) +
  labs(y = "Count of detected features", x = "Samples") +
  sampleQCtheme
# plot.features
ggsave(
  file.path(outdir, "features_count_sample.pdf"),
  plot.features + guides(colour = "none", shape = "none"), width = 6, height = 4
)
plot.features <-
  ggplot(
    data.frame(colData(sce.mini)),
    aes(GroupLabel, total_features)) +
  geom_violin(aes(colour = Infection), alpha=0.25) +
  geom_hline(yintercept = features.M, alpha = 0.5) +
  geom_hline(
    yintercept = features.M + 2 * c(-1,1) * features.MAD, colour = "red",
    linetype = 1:2, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$FeaturesCount,]),
    alpha = 0.6, min.segment.length = unit(0, "lines"), size = rel(2)) +
  scale_colour_manual(values = col.infection) +
  scale_y_continuous(limits=c(0,max(sce.mini$total_features))*1.1) +
  labs(title = "Count of detected features", x=NULL, y=NULL) +
  groupQCtheme
plot.features
ggsave(
  file.path(outdir, "features_count_group.pdf"),
  plot.features + guides(colour = "none", shape = "none"), width = 6, height = 4
)
```

For reference:

* single cells
    - median: `r features.M`
* cut-offs
    - lower bound:
      `r sprintf("%.1f", features.M-2*features.MAD)`

## ERCC content {#ERCCoutliers}

An identical--yet very small--quantity of ERCC spike-in features is added
to each sample prior to library preparation.
As a result, technical variation and issues during library preparation may be
estimated from the proportion of reads assigned to those spike-in features;
an excessive proportion of reads assigned to ERCC spike-in features may
indicate eukaryotic RNA degradation,
while excessive variation may indicate cDNA amplication bias.

Due to the limited amount of variation that results from the generally
low levels of ERCC spike-in features detected,
let us apply a somewhat relaxed cut-off of **5** MADs
to identify as outliers single cells an ERCC content excessively *higher* than
the median of *single cell* samples:

```{r scater_outliers_ercc}
outliers$ERCC <- isOutlier(sce.mini$pct_counts_ERCC, 5, "higher")
table(outliers$ERCC)
```

Let us visualise those outliers, along with blank and bulk samples
for reference:

```{r scater_ggplot_ERCC, echo=FALSE, fig.height=5, warning=FALSE}
ERCC.M <- median(sce.mini$pct_counts_ERCC)
ERCC.MAD <- mad(sce.mini$pct_counts_ERCC)
plot.ERCC <-
  ggplot(data.frame(colData(sce.mini)), aes(Sample, pct_counts_ERCC)) +
  geom_hline(yintercept = ERCC.M, alpha = 0.5) +
  geom_hline(
    yintercept = ERCC.M + 5*c(-1,1)*ERCC.MAD,
    colour = "red", linetype = 2:1, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Time), data.frame(colData(sce.mini))[outliers$ERCC,], alpha = 0.6,
    min.segment.length = unit(0, "lines")) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_colour_manual(values = col.infection) +
  labs(y = "ERCC content (% counts)") +
  sampleQCtheme
# plot.ERCC
ggsave(
  file.path(outdir, "ERCC_sample.pdf"),
  plot.ERCC + guides(colour = "none", shape = "none"),
  width = 6, height = 4
)
plot.ERCC <-
  ggplot(
    data.frame(colData(sce.mini)),
    aes(interaction(Time, Infection, Status, sep=" "), pct_counts_ERCC)) +
  geom_violin(aes(colour = Infection), alpha=0.25) +
  geom_hline(yintercept = ERCC.M, alpha = 0.5) +
  geom_hline(
    yintercept = ERCC.M + 5*c(-1,1)*ERCC.MAD,
    colour = "red", linetype = 2:1, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$ERCC,]),
    alpha = 0.6, min.segment.length = unit(0, "lines"), size = rel(2)) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_colour_manual(values = col.infection) +
  labs(title = "ERCC content (% counts)", y=NULL, x=NULL) +
  groupQCtheme
plot.ERCC
ggsave(
  file.path(outdir, "ERCC_group.pdf"),
  plot.ERCC + guides(colour = "none", shape = "none"),
  width = 6, height = 4
)
```

For reference:

* single cells
    - median: `r sprintf("%.0f%%", ERCC.M)`
* cut-offs
    - upper bound:
      `r sprintf("%.0f%%", ERCC.M+ERCC.MAD*5)`

## Mitochondrial content {#MTcontent}

Similarly to ERCC spike-in features, mitochondrial RNA may be used
as an endogenous source of control features;
the limited count of genes encoded on the mitochondrial chromosome
relative to the human chromosomes implies that,
even at their highest levels of expression,
mitochondrial genes are expected to account for a *low* proportion
of the total read count in each sample.
Consequently, abnormally *high* proportions of reads assigned to mitochondrial
gene features may be indicative of technical issue during RNA capture and
library preparation, or abnormal count of mitochondria within the single cells.

Conversely, an unusually *low* proportion of reads assigned to mitochondrial
gene features may also indicate issues related to RNA capture and
library preparation.

Let us identify as outliers single cells for which the proportion of reads
assigned to mitochondrial gene features is excessively *higher*
than the median of *single cell* samples:

```{r scater_outliers_MT}
outliers$MT <- isOutlier(sce.mini$pct_counts_MT, 3, "higher")
table(outliers$MT)
```

Let us visualise those outliers, along with blank and bulk samples
for reference:

```{r scater_ggplot_MT, echo=FALSE, fig.height=5, warning=FALSE}
MT.M <- median(sce.mini$pct_counts_MT)
MT.MAD <- mad(sce.mini$pct_counts_MT)
plot.MT <-
  ggplot(data.frame(colData(sce.mini)), aes(Sample, pct_counts_MT)) +
  geom_hline(yintercept = MT.M, alpha = 0.4) +
  geom_hline(
    yintercept = MT.M + 3 * c(-1,1) * MT.MAD, colour = "red",
    linetype = 2:1, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Time), data.frame(colData(sce.mini))[outliers$MT,], alpha = 0.6,
    min.segment.length = unit(0, "lines")) +
  scale_colour_manual(values = col.infection) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(y = "MT content (% counts)", x = "Samples") + sampleQCtheme
# plot.MT
ggsave(
  file.path(outdir, "MT_sample.pdf"),
  plot.MT + guides(colour = "none", shape = "none"),
  width = 6, height = 4
)
plot.MT <-
  ggplot(
    data.frame(colData(sce.mini)),
    aes(interaction(Time, Infection, Status, sep=" "), pct_counts_MT)) +
  geom_hline(yintercept = MT.M, alpha = 0.4) +
  geom_hline(
    yintercept = MT.M + 3 * c(-1,1) * MT.MAD, colour = "red",
    linetype = 2:1, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Donor),
    data.frame(colData(sce.mini)[outliers$MT,]),
    alpha = 0.6, min.segment.length = unit(0, "lines"), size = rel(2)) +
  scale_colour_manual(values = col.infection) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title = "MT content (% counts)", x=NULL, y=NULL) +
  groupQCtheme
plot.MT
ggsave(
  file.path(outdir, "MT_group.pdf"),
  plot.MT + guides(colour = "none", shape = "none"),
  width = 6, height = 4
)
```

For reference:

* single cells
    - median: `r sprintf("%.0f%%", MT.M)`
* cut-offs
    - upper bound:
      `r sprintf("%.0f%%", MT.M+MT.MAD*5)`

## Summary

Let us display the count of cells identified as outliers for each of the
filters in this section:

```{r scater.fr}
scater.fr <- FilterRules(setdiff(colnames(outliers), names(fastqc.fr)))
scater.fr
```

```{r scater.fr_table, results='asis', echo=FALSE}
pander::pandoc.table(
  data.frame(
    Fail = summary(evalSeparately(scater.fr, outliers)),
    Pass = summary(!evalSeparately(scater.fr, outliers))
))
```

# Global summary of QC metrics

## By filter

Let us now collect the results of *all* the QC metrics obtained in the previous
sections, and display the count of cells :

```{r all.fr}
all.fr <- FilterRules(colnames(outliers))
all.fr
```

```{r all.fr_table, results='asis', echo=FALSE}
pander::pandoc.table(data.frame(
    Fail = summary(evalSeparately(all.fr, outliers)),
    Pass = summary(!evalSeparately(all.fr, outliers))
))
```

In the table above, none of the cells fails *all* the filters;
this demonstrates the importance of each individual filter, despite the
apparent relatedness and overlap between many of the metrics.

## Heat map summary view

Let us define an order of sample that groups single cells by
experimental phenotypes:

```{r sampleSupervisedOrder}
cells.order <- with(colData(sce.mini), order(Time, Status, Infection,Donor))
```

We may then display the QC metrics failed by each single cell as red cells
in an heat map:

```{r QC_heatmap, fig.height=7, echo=FALSE}
hm_row <- rowAnnotation(
  df = data.frame(colData(sce.mini))[cells.order, c("Time", "Infection", "Status","Donor")],
  col = list(
    Time = col.time,
    Status = col.status,
    Infection = col.infection
  )
)
hm <- Heatmap(
  matrix(c("Pass","Fail")[
    as(evalSeparately(all.fr, outliers)[cells.order,], "matrix") + 1
  ], nrow(outliers), ncol(outliers), dimnames = dimnames(outliers)),
  name = "Filter",
  col = c("Pass" = "lightblue", "Fail" = "red"),
  cluster_rows = FALSE, show_row_names = FALSE, cluster_columns = FALSE,
  split = sce.mini$Time[cells.order]
)
draw(hm + hm_row)
pdf(file.path(outdir, "Heatmap_QC_time.pdf"), height = 6, width = 4)
draw(hm + hm_row)
dev.off()
hm <- Heatmap(
  matrix(c("Pass","Fail")[
    as(evalSeparately(all.fr, outliers)[cells.order,], "matrix") + 1
  ], nrow(outliers), ncol(outliers), dimnames = dimnames(outliers)),
  name = "Filter",
  col = c("Pass" = "lightblue", "Fail" = "red"),
  cluster_rows = FALSE, show_row_names = FALSE, cluster_columns = FALSE,
  split = colData(sce.mini)[cells.order,c("Time","Infection","Status")],
  combined_name_fun = function(x) paste(x, collapse = "\n"),
  row_title_gp = gpar(fontsize = 7)
)
pdf(file.path(outdir, "Heatmap_QC_group.pdf"), height = 8, width = 4)
draw(hm + hm_row)
dev.off()
```

```{r table_outliers, include=FALSE}
# Combine phenotype, QC metrics and filter results
FR.table <- as(evalSeparately(all.fr, outliers), "matrix")
colnames(FR.table) <- paste("Filter", colnames(FR.table), sep = ".")
FR.table <- cbind(DataFrame(Filter_Sum=rowSums(FR.table)), FR.table, colData(sce.mini))
rownames(FR.table) <- colnames(sce.mini)
# Export a table of cells that failed any filter
write.csv(
  subset(FR.table, Filter_Sum > 0),
  file.path(outdir, "Outliers_data.csv"))
write.csv(
  subset(FR.table, Filter_Sum == 0),
  file.path(outdir, "Pass_data.csv"))
write.csv(
  FR.table,
  file.path(outdir, "all_data.csv"))
```

<!-- As HTML pages are built independently, objects need to be saved to disk
and imported as needed by the other R mardown files -->

[^1]: Note that shared theme settings are defined in the
[Themes](00a_themes.html) panel.

```{r ggplot_ERCC_label_legend, include=FALSE, fig.height=5, warning=FALSE}
# ERCC plot with legend for manuscript figures
plot.ERCC <-
  ggplot(data.frame(colData(sce.mini)), aes(Sample, pct_counts_ERCC)) +
  geom_hline(yintercept = ERCC.M, alpha = 0.5) +
  geom_hline(
    yintercept = ERCC.M + 5*c(-1,1)*ERCC.MAD,
    colour = "red", linetype = 2:1, alpha = 0.5) +
  geom_point(aes(colour = Infection, shape = Status), alpha = 0.75) +
  geom_text_repel(
    aes(label = Time), data.frame(colData(sce.mini))[outliers$ERCC,], alpha = 0.6,
    min.segment.length = unit(0, "lines")) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_color_manual(values = col.infection) +
  labs(y = "ERCC content (% counts)") +
  sampleQCtheme
plot.ERCC
ggsave(
  file.path(outdir, "ERCC_legend.pdf"), plot.ERCC,
  width = 6, height = 4)
```

# Selection of samples

## Initial unfiltered approach

For the purpose of this validation analysis,
let us retain _all_ the mini-bulk samples for downstream analyses.
We will however bear in mind that samples from donor `D3` tend to fail
more QC metrics than other donors, in case those samples appear as outliers
in dowbnstream analyses (*e.g.*, dimensionality reduction).

## Exclude donor D3

As a result of findings in downstream steps
(sample QC and dimensionality reduction such as PCA), let us exclude all
samples from donor `D3`, as multiple samples from this donor appear as
outliers.

```{r sce.excludeD3}
sce.excludeD3 <- sce.mini[,sce.mini$Donor != "D3"]
dim(sce.excludeD3)
```

```{r save_sce.excludeD3, include=FALSE}
saveRDS(sce.excludeD3, "rds/sce.minibulks.excludeD3.rds")
```


