---
title: "Differential expression in mini-bulks"
author: "Kevin Rue-Albrecht"
date: "06/01/2018"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
stopifnot(
  require(DESeq2)
)
sce.norm <- readRDS("rds/sce.minibulks.vsd.pca.rds")
outdir <- "005_out"; dir.create(outdir)
dds <- readRDS("rds/DESeq2_dds.rds")
```

# Differential expression tests

In this section, we apply the `DESeq2` workflow on the `DESeqDataSet`
produced in an [earlier](004_reduceDim_v2.html#normalise) section:

```{r DESeq}
dds <- DESeq(dds, test = "Wald", fitType = "parametric")
```

## Setup

To collect the differential expression statistics, let us first define:

* a list to store the result tables returned by *scde*

```{r scde.res_init}
deseq.res <- list()
```

* a function used to annotate the tables of results returned by *scde*:

```{r addGeneName}
addGeneName <- function(x){
  x <- data.frame(
    gene_name = with(rowData(sce.norm), gene_name[match(rownames(x), gene_id)]),
    x
  )
  return(x)
}
```

* a function used to sort the tables by increasing *P*-value

```{r orderTable}
orderTable <- function(x){
  x <- x[order(x$pvalue),]
  return(x)
}
```

* a function to make 'safe' character names from experimental levels

```{r safeLevels}
safeLevels <- function(x){
  gsub("STM-", "", x)
}
```


## List

```{r contrasts.time}
contrasts.2h <- list(
  c("2h_STM-D23580_Infected", "2h_Mock_Uninfected"), # vs. Mock
  c("2h_STM-LT2_Infected", "2h_Mock_Uninfected"),
  c("2h_STM-D23580_Exposed", "2h_Mock_Uninfected"),
  c("2h_STM-LT2_Exposed", "2h_Mock_Uninfected"),
  c("2h_STM-D23580_Infected", "2h_STM-LT2_Infected"), # direct
  c("2h_STM-D23580_Infected", "2h_STM-D23580_Exposed"),
  c("2h_STM-LT2_Infected", "2h_STM-LT2_Exposed"),
  c("2h_STM-D23580_Exposed", "2h_STM-LT2_Exposed")
)
contrasts.4h <- list(
  c("4h_STM-D23580_Infected", "4h_Mock_Uninfected"), # vs. Mock
  c("4h_STM-LT2_Infected", "4h_Mock_Uninfected"),
  c("4h_STM-D23580_Exposed", "4h_Mock_Uninfected"),
  c("4h_STM-LT2_Exposed", "4h_Mock_Uninfected"),
  c("4h_STM-D23580_Infected", "4h_STM-LT2_Infected"), # direct
  c("4h_STM-D23580_Infected", "4h_STM-D23580_Exposed"),
  c("4h_STM-LT2_Infected", "4h_STM-LT2_Exposed"),
  c("4h_STM-D23580_Exposed", "4h_STM-LT2_Exposed")
)
contrasts.6h <- list(
  c("6h_STM-D23580_Infected", "6h_Mock_Uninfected"), # vs. Mock
  c("6h_STM-LT2_Infected", "6h_Mock_Uninfected"),
  c("6h_STM-D23580_Exposed", "6h_Mock_Uninfected"),
  c("6h_STM-LT2_Exposed", "6h_Mock_Uninfected"),
  c("6h_STM-D23580_Infected", "6h_STM-LT2_Infected"), # direct
  c("6h_STM-D23580_Infected", "6h_STM-D23580_Exposed"),
  c("6h_STM-LT2_Infected", "6h_STM-LT2_Exposed"),
  c("6h_STM-D23580_Exposed", "6h_STM-LT2_Exposed")
)
```

## 2h

```{r contrasts.2h}
for (contrastNames in contrasts.2h){
  contrastName <- sprintf("%s-%s", contrastNames[1], contrastNames[2])
  message(contrastName)
  safeNames <- safeLevels(contrastNames)
  stopifnot(all(safeNames %in% dds$Group_Safe))
  deseq.res[[contrastName]] <- data.frame(results(
    dds, contrast = c("Group_Safe", safeNames)
  ))
}
```

## 4h

```{r contrasts.4h}
for (contrastNames in contrasts.4h){
  contrastName <- sprintf("%s-%s", contrastNames[1], contrastNames[2])
  message(contrastName)
  safeNames <- safeLevels(contrastNames)
  stopifnot(all(safeNames %in% dds$Group_Safe))
  deseq.res[[contrastName]] <- data.frame(results(
    dds, contrast = c("Group_Safe", safeNames)
  ))
}
```

## 6h

```{r contrasts.6h}
for (contrastNames in contrasts.6h){
  contrastName <- sprintf("%s-%s", contrastNames[1], contrastNames[2])
  message(contrastName)
  safeNames <- safeLevels(contrastNames)
  stopifnot(all(safeNames %in% dds$Group_Safe))
  deseq.res[[contrastName]] <- data.frame(results(
    dds, contrast = c("Group_Safe", safeNames)
  ))
}
```

```{r write.csv, include=FALSE}
for (contrastName in names(deseq.res)){
  write.csv(
    deseq.res[[contrastName]],
    file.path(outdir, sprintf("DESeq2_%s.csv", contrastName))
  )
}
```

```{r saveRDS, include=FALSE}
saveRDS(deseq.res, "rds/deseq.res.rds")
```

```{r view_DESeq, echo=FALSE, eval=FALSE}
names(deseq.res)
View(addGeneName(deseq.res[["6h_STM-D23580_Infected-6h_Mock_Uninfected"]]))
View(addGeneName(deseq.res[["6h_STM-D23580_Infected-6h_STM-LT2_Infected"]]))
View(addGeneName(deseq.res[["6h_STM-STM-LT2_Infected-6h_Mock_Uninfected"]]))
View(addGeneName(deseq.res[["4h_STM-D23580_Infected-4h_STM-D23580_Exposed"]]))
View(addGeneName(deseq.res[["4h_STM-D23580_Infected-4h_Mock_Uninfected"]]))
```


