---
title: "ERCC spike-in content"
bibliography:
  bibtex.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
stopifnot(
  require(ggplot2),
  requireNamespace("pander")
)
```

<style>
body {
text-align: justify}
</style>

# Objective

The input number of spike-in molecules per cell may be calculated
as described in the
[BASiCS tutorial](#https://github.com/catavallejos/TutorialBASiCS)
(*./Material/TutorialBASiCS2016.html*):

\begin{equation*} \mu_{i} = C_i \times 10^{-18} \times (6.022 \times 10^{23})
\times V \times D \hspace{0.5cm} \mbox{where,} \end{equation*}

* $C_i$ is the concentration of the spike $i$ in the ERCC mix
* $10^{-18}$ is to convert aM (attomolar) to M (molar)
* $6.022 \times 10^{23}$ is the Avogadro number (molecules per mole)
* $V$ is the volume added into each well, here $10\mu{L}$
* $D$ is a dilution factor, here $1:100,000,000$

# Concentrations in ERCC Mix from the manufacturer {#import}

Let us import the concentration of individual spike-in molecules
from
[ThermoFisher](https://www.thermofisher.com/order/catalog/product/4456739):

```{r mixC}
mixC <- read.table(
  "expdata/cms_095046.txt", header = TRUE, sep = "\t",
  stringsAsFactors = FALSE
  )[,c("ERCC.ID","concentration.in.Mix.1..attomoles.ul.")]
rownames(mixC) <- mixC$ERCC.ID
colnames(mixC) <- c("ID","Concentration_attomoles_ul")
```

# Number of individual ERCC molecules added to each cell

Let us calculate from the manufacturer's information and our experimental
protocol:

```{r ERCCmolecules}
mixC$ERCCmolecules <-
  mixC$Concentration_attomoles_ul *
  10^-18 *
  6.022 * 10^23 *
  10 *
  10^-8
```

We may then display the expected count of ERCC Spike-In injected at various
levels in each sample; vertical dashed lines indicate levels of **1**,
**10**, **100**, and **1,000** molecules:

```{r ERCCplotCount, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(mixC) +
  geom_dotplot(aes(ERCCmolecules, fill=log10(ERCCmolecules), group=ID), stackgroups = TRUE) +
  scale_x_log10() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  xlab("Number of molecules injected in sample") +
  # guides(fill="none") + 
  theme_bw() +
  theme(legend.position = "bottom")
```

We may also examine the count individual ERCC Spike-In expected
above different thresholds of $molecules / well$:

```{r threshold_ERCC, results='asis', echo=FALSE}
pander::pandoc.table(data.frame(
  "log10(molecules) cutoff" = 10^(0:3),
  "ERCC features detected" = sapply(0:3, function(x){sum(mixC$ERCCmolecules > 10^x)}),
  check.names = FALSE
  ), rownames = FALSE)
```

<!-- As HTML pages are built independently, objects need to be saved to disk
and imported as needed by the other R mardown files -->

```{r save_sce.pass, include=FALSE}
saveRDS(mixC, "rds/mixC.rds")
```
