---
title: "Highly variable genes (HVGs)"
bibliography:
  bibtex.bib
editor_options: 
  chunk_output_type: console
---

<style>
body {
text-align: justify}
</style>

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
stopifnot(
  require(SingleCellExperiment),
  require(scran),
  require(scater),
  require(RColorBrewer),
  require(ggplot2),
  require(ComplexHeatmap),
  require(goseq),
  requireNamespace("dplyr"),
  requireNamespace("MKmisc"),
  require(venn)
)
sce.norm <- readRDS("rds/sce.norm.rds")
sce.bulk <- readRDS("rds/sce.bulk.SCE.rds")
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,4,2)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.norm$Time)[1:3]
names(col.infection) <- levels(sce.norm$Infection)
names(col.status) <- levels(sce.norm$Status)
outdir <- "05_HVG_out"; dir.create(outdir, showWarnings = FALSE)
```

# Plotting themes

First of all, let us define theme elements used throughout various figures
in the following sections:

```{r exprsViolinTheme}
exprsViolinTheme <- theme(
    panel.grid.major.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical")
```

# Highly variable genes (HVGs)

We identify HVGs to focus on the genes that are driving heterogeneity across
the population of cells.
This requires estimation of the variance in expression for each gene, followed
by decomposition of the variance into biological and technical components.
HVGs are then identified as those genes with the biological component
most significantly greater than the technical component.
This avoids prioritizing genes that are highly variable due to technical
factors such as sampling noise during RNA capture and library preparation.

## Define technical component by mean-variance trend {#trendMeanVarHVGs .tabset}

Ideally, the technical component would be estimated by fitting a mean-variance
trend to the ERCC spike-in features.

Alternatively, the expression level of endogenous features may be used
for this purpose,
**under the assumption** that most endogenous features exhibit mostly
technical variation, in contrast to little biological variation.

In both cases, given the mean abundance of a gene,
the fitted value of the trend can be used
as an estimate of the technical component for that gene.
The biological component of the variance may then be calculated by subtracting
the technical component from the total variance of each gene with the
[scran](http://bioconductor.org/packages/scran) `decomposeVar` function.

Both methods are shown below. However, for this data set,
the use of endogenous features was preferred for the following reasons:

* the limited count of detected ERCC spike-in features adversely affected the
  quality of the fitted trend at intermediate expression levels
  (*i.e.*, mean log-expression in the range \approx 5-10)
* ERCC spike-in features show variability levels generally similar to
  endogenous features.

### Using ERCC spike-in features

Let us recall that the same set of ERCC spike-in features was added
in identical quantity to each sample.
Consequently, ERCC spike-in features should exhibit no biological
variability, *i.e.*, any variance in their counts should be technical
in origin.

Let us fit a trend to the variance of spike-in features
using the [scran](http://bioconductor.org/packages/scran) `trendVar` function;
while accounting
for the minor effect of technical factors described
[earlier](06_cluster_normalised#explanatoryVariables).
We do not block on the key experimental factors `Time`, `Infection`, and
`Status`, as doing so may regress out biological effects of interest.

```{r model.matrix_technical}
dm <- model.matrix(~Plate, data = colData(sce.norm))
colnames(dm)
```

Having defined the design matrix, we fit the mean-dependent trend to the
gene-specific variances in the data set:

```{r trendVar_ercc}
var.fit.ercc <- trendVar(
  sce.norm,  method = "loess", assay.type="logcounts", parametric = TRUE,
  design = dm, loess.args=list(span = 0.4), use.spikes=TRUE
)
names(var.fit.ercc)
```

We may then decompose the gene-specific variance into biological and technical
components using the ERCC spike-in trend as an estimate of technical variance:

```{r decomposeVar_ercc}
var.out.ercc <- decomposeVar(
  sce.norm, var.fit.ercc, assay.type="logcounts", get.spikes=FALSE)
names(var.out.ercc)
```

Finally, we may plot the fitted trend, while highlighting ERCC spike-in
features used to obtain it:

```{r plot_var.out.ercc}
plot(var.out.ercc$mean, var.out.ercc$total, pch = 16, cex = 0.6,
  xlab = "Mean log-expression", ylab = "Variance of log-expression"
)
points(var.fit.ercc$mean, var.fit.ercc$var, col="red", pch=16)
o <- order(var.out.ercc$mean);
lines(var.out.ercc$mean[o], var.out.ercc$tech[o], col="red", lwd=2)
```

While the above figure suggests that the trend fits accurately spike-in
features at extreme values the detection range, it also indicate substantial
scatter around the trend for feature at intermediate levels of detection,
and likely understimates technical variance in both ERCC spike-in and
endogenous features at moderately high levels of detection
(*i.e.*, mean log-expression in the range \approx 5-10).

### Using endogenous features

Similarly to the use of ERCC spike-in features, let us use the
[scran](http://bioconductor.org/packages/scran) `trendVar` method
to estimate technical variation under the assumption stated above,
accouting for the same technical factors of the experimental design:

```{r trendVar_endo}
var.fit.endo <- trendVar(
  sce.norm, method = "loess", parametric = TRUE,
  design = dm, loess.args=list(span = 0.1), assay.type="logcounts", use.spikes=FALSE
)
names(var.fit.endo)
```

Similarly, the estimated variance is decomposed into its technical and
biological components (using only the endogenous features processed above):

```{r decomposeVar_endo}
var.out.endo <- decomposeVar(
  sce.norm, var.fit.endo, assay.type="logcounts", get.spikes=FALSE)
names(var.out.endo)
```

In the figure below, we assess the suitability of the trend fitted to the
endogenous variances by examining whether it is consistent with the spike-in
variances.

```{r plot_var.out.endo}
plot(var.out.endo$mean, var.out.endo$total, pch=16, cex=0.6,
  xlab="Mean log-expression", ylab="Variance of log-expression"
)
o <- order(var.out.endo$mean)
lines(var.out.endo$mean[o], var.out.endo$tech[o], col="dodgerblue", lwd=2)
spike.fit <- trendVar(sce.norm, assay.type="logcounts", use.spikes=TRUE) # To compute spike-in variances.
points(spike.fit$mean, spike.fit$var, col="red", pch=16)
```

Notably, the trend passes through or close to most of the spike-in variances,
suggesting that the assumption
(that most genes have low levels of biological variability) is suitable for
this data set.
Notably,
this strategy exploits the large number of endogenous genes to obtain a stable
trend, with the ERCC spike-in features primarily used as diagnostic features
rather than in the trend fitting itself.

Importantly, if the assumption
(that most genes have low levels of biological variability)
did not hold, one would rather use the trend fitted to the ERCC spike-in
features using the default `use.spikes=TRUE` settings.
Notably,
this would sacrifice stability to reduce systematic errors in the estimate of
the biological component for each gene [@67].

\bioccomment{
The help page for `trendVar` suggests fiddling with arguments for smoothing
such as `span`, `degree`, or `family` or using `semiloess` to produce the
trend. I haved tried the `gaussian` family which barely affects the smoothing
of the ERCC spike-in features, and the `semiloess` trend which does not
converge for ERCC spike-in features.
}

## Model selection

The two methods presented above
(*i.e.*, ERCC spike-in and endogenous features) yield markedly distinct trends.

In particular, the limited count of spike-in features robustly detected
across cells
(*i.e.*, `r sum(grepl("^ERCC-",rownames(sce.norm)))`
ERCC spike-in features were retained in the `sce.norm` object),
scarcer at moderate levels of expression,
may explain to some extent the substantial scatter around the trend at moderate
levels of expression
(*i.e.*, mean log-expression in the range $\approx [2.5 - 7.5]$).
Nevertheless, this trend matches closely observations at low detection levels
dense in ERCC spike-in features, and high detection levels that show
generally limited variability.

In contrast, the use of endogenous features---under the assumption that most
genes exhibit technical variation and little biological variation---yields a
trend that may understimate technical variation at low levels of detection and
overestimate technical variation at high levels of detection, considering
ERCC spike-in for reference. However, at moderate detection levels,
the overall trend provides a more representative average estimate of variance
matching both ERCC spike-in and endogenous features (always under the
assumption that most spike-in and endogenous features only show technical
variation).

Let us adopt a conservative approach and use the latter trend (*i.e.*,
estimated from the endogenous features) as it appears to be more consistent
with the spike-in variances,
and as such supports the assumption that most genes have low
levels of biological variability does hold in this data set.

## Identify variable genes {#identifyHVGs .tabset}

HVGs are defined as genes with biological components that are significantly
*greater than 0* at a false discovery rate (FDR) of **5%**.
These genes are interesting as they are most likely to drive differences in the
expression profiles between cells that belong to different experimental groups,
while more generally explaining a substantial proportion of the
transcriptional variance across all cells in the data set.

In addition to the definition above, let us define HVGs as endogenous features
with an estimated biological component of variance greater than or equal to
**0.5**.
For transformed expression values on the log~2~ scale, this means that the
average difference in true expression between any two cells will be at least
**2**-fold.
This reasoning assumes that the true log-expression
values are *normally distributed* with variance of **0.5** [@67].

### Identification

```{r hvg.out}
hvg.out <- var.out.endo[with(var.out.endo, which(FDR <= 0.05 & bio >= 0.5)),]
hvg.out <- hvg.out[order(hvg.out$bio, decreasing=TRUE),]
```

The above criteria have identified `r nrow(hvg.out)` endogenous genes as
*highly variable*.

\bioccomment{
P-value and FDR of ERCC spike-in features are set to `NA` in the result table.
As a result, any filter on those fields will remove those features.
}

The **100** genes with highest biological component of variance may be
examined in the second tab.

```{r write_hvg.out, include=FALSE}
hvg.out <- cbind(
  gene_name = with(
    as.data.frame(rowData(sce.norm)),
    gene_name[match(rownames(hvg.out), gene_id)]
  ),
  hvg.out
)
write.csv(hvg.out, file.path(outdir, "hvg.out.csv"))
```

## Expression levels {.tabset}

Let us examine the distribution of expression values for the top HVGs across
cells to ensure that the variance estimate is not being dominated by a small
count of extreme outliers.

We may highlight the relationship between those HVGs and experimental factors
such as `Time`, `Infection`, and `Status`.

```{r topHvgIds}
topHvgIds <- rownames(hvg.out)[1:10]
```

### Time

```{r topHvgIds_Time, echo=FALSE}
plotExpression(
  sce.norm, topHvgIds, exprs_values = "logcounts", colour_by = "Time"
  ) +
  scale_x_discrete(labels = with(
    as.data.frame(rowData(sce.norm)),
    gene_name[match(topHvgIds, gene_id)]
  )) +
  # scale_colour_manual(values = col.time) +
  exprsViolinTheme
```

The figure above generally indicates a progressive *up*-regulation of gene
expression over time.

### Infection

```{r topHvgIds_Infection, echo=FALSE, warning=FALSE}
plotExpression(
    sce.norm, topHvgIds, exprs_values = "logcounts", colour_by = "Infection"
  ) +
  scale_x_discrete(labels = with(
    as.data.frame(rowData(sce.norm)),
    gene_name[match(topHvgIds, gene_id)]
  )) +
  scale_colour_manual(values = col.infection) +
  exprsViolinTheme
```

The figure above further links the high variance of those genes to
the differences between bacteria-stimulated and mock-infected cells,
particularly at the later time-points.

### Status

```{r topHvgIds_Status, echo=FALSE, warning=FALSE}
plotExpression(
  sce.norm, topHvgIds, exprs_values = "logcounts", colour_by = "Status"
  ) +
  scale_x_discrete(labels = with(
    as.data.frame(rowData(sce.norm)),
    gene_name[match(topHvgIds, gene_id)]
  )) +
  scale_colour_manual(values = col.status) +
  exprsViolinTheme
```

The figure above reinforces clear-cut differences between cells
stimulated by bacterial infection or exposure and `uninfected` cells,
particularly at the later time-points.

```{r LGALS3BP_violin, include=FALSE}
geneName <- "LGALS3BP"
geneId <- subset(
  as.data.frame(rowData(sce.norm)),
  gene_name == geneName, "gene_id", drop = TRUE)
plotExpression(
  sce.norm, geneId, exprs_values = "logcounts",
  colour_by = "Status"
  ) +
  scale_x_discrete(labels = geneName) +
  scale_colour_manual(values = col.status) +
  theme(legend.position = "bottom", legend.box = "vertical")
```

## GO enrichment {.tabset}

Let us use the [goseq](http://bioconductor.org/packages/goseq) package to
identify the most enriched gene ontologies among the HVGs.
Note that we restrict the results to GO categories associated with at least
**10** genes, for robustness.

### Computation

The initial step of the
[goseq](http://bioconductor.org/packages/goseq) procedure
is to create a named vector that indicates whether each gene (*i.e.*, `names`)
is highly variable (`1`) or not (`0`):

```{r endoFeatures}
endoFeatures <- rownames(sce.norm)[!isSpike(sce.norm, "ERCC")]
hvgs <- as.integer(endoFeatures %in% rownames(hvg.out))
names(hvgs) <- endoFeatures
table(hvgs)
```

Next, [goseq](http://bioconductor.org/packages/goseq) is capable of
accounting for gene length bias during GO testing. In the current case,
the gene length information reported by
[featureCounts](http://bioinf.wehi.edu.au/featureCounts) is used:

```{r geneLengths}
geneLengths <- width(rowRanges(sce.norm[rownames(hvg.out),]))
names(geneLengths) <- rownames(hvg.out)
```

Using the above gene length data, the probability weighting function (PWF)
is calculated for the full gene set in this analysis:

```{r nullp}
pwf <- goseq::nullp(
  hvgs, bias.data = width(rowRanges(sce.norm[names(hvgs),])), plot.fit=TRUE)
```

Finally, the `goseq` function is used to test for enrichment amongst the
HVGs:

```{r GO.out, results='hide'}
GO.out <- goseq::goseq(pwf, "hg38", "ensGene")
```

The ontology field (*i.e.*, `"BP"`, `"MF"`, or `"CC"`) is reformatted as
a factor:

```{r GO.out.trim}
GO.out$ontology <- as.factor(GO.out$ontology)
```

```{r GO_HVGs_all, include=FALSE}
write.csv(GO.out, file.path(outdir, "HVGs_GeneOntolgy.csv"), row.names = FALSE)
```

# Expression heatmap {.tabset}

We may then visualise normalised and scaled expression values in
a heat map manually organised by phenotype, while allowing genes with similar
expression profiles to cluster. The procedure is split below in multiple panels
for visual convenience:

## Preprocessing

Let us first subset the normalised data set to retain only endogenous features,
and also reorder samples by `Time`, `Status`, and `Infection`,
to facilitate interpretation of the expression data in the context of the
present experimental design:

```{r orderSamples}
sce.endo <- sce.norm[!isSpike(sce.norm, "ERCC"),]
sce.endo <- sce.endo[,with(colData(sce.endo), order(Time, Status, Infection))]
dim(sce.endo)
```

We may now select the expression data of the endogenous HVGs identified
[above](#identifyHVGs):

```{r topHVG.norm}
norm.endo.hvg <- assay(sce.endo, "logcounts")[rownames(hvg.out),]
dim(norm.endo.hvg)
```

In preparation for the heat map, let us define an annotation panel
that indicates key phenotype levels for each cell:

```{r h_column}
h_column <- HeatmapAnnotation(
  df = colData(sce.endo)[,c("Time", "Status", "Infection")],
  col = list(
    Time = col.time,
    Status = col.status,
    Infection = col.infection
  )
)
```

Let us first cluster HVGs according to their expression profile across *all*
single cells:

```{r norm.endo.hvg.clust}
hvg.endo.d <- dist(norm.endo.hvg)
hvg.endo.clust <- hclust(hvg.endo.d)
rm(hvg.endo.d)
```

We may then cluster single cells according to their expression profile,
*within each of the experimental groups* (*i.e.*, `Time:Infection:Status`):

```{r sample.order}
sample.order <- c()
expGroups <- data.frame(unique(colData(sce.endo)[,c("Time","Infection","Status")]))
expGroups <- dplyr::arrange(expGroups, Time, Status, Infection)
for (groupIndex in seq_len(nrow(expGroups))){
  time <- expGroups[groupIndex, "Time"]
  infection <- expGroups[groupIndex, "Infection"]
  status <- expGroups[groupIndex, "Status"]
  sample.index <- with(colData(sce.endo), which(
    Time == time & Infection == infection & Status == status
  ))
  hvg.sample.d <- dist(t(norm.endo.hvg[,sample.index]))
  local.order <- sample.index[hclust(hvg.sample.d)$order]
  sample.order <- c(sample.order, local.order)
  rm(sample.index, local.order, time, infection, status)
}
rm(expGroups, groupIndex)
```

We may then produce a heat map that uses the above semi-supervised sample
clustering in combination with the earlier gene clustering, to reveal potential
sub-structure within the otherwise homogeneous experimental groups:

```{r heatmap_semiSupervised}
ht_semisupervised <- Heatmap(
  norm.endo.hvg,
  name = "log-counts", column_title = "Normalised expression",
  top_annotation = h_column,
  row_order = hvg.endo.clust$order, column_order = sample.order,
  cluster_rows = hvg.endo.clust, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
```

We may display the expression level of bulk samples for reference.
($log_2(count+1)$ values).

Order bulks identically to single cells:

```{r orderBulk}
sce.bulk <- sce.bulk[,with(colData(sce.bulk), order(Time, Status, Infection))]
```

Annotate bulks with the same colour code:

```{r bulkColumn}
bulk_column <- HeatmapAnnotation(
  df = colData(sce.bulk)[,c("Time", "Infection")],
  col = list(
    Time = col.time,
    Infection = col.infection
  )
)
```


```{r ht_bulks}
ht_bulks <- Heatmap(
  log2(assay(sce.bulk, "counts")[rownames(norm.endo.hvg),] + 1),
  top_annotation = bulk_column,
  name = "exprs", column_title = "Bulks",
  cluster_rows = hvg.endo.clust, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
```

## Bulks

Alone:

```{r ht_bulks_draw, echo=FALSE}
draw(ht_bulks)
```

Alongside single cells (semi-supervised clustering):

```{r ht_semisupervised_draw, echo=FALSE}
draw(
  Heatmap(
    norm.endo.hvg,
    name = "log-counts", column_title = "Normalised expression",
    top_annotation = h_column,
    row_order = hvg.endo.clust$order, column_order = sample.order,
    cluster_rows = hvg.endo.clust, cluster_columns = FALSE,
    show_row_names = FALSE, show_column_names = FALSE
  ) +
  Heatmap(
    log2(assay(sce.bulk, "counts")[rownames(norm.endo.hvg),] + 1),
    top_annotation = HeatmapAnnotation(
      df = colData(sce.bulk)[,c("Time", "Infection")],
      col = list(
        Time = col.time,
        Infection = col.infection
      ),
      show_legend = FALSE
    ),
    name = "bulks", column_title = "Bulks",
    cluster_rows = hvg.endo.clust, cluster_columns = FALSE,
    show_row_names = FALSE, show_column_names = FALSE,
    width = unit(1, "inches")
  )
)
```

## Gene ontology

Let us retrieve the identifier of genes associated with the most enriched
category
[`r GO.out$term[1]`](http://amigo.geneontology.org/amigo/term/`r GO.out$category[1]`)::

```{r genes.top, message=FALSE}
genes.topCategory <- unique(AnnotationDbi::select(
  org.Hs.eg.db, GO.out$category[1], "ENSEMBL", "GOALL")$ENSEMBL
)
table(rownames(norm.endo.hvg) %in% genes.topCategory)
```

In the same manner, let us also annotated genes relating to **4** more
GO categories relevant to the experimental design:

* *type I interferon signaling pathway*
  ([GO:0060337](http://amigo.geneontology.org/amigo/term/GO:0060337))
* *cell death*
  ([GO:0008219](http://amigo.geneontology.org/amigo/term/GO:0008219))
* *lipid metabolic process*
  ([GO:0006629](http://amigo.geneontology.org/amigo/term/GO:0006629))
* *regulation of sequence-specific DNA binding transcription factor activity*
  ([GO:0051090](http://amigo.geneontology.org/amigo/term/GO:0051090))

```{r genes.ifn, message=FALSE}
genes.ifn <- unique(AnnotationDbi::select(
  org.Hs.eg.db, "GO:0060337", "ENSEMBL", "GOALL")$ENSEMBL)
table(rownames(norm.endo.hvg) %in% genes.ifn)
genes.death <- unique(AnnotationDbi::select(
  org.Hs.eg.db, "GO:0008219", "ENSEMBL", "GOALL")$ENSEMBL)
table(rownames(norm.endo.hvg) %in% genes.death)
genes.lipid <- unique(AnnotationDbi::select(
  org.Hs.eg.db, "GO:0006629", "ENSEMBL", "GOALL")$ENSEMBL)
table(rownames(norm.endo.hvg) %in% genes.lipid)
genes.tf <- unique(AnnotationDbi::select(
  org.Hs.eg.db, "GO:0051090", "ENSEMBL", "GOALL")$ENSEMBL)
table(rownames(norm.endo.hvg) %in% genes.tf)
```

Let us annotate genes associated with the above GO categories on the heat map:

```{r heatmap_GO_annotations}
top_l <- rownames(norm.endo.hvg) %in% genes.topCategory
h_top <- Heatmap(
  top_l + 0, name = GO.out$term[1], col = c("0" = "white", "1" = "purple"),
  column_names_gp = gpar(fontsize = 8),
  show_heatmap_legend = FALSE, width = unit(5, "mm"))
ifn_l <- rownames(norm.endo.hvg) %in% genes.ifn
h_ifn <- Heatmap(
  ifn_l + 0, name = "Type I IFN", col = c("0" = "white", "1" = "red"),
  column_names_gp = gpar(fontsize = 8),
  show_heatmap_legend = FALSE, width = unit(5, "mm"))
death_l <- rownames(norm.endo.hvg) %in% genes.death
h_death <- Heatmap(
  death_l + 0, name = "Cell death", col = c("0" = "white", "1" = "red"),
  column_names_gp = gpar(fontsize = 8),
  show_heatmap_legend = FALSE, width = unit(5, "mm"))
lipid_l <- rownames(norm.endo.hvg) %in% genes.lipid
h_lipid <- Heatmap(
  lipid_l + 0, name = "Lipid metab.", col = c("0" = "white", "1" = "red"),
  column_names_gp = gpar(fontsize = 8),
  show_heatmap_legend = FALSE, width = unit(5, "mm"))
tf_l <- rownames(norm.endo.hvg) %in% genes.tf
h_tf <- Heatmap(
  tf_l + 0, name = "TF activity", col = c("0" = "white", "1" = "red"),
  column_names_gp = gpar(fontsize = 8),
  show_heatmap_legend = FALSE, width = unit(5, "mm"))
```

## Semi-supervised

```{r ht_semisupervised, fig.height=7, echo=FALSE}
draw(ht_semisupervised + h_top + h_ifn + h_death + h_lipid + h_tf)
```

# Pairwise distances between samples

In this section, let us explore three approaches to estimate pairwise distance
between single cell samples using the normalised expression data calculated in
a [previous section](05_normalisation.html#normalize)
for the `r nrow(norm.endo.hvg)` HVGs identified [above](#identifyHVGs).

## Estimate distance {.tabset}

### Euclidian distance

Estimate distance between cells using Euclidian distance:

```{r distEuclidNorm}
d.e <- dist(t(norm.endo.hvg), diag = TRUE, upper = TRUE)
mat.e <- as.matrix(d.e)
```

We may then cluster samples according to their Euclidian distance,
using Ward's criterion to minimize the total variance within each cluster [@67]:

```{r clustEuclidNorm}
h.e <- hclust(d.e, method = "ward.D2")
ord.e <- mat.e[h.e$order, h.e$order]
rm(mat.e)
```

Assemble the heat map:

```{r heatEuclid, fig.height=6}
ht_column <- HeatmapAnnotation(
  df = colData(sce.endo)[h.e$order,c("Time", "Status", "Infection")],
  col = list(
    Time = col.time,
    Status = col.status,
    Infection = col.infection
  )
)
ht_e <- Heatmap(
  ord.e, name = "distance", column_title = "Euclidian distance",
  top_annotation = ht_column,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
```

### Spearman correlation

Estimate distance between cells using Spearman's $\rho$ correlation.

```{r distSpearman}
c.s <- MKmisc::corDist(t(norm.endo.hvg), "spearman", diag = TRUE, upper = TRUE)
mat.c.s <- as.matrix(c.s)
```

Cluster samples:

```{r clustSpearman}
h.c.s <- hclust(c.s)
ord.c.s <- mat.c.s[h.c.s$order, h.c.s$order]
rm(c.s)
```

Assemble the heat map:

```{r heatSpearman}
ht_column <- HeatmapAnnotation(
  df = colData(sce.endo)[h.c.s$order,c("Time", "Status", "Infection")],
  col = list(
    Time = col.time,
    Status = col.status,
    Infection = col.infection
  )
)
ht_s <- Heatmap(
  ord.c.s, name = "distance", column_title = "Spearman absolute correlation distance",
  top_annotation = ht_column,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
```

### Pearson correlation

Estimate distance between cells using Pearson's $r$ correlation:

```{r distPearsonNorm}
c.p <- MKmisc::corDist(t(norm.endo.hvg), diag = TRUE, upper = TRUE)
mat.c.p <- as.matrix(c.p)
```

Cluster samples using Pearson's correlation:

```{r clustPearsonNorm}
h.c.p <- hclust(c.p)
ord.c.p <- mat.c.p[h.c.p$order, h.c.p$order]
rm(c.p)
```

Assemble the heat map:

```{r heatPearson}
hn_column <- HeatmapAnnotation(
  df = colData(sce.endo)[h.c.p$order,c("Time", "Status", "Infection")],
  col = list(
    Time = col.time,
    Status = col.status,
    Infection = col.infection
  )
)
ht_p <- Heatmap(
  ord.c.p, name = "distance", column_title = "Pearson absolute correlation distance",
  top_annotation = hn_column,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
```

## Heat maps {#correlationHeatMaps .tabset}

### Euclidian

```{r ht_e, fig.height=6, echo=FALSE}
draw(ht_e)
```

### Spearman

```{r ht_s, fig.height=6, echo=FALSE}
draw(ht_s)
```

### Pearson

```{r ht_p, fig.height=6, echo=FALSE}
draw(ht_p)
```

<!-- As HTML pages are built independently, objects need to be saved to disk
and imported as needed by the other R mardown files -->

```{r saveRDS, include=FALSE}
saveRDS(hvg.out, "rds/hvg.out.rds")
```

# HVGs within experimental groups

## Identification {#HVGsWithinGroups}

Similarly to the above identification of HVGs across cells (driven primarily
by a strong effect of the `Time` factor), let us identify HVGs within each
experimental group of cells:

```{r HVGs.group, warning=FALSE, message=FALSE, fig.height=9, results='hold'}
par(mfrow=c(5,3))
dVar.groups <- list()
HVGs.group <- list()
for (group in levels(sce.norm$Group)){
  idx.group <- which(sce.norm$Group == group)
  message(group); message(sprintf("# samples: %i", length(idx.group)))
  dm <- model.matrix(~Plate, data = droplevels(colData(sce.norm)[idx.group,]))
  var.fit.endo <- trendVar(
    sce.norm[,idx.group], assay.type="logcounts",
    design = dm, use.spikes = FALSE, span = 0.1
  )
  var.out.endo <- decomposeVar(
    sce.norm[,idx.group], var.fit.endo, assay.type="logcounts", get.spikes=FALSE)
  ## Optional check: to compare spike-in variances.
  plot(var.out.endo$mean, var.out.endo$total, pch=16, cex=0.2,
      xlab="Mean log-expression", ylab="Variance of log-expression",
      main = group)
  o <- order(var.out.endo$mean)
  lines(var.out.endo$mean[o], var.out.endo$tech[o], col="dodgerblue", lwd=2)
  spike.fit <- trendVar(
    sce.norm[,idx.group], assay.type="logcounts", use.spikes=TRUE)
  points(spike.fit$mean, spike.fit$var, col="red", pch=16, cex=0.4)
  var.out.endo <- cbind(
    Symbol = with(
      rowData(sce.endo), gene_name[match(rownames(var.out.endo), gene_id)]),
    var.out.endo
  )
  write.csv(var.out.endo, file.path(outdir, sprintf("HVGs_%s_full.csv", group)))
  dVar.groups[[group]] <- var.out.endo
  hvg.out <- var.out.endo[which(
    var.out.endo$FDR <= 0.05 & var.out.endo$bio >= 0.5
    ),]
  write.csv(
    hvg.out, file.path(outdir, sprintf("HVGs_%s_significant.csv", group)))
  HVGs.group[[group]] <- hvg.out
}
par(mfrow=c(1,1))
```

```{r saveHVGsGroup, include=FALSE}
saveRDS(HVGs.group, "rds/HVGs_Group.rds")
```

In the figure above, note how---in the stimulated cells---the mean-variance
trend shifts to the right over time;
variability increases among moderately to highly expressed genes at later time
points.

Interestingly, at the `6h` time point, all groups of stimulated cells
present a small set of genes that appear to be both
highly expressed (mean log-expression > 10) and highly variable
(markedly above the trend of estimated technical variance for most genes;
blue line).
Those subsets of genes are further investigated [below](#HVGsHE_6h),
as they may represent key markers of polarisation toward distinct
phenotypes (*e.g.*, states of cellular activation).

Let us visualise the count of HVGs identified in each experimental group,
while indicating the count of cells per group in a text label for reference:

```{r countHVGsByGroup, echo=FALSE}
groupNames <- levels(sce.norm$Group)
hvgsPerGroup <- data.frame(
  Group = groupNames,
  HVGs = sapply(groupNames, function(groupName){nrow(HVGs.group[[groupName]])}),
  Time = gsub("(.*)_.*_.*", "\\1", groupNames),
  Infection = gsub(".*_(.*)_.*", "\\1", groupNames),
  Status = gsub(".*_.*_(.*)", "\\1", groupNames),
  Treatment = gsub(".*h_(.*)", "\\1", groupNames),
  Cells = as.numeric(table(sce.norm$Group)[groupNames])
)
ggplot(hvgsPerGroup) +
  geom_col(aes(Treatment, HVGs, alpha = Time, colour = Infection)) +
  theme(
    axis.text.x = element_text(angle=90, hjust = 1, vjust = 0.5),
    axis.ticks.x = element_blank(), panel.background = element_blank(),
    axis.title.x = element_blank()
  ) + geom_label(aes(x = Treatment, y = 0, label = Cells)) +
  facet_grid(. ~ Time)
```

# References
