---
title: "Gene expression levels overlaid onto tSNE view"
editor_options: 
  chunk_output_type: console
---

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
stopifnot(
  require(scater),
  require(scran),
  require(ggplot2),
  require(SummarizedExperiment),
  require(RColorBrewer),
  require(circlize),
  require(ComplexHeatmap)
)
sce.norm <- readRDS("rds/sce.norm.rds")
sce.endo <- sce.norm[!isSpike(sce.norm, "ERCC"),]
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,4,2)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.endo$Time)[1:3]
names(col.infection) <- levels(sce.endo$Infection)
names(col.status) <- levels(sce.endo$Status)
outdir <- "13_violins_out"; dir.create(outdir, showWarnings = FALSE)
pdfWidth <- 7; pdfHeight <- 4
cathepsinWidth <- 6; cathepsinHeight <- 3
pdfName <- "%s_%s.pdf"
sce.endo$TreatmentLabel <- with(colData(sce.endo), factor(
  gsub("_", "\n", Treatment),
  gsub("_", "\n", levels(Treatment))
))
```

# Helper values and functions

Let us first define some values useful to design colour scales:

```{r colours}
range.exprs <- range(assay(sce.endo, "logcounts"))
col.treatment <- RColorBrewer::brewer.pal(9, "Paired")[c(9,3,4,1,2)]
names(col.treatment) <- gsub("_","\n",levels(sce.endo$TreatmentLabel))
```

Let us then define a few convenience functions:

* to fetch the gene ID associated with a gene name

```{r geneNameToID}
geneNameToID <- function(x){
  geneId <- subset(rowData(sce.endo), gene_name == x, "gene_id", drop = TRUE)
  if (length(geneId) > 1){
    warning("Multiple IDs found. Use ID instead")
    return(data.frame(
      gene_name = x,
      gene_id = geneId
    ))
  } else if (length(geneId) == 0){
    stop("gene_name not found")
  }
  return(geneId)
}
```

* to fetch the necessary expression and phenotype data

```{r geneDataByName}
geneDataByName <- function(x){
  geneId <- geneNameToID(x)
  ggData <- data.frame(
    gene = assay(sce.endo, "logcounts")[geneId,],
    colData(sce.endo)[,c("Time","Infection","Status","TreatmentLabel")]
  )
  return(ggData)
}
```

* to draw the figure, faceted by time

```{r plotByNameFacetTime}
plotByNameFacetTime <- function(x){
  geneId <- geneNameToID(x)
  ggData <- geneDataByName(x)
  # return(ggData)
  ggplot(ggData, aes(TreatmentLabel, gene)) +
    facet_grid(Time~.) +
    geom_violin(
      aes(colour = TreatmentLabel, fill = TreatmentLabel),
      alpha = 0.2) +
    geom_jitter(
      aes(colour = TreatmentLabel), height = 0, width = 0.2,
      alpha = 0.5, size = 0.4) +
    scale_color_manual(values = col.treatment) +
    scale_fill_manual(values = col.treatment) +
    labs(
      title = sprintf("%s", x),
      subtitle = sprintf("%s", geneId),
      x = "Treatment",
      y = "log-counts",
      colour = "Treatment",
      fill = "Treatment") +
    scale_y_continuous(limits = range.exprs) +
    theme_bw() +
    theme(
      legend.key.height = unit(1.75, "lines")
    )
}
```

* to draw the figure, for a single time point

```{r plotByName}
plotByNameSubset <- function(
  x,
  time = c("2h","4h","6h"),
  infection = c("Mock","STM-LT2","STM-D23580"),
  status = c("Uninfected","Exposed","Infected")
  ){
  geneId <- geneNameToID(x)
  ggData <- geneDataByName(x)
  ggData <- subset(
    ggData,
    Time %in% time & Infection %in% infection & Status %in% status)
  # return(ggData)
  ggplot(ggData, aes(TreatmentLabel, gene)) +
    geom_violin(
      aes(colour = TreatmentLabel, fill = TreatmentLabel), scale = "width",
      alpha = 0.4) +
    geom_jitter(
      aes(colour = TreatmentLabel), height = 0, width = 0.2,
      alpha = 0.5, size = 0.4) +
    scale_color_manual(values = col.treatment) +
    scale_fill_manual(values = col.treatment) +
    labs(
      title = sprintf("%s (%s)", x, time),
      subtitle = sprintf("%s", geneId),
      x = "Treatment",
      y = "log-counts",
      colour = "Treatment",
      fill = "Treatment") +
    scale_y_continuous(limits = range.exprs) +
    theme_bw() +
    theme(
      legend.key.height = unit(1.75, "lines"),
      axis.text.y = element_text(size = rel(1.5))
    )
}
```

# Genes (alphabetical order) {.tabset}

The above convenience function make it straightforward to produce a figure
for any gene of interest:

## ATF3

```{r ATF3, echo=FALSE, fig.height=7}
geneName <- "ATF3"
plotByNameFacetTime(geneName)
```

```{r ATF3_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(
    geneName, time = "6h",
    infection = c("Mock","STM-LT2","STM-D23580"),
    status = c("Uninfected","Exposed","Infected")
  ),
  height = pdfHeight, width = pdfWidth)
```

## BCL2L1

```{r BCL2L1, echo=FALSE, fig.height=7}
geneName <- "BCL2L1"
plotByNameFacetTime(geneName)
```

```{r BCL2L1_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(
    geneName, time = "6h",
    infection = c("Mock","STM-LT2","STM-D23580"),
    status = c("Uninfected","Exposed","Infected")
  ),
  height = pdfHeight, width = pdfWidth)
```

## BIRC3

```{r BIRC3, echo=FALSE, fig.height=7}
geneName <- "BIRC3"
plotByNameFacetTime(geneName)
```

```{r BIRC3_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(
    geneName, time = "6h",
    infection = c("Mock","STM-LT2","STM-D23580"),
    status = c("Uninfected","Exposed","Infected")
  ),
  height = pdfHeight, width = pdfWidth)
```

## CCL24

```{r CCL24, echo=FALSE, fig.height=7}
geneName <- "CCL24"
plotByNameFacetTime(geneName)
```

```{r CCL24_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "2h")),
  plotByNameSubset(geneName, "2h"),
  height = pdfHeight, width = pdfWidth)
```

## CD1A

```{r CD1A, echo=FALSE, fig.height=7}
geneName <- "CD1A"
plotByNameFacetTime(geneName)
```

```{r CD1A_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameSubset(geneName, time = "4h"),
  height = pdfHeight, width = pdfWidth)
```

## CD83

```{r CD83, echo=FALSE, fig.height=7}
geneName <- "CD83"
plotByNameFacetTime(geneName)
```

```{r CD83_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameSubset(geneName, time = "4h"),
  height = pdfHeight-1, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, time = "6h"),
  height = pdfHeight-1, width = pdfWidth)
```

## CLECL1

```{r CLECL1, echo=FALSE, fig.height=7}
geneName <- "CLECL1"
plotByNameFacetTime(geneName)
```

```{r CLECL1_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf("%s_%s_%s.pdf", geneName, "6h", "uninfected-infected")),
  plotByNameSubset(geneName, time = "6h", status = c("Uninfected", "Infected")),
  height = pdfHeight, width = pdfWidth)
```

## CLEC7A

```{r CLEC7A, echo=FALSE, fig.height=7}
geneName <- "CLEC7A"
plotByNameFacetTime(geneName)
```

```{r CLEC7A_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## CTSD

```{r CTSD, echo=FALSE, fig.height=7}
geneName <- "CTSD"
plotByNameFacetTime(geneName)
```

```{r CTSD_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameSubset(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## CTSL

```{r CTSL, echo=FALSE, fig.height=7}
geneName <- "CTSL"
plotByNameFacetTime(geneName)
```

```{r CTSL_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameSubset(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## CTSS

```{r CTSS, echo=FALSE, fig.height=7}
geneName <- "CTSS"
plotByNameFacetTime(geneName)
```

```{r CTSS_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameSubset(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## CYBB

```{r CYBB, echo=FALSE, fig.height=7}
geneName <- "CYBB"
plotByNameFacetTime(geneName)
```

```{r CYBB_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "2h")),
  plotByNameSubset(geneName, "2h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## EBI3

```{r EBI3, echo=FALSE, fig.height=7}
geneName <- "EBI3"
plotByNameFacetTime(geneName)
```

```{r EBI3_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "all")),
  plotByNameSubset(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## IKBKB

```{r IKBKB, echo=FALSE, fig.height=7}
geneName <- "IKBKB"
plotByNameFacetTime(geneName)
```

```{r IKBKB_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(
    geneName, time = "6h",
    infection = c("Mock","STM-LT2","STM-D23580"),
    status = c("Uninfected","Exposed","Infected")
  ),
  height = pdfHeight, width = pdfWidth)
```

## ICOSLG

```{r ICOSLG, echo=FALSE, fig.height=7}
geneName <- "ICOSLG"
plotByNameFacetTime(geneName)
```

```{r ICOSLG_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## IL10

```{r IL10, echo=FALSE, fig.height=7}
geneName <- "IL10"
plotByNameFacetTime(geneName)
```

```{r IL10_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "all")),
  height = 10, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf("%s_%s_%s.pdf", geneName, "6h", "uninfected-infected")),
  plotByNameSubset(geneName, time = "6h", status = c("Uninfected", "Infected")),
  height = pdfHeight, width = pdfWidth)
```

## IL12B

```{r IL12B, echo=FALSE, fig.height=7}
geneName <- "IL12B"
plotByNameFacetTime(geneName)
```

```{r IL12B_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "all")),
  height = 10, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## IL1B

```{r IL1B, echo=FALSE, fig.height=7}
geneName <- "IL1B"
plotByNameFacetTime(geneName)
```

```{r IL1B_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "all")),
  height = 10, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## IL23A

```{r IL23A, echo=FALSE, fig.height=7}
geneName <- "IL23A"
plotByNameFacetTime(geneName)
```

```{r IL23A_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "all")),
  height = 10, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## JAK1

```{r JAK1, echo=FALSE, fig.height=7}
geneName <- "JAK1"
plotByNameFacetTime(geneName)
```

```{r JAK1_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "all")),
  height = 10, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf("%s_%s_%s.pdf", geneName, "6h", "uninfected-infected")),
  plotByNameSubset(geneName, time = "6h", status = c("Uninfected", "Infected")),
  height = pdfHeight, width = pdfWidth)
```

## LGMN

```{r LGMN, echo=FALSE, fig.height=7}
geneName <- "LGMN"
plotByNameFacetTime(geneName)
```

```{r LGMN_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "2h")),
  plotByNameSubset(geneName, "2h"),
  height = pdfHeight, width = pdfWidth)
```

## MARCH1

```{r MARCH1, echo=FALSE, fig.height=7}
geneName <- "MARCH1"
plotByNameFacetTime(geneName)
```

```{r MARCH1_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
ggsave(
  file.path(outdir, sprintf("%s_%s_%s.pdf", geneName, "6h", "uninfected-violet")),
  plotByNameSubset(geneName, time = "6h", status = c("Uninfected", "Infected")),
  height = pdfHeight, width = pdfWidth)
```

## MS4A4A

```{r MS4A4A, echo=FALSE, fig.height=7}
geneName <- "MS4A4A"
plotByNameFacetTime(geneName)
```

```{r MS4A4A_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameSubset(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
```

## PLAT

```{r PLAT, echo=FALSE, fig.height=7}
geneName <- "PLAT"
plotByNameFacetTime(geneName)
```

```{r PLAT_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "4h")),
  plotByNameSubset(geneName, "4h"),
  height = pdfHeight, width = pdfWidth)
```

## TLR4

```{r TLR4, echo=FALSE, fig.height=7}
geneName <- "TLR4"
plotByNameFacetTime(geneName)
```

```{r TLR4_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(geneName, "6h"),
  height = pdfHeight, width = pdfWidth)
```

## TNFAIP3

```{r TNFAIP3, echo=FALSE, fig.height=7}
geneName <- "TNFAIP3"
plotByNameFacetTime(geneName)
```

```{r TNFAIP3_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(
    geneName, time = "6h",
    infection = c("Mock","STM-LT2","STM-D23580"),
    status = c("Uninfected","Exposed","Infected")
  ),
  height = pdfHeight, width = pdfWidth)
```

## TNIP3

```{r TNIP3, echo=FALSE, fig.height=7}
geneName <- "TNIP3"
plotByNameFacetTime(geneName)
```

```{r TNIP3_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf(pdfName, geneName, "6h")),
  plotByNameSubset(
    geneName, time = "6h",
    infection = c("Mock","STM-LT2","STM-D23580"),
    status = c("Uninfected","Exposed","Infected")
  ),
  height = pdfHeight, width = pdfWidth)
```

## VPS8

```{r VPS8, echo=FALSE, fig.height=7}
geneName <- "VPS8"
plotByNameFacetTime(geneName)
```

```{r VPS8_pdf, include=FALSE}
ggsave(
  file.path(outdir, sprintf("%s_%s_%s.pdf", geneName, "6h", "uninfected-infected")),
  plotByNameSubset(geneName, time = "6h", status = c("Uninfected", "Infected")),
  height = pdfHeight, width = pdfWidth)
```

# Gene sets

## Set 1

```{r GOInames}
GOInames <- c(
  "KLHL2","COPS7B","USP1","USPL1","TRIM39", # autophagy
  "CLASP1","STXBP1","MVB12A","BLOC1S3", # intracellular traffic/ endosomal recycling
  "MARCH1","RNF139","TRIM27","CLECL1","ICOSLG", # Ag presentation
  "IFT57","CASP3", # apoptosis
  "DNAJC13","TFRC" # iron availability
)
GOItable <- data.frame(
  gene_name = GOInames,
  gene_id = with(rowData(sce.endo), as.character(gene_id[match(GOInames, gene_name)])),
  stringsAsFactors = FALSE
)
which_samples <- with(
  colData(sce.endo),
  Time == "6h" & Status %in% c("Uninfected","Infected")
)
tmpList <- lapply(GOItable$gene_id, function(x){
  data.frame(
    gene_id = x,
    logcounts = assay(sce.endo, "logcounts")[x, which_samples],
    TreatmentLabel = colData(sce.endo)[which_samples, "TreatmentLabel"],
    gene_name = with(rowData(sce.endo), gene_name[match(x, gene_id)])
  )
})
tmpDF <- do.call("rbind", tmpList)
#
gg <- ggplot(tmpDF, aes(TreatmentLabel, logcounts)) +
  facet_grid(gene_name ~ ., scales = "free_y") +
  geom_violin(
      aes(colour = TreatmentLabel, fill = TreatmentLabel), scale = "width",
      alpha = 0.4) +
    geom_jitter(
      aes(colour = TreatmentLabel), height = 0, width = 0.2,
      alpha = 0.5, size = 0.4) +
  scale_y_continuous(minor_breaks = NULL, breaks = seq(0, max(tmpDF$logcounts), 2)) +
  scale_color_manual(values = col.treatment, drop = FALSE) +
  scale_fill_manual(values = col.treatment, drop = FALSE) +
  theme_bw() +
  theme(
    strip.text.y = element_text(angle = 0),
    strip.background = element_rect(fill = "white")
  )
ggsave(file.path(outdir, "violins_6h_vertical.pdf"), gg, height = 15, width = 7)
gg <- ggplot(tmpDF, aes(TreatmentLabel, logcounts)) +
  coord_flip() +
  facet_grid(~ gene_name , scales = "free_y") +
  geom_violin(
      aes(colour = TreatmentLabel, fill = TreatmentLabel), scale = "width",
      alpha = 0.4) +
    geom_jitter(
      aes(colour = TreatmentLabel), height = 0, width = 0.2,
      alpha = 0.5, size = 0.4) +
  scale_y_continuous(minor_breaks = NULL, breaks = seq(0, max(tmpDF$logcounts), 2)) +
  scale_x_discrete(position = "top") +
  scale_color_manual(values = col.treatment, drop = FALSE) +
  scale_fill_manual(values = col.treatment, drop = FALSE) +
  theme_bw() +
  theme(
    strip.text.y = element_text(angle = 0),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = rel(2))
  )
ggsave(file.path(outdir, "violins_6h_horizontal.pdf"), gg, height = 5, width = 42)
```

