---
title: "Infected vs. Bystander"
author: "Kevin Rue-Albrecht"
date: "21/07/2018"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r require, include=FALSE}
require(venn)
require(circlize)
require(RColorBrewer)
require(ComplexHeatmap)
require(cba)
outDir <- "21_out"; dir.create(outDir)
```

This notebook examines the genes differentially expressed between infected and bystander MoDCs in either or both STM-LT2 and STM-D23580 infection models.

## Load up the SCE object

```{r}
sce <- readRDS("rds/sce.endo_clusters.rds")
```

## Set up the color maps

Heat map

```{r}
blueredColorRamp2 <- colorRamp2(
    breaks = seq(-3, 3, length.out = 9),
    colors = rev(brewer.pal(9, "RdBu")))
```

Metadata

```{r}
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,4,2)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce$Time)
names(col.infection) <- levels(sce$Infection)
names(col.status) <- levels(sce$Status)
```

## Proces each time point separately

### 2h

```{r}
csv_files <- list(
    LT2 = "07_SCDE_out/SCDE_2h_STM-LT2_Infected-2h_STM-LT2_Exposed.csv",
    D23580 = "07_SCDE_out/SCDE_2h_STM-D23580_Infected-2h_STM-D23580_Exposed.csv"
)
```

Load up the SCDE tables

```{r}
scde_tables <- lapply(csv_files, function(x){read.csv(x, row.names = 1)})
```

Identify the DE genes

```{r}
de_ids <- lapply(scde_tables, function(x){rownames(subset(x, p.value < 0.01))})
de_union <- unique(union(de_ids$LT2, de_ids$D23580))
de_both <- intersect(de_ids$LT2, de_ids$D23580)
de_lt2 <- setdiff(de_ids$LT2, de_ids$D23580)
de_d23 <- setdiff(de_ids$D23580, de_ids$LT2)
venn(de_ids)
```

Identify blocks of DE genes

```{r}
de_groups <- factor(rep(NA, length(de_union)), c("Both", "STM-LT2", "STM-D23580"))
names(de_groups) <- de_union
de_groups[de_both] <- "Both"
de_groups[de_lt2] <- "STM-LT2"
de_groups[de_d23] <- "STM-D23580"
```

Export the combined DE information (for both LT2 and D23580)

```{r}
tableExport <- merge(
    x = scde_tables$LT2, y = scde_tables$D23580, by=0, suffixes = c(".LT2",".D23580")
)
tableExport$Group <- c("None", "STM-LT2", "STM-D23580", "Both")[
    1 + (tableExport$p.value.LT2 < 0.01) + 2*(tableExport$p.value.D23580 < 0.01)
]
tableExport <- subset(tableExport, Group != "None")
write.csv(tableExport, file.path(outDir, "Groups_2h.csv"))
```

Prepare the heat maps

```{r}
matrix_all <- assay(sce[de_union, sce$Time == "2h" & sce$Infection != "Mock"], "logcounts")
matrix_all <- t(scale(t(matrix_all)))
#
rowDist <- dist(matrix_all)
rowClust <- hclust(rowDist, method = "ward.D2")
row.oo <- order.optimal(rowDist, rowClust$merge)
rowClust$merge <- row.oo$merge
rowClust$order <- row.oo$order
#
matrix_all <- matrix_all[rowClust$order, ]
de_groups <- de_groups[rownames(matrix_all)]
```

```{r, fig.height=6}
# subset matrices
matrix_lt2 <- matrix_all[, rownames(subset(colData(sce[, colnames(matrix_all)]), Time == "2h" & Infection == "STM-LT2"))]
matrix_d23 <- matrix_all[, rownames(subset(colData(sce[, colnames(matrix_all)]), Time == "2h" & Infection == "STM-D23580"))]
# optimise column order #
lt2_colDist <- dist(t(matrix_lt2))
lt2_colClust <- hclust(lt2_colDist, method = "ward.D2")
lt2.oo <- order.optimal(lt2_colDist, lt2_colClust$merge)
lt2_colClust$merge <- lt2.oo$merge
lt2_colClust$order <- lt2.oo$order
#
d23_colDist <- dist(t(matrix_d23))
d23_colClust <- hclust(d23_colDist, method = "ward.D2")
d23.oo <- order.optimal(d23_colDist, d23_colClust$merge)
d23_colClust$merge <- d23.oo$merge
d23_colClust$order <- d23.oo$order
# heatmap annotations
ha_lt2 <- HeatmapAnnotation(
    df = colData(sce)[colnames(matrix_lt2), c("Infection", "Status"), drop=FALSE],
    col = list(
        Infection=col.infection,
        Status=col.status)
    )
ha_d23 <- HeatmapAnnotation(
    df = colData(sce)[colnames(matrix_d23), c("Infection", "Status"), drop=FALSE],
    col = list(
        Infection=col.infection,
        Status=col.status)
    )
# label genes
geneLabels <- c(
    "TIMM21",
    "MARCH8", "CCL24", "RAB29",
    "IFITM3"
    )
stopifnot(all(geneLabels %in% rowData(sce)[, "gene_name"]))
geneIds <- with(rowData(sce), gene_id[match(geneLabels, gene_name)])
geneIdx <- match(geneIds, rownames(matrix_lt2))
stopifnot(!any(is.na(geneIdx)))
#
ra <- rowAnnotation(link = row_anno_link(at = geneIdx, labels = geneLabels),
    width = unit(0, "cm") + max_text_width(labels))
#
hm_lt2 <- Heatmap(
    matrix=matrix_lt2, col=blueredColorRamp2, name="Z-score",
    top_annotation = ha_lt2,
    cluster_columns = lt2_colClust, cluster_rows = FALSE,
    split = de_groups,
    gap = unit(5, "mm"),
    show_row_names = FALSE, show_column_names = FALSE)
hm_d23 <- Heatmap(
    matrix=matrix_d23, col=blueredColorRamp2, name="none",
    top_annotation = ha_d23,
    cluster_columns = d23_colClust, cluster_rows = FALSE,
    split = de_groups,
    show_row_names = FALSE, show_column_names = FALSE)
hm <- hm_lt2 + hm_d23 + ra
pdf(file.path(outDir, "Heatmap_2h.pdf"), height = 12, width = 10)
draw(hm)
dev.off()
```

### 4h

```{r}
csv_files <- list(
    LT2 = "07_SCDE_out/SCDE_4h_STM-LT2_Infected-4h_STM-LT2_Exposed.csv",
    D23580 = "07_SCDE_out/SCDE_4h_STM-D23580_Infected-4h_STM-D23580_Exposed.csv"
)
```

Load up the SCDE tables

```{r}
scde_tables <- lapply(csv_files, function(x){read.csv(x, row.names = 1)})
```

Identify the DE genes

```{r}
de_ids <- lapply(scde_tables, function(x){rownames(subset(x, p.value < 0.01))})
de_union <- unique(union(de_ids$LT2, de_ids$D23580))
de_both <- intersect(de_ids$LT2, de_ids$D23580)
de_lt2 <- setdiff(de_ids$LT2, de_ids$D23580)
de_d23 <- setdiff(de_ids$D23580, de_ids$LT2)
venn(de_ids)
```

Identify blocks of DE genes

```{r}
de_groups <- factor(rep(NA, length(de_union)), c("Both", "STM-LT2", "STM-D23580"))
names(de_groups) <- de_union
de_groups[de_both] <- "Both"
de_groups[de_lt2] <- "STM-LT2"
de_groups[de_d23] <- "STM-D23580"
```

Export the combined DE information (for both LT2 and D23580)

```{r}
tableExport <- merge(
    x = scde_tables$LT2, y = scde_tables$D23580, by=0, suffixes = c(".LT2",".D23580")
)
tableExport$Group <- c("None", "STM-LT2", "STM-D23580", "Both")[
    1 + (tableExport$p.value.LT2 < 0.01) + 2*(tableExport$p.value.D23580 < 0.01)
]
tableExport <- subset(tableExport, Group != "None")
write.csv(tableExport, file.path(outDir, "Groups_4h.csv"))
```

Prepare the heat maps

```{r}
matrix_all <- assay(sce[de_union, sce$Time == "4h" & sce$Infection != "Mock"], "logcounts")
matrix_all <- t(scale(t(matrix_all)))
#
rowDist <- dist(matrix_all)
rowClust <- hclust(rowDist, method = "ward.D2")
row.oo <- order.optimal(rowDist, rowClust$merge)
rowClust$merge <- row.oo$merge
rowClust$order <- row.oo$order
#
matrix_all <- matrix_all[rowClust$order, ]
de_groups <- de_groups[rownames(matrix_all)]
```

```{r, fig.height=6}
# subset matrices
matrix_lt2 <- matrix_all[, rownames(subset(colData(sce[, colnames(matrix_all)]), Time == "4h" & Infection == "STM-LT2"))]
matrix_d23 <- matrix_all[, rownames(subset(colData(sce[, colnames(matrix_all)]), Time == "4h" & Infection == "STM-D23580"))]
# optimise column order #
lt2_colDist <- dist(t(matrix_lt2))
lt2_colClust <- hclust(lt2_colDist, method = "ward.D2")
lt2.oo <- order.optimal(lt2_colDist, lt2_colClust$merge)
lt2_colClust$merge <- lt2.oo$merge
lt2_colClust$order <- lt2.oo$order
#
d23_colDist <- dist(t(matrix_d23))
d23_colClust <- hclust(d23_colDist, method = "ward.D2")
d23.oo <- order.optimal(d23_colDist, d23_colClust$merge)
d23_colClust$merge <- d23.oo$merge
d23_colClust$order <- d23.oo$order
# heatmap annotations
ha_lt2 <- HeatmapAnnotation(
    df = colData(sce)[colnames(matrix_lt2), c("Infection", "Status"), drop=FALSE],
    col = list(
        Infection=col.infection,
        Status=col.status)
    )
ha_d23 <- HeatmapAnnotation(
    df = colData(sce)[colnames(matrix_d23), c("Infection", "Status"), drop=FALSE],
    col = list(
        Infection=col.infection,
        Status=col.status)
    )
# label genes
geneLabels <- c(
    "APOE", "APOC1", "CD40", "CTSL", "IL1A", "LRP1", "LTB", "PLAT",
    "CCL1", "CCL7", "EBI3",
    "CD83", "TLR8",
    "RIPK3", "VPS25")
stopifnot(all(geneLabels %in% rowData(sce)[, "gene_name"]))
geneIds <- with(rowData(sce), gene_id[match(geneLabels, gene_name)])
geneIdx <- match(geneIds, rownames(matrix_lt2))
stopifnot(!any(is.na(geneIdx)))
#
ra <- rowAnnotation(link = row_anno_link(at = geneIdx, labels = geneLabels),
    width = unit(0, "cm") + max_text_width(labels))
#
hm_lt2 <- Heatmap(
    matrix=matrix_lt2, col=blueredColorRamp2, name="Z-score",
    top_annotation = ha_lt2,
    cluster_columns = lt2_colClust, cluster_rows = FALSE,
    split = de_groups,
    gap = unit(5, "mm"),
    show_row_names = FALSE, show_column_names = FALSE)
hm_d23 <- Heatmap(
    matrix=matrix_d23, col=blueredColorRamp2, name="none",
    top_annotation = ha_d23,
    cluster_columns = d23_colClust, cluster_rows = FALSE,
    split = de_groups,
    show_row_names = FALSE, show_column_names = FALSE)
hm <- hm_lt2 + hm_d23 + ra
pdf(file.path(outDir, "Heatmap_4h.pdf"), height = 12, width = 10)
draw(hm)
dev.off()
```

### 6h

```{r}
csv_files <- list(
    LT2 = "07_SCDE_out/SCDE_6h_STM-LT2_Infected-6h_STM-LT2_Exposed.csv",
    D23580 = "07_SCDE_out/SCDE_6h_STM-D23580_Infected-6h_STM-D23580_Exposed.csv"
)
```

Load up the SCDE tables

```{r}
scde_tables <- lapply(csv_files, function(x){read.csv(x, row.names = 1)})
```

Identify the DE genes

```{r}
de_ids <- lapply(scde_tables, function(x){rownames(subset(x, p.value < 0.01))})
de_union <- unique(union(de_ids$LT2, de_ids$D23580))
de_both <- intersect(de_ids$LT2, de_ids$D23580)
de_lt2 <- setdiff(de_ids$LT2, de_ids$D23580)
de_d23 <- setdiff(de_ids$D23580, de_ids$LT2)
venn(de_ids)
```

Identify blocks of DE genes

```{r}
de_groups <- factor(rep(NA, length(de_union)), c("Both", "STM-LT2", "STM-D23580"))
names(de_groups) <- de_union
de_groups[de_both] <- "Both"
de_groups[de_lt2] <- "STM-LT2"
de_groups[de_d23] <- "STM-D23580"
```

Export the combined DE information (for both LT2 and D23580)

```{r}
tableExport <- merge(
    x = scde_tables$LT2, y = scde_tables$D23580, by=0, suffixes = c(".LT2",".D23580")
)
tableExport$Group <- c("None", "STM-LT2", "STM-D23580", "Both")[
    1 + (tableExport$p.value.LT2 < 0.01) + 2*(tableExport$p.value.D23580 < 0.01)
]
tableExport <- subset(tableExport, Group != "None")
write.csv(tableExport, file.path(outDir, "Groups_6h.csv"))
```

Prepare the heat maps

```{r}
matrix_all <- assay(sce[de_union, sce$Time == "6h" & sce$Infection != "Mock"], "logcounts")
matrix_all <- t(scale(t(matrix_all)))
#
rowDist <- dist(matrix_all)
rowClust <- hclust(rowDist, method = "ward.D2")
row.oo <- order.optimal(rowDist, rowClust$merge)
rowClust$merge <- row.oo$merge
rowClust$order <- row.oo$order
#
matrix_all <- matrix_all[rowClust$order, ]
de_groups <- de_groups[rownames(matrix_all)]
```

```{r, fig.height=6}
# subset matrices
matrix_lt2 <- matrix_all[, rownames(subset(colData(sce[, colnames(matrix_all)]), Time == "6h" & Infection == "STM-LT2"))]
matrix_d23 <- matrix_all[, rownames(subset(colData(sce[, colnames(matrix_all)]), Time == "6h" & Infection == "STM-D23580"))]
# optimise column order #
lt2_colDist <- dist(t(matrix_lt2))
lt2_colClust <- hclust(lt2_colDist, method = "ward.D2")
lt2.oo <- order.optimal(lt2_colDist, lt2_colClust$merge)
lt2_colClust$merge <- lt2.oo$merge
lt2_colClust$order <- lt2.oo$order
#
d23_colDist <- dist(t(matrix_d23))
d23_colClust <- hclust(d23_colDist, method = "ward.D2")
d23.oo <- order.optimal(d23_colDist, d23_colClust$merge)
d23_colClust$merge <- d23.oo$merge
d23_colClust$order <- d23.oo$order
# heatmap annotations
ha_lt2 <- HeatmapAnnotation(
    df = colData(sce)[colnames(matrix_lt2), c("Infection", "Status"), drop=FALSE],
    col = list(
        Infection=col.infection,
        Status=col.status)
    )
ha_d23 <- HeatmapAnnotation(
    df = colData(sce)[colnames(matrix_d23), c("Infection", "Status"), drop=FALSE],
    col = list(
        Infection=col.infection,
        Status=col.status)
    )
# label genes
geneLabels <- c(
    "CTSC", "CTSL", "CTSB", "C1QA",
    "DEFB1", "HLA-DOA", "NFKB1", "TLR2", "TNFAIP3",
    "APOE", "IRF1", "CLECL1", "MAFB",
    "CIITA", "STAT5B")
stopifnot(all(geneLabels %in% rowData(sce)[, "gene_name"]))
geneIds <- with(rowData(sce), gene_id[match(geneLabels, gene_name)])
geneIdx <- match(geneIds, rownames(matrix_lt2))
stopifnot(!any(is.na(geneIdx)))
#
ra <- rowAnnotation(link = row_anno_link(at = geneIdx, labels = geneLabels),
    width = unit(0, "cm") + max_text_width(labels))
#
hm_lt2 <- Heatmap(
    matrix=matrix_lt2, col=blueredColorRamp2, name="Z-score",
    top_annotation = ha_lt2,
    cluster_columns = lt2_colClust, cluster_rows = FALSE,
    split = de_groups,
    gap = unit(5, "mm"),
    show_row_names = FALSE, show_column_names = FALSE)
hm_d23 <- Heatmap(
    matrix=matrix_d23, col=blueredColorRamp2, name="none",
    top_annotation = ha_d23,
    cluster_columns = d23_colClust, cluster_rows = FALSE,
    split = de_groups,
    show_row_names = FALSE, show_column_names = FALSE)
hm <- hm_lt2 + hm_d23 + ra
pdf(file.path(outDir, "Heatmap_6h.pdf"), height = 12, width = 10)
draw(hm)
dev.off()
```
