---
title: "_004_HVG_heatmap_kmeans"
author: "Kevin Rue-Albrecht"
date: "17/12/2017"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
stopifnot(
    require(ComplexHeatmap),
    require(RColorBrewer),
    require(circlize),
    require(org.Hs.eg.db),
    require(Cairo)
)
sce.norm <- readRDS("rds/sce.norm.tSNE.rds")
# Table of HVGs
hvg.out <- read.csv("05_HVG_out/hvg.out.csv", row.names = 1)
# Prepare colours
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,4,2)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.norm$Time)
names(col.infection) <- levels(sce.norm$Infection)
names(col.status) <- levels(sce.norm$Status)
dirOut <- "_004_20180707_out"; dir.create(dirOut)
```

# Set heat map color scale

```{r}
exprsRange <- range(assay(sce.norm, "logcounts"))
colorMap <- colorRamp2(
  seq(min(exprsRange), max(exprsRange)*0.9, length.out = 9),
  rev(brewer.pal(9, "RdBu"))
)
```

# Order by decreasing biological variance

```{r}
hvg.out <- hvg.out[order(hvg.out$bio, decreasing = TRUE),]
sce.hvg <- sce.norm[rownames(hvg.out),]
sce.hvg
```

# Ensure there are no more spike ins

```{r}
sce.endo <- sce.hvg[!isSpike(sce.hvg),]
sce.endo
```

# Take top N features (by decreasing biological variance)

```{r}
sce.top <- sce.endo[1:200,]
sce.top
```

```{r}
write.csv(rowData(sce.top), file.path(dirOut, "HVGs_top200.csv"))
```


# Make heat map

## Cluster HVGs across all cells

```{r}
hvg.endo.d <- dist(assay(sce.top, "logcounts"))
hvg.endo.clust <- hclust(hvg.endo.d)
rm(hvg.endo.d)
```

## Cluster samples within each experimental group

```{r}
sample.order <- c()
expGroups <- unique(colData(sce.top)[,c("Time", "Infection", "Status")])
expGroups <- dplyr::arrange(data.frame(expGroups), Time, Status, Infection)
for (groupIndex in seq_len(nrow(expGroups))){
  time <- expGroups$Time[groupIndex]
  infection <- expGroups$Infection[groupIndex]
  status <- expGroups$Status[groupIndex]
  sample.index <- with(colData(sce.top), which(
    Time == time & Infection == infection & Status == status
  ))
  hvg.sample.d <- dist(t(assay(sce.top[,sample.index], "logcounts")))
  local.order <- sample.index[hclust(hvg.sample.d)$order]
  sample.order <- c(sample.order, local.order)
  rm(sample.index, local.order, time, infection, status)
}
rm(expGroups, groupIndex)
sce.reorder <- sce.top[,sample.order]
```

## Split samples by time point

```{r}
sce.2h <- sce.reorder[,sce.reorder$Time == "2h"]
sce.4h <- sce.reorder[,sce.reorder$Time == "4h"]
sce.6h <- sce.reorder[,sce.reorder$Time == "6h"]
```

## Draw heat map annotations

```{r}
ha_2h <- HeatmapAnnotation(
  df = colData(sce.2h)[,c("Time", "Status", "Infection")],
  col = list(
    Time = col.time,
    Status = col.status,
    Infection = col.infection
  )
)
ha_4h <- HeatmapAnnotation(
  df = colData(sce.4h)[,c("Time", "Status", "Infection")],
  col = list(
    Time = col.time,
    Status = col.status,
    Infection = col.infection
  ), show_legend = FALSE
)
ha_6h <- HeatmapAnnotation(
  df = colData(sce.6h)[,c("Time", "Status", "Infection")],
  col = list(
    Time = col.time,
    Status = col.status,
    Infection = col.infection
  ), show_legend = FALSE
)
```

## Draw main heat maps

```{r}
ht_2h <- Heatmap(
    assay(sce.2h, "logcounts"), col = colorMap,
    name = "logcounts", column_title = "2h",
    top_annotation = ha_2h,
    row_order = hvg.endo.clust$order, column_order = NULL,
    cluster_rows = hvg.endo.clust, cluster_columns = FALSE,
    show_row_names = FALSE, show_column_names = FALSE, width = unit(3, "inches") #, rect_gp = gpar(col = "black", lty = 1, lwd = 0.05)
)
ht_4h <- Heatmap(
    assay(sce.4h, "logcounts"), col = colorMap,
    name = "logcounts.4h", column_title = "4h",
    top_annotation = ha_4h,
    row_order = hvg.endo.clust$order, column_order = NULL,
    cluster_rows = hvg.endo.clust, cluster_columns = FALSE,
    show_row_names = FALSE, show_column_names = FALSE,
    show_heatmap_legend = FALSE, width = unit(3, "inches")#, rect_gp = gpar(col = "black", lty = 1, lwd = 0.05)
)
ht_6h <- Heatmap(
    assay(sce.6h, "logcounts"), col = colorMap,
    name = "logcounts.6h", column_title = "6h",
    top_annotation = ha_6h,
    row_order = hvg.endo.clust$order, column_order = NULL,
    cluster_rows = hvg.endo.clust, cluster_columns = FALSE,
    show_row_names = FALSE, show_column_names = FALSE,
    show_heatmap_legend = FALSE , width = unit(3, "inches")#, rect_gp = gpar(col = "black", lty = 1, lwd = 0.05)
)
```

```{r}
pdf(file.path(dirOut, "heatmap_panel_time.pdf"), height = 6, width = 12)
draw(ht_2h + ht_4h + ht_6h)
dev.off()
```

# Select GO categories

```{r}
anno.csv <- file.path("_004b_in/HVGs_GeneOntolgy.csv")
anno.data <- read.csv(anno.csv, as.is = TRUE)
anno.subset <- subset(anno.data, show == "x", select=c("category", "term"))
```

Identify the genes in each category that are present in the heat map

```{r}
go2geneId <- list()
for (go_idx in seq_len(nrow(anno.subset))) {
    go2geneId[[anno.subset$term[go_idx]]] <- intersect(
        unique(AnnotationDbi::select(
            org.Hs.eg.db, anno.subset$category[go_idx], "ENSEMBL", "GOALL")$ENSEMBL),
        rownames(sce.top)
    )
}
lengths(go2geneId)
go2symbol <- list()
for (goName in names(go2geneId)) {
    go2symbol[[goName]] <-
        mapIds(org.Hs.eg.db, go2geneId[[goName]], "SYMBOL", "ENSEMBL")
}
lengths(go2symbol) 
# Write a file for Anna
sink(file.path(dirOut, "go2symbol.txt"))
go2symbol
sink()
```

Draw a heat map for each GO category

```{r}
GO_h_list <- list()
for (goName in names(go2geneId)) {
    geneIdx <- rownames(sce.reorder) %in% go2geneId[[goName]]
    GO_h_list[[goName]] <- Heatmap(
        geneIdx + 0, name = goName, col = c("0" = "white", "1" = "purple"),
        column_names_gp = gpar(fontsize = 8),
        show_heatmap_legend = FALSE, width = unit(5, "mm"))
}
```

Add the heat maps together and draw them

```{r}
ht_draw <- ht_2h + ht_4h + ht_6h
for (goName in names(go2geneId)) {
    ht_draw <- ht_draw + GO_h_list[[goName]]
}
pdf(file.path(dirOut, "heatmap_panel_time_GO.pdf"), height = 10, width = 16)
draw(
  ht_draw,
  main_heatmap = 1,
  gap = unit(1, "mm")
)
dev.off()
```

# Select gene names

```{r}
anno.xlsx <- file.path("_004b_in/GENES_cat.xlsx")
anno.data <- xlsx::read.xlsx(anno.xlsx, sheetName = 1, header = TRUE, colClasses = "character" , stringsAsFactors=FALSE)
gene.idx <- match(anno.data$gene_id, rownames(sce.reorder))
data.frame(
    anno.data$gene_id,
    gene.idx
)
ra <- rowAnnotation(
  link = row_anno_link(
    at = gene.idx, labels = anno.data$gene_name,
    labels_gp = gpar(cex=1), link_width = unit(0.5, "inches")
    ), width = max_text_width(labels)
)
ht_draw <- ht_2h + ht_4h + ht_6h
for (goName in names(go2geneId)) {
    ht_draw <- ht_draw + GO_h_list[[goName]]
}
pdf(file.path(dirOut, "heatmap_panel_time_GO_annoLink.pdf"), height = 10, width = 16)
# Cairo(type = 'pdf', file = 'heatmap_panel_time_GO_annoLink.pdf', height = 10, width = 16, units = "in")
draw(
  object = ht_draw + ra,
  gap = unit(1, "mm")
)
dev.off()
```
