---
title: "Heat map of genes DE in STM23580 violet/exposed"
author: "Kevin Rue-Albrecht"
date: "27/01/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
stopifnot(
  require(scater),
  require(scran),
  require(ggplot2),
  require(SummarizedExperiment),
  require(RColorBrewer),
  require(circlize),
  require(ComplexHeatmap)
)
sce.norm <- readRDS("rds/sce.norm.rds")
sce.norm <- sce.norm[,order(sce.norm$Group)]
sce.endo <- sce.norm[!isSpike(sce.norm, "ERCC"),]
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,4,2)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.endo$Time)[1:3]
names(col.infection) <- levels(sce.endo$Infection)
names(col.status) <- levels(sce.endo$Status)
outdir <- "_009_heatmap_out"; dir.create(outdir, showWarnings = FALSE)
scde.res <- readRDS("rds/scde_group.rds")
convert.z.score <- function(x, one.sided = NULL) {
  z <- x$Z
  if(is.null(one.sided)) {
    pval = pnorm(-abs(z));
    pval = 2 * pval
  } else if(one.sided=="-") {
    pval = pnorm(z);
  } else {
    pval = pnorm(-z);
  }
    x <- cbind(
      x,
      p.value = pval
  )
  return(x);
}
addGeneName <- function(x){
  x <- cbind(
    gene_name = with(rowData(sce.endo), gene_name[match(rownames(x), gene_id)]),
    x
  )
  return(x)
}
```

# Color maps

## Colors

```{r yellowblackred}
yellowblackred <- colorRampPalette(c("black","brown","red","orange","yellow"))(20)
```

```{r blackpurpleyellow}
blackpurpleyellow <- colorRampPalette(c("black","purple","yellow"))(11)
```

```{r}
redyellowblue <- rev(brewer.pal(11, "RdYlBu"))[c(3,4,5,7,8,9)]
```

## Range of values

```{r range_logcounts}
scale_range_logcounts <- range(assay(sce.endo, "logcounts")) * 0.9
span_range_logcounts <- max(scale_range_logcounts) - min(scale_range_logcounts)
```

## Color scales

```{r redyellowblue_fun}
redyellowblue_fun <- colorRamp2(
  breaks = seq(-2.5, 2.5, by = 1),
  colors = redyellowblue
)
```

```{r blackpurpleyellow_fun}
scale_range_logcounts <- range(assay(sce.endo, "logcounts")) * 0.9
span_range_logcounts <- max(scale_range_logcounts) - min(scale_range_logcounts)
blackpurpleyellow_fun <- colorRamp2(
  breaks = seq(min(scale_range_logcounts), max(scale_range_logcounts), span_range_logcounts / 10),
  colors = blackpurpleyellow
)
```

# Union of genes DE in STM-D23580 infected/exposed across time point

## All cells

```{r union_DE_STM23580_timepoints.pdf}
#
tmp_contrasts <- grep("STM-D23580_Infected-.*Exposed", names(scde.res), value = TRUE)
tmp_DEgeneList <- lapply(tmp_contrasts, function(x){
  res <- convert.z.score(scde.res[[x]])
  rownames(subset(res, p.value < 0.01))
})
tmp_DEgenes <- unique(unlist(tmp_DEgeneList))
#
samples_2h <- rownames(subset(colData(sce.endo), Time == "2h"))
samples_4h <- rownames(subset(colData(sce.endo), Time == "4h"))
samples_6h <- rownames(subset(colData(sce.endo), Time == "6h"))
#
heat_data_2h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_2h]
heat_data_4h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_4h]
heat_data_6h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_6h]
#
row_clust <- hclust(dist(assay(sce.endo, "logcounts")[tmp_DEgenes,]))
#
ha_2 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_2h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
ha_4 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_4h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
ha_6 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_6h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
#
ht_2 <- Heatmap(
  matrix = heat_data_2h, top_annotation = ha_2,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
ht_4 <- Heatmap(
  matrix = heat_data_4h, top_annotation = ha_4,
  cluster_rows = FALSE, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
ht_6 <- Heatmap(
  matrix = heat_data_6h, top_annotation = ha_6,
  cluster_rows = FALSE, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
ht <- ht_2 + ht_4 + ht_6
pdf(file.path(outdir, "union_DE_STM23580_timepoints.pdf"), width = 10, height = 10)
draw(ht)
dev.off()
```

## Only STM-D23580 cells

```{r union_DE_STM23580only_timepoints.pdf}
#
samples_2h <- rownames(subset(colData(sce.endo), Time == "2h" & Infection == "STM-D23580"))
samples_4h <- rownames(subset(colData(sce.endo), Time == "4h" & Infection == "STM-D23580"))
samples_6h <- rownames(subset(colData(sce.endo), Time == "6h" & Infection == "STM-D23580"))
#
heat_data_2h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_2h]
heat_data_4h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_4h]
heat_data_6h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_6h]
#
ha_2 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_2h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
ha_4 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_4h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
ha_6 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_6h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
#
ht_2 <- Heatmap(
  matrix = heat_data_2h, top_annotation = ha_2,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
ht_4 <- Heatmap(
  matrix = heat_data_4h, top_annotation = ha_4,
  cluster_rows = FALSE, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
ht_6 <- Heatmap(
  matrix = heat_data_6h, top_annotation = ha_6,
  cluster_rows = FALSE, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
ht <- ht_2 + ht_4 + ht_6
pdf(file.path(outdir, "union_DE_STM23580only_timepoints.pdf"), width = 10, height = 10)
draw(ht)
dev.off()
```

# Union of genes DE in STM-D23580/STM-LT2 infected across time point

## All cells

```{r union_DE_violet_timepoints.pdf}
#
tmp_contrasts <- grep("STM-D23580_Infected-.*STM-LT2", names(scde.res), value = TRUE)
tmp_DEgeneList <- lapply(tmp_contrasts, function(x){
  res <- convert.z.score(scde.res[[x]])
  rownames(subset(res, p.value < 0.01))
})
tmp_DEgenes <- unique(unlist(tmp_DEgeneList))
#
samples_2h <- rownames(subset(colData(sce.endo), Time == "2h"))
samples_4h <- rownames(subset(colData(sce.endo), Time == "4h"))
samples_6h <- rownames(subset(colData(sce.endo), Time == "6h"))
#
heat_data_2h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_2h]
heat_data_4h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_4h]
heat_data_6h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_6h]
#
row_clust <- hclust(dist(assay(sce.endo, "logcounts")[tmp_DEgenes,]))
#
ha_2 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_2h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
ha_4 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_4h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
ha_6 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_6h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
#
ht_2 <- Heatmap(
  matrix = heat_data_2h, top_annotation = ha_2,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
ht_4 <- Heatmap(
  matrix = heat_data_4h, top_annotation = ha_4,
  cluster_rows = FALSE, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
ht_6 <- Heatmap(
  matrix = heat_data_6h, top_annotation = ha_6,
  cluster_rows = FALSE, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
ht <- ht_2 + ht_4 + ht_6
pdf(file.path(outdir, "union_DE_violet_timepoints.pdf"), width = 10, height = 10)
draw(ht)
dev.off()
```

## Only infected

### Purple colormap

```{r union_DE_Violetonly_timepoints.pdf}
#
samples_2h <- rownames(subset(colData(sce.endo), Time == "2h" & Status == "Infected"))
samples_4h <- rownames(subset(colData(sce.endo), Time == "4h" & Status == "Infected"))
samples_6h <- rownames(subset(colData(sce.endo), Time == "6h" & Status == "Infected"))
#
heat_data_2h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_2h]
heat_data_4h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_4h]
heat_data_6h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_6h]
#
ha_2 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_2h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
ha_4 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_4h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
ha_6 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_6h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
#
ht_2 <- Heatmap(
  matrix = heat_data_2h, top_annotation = ha_2,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
ht_4 <- Heatmap(
  matrix = heat_data_4h, top_annotation = ha_4,
  cluster_rows = FALSE, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
ht_6 <- Heatmap(
  matrix = heat_data_6h, top_annotation = ha_6,
  cluster_rows = FALSE, cluster_columns = FALSE,
  col = blackpurpleyellow,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
ht <- ht_2 + ht_4 + ht_6
pdf(file.path(outdir, "union_DE_Violetonly_timepoints.pdf"), width = 10, height = 10)
draw(ht)
dev.off()
```

### Vogel colormap

```{r union_DE_Violetonly_timepoints.Vogel.pdf}
#
samples_2h <- rownames(subset(colData(sce.endo), Time == "2h" & Status == "Infected"))
samples_4h <- rownames(subset(colData(sce.endo), Time == "4h" & Status == "Infected"))
samples_6h <- rownames(subset(colData(sce.endo), Time == "6h" & Status == "Infected"))
#
heat_data_2h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_2h]
heat_data_4h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_4h]
heat_data_6h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_6h]
#
ha_2 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_2h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
ha_4 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_4h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
ha_6 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_6h, c("Infection","Status")],
  col = list(Infection = col.infection, Status = col.status)
)
#
ht_2 <- Heatmap(
  matrix = heat_data_2h, top_annotation = ha_2,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = redyellowblue_fun,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
ht_4 <- Heatmap(
  matrix = heat_data_4h, top_annotation = ha_4,
  cluster_rows = FALSE, cluster_columns = FALSE,
  col = redyellowblue_fun,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
ht_6 <- Heatmap(
  matrix = heat_data_6h, top_annotation = ha_6,
  cluster_rows = FALSE, cluster_columns = FALSE,
  col = redyellowblue_fun,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
ht <- ht_2 + ht_4 + ht_6
pdf(file.path(outdir, "union_DE_Violetonly_timepoints_vogel.pdf"), width = 10, height = 10)
draw(ht)
dev.off()
```

# Gene DE D23/LT2 Violet

## Genes of interest

```{r}
GnOI <- c("CTSS", "CTSB", "CTSL", "CTSA", "S100A9", "LGMN", "PSAP", "MS4A4A", "MS4A6A", "CCL7", "CCL26", "APOC1", "MARCH1", "RNF139", "TRIM27", "CLECL1", "CLASP1", "STXBP1", "MVB12A", "DNAJC13", "BLOC1S3", "USP1", "USPL1", "TRIM39", "CLEC16A", "NIT1", "NIT2", "TFRC")
GOIdf <- data.frame(
  gene_name = GnOI,
  gene_id = with(rowData(sce.endo), gene_id[match(GnOI, gene_name)])
)
GnNegativeResult <- c("LAMP1", "EEA", "RAB5A")
```


## 2h

### Violet color map

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["2h_STM-D23580_Infected-2h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_2h <- rownames(subset(colData(sce.endo), Time == "2h" & Status == "Infected"))
#
heat_data_2h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_2h]
#
row_clust <- hclust(dist(heat_data_2h[tmp_DEgenes,]))
#
ha_2 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_2h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_2h))
#
ht_2 <- Heatmap(
  matrix = heat_data_2h, top_annotation = ha_2,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = blackpurpleyellow_fun,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_2h.pdf"), width = 7, height = 10)
draw(ht_2)
dev.off()
```

### Vogel color map

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["2h_STM-D23580_Infected-2h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_2h <- rownames(subset(colData(sce.endo), Time == "2h" & Status == "Infected"))
#
heat_data_2h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_2h]
#
heat_data_2h <- t(scale(t(heat_data_2h)))
#
row_clust <- hclust(dist(heat_data_2h[tmp_DEgenes,]))
#
ha_2 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_2h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_2h))
#
ht_2 <- Heatmap(
  matrix = heat_data_2h, top_annotation = ha_2,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = redyellowblue_fun,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_2h_vogel.pdf"), width = 7, height = 10)
draw(ht_2)
dev.off()
```

## 4h

### Violet color map

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["4h_STM-D23580_Infected-4h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_4h <- rownames(subset(colData(sce.endo), Time == "4h" & Status == "Infected"))
#
heat_data_4h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_4h]
#
row_clust <- hclust(dist(heat_data_4h[tmp_DEgenes,]))
#
ha_4 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_4h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_4h))
GOIdf_4 <- subset(GOIdf, !is.na(row))
ra_4 <- rowAnnotation(
  link = row_anno_link(
    at = GOIdf_4$row,
    labels = as.character(GOIdf_4$gene_name),
    labels_gp = gpar(cex=2/3),
    link_width = unit(0.5, "inches")
  )
)
#
ht_4 <- Heatmap(
  matrix = heat_data_4h, top_annotation = ha_4,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = blackpurpleyellow_fun,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_4h_annolink.pdf"), width = 7, height = 10)
draw(ht_4 + ra_4)
dev.off()
```

### Vogel color map

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["4h_STM-D23580_Infected-4h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_4h <- rownames(subset(colData(sce.endo), Time == "4h" & Status == "Infected"))
#
heat_data_4h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_4h]
#
heat_data_4h <- t(scale(t(heat_data_4h)))
#
row_clust <- hclust(dist(heat_data_4h[tmp_DEgenes,]))
#
ha_4 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_4h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
ht_4 <- Heatmap(
  matrix = heat_data_4h, top_annotation = ha_4,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = redyellowblue_fun,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_4h_vogel.pdf"), width = 7, height = 10)
draw(ht_4)
dev.off()
```

## 6h

### Violet color map

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["6h_STM-D23580_Infected-6h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_6h <- rownames(subset(colData(sce.endo), Time == "6h" & Status == "Infected"))
#
heat_data_6h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_6h]
#
row_clust <- hclust(dist(heat_data_6h[tmp_DEgenes,]))
#
ha_6 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_6h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_6h))
GOIdf_6 <- subset(GOIdf, !is.na(row))
ra_6 <- rowAnnotation(
  link = row_anno_link(
    at = GOIdf_6$row,
    labels = as.character(GOIdf_6$gene_name),
    labels_gp = gpar(cex=2/3),
    link_width = unit(0.5, "inches")
  )
)
#
ht_6 <- Heatmap(
  matrix = heat_data_6h, top_annotation = ha_6,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = blackpurpleyellow_fun,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_6h_annolink.pdf"), width = 7, height = 10)
draw(ht_6 + ra_6)
dev.off()
```

### Vogel color map

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["6h_STM-D23580_Infected-6h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_6h <- rownames(subset(colData(sce.endo), Time == "6h" & Status == "Infected"))
#
heat_data_6h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_6h]
#
heat_data_6h <- t(scale(t(heat_data_6h)))
#
row_clust <- hclust(dist(heat_data_6h[tmp_DEgenes,]))
#
ha_6 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_6h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
ht_6 <- Heatmap(
  matrix = heat_data_6h, top_annotation = ha_6,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = redyellowblue_fun,
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_6h_vogel.pdf"), width = 7, height = 10)
draw(ht_6)
dev.off()
```

## Row-scaled

### 2h

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["2h_STM-D23580_Infected-2h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_2h <- rownames(subset(colData(sce.endo), Time == "2h" & Status == "Infected"))
#
heat_data_2h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_2h]
heat_data_2h <- t(scale(t(heat_data_2h)))
#
row_clust <- hclust(dist(heat_data_2h[tmp_DEgenes,]))
#
ha_2 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_2h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_2h))
#
ht_2 <- Heatmap(
  matrix = heat_data_2h, top_annotation = ha_2,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = colorRampPalette(c("purple","black","yellow"))(11),
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_2h_rowscaled.pdf"), width = 7, height = 10)
draw(ht_2)
dev.off()
```

### 4h

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["4h_STM-D23580_Infected-4h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_4h <- rownames(subset(colData(sce.endo), Time == "4h" & Status == "Infected"))
#
heat_data_4h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_4h]
heat_data_4h <- t(scale(t(heat_data_4h)))
#
row_clust <- hclust(dist(heat_data_4h[tmp_DEgenes,]))
#
ha_4 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_4h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_4h))
GOIdf_4 <- subset(GOIdf, !is.na(row))
ra_4 <- rowAnnotation(
  link = row_anno_link(
    at = GOIdf_4$row,
    labels = as.character(GOIdf_4$gene_name),
    labels_gp = gpar(cex=2/3),
    link_width = unit(0.5, "inches")
  )
)
#
ht_4 <- Heatmap(
  matrix = heat_data_4h, top_annotation = ha_4,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = colorRampPalette(c("purple","black","yellow"))(11),
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_4h_annolink_rowscaled.pdf"), width = 7, height = 10)
draw(ht_4 + ra_4)
dev.off()
```

### 6h

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["6h_STM-D23580_Infected-6h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_6h <- rownames(subset(colData(sce.endo), Time == "6h" & Status == "Infected"))
#
heat_data_6h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_6h]
heat_data_6h <- t(scale(t(heat_data_6h)))
#
row_clust <- hclust(dist(heat_data_6h[tmp_DEgenes,]))
#
ha_6 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_6h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_6h))
GOIdf_6 <- subset(GOIdf, !is.na(row))
ra_6 <- rowAnnotation(
  link = row_anno_link(
    at = GOIdf_6$row,
    labels = as.character(GOIdf_6$gene_name),
    labels_gp = gpar(cex=2/3),
    link_width = unit(0.5, "inches")
  )
)
#
ht_6 <- Heatmap(
  matrix = heat_data_6h, top_annotation = ha_6,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = colorRampPalette(c("purple","black","yellow"))(11),
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_6h_annolink_rowscaled.pdf"), width = 7, height = 10)
draw(ht_6 + ra_6)
dev.off()
```

## Row-scaled (v2)

### 2h

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["2h_STM-D23580_Infected-2h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_2h <- rownames(subset(colData(sce.endo), Time == "2h" & Status == "Infected"))
#
heat_data_2h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_2h]
heat_data_2h <- t(scale(t(heat_data_2h)))
#
row_clust <- hclust(dist(heat_data_2h[tmp_DEgenes,]))
#
ha_2 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_2h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_2h))
#
ht_2 <- Heatmap(
  matrix = heat_data_2h, top_annotation = ha_2,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = colorRamp2(
    c(min(heat_data_2h), c(-1,0,1), max(heat_data_2h)),
    c("purple","gray10","black","gray10","yellow")
  ),
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_2h_rowscaled_offset1.pdf"), width = 7, height = 10)
draw(ht_2)
dev.off()
```

### 4h

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["4h_STM-D23580_Infected-4h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_4h <- rownames(subset(colData(sce.endo), Time == "4h" & Status == "Infected"))
#
heat_data_4h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_4h]
heat_data_4h <- t(scale(t(heat_data_4h)))
#
row_clust <- hclust(dist(heat_data_4h[tmp_DEgenes,]))
#
ha_4 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_4h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_4h))
GOIdf_4 <- subset(GOIdf, !is.na(row))
ra_4 <- rowAnnotation(
  link = row_anno_link(
    at = GOIdf_4$row,
    labels = as.character(GOIdf_4$gene_name),
    labels_gp = gpar(cex=2/3),
    link_width = unit(0.5, "inches")
  )
)
#
ht_4 <- Heatmap(
  matrix = heat_data_4h, top_annotation = ha_4,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = colorRamp2(
    c(min(heat_data_2h), c(-1,0,1), max(heat_data_2h)),
    c("purple","gray10","black","gray10","yellow")
  ),
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_4h_annolink_rowscaled_offset1.pdf"), width = 7, height = 10)
draw(ht_4 + ra_4)
dev.off()
```

### 6h

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["6h_STM-D23580_Infected-6h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_6h <- rownames(subset(colData(sce.endo), Time == "6h" & Status == "Infected"))
#
heat_data_6h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_6h]
heat_data_6h <- t(scale(t(heat_data_6h)))
#
row_clust <- hclust(dist(heat_data_6h[tmp_DEgenes,]))
#
ha_6 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_6h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_6h))
GOIdf_6 <- subset(GOIdf, !is.na(row))
ra_6 <- rowAnnotation(
  link = row_anno_link(
    at = GOIdf_6$row,
    labels = as.character(GOIdf_6$gene_name),
    labels_gp = gpar(cex=2/3),
    link_width = unit(0.5, "inches")
  )
)
#
ht_6 <- Heatmap(
  matrix = heat_data_6h, top_annotation = ha_6,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = colorRamp2(
    c(min(heat_data_2h), c(-1,0,1), max(heat_data_2h)),
    c("purple","gray10","black","gray10","yellow")
  ),
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_6h_annolink_rowscaled_offset1.pdf"), width = 7, height = 10)
draw(ht_6 + ra_6)
dev.off()
```


## Row-scaled blue-white-red

### 2h

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["2h_STM-D23580_Infected-2h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_2h <- rownames(subset(colData(sce.endo), Time == "2h" & Status == "Infected"))
#
heat_data_2h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_2h]
heat_data_2h <- t(scale(t(heat_data_2h)))
#
row_clust <- hclust(dist(heat_data_2h[tmp_DEgenes,]))
#
ha_2 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_2h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_2h))
#
ht_2 <- Heatmap(
  matrix = heat_data_2h, top_annotation = ha_2,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = colorRampPalette(c("blue","white","red"))(11),
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_2h_rowscaled_bluewhitered.pdf"), width = 7, height = 10)
draw(ht_2)
dev.off()
```

### 4h

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["4h_STM-D23580_Infected-4h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_4h <- rownames(subset(colData(sce.endo), Time == "4h" & Status == "Infected"))
#
heat_data_4h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_4h]
heat_data_4h <- t(scale(t(heat_data_4h)))
#
row_clust <- hclust(dist(heat_data_4h[tmp_DEgenes,]))
#
ha_4 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_4h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_4h))
GOIdf_4 <- subset(GOIdf, !is.na(row))
ra_4 <- rowAnnotation(
  link = row_anno_link(
    at = GOIdf_4$row,
    labels = as.character(GOIdf_4$gene_name),
    labels_gp = gpar(cex=2/3),
    link_width = unit(0.5, "inches")
  )
)
#
ht_4 <- Heatmap(
  matrix = heat_data_4h, top_annotation = ha_4,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = colorRampPalette(c("blue","white","red"))(11),
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_4h_annolink_rowscaled_bluewhitered.pdf"), width = 7, height = 10)
draw(ht_4 + ra_4)
dev.off()
```

### 6h

```{r}
#
names(scde.res)
tmp_DEgenes <- rownames(subset(
  convert.z.score(scde.res[["6h_STM-D23580_Infected-6h_STM-LT2_Infected"]]),
  p.value < 0.01)
)
#
samples_6h <- rownames(subset(colData(sce.endo), Time == "6h" & Status == "Infected"))
#
heat_data_6h <- assay(sce.endo, "logcounts")[tmp_DEgenes, samples_6h]
heat_data_6h <- t(scale(t(heat_data_6h)))
#
row_clust <- hclust(dist(heat_data_6h[tmp_DEgenes,]))
#
ha_6 <- HeatmapAnnotation(
  df = colData(sce.endo)[samples_6h, c("Infection"), drop = FALSE],
  col = list(Infection = col.infection)
)
#
GOIdf$row <- match(GOIdf$gene_id, rownames(heat_data_6h))
GOIdf_6 <- subset(GOIdf, !is.na(row))
ra_6 <- rowAnnotation(
  link = row_anno_link(
    at = GOIdf_6$row,
    labels = as.character(GOIdf_6$gene_name),
    labels_gp = gpar(cex=2/3),
    link_width = unit(0.5, "inches")
  )
)
#
ht_6 <- Heatmap(
  matrix = heat_data_6h, top_annotation = ha_6,
  cluster_rows = row_clust, cluster_columns = FALSE,
  col = colorRampPalette(c("blue","white","red"))(11),
  show_row_names = FALSE, show_column_names = FALSE,
  row_names_side = "left"
)
#
pdf(file.path(outdir, "DE_violet_6h_annolink_rowscaled_bluewhitered.pdf"), width = 7, height = 10)
draw(ht_6 + ra_6)
dev.off()
```
