---
title: "Correlation MLE split"
author: "Kevin Rue-Albrecht"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
stopifnot(
  require(broom),
  require(reshape2),
  require(RColorBrewer),
  require(ggplot2)
)
dirOut <- "_012_correlation_out"; dir.create(dirOut)
```

# Load previous data

Differential expression between experimental groups:

```{r scde.groups}
scde.groups <- readRDS("rds/scde_group.rds")
names(scde.groups)
```

Differential expression between unsupervised clusters:

```{r scde.clusters}
scde.clusters <- readRDS("rds/scde_cluster.rds")
names(scde.clusters)
```

`SingleCellExperiment` object:

```{r sce.norm}
sce.norm <- readRDS("rds/sce.norm.tSNE.rds")
sce.norm
```

# Colour codes

```{r colours.infection}
colours.infection <- brewer.pal(3, "Set1")[2:3]
names(colours.infection) <- c("STM-D23580", "STM-LT2")
```

# Select a time point

```{r}
timepoint <- "2h"
```

# Extract DE tables

```{r}
clusters.de <- scde.clusters[[sprintf("%s_cluster2-cluster1", timepoint)]]
head(clusters.de, n=2)
```

```{r}
groups.de.d23 <- scde.groups[[
  sprintf("%s_STM-D23580_Infected-%s_STM-D23580_Exposed", timepoint, timepoint)
]]
head(groups.de.d23, n=2)
```

```{r}
groups.de.lt2 <- scde.groups[[
  sprintf("%s_STM-LT2_Infected-%s_STM-LT2_Exposed", timepoint, timepoint)
]]
head(groups.de.lt2, n=2)
```

# Convert Z-scores to P-values

A helper function:

```{r convert.z.score}
convert.z.score <- function(x, one.sided = NULL) {
  z <- x$Z
  if(is.null(one.sided)) {
    pval = pnorm(-abs(z));
    pval = 2 * pval
  } else if(one.sided=="-") {
    pval = pnorm(z);
  } else {
    pval = pnorm(-z);
  }
  x <- cbind(
    x,
    p.value = pval
  )
  return(x);
}
```

Application of the above function:

```{r apply.convert.z.score}
clusters.de <- convert.z.score(clusters.de)
groups.de.d23 <- convert.z.score(groups.de.d23)
groups.de.lt2 <- convert.z.score(groups.de.lt2)
```

# Annotate with gene name (optional)

```{r addGENENAME}
clusters.de$GENENAME <-
  with(rowData(sce.norm), gene_name[match(rownames(clusters.de), gene_id)])
groups.de.d23$GENENAME <-
  with(rowData(sce.norm), gene_name[match(rownames(groups.de.d23), gene_id)])
groups.de.lt2$GENENAME <-
  with(rowData(sce.norm), gene_name[match(rownames(groups.de.lt2), gene_id)])
```

# Prepare MLE data frame for ggplot

```{r gg.mle}
gg.mle <- data.frame(
  gene_id = c(
    rownames(groups.de.d23),
    rownames(groups.de.lt2)
  ),
  Group = rep(
    c("STM-D23580", "STM-LT2"),
    c(nrow(groups.de.d23), nrow(groups.de.lt2))
  ),
  MLE.group = c(
    groups.de.d23$mle,
    groups.de.lt2$mle
  ),
  MLE.cluster = clusters.de[
    c(
      rownames(groups.de.d23),
      rownames(groups.de.lt2)
    ),
    "mle"
    ],
  P.group = c(
    groups.de.d23$p.value,
    groups.de.lt2$p.value
  ),
  P.cluster = clusters.de[
    c(
      rownames(groups.de.d23),
      rownames(groups.de.lt2)
    ),
    "p.value"
    ],
  gene_name = with(
    rowData(sce.norm),
    gene_name[
      match(c(
        rownames(groups.de.d23),
        rownames(groups.de.lt2)
      ),
      gene_id)
      ]
  )
)
```

```{r ggplot.mle}
ggplot(gg.mle) +
  geom_point(aes(MLE.cluster, -MLE.group, colour = Group), alpha = 0.2) +
  theme_minimal()
```

```{r ggplot.mle.density}
ggplot(gg.mle) +
  stat_density2d(
    aes(MLE.cluster, -MLE.group, fill = ..density..^0.25),
    geom = "tile", contour = FALSE, n = 200) +
  facet_grid(~ Group) +
  scale_fill_continuous(low = "white", high = "dodgerblue4") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

# Prepare regression data for ggplot

```{r reg}
d23.reg <- with(
  data.frame(
    D23580 = groups.de.d23$mle,
    Cluster1v2 = -clusters.de[rownames(groups.de.d23), 'mle']
  ),
  tidy(lm(D23580 ~ Cluster1v2))
)
lt2.reg <- with(
  data.frame(
    LT2 = groups.de.lt2$mle,
    Cluster1v2 = -clusters.de[rownames(groups.de.lt2), 'mle']
  ),
  tidy(lm(LT2 ~ Cluster1v2))
)
all.reg <- rbind(
  cbind(d23.reg, Group = "STM-D23580"),
  cbind(lt2.reg, Group = "STM-LT2")
)
all.reg$term <- rep(c("intercept", "slope"), 2)
```

```{r gg.regression}
gg.regression <- dcast(all.reg, Group ~ term, value.var = "estimate")
gg.regression$x <- 5
gg.regression$y <- -5:-6
```

```{r ggplot.reg}
ggplot(gg.mle) +
  # stat_density2d(
  #   aes(MLE.cluster, -MLE.group, fill = ..density..^0.25),
  #   geom = "tile", contour = FALSE, n = 200) +
  geom_point(aes(MLE.cluster, -MLE.group, colour = Group), alpha = 0.2) +
  geom_abline(
    aes(slope = slope, intercept = intercept, colour = Group),
    gg.regression
  ) +
  # facet_grid(~ Group) +
  scale_fill_continuous(low = "white", high = "dodgerblue4") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

# Subset to genes DE between clusters

```{r gg.sig.cluster}
gg.sig.cluster <- subset(gg.mle, P.cluster < 0.01)
dim(gg.sig.cluster)
```

```{r ggplot.sig}
ggplot(gg.sig.cluster) +
  geom_point(aes(MLE.cluster, -MLE.group, colour = Group), alpha = 0.2) +
  theme_minimal()
```

# Regression on cluster DE genes only

```{r genes.sig.cluster}
genes.sig.cluster <- unique(as.character(gg.sig.cluster$gene_id))
length(genes.sig.cluster)
```

```{r}
d23.reg <- with(
  data.frame(
    D23580 = groups.de.d23[genes.sig.cluster, "mle"],
    Cluster1v2 = -clusters.de[genes.sig.cluster, "mle"]
  ),
  tidy(lm(D23580 ~ Cluster1v2))
)
lt2.reg <- with(
  data.frame(
    LT2 = groups.de.lt2[genes.sig.cluster, "mle"],
    Cluster1v2 = -clusters.de[genes.sig.cluster, "mle"]
  ),
  tidy(lm(LT2 ~ Cluster1v2))
)
all.reg <- rbind(
  cbind(d23.reg, Group = "STM-D23580"),
  cbind(lt2.reg, Group = "STM-LT2")
)
all.reg$term <- rep(c("intercept", "slope"), 2)
```

```{r}
gg.regression <- dcast(all.reg, Group ~ term, value.var = "estimate")
gg.regression$x <- 5
gg.regression$y <- -5:-6
```

# Correlation (Pearson) on genes DE between clusters

```{r cor.sig}
cor.d23 <- tidy(cor.test(
  -groups.de.d23[genes.sig.cluster, "mle"],
  clusters.de[genes.sig.cluster, "mle"]
))
cor.lt2 <- tidy(cor.test(
  -groups.de.lt2[genes.sig.cluster, "mle"],
  clusters.de[genes.sig.cluster, "mle"]
))
```

```{r all.cor}
all.cor <- rbind(
  cbind(cor.d23, Group = "STM-D23580"),
  cbind(cor.lt2, Group = "STM-LT2")
)
all.cor$x <- -5
all.cor$y <- 7:6
all.cor$r2 <- sprintf("italic(R)^2~ %.2f", all.cor$estimate^2)
```

```{r range.mle}
range.mle <- range(
  c(gg.sig.cluster$MLE.group, gg.sig.cluster$MLE.cluster)
)
```

```{r}
gg <- ggplot(gg.sig.cluster) +
  geom_point(aes(-MLE.cluster, MLE.group, colour = Group), alpha = 0.25) +
  geom_abline(
    aes(slope = slope, intercept = intercept, colour = Group),
    gg.regression
  ) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) +
  geom_text(aes(x, y, label = r2, colour = Group), all.cor, parse = TRUE) +
  geom_text(aes(
    x, y,
    label = sprintf("slope: %.2f", slope), colour = Group),
    gg.regression
  ) +
  scale_fill_continuous(low = "white", high = "dodgerblue4") +
  theme_minimal() +
  coord_fixed() +
  scale_x_continuous(limits = 8.5*c(-1,1)) +
  scale_y_continuous(limits = 8.5*c(-1,1)) +
  scale_colour_manual(values = colours.infection) +
  labs(y = "MLE\n(Infected / Exposed)", x = "MLE\n(Cluster 1 / Cluster 2)") +
  facet_wrap(~Group, ncol = 1)
ggsave(
  file.path(dirOut, sprintf("regression_%s_clusters_groups.pdf", timepoint)),
  gg,
  height = 8, width = 5.5
)
gg
```

