---
title: "Highly variable genes (HVGs)"
bibliography:
  bibtex.bib
editor_options: 
  chunk_output_type: console
---

<style>
body {
text-align: justify}
</style>

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
stopifnot(
  require(SingleCellExperiment),
  require(scran),
  require(scater),
  require(RColorBrewer),
  require(ggplot2),
  require(ComplexHeatmap),
  require(goseq),
  requireNamespace("dplyr"),
  requireNamespace("MKmisc"),
  require(venn)
)
sce.norm <- readRDS("rds/sce.norm.rds")
sce.endo <- sce.norm[!isSpike(sce.norm, "ERCC"),]
sce.endo <- sce.endo[,with(colData(sce.endo), order(Time, Status, Infection))]
# sce.bulk <- readRDS("rds/sce.bulk.SCE.rds")
# hvg.out <- readRDS("rds/hvg.out.rds")
HVGs.group <- readRDS("rds/HVGs_Group.rds")
dVar.groups <- readRDS("rds/dVar.groups.rds")
# other
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,4,2)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.norm$Time)[1:3]
names(col.infection) <- levels(sce.norm$Infection)
names(col.status) <- levels(sce.norm$Status)
outdir <- "05_HVG_out"; dir.create(outdir, showWarnings = FALSE)
```

## GO enrichment {.tabset}

For convenience, this section separates the code chunks from the result tables
in multiple panels.

### Preprocessing

Let us use again the [goseq](http://bioconductor.org/packages/goseq) package to
identify the most enriched gene ontologies among the HVGs identified
within each experimental group of cells.
Note that we restrict the results to GO categories associated with at least
**10** genes, for robustness:

```{r GO.withinGroup, message=FALSE, warning=FALSE}
GO.HVGs.group <- list()
for (groupName in names(HVGs.group)){
  hvgs <- as.integer(rownames(sce.endo) %in% rownames(HVGs.group[[groupName]]))
  names(hvgs) <- rownames(sce.endo)
  table(hvgs)
  geneLengths <- read.delim(
    "../expdata/counts/WTCHG_305264_201201", row.names = 1)[,1, drop = FALSE]
  pwf <- nullp(hvgs, bias.data=geneLengths[names(hvgs),], plot.fit=FALSE)
  GO.out <- goseq(pwf, "hg38", "ensGene")
  GO.out$ontology <- as.factor(GO.out$ontology)
  write.csv(
    GO.out, file.path(outdir, sprintf("GO_HVGs_%s.csv", groupName)),
    row.names = FALSE)
  GO.out.trim <- dplyr::arrange(
    subset(
      GO.out,
      p.adjust(over_represented_pvalue, "BH") < 0.05 & numInCat >= 10
    ),
    over_represented_pvalue
  )
  GO.HVGs.group[[groupName]] <- GO.out.trim
}
```

## TNFAIP2 over time {.tabset}

In this section, let us examine more closely how the distribution
of *TNFAIP2* expression levels evolves over time within each
experimental group of stimulated cells:

```{r TNFAIP2normExprsMelt, include=FALSE}
TNFAIP2normExprsMelt <- reshape2::melt(
  assay(sce.norm, "logcounts")[
    subset(rowData(sce.endo), gene_name=="TNFAIP2", "gene_id", drop = TRUE),
    sce.norm$Infection != "Mock", drop=FALSE
  ],
  varnames = c("Feature","Sample"), value.name = "logcounts")
TNFAIP2normExprsMelt <- merge(
  TNFAIP2normExprsMelt, colData(sce.norm)[,c("Sample","Treatment","Time")])
TNFAIP2normExprsMelt <- data.frame(merge(
  TNFAIP2normExprsMelt, data.frame(
    Feature = rownames(sce.norm),
    gene_name = rowData(sce.norm)[,"gene_name"])
  ))
```

```{r TNFAIP2}
ggplot(TNFAIP2normExprsMelt) +
  geom_density(aes(logcounts, colour=Time), size = 2) +
  scale_color_manual(values = col.time) +
  facet_wrap(~Treatment) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

## Technical noise fit

We may also plot the biological component of variance against the mean
normalised expression level (log-count) of each gene in each group.

For this, let us first estimate the technical coefficient of variation as a
function of the mean using the [scran](http://bioconductor.org/packages/scran)
`technicalCV2` function for each group of cells:

```{r CV2data, fig.height=9}
CV2data <- data.frame()
for (groupName in names(dVar.groups)){
  techCV2 <- technicalCV2(
    sce.norm[,sce.norm$Group == groupName],
    spike.type = "ERCC", assay.type="counts"
  )
  geneNames <- rownames(techCV2)
  CV2data <- rbind(CV2data, data.frame(
    CV2 = techCV2$cv2,
    meanScaledCount = techCV2$mean,
    trend = techCV2$trend,
    HVG = geneNames %in% rownames(HVGs.group[[groupName]]),
    Time = gsub("(.*h)_.*", "\\1", groupName),
    Infection = gsub(".*h_(.*)_.*", "\\1", groupName),
    Status = gsub(".*h_.*_(.*)", "\\1", groupName),
    Feature = geneNames
  ))
}
```

We may then display the estimated $CV^2$ against the average scaled read count,
while indicating the fitted variance-mean dependence (solid red curve), and
the HVGs detected [earlier](#HVGsWithinGroups)

```{r CV2plot, warning=FALSE, fig.height=7}
ggplot(CV2data) +
  geom_point(aes(meanScaledCount, CV2, colour = HVG), size = 0.2, alpha = 0.3) +
  geom_smooth(
    aes(meanScaledCount, CV2, fill = HVG), colour = "dimgray",
    size = 0.5, alpha = 0.5) +
  scale_x_log10() + scale_y_log10() +
  facet_grid(Infection + Status ~ Time) +
  coord_fixed() +
  labs(x = "Average scaled read count", y = expression(CV^{2})) +
  theme_minimal() +
  theme(
    legend.text = element_text(size=7), legend.title = element_text(size=7)
  )
```


```{r vennExport, include=FALSE}
# Export Venn diagram figures
for (time in levels(sce.norm$Time)){
  groupNames <- grep(time, names(HVGs.group), value = TRUE)
  hvgsList <- lapply(HVGs.group[groupNames], rownames)
  names(hvgsList) <- gsub(paste0(time,"_"), "", names(hvgsList))
  #venn.diagram(hvgsList, filename = sprintf("HVGs/%s_venn.tiff", time))
  pdf(
    file.path(outdir, sprintf("HVGs_%s_venn.pdf", time)),
    width = 6, height = 6)
  venn(hvgsList)
  dev.off()
}
```

```{r HVGs4hStim, warning=FALSE, include=FALSE}
# [earlier tSNE results](06_cluster_normalised#tSNEstatus)
# larger differences between D23 exp/inf relative to LT2
idx.4hStim <- which(sce.norm$Time == "4h" & sce.norm$Infection != "Mock")
dm <- model.matrix(~Plate, data = colData(sce.norm)[idx.4hStim,])
var.fit.endo <- trendVar(
  sce.norm[,idx.4hStim], assay.type="logcounts", method = "loess", design = dm,
  use.spikes = FALSE, span = 0.1
)
var.out.endo.4h <- decomposeVar(sce.norm[,idx.4hStim], var.fit.endo)
plot(
  var.out.endo.4h$mean, var.out.endo.4h$total, pch=16, cex=0.2,
  xlab="Mean log-expression", ylab="Variance of log-expression",
  main = "4h stimulated"
)
o <- order(var.out.endo.4h$mean)
lines(var.out.endo.4h$mean[o], var.out.endo.4h$tech[o], col="dodgerblue", lwd=2)
spike.fit <- trendVar(sce.norm[,idx.4hStim], use.spikes=TRUE)
points(spike.fit$mean, spike.fit$var, col="red", pch=16)
var.out.endo.4h <- cbind(
  Symbol=with(
      rowData(sce.norm), gene_name[match(rownames(var.out.endo.4h),gene_id)]),
  var.out.endo.4h
)
write.csv(
  var.out.endo.4h,
  file.path(outdir, sprintf("HVGs_4h_stimulated_full.csv", group)))
hvg.out <- var.out.endo.4h[which(
  var.out.endo.4h$FDR <= 0.05 & var.out.endo.4h$bio >= 0.5
  ),]
write.csv(
  hvg.out,
  file.path(outdir, sprintf("HVGs_4h_stimulated_significant.csv", group)))
sce.endo.4hStim <- sce.norm[rownames(hvg.out),idx.4hStim]
#sce.endo.4hStim <- sce.endo[,with(as.data.frame(colData(sce.endo)),order(Time,Status,Infection))]
dim(sce.endo.4hStim)
norm.endo.4hStim <- assay(sce.endo.4hStim, "logcounts")
sce.endo.4hStim <-
  sce.endo.4hStim[,with(colData(sce.endo.4hStim), order(Infection,Status))]
dim(norm.endo.4hStim)
h_column <- HeatmapAnnotation(
  df = data.frame(colData(sce.endo.4hStim)[,c("Status", "Infection")]),
  col = list(
    Status = col.status,
    Infection = col.infection
  )
)
ht_supervised <- Heatmap(
  norm.endo.4hStim,
  name = "logcounts", column_title = "Normalised expression",
  top_annotation = h_column,
  # row_order = hvg.endo.clust$order,
  # cluster_rows = hvg.endo.clust, 
  cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
draw(ht_supervised)
```

```{r euclid4h, include=FALSE}
d.e <- dist(t(norm.endo.4hStim), diag = TRUE, upper = TRUE)
mat.e <- as.matrix(d.e)
h.e <- hclust(d.e, method = "ward.D2")
ord.e <- mat.e[h.e$order, h.e$order]
rm(mat.e)
ht_column <- HeatmapAnnotation(
  df = as.data.frame(colData(sce.endo.4hStim)[h.e$order,c("Status", "Infection")]),
  col = list(
    Status = col.status,
    Infection = col.infection
  )
)
Heatmap(
  ord.e, name = "distance", column_title = "Euclidian distance",
  top_annotation = ht_column,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
```


```{r spearman4h, include=FALSE}
c.s <- MKmisc::corDist(t(norm.endo.4hStim), "spearman",diag=TRUE,upper=TRUE)
mat.c.s <- as.matrix(c.s)
h.c.s <- hclust(c.s)
ord.c.s <- mat.c.s[h.c.s$order, h.c.s$order]
rm(c.s)
ht_column <- HeatmapAnnotation(
  df = as.data.frame(colData(sce.endo.4hStim)[h.c.s$order,c("Status", "Infection")]),
  col = list(
    Status = col.status,
    Infection = col.infection
  )
)
Heatmap(
  ord.c.s, name = "distance", column_title = "Spearman absolute correlation distance",
  top_annotation = ht_column,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
```
