---
title: "Differential expression between clusters using *scde* (b)"
bibliography:
  bibtex.bib
editor_options: 
  chunk_output_type: console
---

<style>
body {
text-align: justify}
</style>

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
stopifnot(
  require(scater),
  require(scran),
  require(ggplot2),
  require(ComplexHeatmap),
  require(dplyr),
  require(ggrepel),
  require(RColorBrewer),
  require(goseq),
  require(scde),
  requireNamespace("dplyr")
)
sce.endo <- readRDS("rds/sce.endo_clusters.rds")
o.ifm.2h <- readRDS("rds/o.ifm.2h.rds")
o.ifm.4h <- readRDS("rds/o.ifm.4h.rds")
o.ifm.6h <- readRDS("rds/o.ifm.6h.rds")
col.time <- brewer.pal(9, "Set3")[c(2,7:8)]
col.infection <- brewer.pal(12, "Paired")[c(9,4,2)]
col.status <- brewer.pal(12, "Paired")[c(9,1,7)]
names(col.time) <- levels(sce.endo$Time)[1:3]
names(col.infection) <- levels(sce.endo$Infection)
names(col.status) <- levels(sce.endo$Status)
outdir <- "10_SCDE_cluster_out"; dir.create(outdir, showWarnings = FALSE)
scde.res <- readRDS("rds/scde_cluster.rds")
goseq.res <- readRDS("rds/goseq_cluster_res.rds")
# Subset by time point
sce.2h <- sce.endo[,sce.endo$Time == '2h']
sce.4h <- sce.endo[,sce.endo$Time == '4h']
sce.6h <- sce.endo[,sce.endo$Time == '6h']
# Filter detected genes
filterCounts <- function(m, counts = 10, cells = 10){
  apply(m, 1, function(e){
    return(sum(e >= counts) >= cells)
  })
}
# Reorder SCESet to match error models and filter genes for testing
sce.ifm.2h <- sce.2h[,rownames(o.ifm.2h)]
sg.ifm.2h <- droplevels(sce.ifm.2h$Group)
keep.2h <- filterCounts(counts(sce.ifm.2h))
cd.ifm.2h <- counts(sce.ifm.2h)[keep.2h,]; storage.mode(cd.ifm.2h) <- 'integer'
sce.ifm.4h <- sce.4h[,rownames(o.ifm.4h)]
sg.ifm.4h <- droplevels(sce.ifm.4h$Group)
keep.4h <- filterCounts(counts(sce.ifm.4h))
cd.ifm.4h <- counts(sce.ifm.4h)[keep.4h,]; storage.mode(cd.ifm.4h) <- 'integer'
sce.ifm.6h <- sce.6h[,rownames(o.ifm.6h)]
sg.ifm.6h <- droplevels(sce.ifm.6h$Group)
keep.6h <- filterCounts(counts(sce.ifm.6h))
cd.ifm.6h <- counts(sce.ifm.6h)[keep.6h,]; storage.mode(cd.ifm.6h) <- 'integer'
# Prior of expression
o.prior.2h <- scde.expression.prior(
  models = o.ifm.2h, counts = cd.ifm.2h, show.plot = FALSE
)
o.prior.4h <- scde.expression.prior(
  models = o.ifm.4h, counts = cd.ifm.4h, show.plot = FALSE
)
o.prior.6h <- scde.expression.prior(
  models = o.ifm.6h, counts = cd.ifm.6h, show.plot = FALSE
)
# Functions
convert.z.score <- function(x, one.sided = NULL) {
  z <- x$Z
  if(is.null(one.sided)) {
    pval = pnorm(-abs(z));
    pval = 2 * pval
  } else if(one.sided=="-") {
    pval = pnorm(z);
  } else {
    pval = pnorm(-z);
  }
    x <- cbind(
      x,
      p.value = pval
  )
  return(x);
}   
addGeneName <- function(x){
  x <- cbind(
    gene_name = with(rowData(sce.endo), gene_name[match(rownames(x), gene_id)]),
    x
  )
  return(x)
}
orderResults <- function(x){
  x <- x[with(x, order(abs(Z), decreasing = TRUE)),]
  return(x)
}
sig.levels <- c(0.05, 0.01)
volcano.sig <- data.frame(
  p.value = sig.levels,
  level = as.character(sig.levels)
)
volcano.mle <- function(x, sub = NULL){
  varName <- deparse(substitute(x))
  x <- convert.z.score(x)
  gg <- ggplot(x, aes(mle, -log10(p.value))) +
    geom_point(aes(colour = (cZ != 0))) +
    geom_hline(aes(yintercept=-log10(p.value),linetype=level), volcano.sig) +
    ggtitle(varName, sub)
  print(gg)
  return(x)
}
```

<!-- Cleaner code based on 05_SCDE
Remove ERCC spike-ins
!! Split cells by time
!! Retain genes detected at each time (10 counts, 5 cells, 1 group)
Identify cell clusters @ time
<!-- Error model: {cells @ time; genes @ time}
Prior @ time
DE: {group on all cells with NA; no batch correction; P-value < 0.01}
-->

# Gene ontology

Let us use the [goseq](http://bioconductor.org/packages/goseq) package to
identify the most enriched gene ontologies among the various lists of DE genes.
Note that we restrict the results to GO categories associated with at least
**10** genes, for robustness.

Let us first define:

* a list to store the GO enrichment tables returned by *goseq*

```{r goseq.res_init, eval=FALSE}
goseq.res <- list()
```

* the gene length information reported by *featureCounts*:

```{r geneLengths}
geneLengths <- width(sce.endo)
names(geneLengths) <- rownames(sce.endo)
```

Let us identify over- and under-represented GO categories in DE genes
identified in each contrast,
using as background the list of genes detected at each time point:

## Estimate enrichment {.tabset}

### 2h (2 vs 1)

```{r GO_2h_DE, eval=FALSE}
contrastName <- "2h_cluster2-cluster1"
bg.2h <- rownames(cd.ifm.2h)
scde.table <- scde.res[[contrastName]]
scde.table <- convert.z.score(scde.table)
scde.table <- subset(scde.table, p.value < 0.01)
de.genes <- (bg.2h %in% rownames(scde.table))
names(de.genes) <- bg.2h; table(de.genes)
pwf <- nullp(de.genes, bias.data = geneLengths[names(de.genes)])
go.res <- goseq(pwf, "hg38", "ensGene")
goseq.res[[contrastName]] <- go.res
```

### 4h (2 vs 1)

```{r GO_4h_DE, eval=FALSE}
contrastName <- "4h_cluster2-cluster1"
bg.4h <- rownames(cd.ifm.4h)
scde.table <- scde.res[[contrastName]]
scde.table <- convert.z.score(scde.table)
scde.table <- subset(scde.table, p.value < 0.01)
de.genes <- (bg.4h %in% rownames(scde.table))
names(de.genes) <- bg.4h; table(de.genes)
pwf <- nullp(de.genes, bias.data = geneLengths[names(de.genes)])
go.res <- goseq(pwf, "hg38", "ensGene")
goseq.res[[contrastName]] <- go.res
```

### 6h (2 vs 1)

```{r GO_6h_DE, eval=FALSE}
contrastName <- "6h_cluster2-cluster1"
bg.6h <- rownames(cd.ifm.6h)
scde.table <- scde.res[[contrastName]]
scde.table <- convert.z.score(scde.table)
scde.table <- subset(scde.table, p.value < 0.01)
de.genes <- (bg.6h %in% rownames(scde.table))
names(de.genes) <- bg.6h; table(de.genes)
pwf <- nullp(de.genes, bias.data = geneLengths[names(de.genes)])
go.res <- goseq(pwf, "hg38", "ensGene")
goseq.res[[contrastName]] <- go.res
```

```{r goseq.res, echo=FALSE, eval=FALSE}
saveRDS(goseq.res, "rds/goseq_cluster_res.rds")
for (contrastName in names(goseq.res)){
  goseq.table <- goseq.res[[contrastName]]
  csvFile <- sprintf("SCDE_cluster_direct_GO_%s.csv", contrastName)
  write.csv(goseq.table, file.path(outdir, csvFile))
}
```

```{r include=FALSE}
# testing
geneName <- "IL1A"
x_time <- "4h"
geneId <- subset(rowData(sce.endo), gene_name == geneName, "gene_id", drop = TRUE)
idxCells <- which(sce.endo$Time %in% x_time)
ggdata <- data.frame(
  logcounts = assay(sce.endo, "logcounts")[geneId, idxCells],
  cluster = colData(sce.endo)[idxCells, sprintf("quickCluster.%s", x_time)],
  colData(sce.endo)[idxCells, c("Time","Infection","Status")]
)
gg <- ggplot(ggdata, aes(cluster, logcounts, color = Status)) +
  geom_violin(aes(group = cluster), alpha = 0.3) +
  geom_jitter(width = 0.2, height = 0, size = 3, alpha = 0.75) +
  scale_color_manual(values = col.status) +
  labs(title = sprintf("%s | %s", geneName, x_time)) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
ggsave(file.path(outdir, sprintf("%s_%s_norm.pdf", geneName, x_time)), gg, width = 6, height = 4)
```
