---
title: Quality Control
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Quality control of endogenous features in single cells that passed quality
  control.
vignette: >
  %\VignetteIndexEntry{Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2:
    toc_newpage: true
bibliography:
  QC.bib
---

# Prior work

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks) was obtained using
[Kallisto](https://pachterlab.github.io/kallisto/)
as follows:

```
kallisto quant \
    --index=$kallistoIdx \
    --output-dir=$kallistoDir/$cell \
    --bootstrap-samples=100 \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```

Subsequently:

* The quantitation data was imported into the _R_ environment.
* Phenotype data was attached imported and attached to the `SCESet`
* Transcript-level expression data was aggregated at the gene level
* Gene identifiers were made more intuitive by prepending gene symbol where
  possible
* Outliers were identified and removed among single cells,
  using blank and bulk samples as reference for certain metrics.
* The remaining single cells were serialised as a `SCESet` and written to file.

# Settings

Packages required for this vignette:

```{r packages, message=FALSE}
library(scater)
library(ensembldb)
library(EnsDb.Hsapiens.v79)
library(scran)
library(reshape2)
```

Let us set the working directory to execute the following code chunks:

```{r workdir}
workdir <- ifelse(interactive(), ".", "..")
```

Let us set a few paths relevant to the analysis:

```{r paths}
folder.rds <- file.path(workdir, "rds")
folder.expdata <- file.path(workdir, "expData")
```

# Import preprocessed data

Here, we simply import the preprocessed data:

```{r DC_all}
DC_sc <- readRDS(file.path(folder.rds, "DC_goodCells.rds"))
```

# Overwrite QC metrics

In the previous step, QC metrics were calculated using _all_ single cells
in the data set. This has not updated certain metrics depending on the number
of cells, such as `n_cells_exprs`
(The number of cells for which the expression level of the feature is above the
detection limit).
To update those metrics, let us call once more the `calculateQCMetrics`
method:

```{r calculateQCMetrics}
MT_ids <- genes(
  EnsDb.Hsapiens.v79, "gene_id", SeqnameFilter("MT", "=")
  )$gene_id
DC_sc <- calculateQCMetrics(
  DC_sc,
  feature_controls = list(
    ERCC = grepl("ERCC-[[:digit:]]", featureNames(DC_sc)),
    MT = fData(DC_sc)$feature_id %in% MT_ids),
  nmads = 3
)
```

# Visualise gene average expression and cell detection level

In order to select expressed features appropriate for downstream analyses,
let us visualise, for each measured feature (including ERCC spike-in features),
the count of cells expressing each feature against the log~10~-transformed
average estimated counts.

First, let us calculate the average estimate counts for each gene:

```{r avgCounts}
ave.counts <- rowMeans(counts(DC_sc))
```

We may then identify features showing an average estimated count above `1`:

```{r avgCountsGe1}
keep <- ave.counts >= 1
table(keep)
```

Let us visualise the distribution of average estimated counts,
and mark the cut-off value of `1` used to remove features associated
with excessive dropout events:

```{r histAvgCount}
hist(
  log10(ave.counts), breaks=100, main="", col="grey80",
  xlab=expression(Log[10]~"average count")
)
abline(v=log10(1), col="blue", lwd=2, lty=2)
```

In this plot, the cut-off line crosses the peak, rather than the more flat
region on the left of the peak, as suggested in [@67].
However, the fact that this data set includes more cells allows the
average estimated counts to reach lower values
(_e.g._ a single estimated count in a single cell, divided by `400` cells
produced a value of `0.0025`, while in `100` cells it produces a minimal value
of `0.01`).

It is therefore important to represent the average estimated count value
against the number of cells in which each feature is detected,
and mark the cutoff value of `1` as above:

```{r smoothScatter}
numcells <- nexprs(DC_sc, byrow=TRUE)
setSpike(DC_sc) <- "ERCC"
smoothScatter(
  log10(ave.counts), numcells,
  xlab = expression(Log[10]~"average count"),
  ylab = "Number of expressing cells"
)
points(
  log10(ave.counts[isSpike(DC_sc)]),
  numcells[isSpike(DC_sc)], col = "red", pch = 16, cex = 0.5)
abline(v = 0, col = "orange")
rm(numcells)
```

# Normalisation

This section applies the normalisation procedure as presented in [@67]
for the haematopoietic stem cells (HSCs) data set.

## Cell-specific biases

Let us compute size factors using the `r Biocpkg("scran")` `computeSumFactors`
method:

```{r computeSumFactors}
DC_sc <- computeSumFactors(DC_sc, sizes=c(20, 40, 60, 80))
summary(sizeFactors(DC_sc))
```

The range of size factors is similar to that of the HSCs data set.
This may indicate that the outlier detection procedure applied in the previous
vignette has successfully removed extreme outliers, leaving an homogeneous
data set (as homogeneous as can bean experimental design including three
time points, uninfected cells, and cells exposed to or infected by two
different bacterial strains).

```{r plotsizeFactors}
plot(sizeFactors(DC_sc), DC_sc$total_counts/1e6, log="xy",
    ylab="Library size (millions)", xlab="Size factor")
cor.test(sizeFactors(DC_sc), DC_sc$total_counts/1e6, alternative = "greater")
```

In contrast to [@67], the size factors computed for this data set do not
show an important scatter around the trend of correlation.
This suggests that the systematic differences between cells are _not_ primarily
driven by differences in capture efficiency or sequencing depth.
Instead, this observation hints to the existence of differential expression
between single cells, which is expected from the experimental design.

## Separate size factors for spike-in transcripts

Size factors computed from the counts for endogenous genes are usually not
appropriate for normalising the counts for spike-in transcripts.
To ensure normalisation is performed correctly, we compute a separate set of
size factors for the spike-in set.

```{r computeSpikeFactors}
DC_sc <- computeSpikeFactors(DC_sc, type="ERCC", general.use=FALSE)
```

## Applying the size factors to normalize gene expression

As explained in [@67], the (estimated) count data are used to compute
normalised log-expression values for use in downstream analyses.
Each value is de ned as the log-ratio of each count to the size factor for the
corresponding cell, after adding a prior count of `1` to avoid unde ned values
at zero counts.
Division by the size factor ensures that any cell-specific biases are removed.

Spike-in-specific size factors are automatically applied to normalize the
spike-in transcripts separately from the endogenous genes.

```{r normalize}
DC_sc <- normalize(DC_sc)
```

# Visualise

## PCA

Let us obtain a first overview of the normalised data set using a
principal component analysis.

```{r prcompSamples}
dc_pca.samples <- prcomp(t(assayData(DC_sc)[["norm_exprs"]]))
```

We may then use colours and shapes to mark experimental groups of single
cells, and identify the primary experimental factors that correlate with
the first two components of highest variance in the data set.
We may also add 2D-density lines estimated from the 2D distribution
of single cells obtained at the three different time points (irrespective of
`Stimulation` or `Infection`):

```{r pcaSamples, fig.height=10}
pca.samples <- cbind(dc_pca.samples$x, pData(DC_sc))
dim(pca.samples)
multiplot(
  ggplot(pca.samples, aes(x = PC1, y = PC2, shape = Time)) +
    geom_point(aes(colour = Status)) +
    geom_density_2d(aes(linetype = Time)),
  ggplot(pca.samples, aes(x = PC1, y = PC2, shape = Time)) +
    geom_point(aes(colour = Infection)) +
    geom_density_2d(aes(linetype = Time)),
  cols = 1
)
```

Note how 2D density lines emphasise the increasing divergence between
uninfected and stimulated cells over time; at `6` hours post-infection,
the uninfected cells are given their own set of contours by the 2D kernel
density estimation.

Similarly to principal component analysis prior to outlier detection and
removal, `Time` remains the experimental factor most correlated with `PC1`,
explaining most of the variance in the filtered data set.
In contrast, `PC2` now separates primarily uninfected cells from stimulated
cells;
principal component analysis of the data set before outlier detection showed
a clear correlation of `PC2` with library size,
owing to presence of important outliers for their excessively low library size,
low total number of features detected, or high content of reads originating
from mitochondrial genes.

```{r multiPCA}
multiplot(
  ggplot(pca.samples, aes(x = PC1, y = PC2, shape = Time)) +
    geom_point(aes(colour = Status)),
  ggplot(pca.samples, aes(x = PC1, y = PC3, shape = Time)) +
    geom_point(aes(colour = Infection)),
  ggplot(pca.samples, aes(x = PC2, y = PC3, shape = Time)) +
    geom_point(aes(colour = Infection)),
  ggplot(pca.samples, aes(x = PC1, y = PC4, shape = Time)) +
    geom_point(aes(colour = Infection)),
  cols = 2
)
```

# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
