---
title: Quality Control
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Quality control of endogenous features in single cells that passed quality
  control.
vignette: >
  %\VignetteIndexEntry{Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2:
    toc_newpage: true
---

# Prior work

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks) was obtained using
[Kallisto](https://pachterlab.github.io/kallisto/)
as follows:

```
kallisto quant \
    --index=$kallistoIdx \
    --output-dir=$kallistoDir/$cell \
    --bootstrap-samples=100 \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```

Subsequently:

* The quantitation data was imported into the _R_ environment.
* Phenotype data was attached imported and attached to the `SCESet`
* Transcript-level expression data was aggregated at the gene level
* Gene identifiers were made more intuitive by prepending gene symbol where
  possible
* Outliers were identified and removed among single cells,
  using blank and bulk samples as reference for certain metrics.
* The remaining single cells were serialised as a `SCESet` and written to file.

# Settings

Packages required for this vignette:

```{r packages, message=FALSE}
library(scater)
library(ensembldb)
library(EnsDb.Hsapiens.v79)
library(scran)
library(reshape2)
```

Let us set the working directory to execute the following code chunks:

```{r workdir}
workdir <- ifelse(interactive(), ".", "..")
```

Let us set a few paths relevant to the analysis:

```{r paths}
folder.rds <- file.path(workdir, "rds")
folder.expdata <- file.path(workdir, "expData")
```

# Import preprocessed data

Here, we simply import the preprocessed data:

```{r DC_all}
DC_sc <- readRDS(file.path(folder.rds, "DC_goodCells.rds"))
```

# Overwrite QC metrics

In the previous step, QC metrics were calculated using _all_ single cells
in the data set. This has not updated certain metrics depending on the number
of cells, such as `n_cells_exprs`
(The number of cells for which the expression level of the feature is above the
detection limit).
To update those metrics, let us call once more the `calculateQCMetrics`
method:

```{r calculateQCMetrics}
MT_ids <- genes(
  EnsDb.Hsapiens.v79, "gene_id", SeqnameFilter("MT", "=")
  )$gene_id
DC_sc <- calculateQCMetrics(
  DC_sc,
  feature_controls = list(
    ERCC = grepl("ERCC-[[:digit:]]", featureNames(DC_sc)),
    MT = fData(DC_sc)$feature_id %in% MT_ids),
  nmads = 3
)
```

# Visualise gene average expression and cell detection level

In order to select expressed features appropriate for downstream analyses,
let us visualise, for each measured feature (including ERCC spike-in features),
the count of cells expressing each feature against the log~10~-transformed
average estimated counts.

First, let us calculate the average estimate counts for each gene:

```{r avgCounts}
ave.counts <- rowMeans(counts(DC_sc))
```

We may then identify features showing an average estimated count above `1`:

```{r avgCountsGe1}
keep <- ave.counts >= 1
table(keep)
```

Let us visualise the distribution of average estimated counts,
and mark the cut-off value of `1` used to remove features associated
with excessive dropout events:

```{r histAvgCount}
hist(
  log10(ave.counts), breaks=100, main="", col="grey80",
  xlab=expression(Log[10]~"average count")
)
abline(v=log10(1), col="blue", lwd=2, lty=2)
```

In this plot, the cut-off line crosses the peak, rather than the more flat
region on the left of the peak, as suggested .
However, the fact that this data set includes more cells allows the
average estimated counts to reach lower values
(_e.g._ a single estimated count in a single cell, divided by `400` cells
produced a value of `0.0025`, while in `100` cells it produces a minimal value
of `0.01`).

It is therefore important to represent the average estimated count value
against the number of cells in which each feature is detected,
and mark the cutoff value of `1` as above:

```{r smoothScatter}
numcells <- nexprs(DC_sc, byrow=TRUE)
setSpike(DC_sc) <- "ERCC"
smoothScatter(
  log10(ave.counts), numcells,
  xlab = expression(Log[10]~"average count"),
  ylab = "Number of expressing cells"
)
points(
  log10(ave.counts[isSpike(DC_sc)]),
  numcells[isSpike(DC_sc)], col = "red", pch = 16, cex = 0.5)
abline(v = 0, col = "orange")
rm(numcells)
```

# Normalisation

```{r computeSumFactors}
DC_sc <- computeSumFactors(DC_sc, sizes=c(20, 40, 60, 80))
summary(sizeFactors(DC_sc))
```

```{r plotsizeFactors}
plot(sizeFactors(DC_sc), DC_sc$total_counts/1e6, log="xy",
    ylab="Library size (millions)", xlab="Size factor")
```

```{r computeSpikeFactors}
DC_sc <- computeSpikeFactors(DC_sc, type="ERCC", general.use=FALSE)
```

```{r normalize}
DC_sc <- normalize(DC_sc)
```

# Visualise

## PCA

```{r prcompSamples}
dc_pca.samples <- prcomp(t(assayData(DC_sc)[["norm_exprs"]]))
```

```{r pcaSamples}
pca.samples <- cbind(dc_pca.samples$x, pData(DC_sc))
dim(pca.samples)
multiplot(
  ggplot(pca.samples, aes(x = PC1, y = PC2, shape = Time)) +
    geom_point(aes(colour = Status)) +
    geom_density_2d(aes(linetype = Time)),
  ggplot(pca.samples, aes(x = PC1, y = PC2, shape = Time)) +
    geom_point(aes(colour = Infection)) +
    geom_density_2d(aes(linetype = Time)),
  cols = 1
)
```

```{r, include=FALSE}
multiplot(
  ggplot(pca.samples, aes(x = PC1, y = PC2, shape = Time)) +
    geom_point(aes(colour = Status)),
  ggplot(pca.samples, aes(x = PC1, y = PC3, shape = Time)) +
    geom_point(aes(colour = Infection)),
  ggplot(pca.samples, aes(x = PC2, y = PC3, shape = Time)) +
    geom_point(aes(colour = Infection)),
  ggplot(pca.samples, aes(x = PC1, y = PC4, shape = Time)) +
    geom_point(aes(colour = Infection)),
  cols = 2
)
```



```{r}
ggplot(pca.samples.long) +
  geom_point(aes(x = Coordinate, y = Coordinate, colour = Status, shape = Time)) +
  facet_grid(PC ~ PC)
```



# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
