---
title: Distance matrix
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Visualise pairwise distance between single cells using normalised gene
  expression.
vignette: >
  %\VignetteIndexEntry{Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2:
    toc_newpage: true
bibliography:
  QC.bib
---

# Prior work

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks) was obtained using
[Kallisto](https://pachterlab.github.io/kallisto/)
as follows:

```
kallisto quant \
    --index=$kallistoIdx \
    --output-dir=$kallistoDir/$cell \
    --bootstrap-samples=100 \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```

Subsequently:

* The quantitation data was imported into the _R_ environment.
* Phenotype data was attached imported and attached to the `SCESet`
* Transcript-level expression data was aggregated at the gene level
* Gene identifiers were made more intuitive by prepending gene symbol where
  possible
* Outliers were identified and removed among single cells,
  using blank and bulk samples as reference for certain metrics.
* Endogenous and spike-in features were filtered to retain only those
  detected in the data set
* The remaining single cells and features were serialised as a `SCESet` and
  written to file.

# Settings

Packages required for this vignette:

```{r packages, message=FALSE}
library(scater)
library(matrixStats)
library(MKmisc)
library(ComplexHeatmap)
library(RColorBrewer)
```

Let us set the working directory to execute the following code chunks:

```{r workdir}
workdir <- ifelse(interactive(), ".", "..")
```

Let us set a few paths relevant to the analysis:

```{r paths}
folder.rds <- file.path(workdir, "rds")
folder.expdata <- file.path(workdir, "expData")
```

# Import preprocessed data

Let us import the normalised data:

```{r DC_all}
sc_norm <- readRDS(file.path(folder.rds, "normalised.rds"))
```

Let us immediately subset the data set to retain only endogenous features:

```{r excludeERCC}
endo_sc <- sc_norm[!fData(sc_norm)$is_feature_control_ERCC,]
```

Let us also reorder samples by `Time`, `Status`, and `Infection`:

```{r orderSamples}
endo_sc <- endo_sc[,order(endo_sc$Time, endo_sc$Status, endo_sc$Infection)]
```

Let us also order genes by decreasing variance:

```{r}
rv <- rowVars(assayData(endo_sc)[["norm_exprs"]])
endo_sc <- endo_sc[order(rv, decreasing = TRUE),]
rm(rv)
```

Let us focus on the most variable genes, centre and scale their expression
values across all samples:

```{r}
top300 <- assayData(endo_sc)[["norm_exprs"]][1:300,]
top300.scaled <- t(scale(t(top300)))
```

# Display expression data as supervised clusters

Let us prepare colours to indicate phenotypes:

```{r palette9}
col9 <- brewer.pal(10, "Set3")
```

We may then visualised normalised and scaled expression values in
a heat map manually organised by phenotype, while allowing genes with
similar expression profiles to cluster:

```{r heatManual}
h_column = HeatmapAnnotation(
  df = pData(endo_sc)[,c("Time", "Status", "Infection")],
  col = list(
    Time = c("2h" = col9[1], "4h" = col9[2], "6h" = col9[3]),
    Status=c("uninfected"=col9[4],"exposed"=col9[5],"infected"=col9[6]),
    Infection = c("Mock" = col9[7], "LT2" = col9[8], "D23580" = col9[10])
  )
)
ht = Heatmap(
  top300,
  name = "norm_exprs", column_title = "Normalised expression",
  top_annotation = h_column,
  cluster_rows = TRUE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE,
  show_row_dend = TRUE
)
ht.scaled = Heatmap(
  top300.scaled,
  name = "scaled_exprs", column_title = "Scaled expression",
  top_annotation = h_column,
  cluster_rows = TRUE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE,
  show_row_dend = TRUE
)
```

```{r heatManualNorm, fig.height=8}
draw(ht)
```

```{r heatManualScaled, fig.height=8}
draw(ht.scaled)
```

In this case, the heat map that displays normalised (non-scaled) expression
is more informative, as it gives a better impression of the _relative_
expression of different genes within and between samples that were manually
clustered to emphasise differences and similiarities between experimental
groups.

In contrast, scaled expression values are more relevant in following sections
to perform unsupervised clustering of samples while preventing distance metrics
such as the Euclidian distance from being overwhelmed by highly expressed genes.

# Pairwise distances between samples

## Euclidian distance between most variable genes

### Estimate distances

Calculate the Euclidian distance between samples using the
scaled normalised expression values calculated earlier:

```{r distEuclidScaled}
d.e.s <- dist(t(top300.scaled), diag = TRUE, upper = TRUE)
mat.e.s <- as.matrix(d.e.s)
```

For demonstration purpose, do the same with normalised (unscaled) values:

```{r distEuclidNorm}
d.e.n <- dist(t(top300), diag = TRUE, upper = TRUE)
mat.e.n <- as.matrix(d.e.n)
```

Cluster samples according to their Euclidian distance for scaled values:

```{r clustEuclidScaled}
h.e.s <- hclust(d.e.s)
ord.e.s <- mat.e.s[h.e.s$order, h.e.s$order]
rm(mat.e.s)
```

Again, do the same for unscaled values:

```{r clustEuclidNorm}
h.e.n <- hclust(d.e.n)
ord.e.n <- mat.e.n[h.e.n$order, h.e.n$order]
rm(mat.e.n)
```

### Display heat maps

Draw the heatmaps using Euclidian distance for normalised or
normalised-scaled data:

```{r heatEuclid}
hs_column = HeatmapAnnotation(
  df = pData(endo_sc)[h.e.s$order,c("Time", "Status", "Infection")],
  col = list(
    Time = c("2h" = col9[1], "4h" = col9[2], "6h" = col9[3]),
    Status=c("uninfected"=col9[4],"exposed"=col9[5],"infected"=col9[6]),
    Infection = c("Mock" = col9[7], "LT2" = col9[8], "D23580" = col9[10])
  )
)
hs = Heatmap(
  ord.e.s, name = "norm_scaled", column_title = "Normalised scaled",
  top_annotation = hs_column,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
hn_column = HeatmapAnnotation(
  df = pData(endo_sc)[h.e.n$order,c("Time", "Status", "Infection")],
  col = list(
    Time = c("2h" = col9[1], "4h" = col9[2], "6h" = col9[3]),
    Status=c("uninfected"=col9[4],"exposed"=col9[5],"infected"=col9[6]),
    Infection = c("Mock" = col9[7], "LT2" = col9[8], "D23580" = col9[10])
  )
)
hn = Heatmap(
  ord.e.n, name = "norm", column_title = "Normalised unscaled",
  top_annotation = hn_column,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
```

```{r heatEuclidNorm, fig.height=8}
draw(hn)
```

```{r heatEuclidScaled, fig.height=8}
draw(hs)
```

## Correlation between the most variable genes

### Spearman

Estimate distance between sample using Spearman's (rank) correlation.

\bioccomment{
The rank correlation is insensitive to the scaling process (the order
of samples are preserved by the scaling procedure). Therefore, there is
no need to compare the identical heat maps produced for the two data sets.
}

```{r distSpearman}
c.s <- corDist(t(top300), "spearman", diag = TRUE, upper = TRUE)
mat.c.s <- as.matrix(c.s)
```

Cluster samples:

```{r clustSpearman}
h.c.s <- hclust(c.s)
ord.c.s <- mat.c.s[h.c.s$order, h.c.s$order]
rm(c.s)
```

### Pearson

Estimate distance between sample using Pearson's correlation.
Let us compare the effect of scaling the expression data (by gene)
on the correlation between samples.

We start calculating the sample correlation using the unscaled normalised data:

```{r distPearsonNorm}
c.p.n <- corDist(t(top300), diag = TRUE, upper = TRUE)
mat.c.p.n <- as.matrix(c.p.n)
```

We then do the same with the scaled data:

```{r distPearsonScaled}
c.p.s <- corDist(t(top300.scaled), diag = TRUE, upper = TRUE)
mat.c.p.s <- as.matrix(c.p.s)
```

Cluster samples using Pearson's correlation for the unscaled data:

```{r clustPearsonNorm}
h.c.p.n <- hclust(c.p.n)
ord.c.p.n <- mat.c.p.n[h.c.p.n$order, h.c.p.n$order]
rm(c.p.n)
```

And the same for the scaled data:

```{r clustPearsonScaled}
h.c.p.s <- hclust(c.p.s)
ord.c.p.s <- mat.c.p.s[h.c.p.s$order, h.c.p.s$order]
rm(c.p.s)
```

# Heat maps


Draw the heat map using Spearman's _rho_ correlation:

```{r heatSpearman, fig.height=8}
ht_column = HeatmapAnnotation(
  df = pData(endo_sc)[h.c.s$order,c("Time", "Status", "Infection")],
  col = list(
    Time = c("2h" = col9[1], "4h" = col9[2], "6h" = col9[3]),
    Status=c("uninfected"=col9[4],"exposed"=col9[5],"infected"=col9[6]),
    Infection = c("Mock" = col9[7], "LT2" = col9[8], "D23580" = col9[10])
  )
)
ht = Heatmap(
  ord.c.s, name = "1 - r", column_title = "Spearman distance",
  top_annotation = ht_column,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
draw(ht)
```

Draw the heat maps using Pearson's _r_ correlation:

```{r heatPearson}
hs_column = HeatmapAnnotation(
  df = pData(endo_sc)[h.c.p.s$order,c("Time", "Status", "Infection")],
  col = list(
    Time = c("2h" = col9[1], "4h" = col9[2], "6h" = col9[3]),
    Status=c("uninfected"=col9[4],"exposed"=col9[5],"infected"=col9[6]),
    Infection = c("Mock" = col9[7], "LT2" = col9[8], "D23580" = col9[10])
  )
)
hs = Heatmap(
  ord.c.p.s, name = "scaled (1 - r)", column_title = "Normalised scaled",
  top_annotation = hs_column,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
hn_column = HeatmapAnnotation(
  df = pData(endo_sc)[h.c.p.n$order,c("Time", "Status", "Infection")],
  col = list(
    Time = c("2h" = col9[1], "4h" = col9[2], "6h" = col9[3]),
    Status=c("uninfected"=col9[4],"exposed"=col9[5],"infected"=col9[6]),
    Infection = c("Mock" = col9[7], "LT2" = col9[8], "D23580" = col9[10])
  )
)
hn = Heatmap(
  ord.c.p.n, name = "norm (1 - r)", column_title = "Normalised unscaled",
  top_annotation = hn_column,
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE
)
```

```{r heatPearsonNorm, fig.height=8}
draw(hn)
```

Heat map based on the normalised-scaled data:

```{r heatPearsonScaled, fig.height=8}
draw(hs)
```

# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
