---
title: Correlation with align-and-count pipeline
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Compare the expression values obtained with _Kallisto_ with those obtained
  with _HISAT2_ and _featureCounts_.
vignette: >
  %\VignetteIndexEntry{Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2:
    toc_newpage: true
bibliography:
  QC.bib
---

# Prior work

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks) was obtained using
[Kallisto](https://pachterlab.github.io/kallisto/)
as follows:

```
kallisto quant \
    --index=$kallistoIdx \
    --output-dir=$kallistoDir/$cell \
    --bootstrap-samples=100 \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```

Subsequently:

* The quantitation data was imported into the _R_ environment.
* Phenotype data was attached imported and attached to the `SCESet`
* Transcript-level expression data was aggregated at the gene level
* Gene identifiers were made more intuitive by prepending gene symbol where
  possible
* Outliers were identified and removed among single cells,
  using blank and bulk samples as reference for certain metrics.
* Endogenous and spike-in features were filtered to retain only those
  detected in the data set
* The remaining single cells and features were serialised as a `SCESet` and
  written to file.

# Settings

Packages required for this vignette:

```{r packages, message=FALSE}
library(scater)
```

Let us set the working directory to execute the following code chunks:

```{r workdir}
workdir <- ifelse(interactive(), ".", "..")
```

Let us set a few paths relevant to the analysis:

```{r paths}
folder.rds <- file.path(workdir, "rds")
folder.expdata <- file.path(workdir, "expData")
```

# Import preprocessed data

Let us import the normalised data:

```{r DC_all}
sc_norm <- readRDS(file.path(folder.rds, "normalised.rds"))
```

Let us immediately subset the data set to retain only endogenous features:

```{r excludeERCC}
endo_sc <- DC_sc[!fData(DC_sc)$is_feature_control_ERCC,]
```

# 

# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
