---
title: Comparison of pseudo-aligned and aligned data sets
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Comparison of pseudo-aligned data obtained using _Kallisto_ to aligned data
  obtained using _Trim Galore!_, _HISAT2_, and _featureCounts_.
vignette: >
  %\VignetteIndexEntry{Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2:
    toc_newpage: true
---
  
# Prior work

## Kallisto

Prior to this document:
* Quantitation data was obtained using
  [Kallisto](https://pachterlab.github.io/kallisto/)
* The quantitation data was imported into the _R_ environment as a `SCESet`.
* Phenotype data was attached to the `SCESet`.
* Transcript-level expression data was aggregated into gene-level
expression estimates.
* Gene identifiers were made more intuitive by prepending gene symbol where
    possible 
* Outlier samples were identified and removed from the `SCESet`
* Detected features were selected
  + To compare the pseudo-aligned and aligned data sets, we consider in this
    vignette _all_ features measured in both platforms,
    even those below detection level in either of them.
* The resulting `SCESet` was serialised and written to file.

The pipeline is described in more details across vignettes _01_ and _02_.
  
## HISAT2

An alignment-based pipeline was implemented using:

* [Trim Galore!](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore)
* [HISAT2](https://ccb.jhu.edu/software/hisat2/index.shtml)
* [featureCounts](http://bioinf.wehi.edu.au/featureCounts)

The pipeline is described in more details in _Vignette 05_.

# Settings

Packages required for this vignette:

```{r packages, message=FALSE}
library(scater)
library(edgeR)
library(dplyr)
library(ggplot2)
library(EnsDb.Hsapiens.v79)
```

Let us set the working directory to execute the following code chunks:

```{r workdir}
workdir <- ifelse(interactive(), ".", "..")
```

Let us set a few paths relevant to the analysis:

```{r paths}
folder.rds <- file.path(workdir, "rds")
```

# Import preprocessed data

## Kallisto

Here, we simply import the preprocessed data for all samples,
including bulks and blanks:

```{r DC_all}
kallisto_all <- readRDS(file.path(folder.rds, "all_EnsDb.rds"))
```

Let us examine a summary of the experimental phenotype data in the full
data set:

```{r expdata}
expdata <- droplevels(
  pData(kallisto_all)[,c("Infection","Status","Time","Lane","Plate")]
)
expdata$Lane <- as.factor(expdata$Lane)
summary(expdata)
```

## Alignment pipeline

```{r readDGE}
expdata$files <- file.path(sprintf(
  "WTCHG_%s_%s", expdata$Lane, rownames(expdata)
))
expdata$group <- paste(
  expdata$Time, expdata$Infection, expdata$Status,
  sep = "_"
)
```

Let us import the data previously serialised and saved to file:

```{r}
dge_all <- readRDS(file.path(folder.rds, "dge_all.rds"))
```

# Overview of alignment pipeline

# Bulks

Let us compare the estimated counts produced by _Kallisto_ to the count
obtained through the alignment pipeline in each `BULK` sample.

First, let us retain only the features identified in both platforms:

```{r}
features.overlap <- intersect(rownames(dge_all), fData(kallisto_all)$feature_id)
kallisto.overlap <- kallisto_all[
  match(features.overlap, fData(kallisto_all)$feature_id),
  ]
dge.overlap <- dge_all[features.overlap,]
stopifnot(all(fData(kallisto.overlap)$feature_id == rownames(dge.overlap)))
rm(kallisto_all, dge_all)
```

Then, let us identify this comparison to only endogenous features:

```{r endogenous}
ercc <- grepl("^ERCC-", features.overlap)
table(ercc)
```

Let us then collect for each experimental group, the estimated counts
obtained using _Kallisto_) and the counts using the alignment pipeline:

```{r, message=FALSE}
comp.bulks <- data.frame(
  Time = character(),
  Infection = character(),
  Kallisto = integer(),
  THF = integer(),
  ERCC = logical()
)
for (time in c("2h", "4h", "6h")){
  for (infection in c("Mock", "D23580", "LT2")){
    message(sprintf("%s - %s", time, infection))
    kallisto <- counts(kallisto.overlap[
      ,
      kallisto.overlap$Time == time &
        kallisto.overlap$Infection == infection &
        kallisto.overlap$Status == "BULK"
    ])[,1]
    thf = dge.overlap$counts[
      ,
      dge.overlap$samples$Time == time &
        dge.overlap$samples$Infection == infection &
        dge.overlap$samples$Status == "BULK"
    ]
    new.df <- data.frame(
      Time = time,
      Infection = infection,
      Kallisto = kallisto,
      THF = thf,
      ERCC = ercc
    )
    comp.bulks <- rbind(comp.bulks, new.df)
  }
}
rm(kallisto, thf, time, infection, new.df)
```

Let us now plot the log~10~-transformed
estimated counts obtained using _Kallisto_ against
the counts obtained from the alignment pipeline
(_Trim Galore!_-_HISAT2_-_featureCounts_; THF)
after adding one count to all features:

```{r}
ggplot(comp.bulks, aes(log10(Kallisto + 1), log10(THF + 1))) +
  geom_hex() +
  facet_grid(Infection ~ Time) +
  geom_point(
    aes(log10(Kallisto + 1), log10(THF + 1)),
    subset(comp.bulks, ERCC),
    colour = "red", size = 0.3)
```

For reference, let us show display the data without log-scaling the axes:

```{r}
ggplot(comp.bulks, aes(Kallisto + 1, THF + 1)) +
  geom_hex() +
  facet_grid(Infection ~ Time) +
  geom_point(
    aes(Kallisto + 1, THF + 1),
    subset(comp.bulks, ERCC),
    colour = "red", size = 0.3)
```

# Single cells

To compare the output of _Kallisto_ and the alignment-based pipeline on
single cells, let us focus on the subset of cells that passed filtering:

```{r}
kallisto_cells <- readRDS(file.path(folder.rds, "goodCells.rds"))
```

Let us retain only the features also measured by the alignment pipeline:

```{r}
kallisto.overlap <- kallisto_cells[
  match(features.overlap, fData(kallisto_cells)$feature_id),
  ]
```

To compare the performance of both approaches in a representative sample of
experimental groups, let us randomly select a single cell from each
`Time:Infection` condition, and colour the data points according the cell
`Status` (`uninfected`, `exposed`, `infected`):

```{r}
set.seed(1985)
comp.sc <- data.frame(
  Time = character(),
  Infection = character(),
  Status = character(),
  Kallisto = numeric(),
  THF = numeric(),
  ERCC = logical()
)
for (time in c("2h", "4h", "6h")){
   for (infection in c("Mock", "D23580", "LT2")){
    sname <- as.character(sample(subset(
      pData(kallisto.overlap),
      Time == time & Infection == infection,
      select = sample,
      drop = TRUE), 1))
    message(sname)
    kallisto <- counts(kallisto.overlap[,sname])[,1]
    thf = dge.overlap$counts[,sname]
    new.df <- data.frame(
      Time = time,
      Infection = infection,
      Status = subset(
        pData(kallisto.overlap), sample == sname, select = Status, drop = TRUE
      ),
      Kallisto = kallisto,
      THF = thf,
      ERCC = ercc
    )
    comp.sc <- rbind(comp.sc, new.df)
  }
}
```

Let us now plot the log~10~-transformed
estimated counts obtained using _Kallisto_ against
the counts obtained from the alignment pipeline
(_Trim Galore!_-_HISAT2_-_featureCounts_; THF)
after adding one count to all features:

```{r}
ggplot(comp.sc, aes(log10(Kallisto + 1), log10(THF + 1))) +
  geom_hex(aes(colour = Status)) +
  facet_grid(Infection ~ Time) +
  geom_point(
    aes(log10(Kallisto + 1), log10(THF + 1)),
    subset(comp.sc, ERCC),
    colour = "red", size = 0.3)
```

Let us show the same data without colour:

```{r}
ggplot(comp.sc, aes(log10(Kallisto + 1), log10(THF + 1))) +
  geom_hex() +
  facet_grid(Infection ~ Time) +
  geom_point(
    aes(log10(Kallisto + 1), log10(THF + 1)),
    subset(comp.sc, ERCC),
    colour = "red", size = 0.3)
```


# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
