---
title: Preprocessing
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Preprocessing of _Kallisto_ quantitation data, including aggregation
  of transcript-level expression data into gene-level estimates, and
  annotation of gene identifiers with gene symbol annotation.
vignette: >
  %\VignetteIndexEntry{Quality control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2
---

# Settings

```{r check, include=FALSE}
stopifnot(
    requireNamespace("scater"),
    requireNamespace("biomaRt")
)
```

Define of folders where to find and write data:

```{r settingsOut}
if (interactive()){
    workdir <- "."
} else {
    workdir <- ".."
}
folder.quant <- file.path(workdir, "quant")
folder.expData <- file.path(workdir, "expData")
folder.rds <- file.path(workdir, "rds")
folder.qc <- file.path(workdir, "qc")
```


# Pre-process data

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks) was obtained using _Kallisto_ 
as follows:

```
kallisto quant \
    --index=$kallistoIdx \
    --output-dir=$kallistoDir/$cell \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```


# Import data

First, let us detect the _Kallisto_ quantitation folder for all samples:

```{r kallistoDir}
kallisto.dirs <- list.files(folder.quant)
length(kallisto.dirs)
```

Now, let us import the quantitation data, using the trimmed folder name
as a unique identifier for each sample:

```{r readKallistoResults, message=FALSE}
library(scater)
sc_DC <- readKallistoResults(
    samples = gsub("WTCHG_", "", kallisto.dirs),
    directories = file.path(folder.quant, kallisto.dirs)
)
dim(sc_DC)
```

Then, let us attach experimental covariates to the `SCESet`, in the `phenoData`
slot:

```{r expData}
samplesData <- read.csv(
    file.path(folder.expData, "samples.csv"),
    row.names = 1)
pData(sc_DC) <- cbind(
    samplesData,
    pData(sc_DC)
)
rm(samplesData)
varLabels(sc_DC)
```

**Note:** To match each row of the CSV file to the corresponding sample in the
`SCESet`, the first column of the CSV file contains the same unique sample
identifier supplied to the `readKallistoResults` method in the previous
code chunk.

# Collapse transcript to genes

The index file supplied to Kallisto was generated from the file
`Homo_sapiens.GRCh38.cdna.all.fa`, which supplied sequence information
annotated to versioned Ensembl transcript identifiers of the form
`ENST00000448914.1`.
Before gene annotations may be retrieved from the Ensembl Biomart, the
version identifier `.x` of each transcript must be trimmed:

```{r trimTxId}
fData(sc_DC)$feature_id <- gsub(
    "\\.[[:digit:]]$", "", fData(sc_DC)$feature_id
)
```

Only then, the `SCESet` may be updated with gene annotations fetched from
the Ensembl Biomart using the `getBMFeatureAnnos` method (a wrapper around the
`r Biocpkg("biomaRt")` package):

```{r getBMFeatureAnnos}
sc_DC <- getBMFeatureAnnos(
    sc_DC,
    attributes = c(
        "ensembl_transcript_id", "ensembl_gene_id", "hgnc_symbol",
        "chromosome_name", "transcript_biotype", "transcript_start",
        "transcript_end", "transcript_count"),
    feature_symbol = "hgnc_symbol",
    dataset = "hsapiens_gene_ensembl")
names(fData(sc_DC))
```

Once the `SCESet` contains gene annotations, transcript-level expression
estimates produced by _Kallisto_ can be collapsed to gene-level estimates using the `summariseExprsAcrossFeatures` method:

```{r summariseExprsAcrossFeatures}
sc_DC_gene <- summariseExprsAcrossFeatures(
    sc_DC, exprs_values="tpm", summarise_by="feature_id")
names(fData(sc_DC_gene))
colnames(fData(sc_DC_gene)) <-
    gsub("exprs_collapsed_to", "ensembl_gene_id", names(fData(sc_DC_gene)))
rm(sc_DC)
```

# Add gene symbols to featureNames to make them more intuitive

First, let us fetch from the Ensembl Biomart a mapping table between
the Ensembl gene identifier used to collapse gene expression, and the
associated official HGNC gene symbol:

```{r biomartGeneId2GeneSymbol}
library(biomaRt)
mart <- useMart(
    biomart = "ENSEMBL_MART_ENSEMBL",
    dataset = "hsapiens_gene_ensembl", host = "www.ensembl.org"
)
bm.map <- getBM(c("ensembl_gene_id", "hgnc_symbol"), mart = mart)
rm(mart)
```

Using the mapping table, let us store the (first) gene symbol associated with
each gene identifier as an additional column of the `featureData` slot:

```{r storeGeneSymbol}
fData(sc_DC_gene)$hgnc_symbol <- bm.map[
    match(fData(sc_DC_gene)$ensembl_gene_id, bm.map$ensembl_gene_id),
    "hgnc_symbol"
]
rm(bm.map)
```

The downloaded HGNC gene symbols information contain a mixture of `NA` values
and empty character names `""`. Let us convert empty gene names `""` to
more consistent and convenient `NA` values:

```{r empty2NA}
fData(sc_DC_gene)$hgnc_symbol[fData(sc_DC_gene)$hgnc_symbol == ""] <- NA
```

Now, we may rename features by prepending their gene unique gene identifier
with their official gene symbol, when available:

```{r updateFeatureNames}
featureNames(sc_DC_gene) <- ifelse(
    is.na(fData(sc_DC_gene)$hgnc_symbol),
    as.character(fData(sc_DC_gene)$ensembl_gene_id),
    paste(
        fData(sc_DC_gene)$hgnc_symbol,
        fData(sc_DC_gene)$ensembl_gene_id,
        sep = "_"
    )
)
head(fData(sc_DC_gene), n = 3)
```

# Save the updated SCESet to file

```{r saveRDS}
saveRDS(sc_DC_gene, file.path(folder.rds, "sc_DC_gene.rds"))
```

# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
