---
title: Quality control
author:
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: Oxford University
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Quality control based for all 384 samples, including blanks, bulks and
  single cells.
vignette: >
  %\VignetteIndexEntry{Quality control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2
---

# Settings

Define of folders where to find data:

```{r, eval=FALSE}
# Location of Kallisto quantitation folders
folder.quant <- "quant"
# Location of experimental data (e.g., phenotypes)
folder.expData <- "expData"
```

and where to write data:

```{r, settingsOut}
# Location of RDS files (Serialized Single Objects) 
folder.rds <- "rds"
# Location of QC files
folder.qc <- "qc"
```

# Data pre-processing

Quantitation data imported into the _R_ environment was subsequently:

* transcript-level expression data was aggregated into gene-level estimates
    (*i.e.* `ensembl_transcript_id` `->` `ensembl_gene_id`)
* Gene identifiers were made more intuitive by prepending gene symbol were
    possible

For the sake of time, the code is only presented here, but not run.
Pre-processed data will be immediately imported in the following section.

## Import

First, _Kallisto_ quantitation folders were detected:

```{r, eval=FALSE}
kallisto.dirs <- list.files(folder.quant)
length(kallisto.dirs)
```

Then, data was imported:

```{r, eval=FALSE}
sc_DC <- readKallistoResults(
    samples = gsub("WTCHG_", "", kallisto.dirs),
    directories = file.path(folder.quant, kallisto.dirs)
)
```



# Import pre-processed data {#Preprocessed}

```{r, include=FALSE}
sc_DC_gene <- readRDS(file.path(folder.rds, "sc_DC_gene.rds"))
```


# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
