---
title: Quality control
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Quality control based for all 384 samples, including blanks, bulks and
  single cells.
vignette: >
  %\VignetteIndexEntry{Quality control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2:
    toc_newpage: true
---


# Settings

```{r check, include=FALSE}
stopifnot(
    requireNamespace("scater"),
    requireNamespace("S4Vectors"),
    requireNamespace("reshape2"),
    requireNamespace("knitr")
)
```

Define of folders where to find and write data:

```{r settings}
if (interactive()){
    workdir <- "."
} else {
    workdir <- ".."
}
folder.rds <- file.path(workdir, "rds")
folder.qc <- file.path(workdir, "qc")
```

\bioccomment{
Paths are relative the location of the vignette file; if code chunks
are run interactively, paths must be relative to the working directory 
}

# Pre-process data

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks) was obtained using _Kallisto_ 
as follows:

```
kallisto quant \
    --index=$kallistoIdx \
    --output-dir=$kallistoDir/$cell \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```

The resulting quantitation data was imported into the _R_ environment as a
`SCESet` using the `r Biocpkg("scater")` `readKallistoResults` method,
and pre-processed as follows:

* Phenotype data was imported from a CSV file and attached to the `SCESet`.
* Transcript-level expression data was aggregated into gene-level
    expression estimates using the `r Biocpkg("scater")` `getBMFeatureAnnos`
    and `summariseExprsAcrossFeatures` methods.
* Gene identifiers were made more intuitive by prepending gene symbol where
    possible (*e.g.* `ENSG00000277943` to `CCL4_ENSG00000277943`).
* The resulting `SCESet` was serialised and written to file.

Here, we simply import the pre-processed data:

```{r preprocessed, message=FALSE}
library(scater)
sc_DC_gene <- readRDS(file.path(folder.rds, "sc_DC_gene.rds"))
```

Let us examine a summary of the experimental phenotype data:

```{r expdata}
expdata <- pData(sc_DC_gene)[,1:7]
expdata$Lane <- as.factor(expdata$Lane)
expdata$Identifier.number <- as.character(expdata$Identifier.number)
summary(expdata)
```

# Calculate QC metrics

First, let us annotate the `SCESet` with various QC metrics
using control features (*i.e.* ERCC spike-in)
and samples (*i.e.* blanks, bulks):

```{r ctrlFeaturesSamples, message=FALSE}
library(S4Vectors)
filter.features.ctrl <- FilterRules(list(
        ERCC = function(x) grepl("ERCC-[[:digit:]]", featureNames(x))
))
filter.samples.ctrl <- FilterRules(list(
        BULK = function(x) x$Status == "BULK",
        Blank = function(x) x$Status == "Blank"
))
summary(evalSeparately(filter.features.ctrl, sc_DC_gene))
summary(evalSeparately(filter.samples.ctrl, pData(sc_DC_gene)))
sc_DC_gene <- calculateQCMetrics(sc_DC_gene,
    list(ERCC = eval(filter.features.ctrl["ERCC"], sc_DC_gene)),
    list(
        Bulk = eval(filter.samples.ctrl["BULK"], pData(sc_DC_gene)),
        Blank = eval(filter.samples.ctrl["Blank"], pData(sc_DC_gene))
    )
)
```

\bioccomment{
Mitochondrial genes might be used as control features in the future.
}

# Library complexity

Let us examine the library complexity across `Status` and `Infection`:

```{r libraryComplexity, fig.height=7.5}
plot(
    sc_DC_gene, block1 = "Infection", block2 = "Status",
    colour_by = "Time", nfeatures = 300, exprs_values = "counts"
)
```

First, from a QC perspective, some reassuring observations:

* Blanks show a very low library complexity
  (few features make up most of the libraries)
* Bulks show a higher complexity
* Single cells generally show a complexity comparable to that of the `BULK`
    control samples

Now, from a biological perspective:

* A small number of single cells (e.g. `exposed:4h:LT2`, `infected:6h:LT2`)
    show suspiciously low library complexity.
    Those single cells must be treated as outliers and removed from
    subsequent analyses.
* Library complexity seems to decrease with `Time` and `Infection`.
    This **may** indicate a *polarisation* of single cells toward particular
    phenotype(s).
    + `Mock` (*i.e.* `uninfected`) single cells tend to show a consistent,
        relatively high library complexity
    + Both `LT2` and `D23580` infections show a gradual decrease of library
        complexity over time. This observation **might** be more pronounced
        for `D23580`.

## Filter low-complexity samples

As a consequence of the above observation,
let us define a reasonable cut-off value to exclude outliers based on
library complexity.
For instance, let us visualise the percentage of estimated counts accounted
for by the top `50` and `200` features in each sample against the median
± `3` standard deviations across all samples.
We also store the number of standard deviations for each metric, for each
sample as additional columns in the `phenoData` slot of the `SCESet`[^1]:

[^1]: The verbose `ggplot2` code is hidden for convenience.

```{r calcPctCountSD}
pct_count_50.med <- median(sc_DC_gene$pct_counts_top_50_features)
pct_count_50.sd <- sd(sc_DC_gene$pct_counts_top_50_features)
sc_DC_gene$top_50_n_sd <-
    (sc_DC_gene$pct_counts_top_50_features - pct_count_50.med) /
    pct_count_50.sd
pct_count_200.med <- median(sc_DC_gene$pct_counts_top_200_features)
pct_count_200.sd <- sd(sc_DC_gene$pct_counts_top_200_features)
sc_DC_gene$top_200_n_sd <-
    (sc_DC_gene$pct_counts_top_200_features - pct_count_200.med) /
    pct_count_200.sd
```

```{r pctCountSD, message=FALSE, echo=FALSE}
library(ggplot2)
library(reshape2)
ggplot() +
    geom_hline( # median
        aes(yintercept = y),
        data.frame(
            top_features = c(
                "pct_counts_top_50_features",
                "pct_counts_top_200_features"
            ),
            y = c(pct_count_50.med, pct_count_200.med)
        )
    ) +
    geom_hline( # standard deviation cut-offs
        aes(yintercept = y),
        data.frame(
            top_features = c(
                rep("pct_counts_top_50_features", 2),
                rep("pct_counts_top_200_features", 2)
            ),
            y = c(
                c(pct_count_50.med, pct_count_50.med) + pct_count_50.sd*c(3,-3),
                c(pct_count_200.med, pct_count_200.med) + pct_count_200.sd*c(3,-3)
            )
        ),
        colour = "red"
    ) + 
    geom_point(
        aes(
            x = Identifier.number, y = pct_counts,
            colour = Status, shape = Infection
        ),
        melt(
            pData(sc_DC_gene),
            id.vars = c("Identifier.number", "Status", "Infection"),
            measure.vars = c(
                "pct_counts_top_50_features",
                "pct_counts_top_200_features"
            ),
            variable.name = "top_features",
            value.name = "pct_counts"
        ),
        alpha = 0.4
    ) +
    facet_grid(top_features ~ .)
```

In light of this information, let us create filter rules to **exclude**
samples for which any of the following conditions is `TRUE`:

* the top `50` features make up more than the median value plus `3`
    standard deviations.
* the top `200` features make up more than the median value plus `3`
    standard deviations.

We may then examine more closely the number and nature of those samples:

```{r complexityFilterMatrix}
filter.samples <- FilterRules(list(
    top_50_3_sd = function(x) x$top_50_n_sd < 3,
    top_200_3_sd = function(x) x$top_200_n_sd < 3
))
summary(evalSeparately(filter.samples, pData(sc_DC_gene)))
```

```{r kableTop50, echo=FALSE}
library(knitr)
kable(
    pData(sc_DC_gene)[
        !eval(filter.samples["top_50_3_sd"], pData(sc_DC_gene)), 1:7
    ]
)
```

Table: Samples failing the `top_50_3_sd` filter rule.

```{r kableTop200, echo=FALSE}
kable(
    pData(sc_DC_gene)[
        !eval(filter.samples["top_200_3_sd"], pData(sc_DC_gene)), 1:7
    ]
)
```

Table: Samples failing the `top_200_3_sd` filter rule.

The `top_50_3_sd` rule identifies `5` outliers (`379` samples passed the
filter):

* the two `Blank` samples
* the two single cells visually identified in the previous section
* one more single cell which can be confirmed visually in the previous section:
    `D23580:6h:exposed`.

The `top_200_3_sd` rule is slightly more stringent, and identifies a total of
`8` samples (`376` samples passed the filter),
including all `5` samples identified by the `top50_75` rule
(the combined rules have the same effect as the `top_200_3_sd` rule alone).

# Association of principal components with experimental covariates

## Expression-based PCA

Using the expression data for the `500` most abundant features,
let us visualise the three principal components that capture the largest
variance in the data set:

```{r pcaExpr384}
sc_DC_gene <- plotPCA(
    sc_DC_gene, ntop = 500, ncomponents = 3,
    exprs_values = "exprs", colour_by = "Status", shape_by = "Time",
    return_SCESet = TRUE
)
```

In combination with the figures shown in [Appendix 1](#Appendix1),
and the following section,
the above figure highlights that _in the complete data set_:

* `PC1` negatively correlates with `Time`, and separates to some extent
    `uninfected` cells from `infected` or `exposed`.
* `PC2` negatively correlates with QC metrics related to the number of
    detected features
    (*i.e.* `pct_dropout`, `log10_total_features`, `pct_exprs_top_50_features`).
    Higher coordinates on `PC2` indicate lower library complexity
    (*e.g.* `Blank` samples), and lower coordinates indicate high library
    complexity (*e.g.* `BULK` samples).
* `PC3` primarily separates `uninfected` cells from those `infected` or
    `exposed`.
* Together, `PC1` and `PC3` cluster cells according to `Time` and `Status`,
    with:
    + `uninfected` cells in the top-right corner.
    + `infected` and `exposed` cells at `2h` in the top-left corner.
    + `infected` and `exposed` cells at `4h` in the middle-left corner.
    + `infected` and `exposed` cells at `6h` in the bottom-right corner.

```{r pcaImportant384, include=FALSE}
pc.384.time <- findImportantPCs(sc_DC_gene, "Time")
pc.384.infection <- findImportantPCs(sc_DC_gene, "Infection")
pc.384.status <- findImportantPCs(sc_DC_gene, "Status")
pc.384.pct_dropout <- findImportantPCs(sc_DC_gene, "pct_dropout")
pc.384.pct_dropout <- pc.384.pct_dropout +
    geom_point(
        aes(colour = Status),
        data.frame(
            pc.384.pct_dropout$data,
            Status = pData(sc_DC_gene)[
                pc.384.pct_dropout$data$Var1,
                "Status"
            ]
        ),
        inherit.aes = TRUE
    )
```

## Plate-level overview

Let us examine the coordinate of each sample on the three principal
components capturing the largest amount of variance, to emphasise the absence
of batch effect between the different plates:

```{r PCplates, echo=FALSE, fig.height=12}
sc_DC_gene$Row <- factor(
    gsub("^([[:alpha:]])([[:digit:]]+)$", "\\1", sc_DC_gene$Well),
    LETTERS[8:1] # Facilitates later geom_tile
)
sc_DC_gene$Column <- factor(
    gsub("^([[:alpha:]])([[:digit:]]+)$", "\\2", sc_DC_gene$Well),
    1:12
)
multiplot(
    ggplot(cbind(pData(sc_DC_gene), reducedDimension(sc_DC_gene))) +
    geom_tile(aes(x=Column, y=Row, fill=PC1)) +
    facet_wrap(~ Plate),
    ggplot(cbind(pData(sc_DC_gene), reducedDimension(sc_DC_gene))) +
    geom_tile(aes(x=Column, y=Row, fill=PC2)) +
    facet_wrap(~ Plate),
    ggplot(cbind(pData(sc_DC_gene), reducedDimension(sc_DC_gene))) +
    geom_tile(aes(x=Column, y=Row, fill=PC3)) +
    facet_wrap(~ Plate)
)
```


## Outlier detection

Let us identify outlier samples using `r Biocpkg("scater")` and
the `r CRANpkg("mvoutlier")` package and
common QC metrics that relate to library complexity and feature detection,
and save the result in the `SCESet`:

```{r outlierDetection, results='hide'}
sc_DC_gene <- plotPCASCESet(
        sc_DC_gene, ntop = 500, ncomponents = 3,
        exprs_values = "exprs", shape_by = "Status",
        detect_outliers = TRUE, pca_data_input = "pdata",
        selected_variables = c(
            "pct_counts_top_100_features",
            "total_features",
            "pct_counts_feature_controls",
            "n_detected_feature_controls",
            "log10_counts_endogenous_features",
            "log10_counts_feature_controls"
        ),
        return_SCESet = TRUE,
        draw_plot = FALSE
    )
```

In particular, the known properties of `Blank` and `Bulk` samples
are useful landmarks to identify such outliers in a PCA plot.
Let us compare the outliers detected to the known `Status`:

```{r outlierPCAPlot, echo=FALSE, fig.height=12}
sc_DC_gene$top_200_3_sd <- sc_DC_gene$top_200_n_sd > 3
multiplot(
    plotPCASCESet(
        sc_DC_gene, ntop = 500, ncomponents = 3,
        exprs_values = "exprs", colour_by = "Status", shape_by = "Time",
        detect_outliers = FALSE, pca_data_input = "pdata",
        selected_variables = c(
            "pct_counts_top_100_features",
            "total_features",
            "pct_counts_feature_controls",
            "n_detected_feature_controls",
            "log10_counts_endogenous_features",
            "log10_counts_feature_controls"
        )
    ),
    plotPCASCESet(
        sc_DC_gene, ntop = 500, ncomponents = 3,
        exprs_values = "exprs", shape_by = "Time",
        detect_outliers = TRUE, pca_data_input = "pdata",
        selected_variables = c(
            "pct_counts_top_100_features",
            "total_features",
            "pct_counts_feature_controls",
            "n_detected_feature_controls",
            "log10_counts_endogenous_features",
            "log10_counts_feature_controls"
        )
    )
)
rm()
```

Among the resulting principal components, `PC1` strongly correlates with
metrics that relate to the library complexity, with `BULK` samples
at the lowest coordinates, and `Blank` at the highest coordinates.

In addition, [Appendix 2](#Appendix2) reveals how the earlier filter rules
`top_50_3_sd` and `top_200_3_sd` corroborate the most extreme outliers
detected by the multivariate PCA approach applied in this section.

Let us create a fiter rule to retain samples **not** detected as outliers by
this PCAapproach:

```{r filterOutlier}
filter.samples.outliers <- FilterRules(list(
    notOutlierPCA = function(x) x$outlier == FALSE
))
summary(evalSeparately(filter.samples.outliers, pData(sc_DC_gene)))
```

We may then examine more closely the number and nature of those samples:

```{r kablePcaOutlier, echo=FALSE}
kable(
    pData(sc_DC_gene)[
        !eval(filter.samples.outliers["notOutlierPCA"], pData(sc_DC_gene)), 1:7
    ]
)
```

Table: Samples failing the `notOutlierPCA` filter rule.

# Identify most abundant features

The following figure confirms that ERCC spike-ins are the most abundant
features detected in the two `Blank` samples, accounting for over `98%` of the
expression data in those samples.
Conversely, the most abundant features among the nine `BULK` samples and
`373` single cells tend to be known associates of immune cells
(*e.g.* `CD74`, `CCL4`, `ISG15`).

```{r highestExpression, echo=FALSE, out.width='90%', fig.height = 8}
multiplot(
    plotQC(
        sc_DC_gene[,!eval(filter.samples.ctrl["Blank"], pData(sc_DC_gene))],
        type = "highest-expression", exprs_values = "counts"
    ), #+ ggtitle("Bulks and single cells")
    plotQC(
        sc_DC_gene[,eval(filter.samples.ctrl["Blank"], pData(sc_DC_gene))],
        type = "highest-expression", exprs_values = "counts"
    ), #  + ggtitle("Blanks")
    cols = 2
)
```

# Identify robust detected features

## Using bulk samples

Next, to focus on biologically relevant and robust features,
let us identify all genes that are detected in at least `2/3` of
the bulk samples (*i.e.* `6` or more of the `9` bulks).
For reference, let us compare that number to:

* the count of features detected in at least any one sample
* the count of features detected in at least three `BULK` samples
* the count of features with non-zero variance

```{r rulesDetected}
filter.features <- FilterRules(
    list(
        detected1sample = function(x){
            rowSums(exprs(x) > 0) > 0
        },
        detected3bulks = function(x){
            rowSums(exprs(x[,x$Status == "BULK"]) > 0) >= 3
        },
        detected6bulks = function(x){
            rowSums(exprs(x[,x$Status == "BULK"]) > 0) >= 6
        },
        variance = function(x){
            apply(exprs(x), 1, function(x) {var(x) > 0})
        }
    )
)
summary(evalSeparately(filter.features, sc_DC_gene))
summary(evalSeparately(
    c(filter.features, filter.features.ctrl),
    sc_DC_gene
))
```

We observe that:

* Approximately a quarter of the features
    (including only `10` of the `92` ERCC spike-ins)
    are detected in at least six of the nine `BULK` samples,
* All the features that pass the most stringent `detected6bulks` rule
    show non-zero variance (the combined rules have the same effect as the
    `detected6bulks` rule alone).

## Using single cells

For comparison, let us examine the number of features detected as a function
of the number of single cells in which each feature was detected:

```{r featuresDetectedSC, echo=FALSE, out.width="100%"}
# For each gene, number of cells in which it was detected
feature_N_cells <- rowSums(
    exprs(sc_DC_gene[,!sc_DC_gene$is_cell_control])
)
# From one cell to all single cells, how many genes are detected at each level
tmp <- data.frame(
    N_cells = seq_len(
        sum(!sc_DC_gene$is_cell_control)
    )
)
tmp$N_features <- sapply(tmp$N_cells, function(x) sum(feature_N_cells >= x))
ggplot(tmp) +
    geom_point(aes(x=N_cells, y=N_features), size = 0.5) +
    scale_y_continuous(limits = c(0, nrow(sc_DC_gene)))
rm(feature_N_cells, tmp)
```



# Filter data set

## Overview of filter rules

### Samples

Let us summarise the result of the active filter rules applied to single
cells (*i.e.* ignoring `BULK` and `Blank` samples):

```{r heatmapFilterSamples, echo=FALSE}
tmp <- evalSeparately(
    c(filter.samples, filter.samples.outliers),
    pData(sc_DC_gene)[!sc_DC_gene$is_cell_control,]
)
summary(tmp)
tmp <- melt(
    cbind(
        Cell = seq_len(nrow(tmp)),
        as.data.frame(tmp)
    ),
    id.vars = "Cell",
    variable.name = "Filter",
    value.name = "Pass"
)
ggplot(tmp) +
    geom_tile(aes(x=Filter, y=Cell, fill=Pass))
rm(tmp)
```

## Features

\fixme{
In progress.
}

Let us summarise the result of the active filter rules applied to features
excluding the ERCC control features:

```{r filterFeaturesSummary}
summary(evalSeparately(
    c(filter.features),
    sc_DC_gene[!fData(sc_DC_gene)$is_feature_control_ERCC]
))
```

## Apply filters

Let us now subset the `SCESet` to retain only
_single cells_ (*i.e.* no `BULK` or `Blank` control sample)
and _gene features_ (*i.e.* no ERCC control feature)
that pass all the outlier filter rules defined above:

```{r filterSCESet}
sc_DC_filtered <- sc_DC_gene[
    eval(c(filter.features), sc_DC_gene) &
        !fData(sc_DC_gene)$is_feature_control,
    eval(c(filter.samples, filter.samples.outliers), pData(sc_DC_gene)) &
        !sc_DC_gene$is_cell_control
]
dim(sc_DC_filtered)
```

# Gene expression 

Currently, this plot displays a random set of `6` genes selected
only on the basis that their name starts with either `CCL`, `CXCL`, or `IL`.

```{r plotExpression}
plotExpression(
    sc_DC_filtered,
    rownames(sc_DC_filtered)[sample(
        grep("^((CCL)|(CXCL)|(IL))", rownames(sc_DC_filtered)),
        6
    )],
    x = "Status", exprs_values = "exprs", colour = "Infection"
) + theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)
)
```

\newpage

# Appendix

## Association of principal components with experimental covariates {#Appendix1}

```{r pcImportantTime, out.width='80%' , fig.width = 8, fig.height = 12}
pc.384.time
```
Most important principal components associated with `Time` phenotype.
\newpage
```{r pcImportantStatus, out.width='80%' , fig.width = 8, fig.height = 12}
pc.384.status
```
Most important principal components associated with `Status` phenotype.
\newpage
```{r pcImportantInfection, out.width='80%' , fig.width = 8, fig.height = 12}
pc.384.infection
```
Most important principal components associated with `Infection` phenotype.
\newpage

```{r pcImportantDropout, out.width='80%' , fig.width = 8, fig.height = 10}
pc.384.pct_dropout
```
Most important principal components associated with `pct_dropout` phenotype
(*i.e.* proportion of features below detection level).
\newpage


## PCA-based detection of outlier samples {#Appendix2}

```{r, echo=FALSE, fig.height=12}
multiplot(
    plotPCASCESet(
        sc_DC_gene, ntop = 500, ncomponents = 3,
        exprs_values = "exprs", shape_by = "Time",
        detect_outliers = TRUE, pca_data_input = "pdata",
        selected_variables = c(
            "pct_counts_top_100_features",
            "total_features",
            "pct_counts_feature_controls",
            "n_detected_feature_controls",
            "log10_counts_endogenous_features",
            "log10_counts_feature_controls"
        )
    ),
    plotPCASCESet(
        sc_DC_gene, ntop = 500, ncomponents = 3,
        exprs_values = "exprs", colour_by = "top_200_3_sd", shape_by = "Time",
        detect_outliers = FALSE, pca_data_input = "pdata",
        selected_variables = c(
            "pct_counts_top_100_features",
            "total_features",
            "pct_counts_feature_controls",
            "n_detected_feature_controls",
            "log10_counts_endogenous_features",
            "log10_counts_feature_controls"
        )
    )
)

```

# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
