---
title: Quality control
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Quality control based for all 384 samples, including blanks, bulks and
  single cells.
vignette: >
  %\VignetteIndexEntry{Quality control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2:
    toc_newpage: true
---


# Settings

```{r check, include=FALSE}
stopifnot(
    requireNamespace("scater"),
    requireNamespace("S4Vectors"),
    requireNamespace("reshape2")
)
```

Define of folders where to find and write data[^1]:

[^1]: Paths are relative the location of the vignette file; if code chunks
are run interactively, paths must be relative to the working directory 

```{r settings}
# Location of RDS files (Serialized Single Objects) 
folder.rds <- "../rds"
# Location of QC files
folder.qc <- "../qc"
```

# Pre-process data

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks)
was imported into the _R_ environment as a `SCESet` using the
`r Biocpkg("scater")` `readKallistoResults` method,
and pre-processed as follows:

* Phenotype data was imported from a CSV file and attached to the `SCESet`.
* Transcript-level expression data was aggregated into gene-level
    expression estimates using the `r Biocpkg("scater")` `getBMFeatureAnnos`
    and `summariseExprsAcrossFeatures` methods.
* Gene identifiers were made more intuitive by prepending gene symbol where
    possible.
* The resulting `SCESet` was serialised and written to file.

Here, we simply import the pre-processed data:

```{r preprocessed, message=FALSE}
library(scater)
sc_DC_gene <- readRDS(file.path(folder.rds, "sc_DC_gene.rds"))
```

Let us examine a summary of the experimental phenotype data:

```{r expdata}
expdata <- pData(sc_DC_gene)[,1:7]
expdata$Lane <- as.factor(expdata$Lane)
expdata$Identifier.number <- as.character(expdata$Identifier.number)
summary(expdata)
```

# Calculate QC metrics

Let us calculate various QC[^2] metrics using control features
(*i.e.* ERCC spike-in) and samples (*i.e.* blanks, bulks):

[^2]: Quality control.

```{r identifyCtrls}
idx.ERCC <- which(grepl("ERCC-[[:digit:]]", featureNames(sc_DC_gene)))
length(idx.ERCC)
idx.BULK <- which(sc_DC_gene$Status == "BULK")
length(idx.BULK)
idx.Blank <- which(sc_DC_gene$Status == "Blank")
length(idx.Blank)
sc_DC_gene <- calculateQCMetrics(
    sc_DC_gene,
    list(ERCC = idx.ERCC),
    list(
        Bulk = idx.BULK,
        Blank = idx.Blank
    )
)
```

\bioccomment{
Mitochondrial genes might be used as control features in the future.
}

# Filter detected features

First, let us discard all genes that are not detected in at least `2/3` of the
bulk samples (*i.e.* `6` or more of the `9` bulks).

Let us first define the filter rules, and examine the result of their
evaluation on the `SCESet` object:

```{r filterOnBulks, message=FALSE}
library(S4Vectors)
filter.features <- FilterRules(
    exprs = list(
        detected = function(x){
            rowSums(exprs(x[,x$Status == "BULK"]) > 0) >= 6
        },
        variance = function(x){
            apply(exprs(x), 1, function(x) {var(x) > 0})
        }
    )
)
fm <- evalSeparately(filter.features, sc_DC_gene)
summary(fm)
idx.f.detected <- which(eval(filter.features, sc_DC_gene))
rm(fm)
```

Approximately a quarter of the genes are detected in at least two thirds
of the `BULK` samples, and all those genes show non-zero variance
(the combined rules have the same effect as the `detected` rule alone).

Therefore, let us deactivate the unnecessary `variance` rule,
before applying the remaining active rule `detected` to subset the `SCESet`:

```{r genesDetectedBulks}
active(filter.features)["variance"] <- FALSE
sc_DC_gene <- subsetByFilter(sc_DC_gene, filter.features)
dim(sc_DC_gene)
```

# Library complexity

Using the previously filtered `SCESet`,
let us examine the library complexity across `Status` and `Infection`:

```{r libraryComplexity}
plot(
    sc_DC_gene, block1 = "Status", block2 = "Infection",
    colour_by = "Time", nfeatures = 300, exprs_values = "counts")
```

First, from a QC perspective, some reassuring observations:

* Blanks show a very low library complexity
  (few features make up most of the libraries)
* Bulks show a higher complexity
* Single cells generally show a complexity comparable to that of the `BULK`
    control samples

Now, from a biological perspective:

* A small number of single cells (e.g. `exposed:4h:LT2`, `infected:6h:LT2`)
    show suspiciously low library complexity.
    Those single cells must be treated as outliers and removed from
    subsequent analyses.
* Library complexity seems to decrease with `Time` and `Infection`.
    This **may** indicate a *polarisation* of single cells toward particular
    phenotype(s).
    + `Mock` (*i.e.* `uninfected`) single cells tend show a consistent,
        relatively high library complexity
    + Both `LT2` and `D23580` infections show a gradual decrease of library
        complexity. This observation **might** be more pronounced for `D23580`.

## Filter low-complexity samples

Using the above observation,
let us define a reasonable cut-off value to identify outliers based on
library complexity:

```{r lowComplexityCutoff, message=FALSE, echo=FALSE}
library(reshape2)
ggplot(
    melt(
        pData(sc_DC_gene),
        id.vars = c("Identifier.number", "Status", "Infection"),
        measure.vars = c(
            "pct_counts_top_50_features",
            "pct_counts_top_200_features"
        ),
        variable.name = "top_features",
        value.name = "pct_counts"
    ),
    aes(Identifier.number, pct_counts)
) +
    geom_point(aes(colour = Status, shape = Infection)) +
    facet_grid(top_features ~ .) +
    geom_hline(yintercept = c(75, 85), colour = "red") +
    annotate("text", x = 215E3, y = 75+3, label = 75, colour = "red") +
    annotate("text", x = 215E3, y = 85+3, label = 85, colour = "red")
```

In light of this information, let us **exclude** samples for which any of the
following conditions is `TRUE`:

* the top `50` features make up more than `75%` of the library
* the top `200` features make up more than `85%` of the library

```{r complexityFilterMatrix}
filter.samples <- FilterRules(
    exprs = list(
        top50_75 = function(x){
            x$pct_counts_top_50_features < 75
        },
        top200_85 = function(x){
            x$pct_counts_top_200_features < 85
        }
    )
)
fm <- evalSeparately(filter.samples, pData(sc_DC_gene))
summary(fm)
rm(fm)
```

The `top50_75` rule identifies `5` outliers (`379` samples passed the filter):

* the two `Blank` samples
* the two single cells visually identified in the previous section
* one more single cell which can be confirmed visually in the previous section:
    `D23580:6h:exposed`.

The `top200_85` rule is slightly more stringent, and identifies a total of `9`
samples (`375` samples passed the filter),
including all `5` samples identified by the `top50_75` rule
(the combined rules have the same effect as the `top200_85` rule alone).

Let us apply the more stringent `top200_85` rule to the data set:

```{r filterLowComplexity}
active(filter.samples)["top50_75"] <- FALSE
idx.s.lowCplx <- which(!eval(filter.samples, pData(sc_DC_gene)))
sc_DC_gene <- sc_DC_gene[,-idx.s.lowCplx]
dim(sc_DC_gene)
```

# Gene expression 

Currently, this plot displays a random set of `6` genes selected
only on the basis that their name starts with either `CCL`, `CXCL`, or `IL`.

```{r plotExpression}
plotExpression(
    sc_DC_gene, rownames(sc_DC_gene)[
        sample(
            grep("^((CCL)|(CXCL)|(IL))", rownames(sc_DC_gene)),
            6
        )
        ],
    x = "Status", exprs_values = "exprs", colour = "Infection"
) + theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)
)
```

# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
