---
title: Quality Control
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Revised quality control using all 384 samples, including blanks, bulks
  and single cells.
vignette: >
  %\VignetteIndexEntry{Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2:
    toc_newpage: true
---

# Pre-processed data

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks) was obtained using _Kallisto_ 
as follows:

```
kallisto quant \
    --index=$kallistoIdx \
    --output-dir=$kallistoDir/$cell \
    --bootstrap-samples=100 \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```

# Settings

Packages required for this vignette:

```{r packages, include=FALSE}
library(scater)
library(S4Vectors)
library(reshape2)
library(knitr)
library(dplyr)
library(BiocParallel)
library(EnsDb.Hsapiens.v79)
```

Let us set the working directory to execute the following code chunks:

```{r workdir}
workdir <- ifelse(interactive(), ".", "..")
```

Let us set a few paths relevant to the analysis:

```{r paths, include=FALSE}
folder.rds <- file.path(workdir, "rds")
```

# Import preprocessed data

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks) was obtained using
[Kallisto](https://pachterlab.github.io/kallisto/)
as follows:

```
kallisto quant \
    --index=$kallistoIdx \
    --output-dir=$kallistoDir/$cell \
    --bootstrap-samples=100 \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```

Subsequently:

* The quantitation data was imported into the _R_ environment as a
  `SCESet` using the `r Biocpkg("scater")` `readKallistoResults` method,
* Phenotype data was imported from a CSV file and attached to the `SCESet`,
* Transcript-level expression data was aggregated into gene-level
    expression estimates using the `r Biocpkg("EnsDb.Hsapiens.v79")` package
    and `r Biocpkg("scater")` `summariseExprsAcrossFeatures` method.
* Gene identifiers were made more intuitive by prepending gene symbol where
    possible (*e.g.* `ENSG00000277943` to `CCL4_ENSG00000277943`).
* The resulting `SCESet` was serialised and written to file.

Here, we simply import the preprocessed data:

```{r preprocessed, message=FALSE}
library(scater)
sc_DC_gene <- readRDS(file.path(folder.rds, "sc_DC_EnsDb.rds"))
```

Let us examine a summary of the experimental phenotype data:

```{r expdata}
expdata <- pData(sc_DC_gene)[,1:7]
expdata$Lane <- as.factor(expdata$Lane)
expdata$Identifier.number <- as.character(expdata$Identifier.number)
summary(expdata)
```

# Calculate QC metrics

## Identify control samples

Using the phenotype data, let us create `FilterRules` that identify
the control samples, namely `BULK` and `Blank` samples:

```{r controlSamplesRules}
filter.samples.ctrl <- FilterRules(list(
    BULK = function(x) x$Status == "BULK",
    Blank = function(x) x$Status == "Blank"
))
summary(evalSeparately(filter.samples.ctrl, pData(sc_DC_gene)))
```

## Identify control features 

### Identify ERCC spike-in features

ERCC spike-in features can be easily identified from their identifier:

```{r ERCCrules}
filter.features.ctrl <- FilterRules(list(
    ERCC = function(x) grepl("ERCC-[[:digit:]]", featureNames(x))
))
summary(evalSeparately(filter.features.ctrl, sc_DC_gene))
```

### Identify mitochondrial genes

```{r}
filter.features.ctrl <- c(
  filter.features.ctrl, FilterRules(list(
    MT = function(x) fData(x)$feature_id %in%
      genes(EnsDb.Hsapiens.v79, "gene_id", SeqnameFilter("MT", "="))$gene_id
  ))
)
summary(evalSeparately(filter.features.ctrl, sc_DC_gene))
```


### Calculate QC metrics

Once control features and samples were defined in the above sections,
the `r Biocpkg("scater")` `calculateQCMetrics` method may be used
to annotate the `SCESet` with various QC metrics:

```{r ctrlFeaturesSamples, message=FALSE}
sc_DC_gene <- calculateQCMetrics(sc_DC_gene,
    list(
      ERCC = eval(filter.features.ctrl["ERCC"], sc_DC_gene),
      MT = eval(filter.features.ctrl["MT"], sc_DC_gene)),
    list(
        Bulk = eval(filter.samples.ctrl["BULK"], pData(sc_DC_gene)),
        Blank = eval(filter.samples.ctrl["Blank"], pData(sc_DC_gene))
    )
)
```



# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
