---
title: Quality Control
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Quality control using all 384 samples, including blanks, bulks
  and single cells.
vignette: >
  %\VignetteIndexEntry{Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2:
    toc_newpage: true
---

# Prior work

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks) was obtained using
[Kallisto](https://pachterlab.github.io/kallisto/)
as follows:

```
kallisto quant \
    --index=$kallistoIdx \
    --output-dir=$kallistoDir/$cell \
    --bootstrap-samples=100 \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```

Subsequently:

* The quantitation data was imported into the _R_ environment as a
  `SCESet` using the `r Biocpkg("scater")` `readKallistoResults` method,
* Phenotype data was imported from a CSV file and attached to the `SCESet`,
* Transcript-level expression data was aggregated into gene-level
  expression estimates using the `r Biocpkg("EnsDb.Hsapiens.v79")` package
  (GRCh38) and `r Biocpkg("scater")` `summariseExprsAcrossFeatures` method.
* Gene identifiers were made more intuitive by prepending gene symbol where
  possible (*e.g.* `ENSG00000277943` to `CCL4_ENSG00000277943`).
* The resulting `SCESet` was serialised and written to file.

# Settings

Packages required for this vignette:

```{r packages, message=FALSE}
library(scater)
library(S4Vectors)
library(reshape2)
library(knitr)
library(dplyr)
library(BiocParallel)
library(EnsDb.Hsapiens.v79)
library(pander)
```

Let us set the working directory to execute the following code chunks:

```{r workdir}
workdir <- ifelse(interactive(), ".", "..")
```

Let us set a few paths relevant to the analysis:

```{r paths}
folder.rds <- file.path(workdir, "rds")
```

# Import preprocessed data

Here, we simply import the preprocessed data:

```{r DC_all}
DC_all <- readRDS(file.path(folder.rds, "sc_DC_EnsDb.rds"))
```

Let us create a copy of the data set, subsetted to only single cells:

```{r DC_sc}
DC_sc <- DC_all[, !DC_all$Status %in% c("Blank", "BULK")]
```

Let us examine a summary of the experimental phenotype data in the full
data set:

```{r expdata}
expdata <- droplevels(
  pData(DC_all)[,c("Infection","Status","Time","Lane","Plate","Well")]
)
expdata$Lane <- as.factor(expdata$Lane)
summary(expdata)
```

# Calculate QC metrics

In the following sections,
let us calculate QC metrics across _all_ samples (including blanks and bulks),
and across single cells _only_.

## Identify mitochondrial genes

To calculate QC metrics related to mitochondrial genes, let us first obtain
the gene identifier of those genes from the `EnsDb.Hsapiens.v79` package:

```{r MTrules}
MT_ids <- genes(
  EnsDb.Hsapiens.v79, "gene_id", SeqnameFilter("MT", "=")
  )$gene_id
length(MT_ids)
```

## Calculate QC metrics in the complete data set

Let us calculate several QC metrics based on control samples and features
available in the complete data set:

```{r calculateQCMetrics384, message=FALSE}
DC_all <- calculateQCMetrics(
  DC_all,
  feature_controls = list(
    ERCC = grepl("ERCC-[[:digit:]]", featureNames(DC_all)),
    MT = fData(DC_all)$feature_id %in% MT_ids),
  cell_controls = list(
    Bulk = DC_all$Status == "BULK",
    Blank = DC_all$Status == "Blank"
  )
)
```

## Calculate QC metrics across single cells only

In the data set that contains single cells _only_, we may calculate
QC metrics based on the control features:

```{r calculateQCMetricsSC, message=FALSE}
DC_sc <- calculateQCMetrics(
  DC_sc,
  feature_controls = list(
    ERCC = grepl("ERCC-[[:digit:]]", featureNames(DC_sc)),
    MT = fData(DC_sc)$feature_id %in% MT_ids),
  nmads = 3
)
```

Let us examine the list of QC metrics calculated for the _single cells only_
data set:

```{r QCMetrics}
varLabels(DC_sc)
```

# Global quality control for the experiment

## Most abundant features {#highestExpression}

The following figure confirms that ERCC spike-ins are the most abundant
features detected in the two `Blank` samples (right panel),
accounting for over `98%` of the expression data in those samples.
Conversely, the most abundant features among the _combined_ data set of
nine `BULK` samples and `373` single cells tend to be known associates of
immune cells (*e.g.* `CD74`, `CCL4`, `ISG15`; left panel).

```{r highestExpression, echo=FALSE, out.width='90%', fig.height = 10}
multiplot(
    plotQC(
        DC_all[,!DC_all$is_cell_control_Blank],
        type = "highest-expression", exprs_values = "tpm"
    ), # + ggtitle("Bulks and single cells")
    plotQC(
        DC_all[,DC_all$is_cell_control_Blank],
        type = "highest-expression", exprs_values = "tpm"
    ), # + ggtitle("Blanks")
    cols = 2
)
```

## Library complexity {#LibraryComplexity}

Let us examine library complexity across `Status`, `Infection`, and `Time`:

```{r libraryComplexity, fig.height=7.5}
plot(
    DC_all, block1 = "Infection", block2 = "Status",
    colour_by = "Time", nfeatures = 300, exprs_values = "counts"
)
```

First, from a QC perspective, some reassuring observations:

* Blanks show a very low library complexity (which reaffirms the
  [earlier](#highestExpression) observation that the `50` most highly expressed
  features account for more than `98%` of TPM values in `Blank` samples).
* Bulks show a generally higher complexity, as is expected for the combined
  tanscriptomes of many cells.
* Single cells generally show a complexity profile comparable to that of `BULK`
  control samples

Now, from a biological perspective:

* A small number of single cells (e.g. `exposed:LT2:4h`, `infected:LT2:6h`,
  `exposed:D23580:6h`) show suspiciously low library complexity.
  Those single cells must henceforth be treated as outliers and removed from
  subsequent analyses.
* Library complexity seems to decrease with `Time` in `exposed` and `infected`
  samples.
  This **may** indicate a polarisation of treated single cells toward
  particular phenotype(s).
  + `Mock` (*i.e.* `uninfected`) single cells tend to show a consistent,
    relatively high library complexity
  + Both `LT2` and `D23580` infections show a gradual decrease of library
    complexity over time. This observation **might** be more pronounced
    for `D23580`.

# Identify outliers

## Overview of built-in `scater` filters {#builtinFilters}

Let us illustrate the result of the `r Biocpkg("scater")` `calculateQCMetrics`
method by displaying whether each of the `373` single cells passes each
built-in filter. Non-default cut-off values are listed below:

* Number of median absolute deviation (MADs) `nmads=3` (instead of `5`)
  on the following QC metrics:
    + `total_counts` (_i.e._, total number of counts for the cell)
    + `total_features` (_i.e._, number of features with non-zero expression)
  
To facilitate comparison, the filters are listed by order of outliers detected:

```{r scaterFiltersHeatmap, echo=FALSE, fig.height=8}
QC_long <- melt(
  pData(DC_sc),
  id.vars = "sample",
  measure.vars = grep("^filter_on", varLabels(DC_sc), value = TRUE),
  variable.name = "Filter",
  value.name = "Outlier"
)
QC_long$Filter <- with(QC_long,reorder(Filter, Outlier, sum))
ggplot(QC_long) +
  geom_tile(aes(x = as.character(sample), y = Filter, fill = !Outlier)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y= element_text(size = rel(0.8))
  ) +
  labs(x = "Sample") +
  scale_fill_discrete(name = "Pass filter")
rm(QC_long)
```

## Total counts per sample {#totalCount}

Let us illustrate the single cells identified as outliers due to their 
`total_count` metric (_i.e._, total number of counts for the cell) being
_lower_ than `3` median absolute deviation (MADs) below the median value
(both median and MAD being calculated using single cell samples _only_).
The median _plus_ `3` MADs is indicated with a dashed line, for information
purpose only:

```{r totalCounts, echo=FALSE}
metric <- DC_sc$total_counts
M <- median(metric)
SD <- mad(metric, center = M)
ggplot(pData(DC_all)) +
  geom_point(aes(
    x = seq_len(nrow(pData(DC_all))),
    y = total_counts,
    colour = Status,
    shape = Infection
  )) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(x = "Sample") +
  geom_hline(
    yintercept = M + 3*SD*(-1:1), colour = c(2,1,2), linetype = c(1,1,2)
  )
```

Let us count _single cells_ identified as outliers (those numbers
do not consider the `Blank` and `BULK` samples displayed in the figure
for reference only):

```{r totalCountTable}
table(metric < (M-(3*SD)))
rm(metric, M, SD)
```

Note that only single cells **below** the median minus `3` MADs are detected
as outliers, given their excessively low library size.

## Total features per sample

Let us illustrate the single cells identified as outliers due to their 
`total_features` metric (_i.e._, number of features with non-zero expression)
being _lower_ than `3` median absolute deviation (MADs) below the median value
(both median and MAD being calculated using single cell samples _only_).
The median _plus_ `3` MADs is indicated with a dashed line, for information
purpose only:

```{r totalFeatures, echo=FALSE}
metric <- DC_sc$total_features
M <- median(metric)
SD <- mad(metric, center = M)
ggplot(pData(DC_all)) +
  geom_point(aes(
    x = seq_len(nrow(pData(DC_all))),
    y = total_features,
    colour = Status,
    shape = Infection
  )) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )  +
  labs(x = "Sample") +
  geom_hline(
    yintercept = M + 3*SD*(-1:1), colour = c(2,1,2), linetype = c(1,1,2)
  )
```

Note how only `BULK` samples show exceed the median values plus `3` MADs
(calculated on single cells _only_). In contrast, `3` single cells
display a markedly _low_ total of features detected, comparable to that of
`Blank` samples. Notably, those three samples are also detected by the
`filter_on_total_counts` QC in the [previous section](#totalCount), as
visible in the heatmap presented [earlier](#builtinFilters) to summarise
built-in `scater` filters.

Let us show how many _single cells_ are identified as outliers (those numbers
do not consider the `Blank` and `BULK` samples also displayed in the figure
for information purpose only):

```{r totalFeaturesTable}
table((metric-M)/SD < -3)
rm(metric, M, SD)
```

Note that only single cells **below** the median minus `3` MADs are detected
as outliers, given their excessively _low_ total number of features detected.

## Library complexity {#ComplexityOutliers}

In addition to the total number of features detected in each sample,
another critical QC metric si the proportion of library accounted for by
the top `N` features.
Different levels of `N` may help identify blank-like features of very low
complexity (very few features account for virtually the entire library), or
unusual libraries in the data set at larger values of `N`.
Typically, libraries showing excessively _low_ complexity are problematic,
indicating partial or biased coverage of the transcriptome.

Let us illustrate display the proportion of libraries accounted for by the
top `50`, `100`, and `200` features.
Single cells higher than the median _plus_ `3` MADs are considered outliers
due to their excessively _low_ library complexity for the given number of
features. The median _minus_ `3` MADs is indicated with a dashed line, for information purpose only:

```{r complexity50100200, echo=FALSE, fig.height=10}
res_table <- data.frame()
plots <- lapply(c(50, 100, 200), function(Nfeatures){
    metricName <- sprintf("pct_tpm_top_%i_features", Nfeatures)
    metric <- pData(DC_sc)[,metricName]
    M <- median(metric)
    MAD <- mad(metric, center = M)
    res_table[as.character(Nfeatures), "M"] <<- M
    res_table[as.character(Nfeatures), "MAD"] <<- MAD
    ggplot(pData(DC_all)) +
      geom_point(aes_string(
        x = seq_len(nrow(pData(DC_all))),
        y = metricName,
        colour = "Status",
        shape = "Infection"
      )) +
      theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
      )  +
      labs(x = "Sample") +
      geom_hline(
        yintercept = M + 3*MAD*(-1:1), colour = c(2,1,2), linetype = c(2,1,1)
      )
  })
multiplot(plotlist = plots)
rm(plots)
```

Let us create additional columns in the `phenoData` slot to mark
_single cells_ identified as outliers by those at those counts of features:

```{r complexityTable}
DC_sc$filter_on_pct_tpm_top50 <- DC_sc$pct_tpm_top_50_features >
  res_table["50", "M"] + 3 * res_table["50", "MAD"]
DC_sc$filter_on_pct_tpm_top100 <- DC_sc$pct_tpm_top_100_features >
  res_table["100", "M"] + 3 * res_table["100", "MAD"]
DC_sc$filter_on_pct_tpm_top200 <- DC_sc$pct_tpm_top_200_features >
  res_table["200", "M"] + 3 * res_table["200", "MAD"]
rm(res_table)
```

Let us show how many _single cells_ are identified as outliers (those numbers
do not consider the `Blank` and `BULK` samples also displayed in the figure
for information purpose only):

```{r evalComplexityFilters}
summary(evalSeparately(
  FilterRules(list(
    top50 = expression(filter_on_pct_tpm_top50),
    top100 = expression(filter_on_pct_tpm_top100),
    top200 = expression(filter_on_pct_tpm_top200)
  )),
  pData(DC_sc)
))
```

## Spike-in ERCC control feature

Let us illustrate the single cells identified as outliers due to
_over_-representation of spike-in ERCC control features.
For this purpose, let us display the `pct_tpm_feature_controls_ERCC` metric,
the default cut-off value of `80%`, and the more stringent cut-off of `5%`:

```{r pct_tpm_feature_controls_ERCC_80, echo=FALSE}
ggplot(pData(DC_all)) +
  geom_point(aes(
    x = seq_len(nrow(pData(DC_all))),
    y = pct_tpm_feature_controls_ERCC,
    colour = Status,
    shape = Infection
  )) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )  +
  labs(x = "Sample") +
  geom_hline(
    yintercept = c(5, 80), colour = "red"
  )
```

Notably, one single cell displays ERCC detection levels accounting for more
than `80%` of TPM in the data set---in a similar fashion to `Blank` control
samples---flagging this single cell as a clear outlier, in addition
to the `total_counts` and `total_features` metrics described
previous sections. In contrast, `BULK` samples show very low detection levels
for spike-ins features, as can be expected considering the larger relative
quantity of endogenous RNA in bulk samples.

Finally, a small number of single cells show detection levels of spike-in
control features markedly higher than the vast majority of single cell samples
(80% of single cells are found in the range `[0.87-2.54%]`). For this reason,
let us define single cells as outliers if their ERCC detection levels exceed
`5%` of TPM:

```{r pctERCCTable}
DC_sc$filter_on_ERCC_5pct <- DC_sc$pct_tpm_feature_controls_ERCC > 5
table(DC_sc$filter_on_ERCC_5pct)
```

Note that only single cells **above** `5%` are defined
as outliers, given their excessively _high_ total number of ERCC spike-in
detected.

## Mitochondrial endogenous features

Let us visualise the percentage of TPM assigned to mitochondrial genes in all
samples (including bulks and blanks).
We also display the median value, and `3` median absolute deviations (MADs)
from the median, calculated using single cells _only_ :

```{r MTpctTPMplot}
pct.mt.sc <- DC_sc$pct_tpm_feature_controls_MT
M <- median(pct.mt.sc)
SD <- mad(pct.mt.sc, center = M)
ggplot(pData(DC_all)) +
  geom_point(aes(
    x = seq_len(nrow(pData(DC_all))),
    y = pct_tpm_feature_controls_MT,
    colour = Status,
    shape = Infection
  )) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )  +
  labs(x = "Sample", y = "Percentage TPM in MT genes") +
  geom_hline(yintercept = M + 3*SD*(-1:1), colour = c(2,1,2))
table((abs(pct.mt.sc-M)/SD) > 3)
```

Again, all samples are displayed in the figure, including `Blank` and `BULK`
samples for reference.
Notably, `Blank` samples show virtually complete absence of mitochondrial gene
expression (< 0.1% of TPM), along with one single cell (`exposed:LT2`)
which must therefore be considered as an outlier and excluded from downstream
analyses.

In contrast, all `BULK` samples show expression level of mitochondrial genes
markedly higher than the majority of single cells, as one may expect from a
population of cells that contain various counts of mitochondria. Setting an
arbitrary cut-off of `3` MADs from the median value (calculated on single cells
_only_), `8` single cells show mitochondrial gene expression levels higher
than the median values of single cells plus `3` MADs; those single cells are
therefore also considered outliers, and marked as such in the `SCESet`:

```{r MTpctTPMfilter}
DC_sc$filter_on_MT_3_MAD <- (abs(pct.mt.sc-M)/SD) > 3
rm(M, SD, pct.mt.sc)
```

## Principal components analysis of selected QC metrics

Let us detect outliers based on a PCA of selected sample-level QC metrics:

```{r plotPCASCESet}
DC_sc <- plotPCASCESet(
  DC_sc, 300, 3,
  colour_by = "Status", shape_by = "Time",
  return_SCESet = TRUE, draw_plot = TRUE,
  pca_data_input = "pdata", selected_variables = c(
    "pct_counts_top_100_features",
    "total_features",
    "pct_counts_feature_controls",
    "n_detected_feature_controls",
    "log10_counts_endogenous_features",
    "log10_counts_feature_controls"
  ),
  detect_outliers = TRUE
)
```

Note that outliers detected by the PCA are identified by the new
column `outlier` in the `phenoData` slot of the `SCESet`.
Let us rename this column to `filter_on_PCA_QC` to match the other
columns identifying outliers:

```{r outliersPCA}
varLabels(DC_sc) <- gsub("^outlier$", "filter_on_PCA_QC", varLabels(DC_sc))
table(DC_sc$filter_on_PCA_QC)
```

## Final overview of sample filters

Let us display the combined built-in filters implemented in
`r Biocpkg("scater")` and those specifically calculated in this analysis.
To facilitate comparison, the filters are ordered by count of outliers
detected:

```{r finalFiltersHeatmap, echo=FALSE, fig.height=8}
QC_long <- melt(
  pData(DC_sc),
  id.vars = "sample",
  measure.vars = grep("^filter_on", varLabels(DC_sc), value = TRUE),
  variable.name = "Filter",
  value.name = "Outlier"
)
QC_long$Filter <- with(QC_long,reorder(Filter, Outlier, sum))
ggplot(QC_long) +
  geom_tile(aes(x = as.character(sample), y = Filter, fill = !Outlier)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y= element_text(size = rel(0.8))
  ) +
  labs(
    x = "Sample"
  ) +
  scale_fill_discrete(name = "Pass filter")
rm(QC_long)
```

Note how all outliers identified by the PCA based on QC metrics are also
detected by at least one other filter, highlighting the robustness 
and sensitivity of the PCA method.
It is important to bear in mind that the sensitivity of the PCA approach
is largely controlled by the selection of QC metrics provided to the method.

Notably several samples were identified as outliers due only to their
low library size (_i.e._ `total_counts`). In other words, those samples are
excluded from downstream analyses solely on the basis of risking insufficient
sequencing depth, while showing otherwise normal QC metrics.

Let us count the number of single cells according to the count of filters failed:

```{r countCellsFilterFailed, results='asis', echo=FALSE}
count_failed <- pData(DC_sc)[,grep("^filter_on", varLabels(DC_sc))]
count_failed <- rowSums(count_failed)
count_failed <- as.data.frame(table(subset(count_failed, count_failed > 0)))
colnames(count_failed) <- c("Filters failed", "Cells")
pandoc.table(count_failed)
rm(count_failed)
```

And the number of single cells that fail at least one filter:

```{r finalFilterCounts}
DC_sc$outlier <- rowSums(
  dplyr::select(pData(DC_sc), grep("^filter", varLabels(DC_sc)))) > 0
table(DC_sc$outlier)
```

Let us examine the phenotype information of those samples:

```{r phenoOutliers, results='asis', echo=FALSE}
pander::pandoc.table(subset(
  pData(DC_sc), outlier, c("Infection","Status","Time","Lane","Plate","Well"))
)
```

# Miscellaneous

Below is a PCA based on the expression level of the `300` mot highly
expressed genes across the single cells:

```{r PCAexprTimeStatus}
plotPCASCESet(DC_sc, 300, 3, colour_by = "Status", shape_by = "Time")
```

Note how `PC1` tends to associate with `Time`,
while `PC3` tends to associate with `Status`.

We may show the same PCA highlighting only the outliers identified in the
previous sections:

```{r PCAexprOutliers}
plotPCASCESet(
  DC_sc, 300, 3, colour_by = "outlier", shape_by = "Time"
)
```

Most of those outlier single cells appear clearly on `PC2`---which do not
clearly associate with any experimental covariate---underlying the need for
those samples to be removed from the data set before proceeding to downstream
analyses.


\fixme{
Remove samples.
}

# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
