---
title: Sleuth
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Preprocessing of _Kallisto_ quantitation data using _sleuth_.
vignette: >
  %\VignetteIndexEntry{Quality control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2
---

<!--
The following code chunks set the scene.
However, they are not directly relevant to the analysis,
and therefore are hidden from the compiled PDF.
-->

# Pre-processed data

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks) was obtained using _Kallisto_ 
as follows:

```
kallisto quant \
    --index=$kallistoIdx \
    --output-dir=$kallistoDir/$cell \
    --bootstrap-samples=100 \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```

# Settings

Packages required for this vignette:

```{r packages, include=FALSE}
library(sleuth)
```

```{r workdir, include=FALSE}
if (interactive()){
    workdir <- "."
} else {
    workdir <- ".."
}
```

```{r paths, include=FALSE}
folder.kallisto <- file.path(workdir, "boot")
folder.expData <- file.path(workdir, "expData")
folder.rds <- file.path(workdir, "rds")
```

# Import data

List the result folders produced by _Kallisto_ (one per sample):

```{r kallistoDirs}
kallisto.dirs <- dir(folder.kallisto, pattern = "WTCHG")
length(kallisto.dirs)
```

Let us derive a unique sample identifier from the folder name of each sample:

```{r sampleIds}
sampleIDs <- gsub("WTCHG_[[:digit:]]*_", "", kallisto.dirs)
length(unique(sampleIDs))
```

Let us load an auxillary table that describes the experimental design:

```{r experimentalDesign}
samplesData <- read.csv(file.path(folder.expData, "samples.csv"))
colnames(samplesData)
```

Now, we must enter the directories into a column in the table describing the
experiment. This column must be labeled path, otherwise sleuth will throw an
error.

```{r experimentalPaths}
samplesData <- dplyr::mutate(
  samplesData,
  path = file.path(folder.kallisto, sprintf(
    "WTCHG_%i_%i", Lane, sample)))
```

Let us check that the order is correct:

```{r samplesDataShow, results='asis'}
pander::pandoc.table(head(samplesData))
```

# 

Let us obtain a table mapping transcipts to gene identifiers:

```{r tx2gene}
library(EnsDb.Hsapiens.v79)
columns(EnsDb.Hsapiens.v79)
tx2gene <- select(
  EnsDb.Hsapiens.v79, keys(EnsDb.Hsapiens.v79), c("TXID", "GENEID", "ENTREZID"),
  "GENEID")
colnames(tx2gene)
tx2gene$target_id <- tx2gene$TXID
```


```{r sleuthPrep}
so <- sleuth_prep(
  samplesData, ~ Infection + Status + Time,
  target_mapping = tx2gene, aggregation_column = "GENEID")
```


# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
