---
title: Preprocessing
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Preprocessing of _Kallisto_ quantitation data using the `tximport`
  package, including aggregation of transcript-level expression data into
  gene-level estimates. 
vignette: >
  %\VignetteIndexEntry{Quality control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2
---

# Pre-processed data

Prior to this document, quantitation data for all 384 samples
(including single cells, blanks and bulks) was obtained using _Kallisto_ 
as follows:

```
kallisto quant \
    --index=$kallistoIdx \
    --output-dir=$kallistoDir/$cell \
    --bootstrap-samples=100 \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```

# Settings

Packages required for this vignette:

```{r packages, message=FALSE}
library(tximport)
library(readr)
library(EnsDb.Hsapiens.v79)
```

Let us set the working directory to execute the following code chunks:

```{r workdir}
workdir <- ifelse(interactive(), ".", "..")
```

Let us set a few paths relevant to the analysis:

```{r paths}
folder.kallisto <- file.path(workdir, "boot")
folder.expData <- file.path(workdir, "expData")
folder.rds <- file.path(workdir, "rds")
```

Create folders as necessary:

```{r}
if (!dir.exists(folder.rds)){
  dir.create(folder.rds)
}
```

# Import transcript-level estimates

## Phenotypes

Let us load an auxillary table that describes the experimental design:

```{r samplesData}
samplesData <- read.csv(file.path(folder.expData, "samples.csv"))
colnames(samplesData)
```

Now, let create a vector of filenames using the phenotype information:

```{r files}
files <- file.path(
  folder.kallisto, sprintf(
    "WTCHG_%i_%i", samplesData$Lane, samplesData$sample
  ),
  "abundance.tsv"
)
names(files) <- samplesData$sample
all(file.exists(files))
```

## Table mapping transcripts to gene identifiers

```{r tx2gene}
edb <- EnsDb.Hsapiens.v79
k <- keys(edb, keytype = "GENEID")
df <- select(edb, keys = k, keytype = "GENEID", columns = "TXNAME")
tx2gene <- df[, 2:1]  # tx ID, then gene ID
```

Let us also add the spike-in features in this mapping table:

```{r addERCC}
erccIds <- scan(file.path(folder.expData, "ERCC.txt"), "character")
tx2gene <- rbind(tx2gene, data.frame(TXNAME = erccIds, GENEID = erccIds))
```

## Import expression levels aggregated by gene identifier

```{r tximport, message=FALSE, results='asis'}
txi <- tximport(
  files, type = "kallisto", tx2gene = tx2gene, reader = read_tsv,
  ignoreTxVersion = TRUE
)
names(txi)
pander::pandoc.table(txi$counts[1:5,1:4])
```

# Save the resulting Large List

```{r saveRds}
saveRDS(txi, file.path(folder.rds, "txi.rds"))
```


# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
