---
title: "ERCC spike-in content"
bibliography:
  bibtex.bib
---

```{r checkPkgs, child="_checkLibraries.Rmd", include=FALSE}
```

```{r setup, include=FALSE}
library(DT)
library(ggplot2)
```

# Objective

As an input, `BASiCS` requires the input number of spike-in molecules per cell,
which may be calculated as described in the
[BASiCS tutorial](#https://github.com/catavallejos/TutorialBASiCS)
(file *./Material/TutorialBASiCS2016.html*):

\begin{equation*} \mu_{i} = C_i \times 10^{-18} \times (6.022 \times 10^{23})
\times V \times D \hspace{0.5cm} \mbox{where,} \end{equation*}

* $C_i$ is the concentration of the spike $i$ in the ERCC mix
* $10^{-18}$ is to convert aM (attomolar) to M (molar)
* $6.022 \times 10^{23}$ is the Avogadro number (molecules per mole)
* $V$ is the volume added into each well, here $10\mu{L}$
* $D$ is a dilution factor, here $1:100,000,000$

# Concentrations in ERCC Mix from the manufacturer

Let us import the concentration of individual spike-in molecules
from the *ThermoFisher*
[website](https://www.thermofisher.com/order/catalog/product/4456739):

```{r}
mixC <- read.table(
  "expdata/cms_095046.txt", header = TRUE, sep = "\t",
  stringsAsFactors = FALSE
  )[,c("ERCC.ID","concentration.in.Mix.1..attomoles.ul.")]
rownames(mixC) <- mixC$ERCC.ID
colnames(mixC) <- c("ID","Concentration_attomoles_ul")
datatable(
  mixC,
  options = list(pageLength = 10, searching = TRUE), filter = "top",
  rownames = FALSE
)
```

# Number of individual ERCC molecules added to each cell

Let us calculate from the manufacturer's information and our experimental
protocol:

```{r, message=FALSE}
mixC$ERCCmolecules <-
  mixC$Concentration_attomoles_ul *
  10^-18 *
  6.022 * 10^23 *
  10 *
  10^-8
ggplot(mixC) + geom_histogram(aes(ERCCmolecules)) + scale_x_log10() +
  xlab("Number of molecules injected in sample") +
  geom_vline(xintercept = 10^(0:3), colour = "red", linetype = "dashed")
```

Let us examine the count of individual ERCC Spike-Ins that are expected
at or above different thresholds of $molecules / well$:

```{r}
DT::datatable(data.frame(
  Threshold = 10^(0:3),
  ERCC = sapply(0:3, function(x){sum(mixC$ERCCmolecules >= 10^x)})
), rownames = FALSE)
```

<!-- As HTML pages are built independently, objects need to be saved to disk
and imported as needed by the other R mardown files -->

```{r save_sce.pass, include=FALSE}
saveRDS(mixC, "rds/mixC.rds")
```
