---
title: "Import of RNA-Sequencing read counts and associated metadata"
---

```{r checkPkgs, child="_checkLibraries.Rmd", include=FALSE}
```

<!-- Enable caching -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(xlsx)
library(edgeR)
library(Biobase)
library(scater)
library(rtracklayer)
library(DT)
```

# Phenotype information

Let us import experimental information:

```{r importPheno}
pheno <- read.xlsx("expdata/samples.xlsx", sheetName = "pheno")
datatable(
  pheno,
  options = list(pageLength = 10, searching = TRUE), filter = "top",
  rownames = FALSE
)
```

To facilitate later use, let us retype some fields, reorder some factor levels,
and add an additional field `Sample` field that contains a shortened unique
identifier derived from the count file name for each sample:

```{r editPheno}
pheno$File <- as.character(pheno$File)
pheno$Infection <- factor(pheno$Infection, c("Mock","D23580","LT2","Blank"))
pheno$Status <-
  factor(pheno$Status, c("uninfected","exposed","infected","Blank","BULK"))
pheno$Lane <- as.factor(pheno$Lane)
pheno$Sample <- gsub("WTCHG_[[:digit:]]+_", "", pheno$File) # character
pheno$sampleAnnotated <-
  with(pheno, paste(Time, Infection, Status, Sample, sep = "_")) # character
pheno$Treatment <-
  droplevels(interaction(pheno$Infection, pheno$Status, sep = "_")) # factor
pheno$Group <-
  droplevels(interaction(pheno$Time, pheno$Treatment, sep = "_")) # factor
```

# Counts {#Counts}

To import count data that is stored in a separate file for each sample,
let us use the [edgeR](http://bioconductor.org/packages/edgeR)
`readDGE` method and the above `data.frame` of phenotype information:

```{r readDGE}
RG <- readDGE(
  pheno$File,
  "counts",
  c(1, 3),
  paste(pheno$Time, pheno$Infection, pheno$Status, sep = "_"),
  pheno$Sample
)
dim(RG)
```

# `SCESet`

Let us assemble the phenotype information and count data above into
[scater](http://bioconductor.org/packages/scater) `SCESet` objects:

* one `SCESet` that contains all samples together
* one for the blank samples
* one for the bulk samples
* one for the single cells

**Note:**
This distinction will be important later, to compute QC metrics within
particular sets of samples.

First, let us set the `rownames` of the phenotypes information `data.frame` to
the unique sample identifier for consistency with the sample identifier
associated with the count data imported [above](#Counts):

```{r renamePheno}
rownames(pheno) <- pheno$Sample
```

We may then create the four data sets:

```{r sce.all}
pd.all <- new("AnnotatedDataFrame", data = pheno)
sce.all <- newSCESet(countData = RG$counts, phenoData = pd.all)
dim(sce.all)
```
```{r sce.blank}
idx.blank <- pheno$Status == "Blank"
pd.blank <- new("AnnotatedDataFrame", data = droplevels(pheno[idx.blank,]))
sce.blank <- newSCESet(countData = RG$counts[,idx.blank], phenoData = pd.blank)
dim(sce.blank)
```
```{r sce.bulk}
idx.bulk<- pheno$Status == "BULK"
pd.bulk <- new("AnnotatedDataFrame", data = droplevels(pheno[idx.bulk,]))
sce.bulk <- newSCESet(countData = RG$counts[,idx.bulk], phenoData = pd.bulk)
dim(sce.bulk)
```
```{r sce.sc}
idx.sc <- pheno$Status %in% c("exposed","infected","uninfected")
pd.sc <- new("AnnotatedDataFrame", data = droplevels(pheno[idx.sc,]))
sce.sc <- newSCESet(countData = RG$counts[,idx.sc], phenoData = pd.sc)
dim(sce.sc)
```

# Gene annotations

To ensure coherence between gene annotations and gene expression data,
let us also import the `gene` records from the same GTF file that was supplied
to *featureCounts* as `GRanges` of the
[GenomicRanges](http://bioconductor.org/packages/GenomicRanges) package:

```{r import.gff}
gtfData <- import.gff("expdata/genes_with_ERCC.gtf", feature.type = "gene")
names(as.data.frame(gtfData))
```

**Note:** the above code does *not* import records that relate to ERCC
spike-in features, as these are annotated with `exon` *type* (GTF column 2)
from the `ERCC` *soure* (GTF column 3).

<!-- As HTML pages are built independently, objects need to be saved to disk
and imported as needed by the other R mardown files -->

```{r saveRDS, include=FALSE}
saveRDS(sce.all, "rds/sce.all.rds")
saveRDS(sce.blank, "rds/sce.blank.rds")
saveRDS(sce.bulk, "rds/sce.bulk.rds")
saveRDS(sce.sc, "rds/sce.sc.rds")
saveRDS(gtfData, "rds/gtfData.rds")
```
