---
title: "Comparison HISAT2+featureCounts vs. Kallisto"
---

```{r loadKallisto, echo=FALSE}
knitr::opts_chunk$set(cache = TRUE)
kallisto_all <- readRDS("../../01-kallisto/rds/all_EnsDb.rds")
sce.all <- readRDS("rds/sce.all.rds")
```

# Preprocessing

```{r kallisto_all_fix, include=FALSE}
# Quick fix: update phenodata in Kallisto to match current ones
kallisto_all$Infection <- factor(
  kallisto_all$Infection,
  levels = c("Blank", "D23580", "LT2", "Mock"),
  labels = c("Blank", "STM-D23580", "STM-LT2", "Mock"))
kallisto_all$Status <- factor(
  kallisto_all$Status,
  levels = c("Blank", "BULK", "exposed", "infected", "uninfected"),
  labels = c("Blank", "Bulk", "Exposed", "Violet +", "Uninfected"))
```

Let us identify features present in both the _Kallisto_ and the
_HISAT2 & featureCounts_ (hereafter, "_HF_") pipelines:

```{r overlapKallistoAligned}
features.overlap <-
  intersect(fData(kallisto_all)[,"feature_id"], featureNames(sce.all))
```

Let us reorder features in the _Kallisto_  to match the order of the
_HF_ pipeline, and retain only features observed in both data sets:

```{r kallisto.overlap}
kallisto.overlap <- kallisto_all[
  match(features.overlap, fData(kallisto_all)$feature_id),
]
sce.overlap <- sce.all[features.overlap,]
rm(kallisto_all, sce.all)
```

Let us identfy ERCC features for specific visualisation:

```{r erccIds}
ercc <- grepl("^ERCC-", features.overlap)
table(ercc)
```

# Comparative read counts at bulk level

Let us collect the count of sequenced fragments produce by each pipeline
for each gene feature:

```{r prepareBulkPlot, message=FALSE}
comp.bulks <- data.frame(
  Time = character(),
  Infection = character(),
  Kallisto = integer(),
  HF = integer(),
  ERCC = logical()
)
for (time in c("2h", "4h", "6h")){
  for (infection in c("Mock", "STM-D23580", "STM-LT2")){
    message(sprintf("%s - %s", time, infection))
    kallisto <- counts(kallisto.overlap[
      ,
      kallisto.overlap$Time == time &
        kallisto.overlap$Infection == infection &
        kallisto.overlap$Status == "Bulk"
    ])[,1]
    hf = counts(sce.overlap)[
      ,
      sce.overlap$Time == time &
        sce.overlap$Infection == infection &
        sce.overlap$Status == "Bulk"
    ]
    new.df <- data.frame(
      Time = time,
      Infection = infection,
      Kallisto = kallisto,
      HF = hf,
      ERCC = ercc
    )
    comp.bulks <- rbind(comp.bulks, new.df)
  }
}
rm(kallisto, hf, time, infection, new.df)
```

We may then compare the estimated counts produced by _Kallisto_ to
the counts produced by the _HF_ pipeline,
while indicating the identity relationship by a green dashed line:

```{r bulkPlot, echo=FALSE}
ggplot(comp.bulks, aes(log10(Kallisto + 1), log10(HF + 1))) +
  geom_hex() +
  geom_abline(
    slope=1, intercept=0, alpha=0.5, colour="green", linetype="dashed") +
  coord_fixed(ratio = 1) +
  facet_grid(Infection ~ Time) +
  geom_point(
    aes(log10(Kallisto + 1), log10(HF + 1)),
    subset(comp.bulks, ERCC),
    colour = "red", size = 0.5) +
  labs(
    y="HISAT2 & featureCounts", x="Kallisto",
    title=expression(log[10]*(counts+1))
  ) +
  guides(fill = "none") + theme_bw()
ggsave("NGS-SIG/Kallisto-HISAT2_bulk.pdf", width = 5, height = 4)
ggsave("02_out/Kallisto-HISAT2_bulk.pdf", width = 5, height = 4)
```

# Comparative read counts in random single cells

```{r prepareSCplot, message=FALSE}
set.seed(1985)
comp.sc <- data.frame(
  Time = character(),
  Infection = character(),
  Status = character(),
  Kallisto = numeric(),
  THF = numeric(),
  ERCC = logical()
)
cells.df <- data.frame(
  Cell = character(),
  Time = character(),
  Infection  = character()
)
for (time in c("2h", "4h", "6h")){
   for (infection in c("Mock", "STM-D23580", "STM-LT2")){
     message(sprintf("%s - %s", time, infection))
    sname <- as.character(sample(subset(
      pData(kallisto.overlap),
      Time == time & Infection == infection & Status != 'Bulk',
      select = sample,
      drop = TRUE), 1))
    kallisto <- counts(kallisto.overlap[,sname])[,1]
    hf = counts(sce.overlap)[,sname]
    new.df <- data.frame(
      Time = time,
      Infection = infection,
      Status = subset(
        pData(kallisto.overlap), sample == sname, select = Status, drop = TRUE
      ),
      Kallisto = kallisto,
      HF = hf,
      ERCC = ercc
    )
    comp.sc <- rbind(comp.sc, new.df)
    newCell <- data.frame(
      Cell = sprintf("Cell %s", sname),
      Time = time,
      Infection = infection
    )
    cells.df <- rbind(cells.df, newCell)
  }
}
```

```{r SCplot, echo=FALSE}
cells.df$Y = 5
cells.df$X = 0
ggplot(comp.sc, aes(log10(Kallisto + 1), log10(HF + 1))) +
  geom_hex(aes(colour = Status)) +
  geom_abline(
    slope=1, intercept=0, alpha=0.75, linetype="dashed") +
  coord_fixed(ratio = 1) +
  facet_grid(Infection ~ Time) +
  geom_point(
    aes(log10(Kallisto + 1), log10(HF + 1)),
    subset(comp.sc, ERCC),
    colour = "red", size = 0.5) +
  ggrepel::geom_label_repel(
    aes(X, Y, label = Cell), cells.df, min.segment.length = unit(3, "lines"),
    size = 2, alpha = 0.5) +
  labs(
    y="HISAT2 & featureCounts", x="Kallisto",
    title=expression(log[10]*(counts+1))
  ) +
  scale_fill_continuous(name = "Density") +
  guides(fill = "none") + theme_bw()
ggsave("NGS-SIG/Kallisto-HISAT2_SC9.pdf", width = 6, height = 4)
ggsave("02_out/Kallisto-HISAT2_SC9.pdf", width = 6, height = 4)
```
