---
title: "Normalisation"
---

```{r checkPkgs, child="_checkLibraries.Rmd", include=FALSE}
```

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
library(scran)
library(scater)
library(ggplot2)
sce.filtered <- readRDS("rds/sce.filtered.rds")
```

# Sum factors

Having identified the cells of good quality and the features robustly
detected across those cells, we may now normalise expression levels between
cells using an approach suitable for single cell data sets.

Considering the multifactorial experimental design of this data set, and the
expected differential expression in `exposed` and `infected` cells,
let us cluster similar cells together prior to normalising the cells in each
cluster using the deconvolution method implemented in the
[scran](http://bioconductor.org/packages/scran) package:

```{r quickCluster}
clusters <- quickCluster(sce.filtered)
```

In this case, none of the cells were assigned to any cluster, which suggests
that despite the expected differences between experimental groups, the data set
is sufficiently homogenous for the normalisation procedure to be applied to
all cells equally:

```{r computeSumFactors}
sce.filtered <- computeSumFactors(sce.filtered, clusters = clusters)
summary(sizeFactors(sce.filtered))
```

In this case, the size factors are somewhat correlated with library size,
while significant scatter is visible around the trend;
augmenting the figure with experimental phenotype information emphasises
how the effect of library size on the size factor of individual cells is
subject to factors related to the experimental design of the experiment.

In particular, the figure below emphasises how the effect of library size is
more important in `uninfected` cells, and at the earliest time point
(*i.e.*, `2h`), when the biological effect of infection may not have fully
affect the transcriptome of most cells yet.

In contrast, at later time points (*i.e.*, `4h` and `6h`), trends indicate a
markedly weaker effect of library size on size factor.
This suggest that other sources of systematic differences between cells
arose, and that differential expression is to be expected
between experimental groups at those time points:

```{r sizeFactor_librarySize}
ggplot(
  pData(sce.filtered), aes(total_counts / 1E6, size_factor)) +
  geom_point(aes(colour = Time, shape = Status)) +
  labs(x = "Library size (millions)", y = "Size factor") +
  stat_smooth(
    aes(colour = Time, linetype = Status),
    method = "lm", se = TRUE, size = 0.5, alpha = 0.1
  )
```

Let us examine further the effect of experimental covariates on the size
factors:

```{r sizeFactor_expFactors}
ggplot(
  pData(sce.filtered),
  aes(interaction(Infection, Status, Time, sep = " - "), size_factor)) +
  geom_boxplot(aes(alpha = Time, fill = Infection, linetype = Status)) +
  coord_flip() +
  labs(y = "Size factor", x = "")
```

In a different perspective, the above figure emphasises again how
size factors are affected by experimental factors in this data set;
in particular at the later time points, size factors for cells infected by
or exposed to bacteria display markedly distinct size factors,
irrespective of library size.
This suggests that `Infection`, and `Status` drive more important
systematic differences between cells than library size at those time points,
having likely reshaped the transcriptional profile of stimulated cells.

# Spike factors

To ensure normalization is performed correctly, let us compute a separate set
of size factors for the spike-in set. This assumes
that none of the spike-in transcripts are differentially expressed.

```{r computeSpikeFactors}
setSpike(sce.filtered) <- "ERCC"
sce.filtered <- computeSpikeFactors(
  sce.filtered, type = "ERCC", general.use = FALSE
)
```

We may then examine the spike factors calculated:

```{r spikeFactors_ggplot}
ggplot(pData(sce.filtered), aes(total_counts / 1E6, size_factor_ERCC)) +
  geom_point(aes(colour = Time, shape = Status)) +
  stat_smooth(method = "lm") +
  labs(x = "Library size", y = "Spike factor")
```

In contrast to size factors calculated using endogenous features, spike factors
calculated using ERCC spike-in display a much weaker correlation with
library size, and no trend with experimental factors.
This reinforces the absence of extreme outliers after the removal
of cells with abnormal ERCC content in an [earlier section](#ERCCoutliers);
moreover, it suggests that counts assigned to individual ERCC spike-in features
are similar across all the cells retained after quality control.

# Apply size factors to normalise

The count data are used to compute normalized log-expression values for use in
downstream analyses.
Each value is defined as the log-ratio of each count to the size factor for the
corresponding cell, after adding a prior count of `1` to avoid undefined values
at zero counts.
Division by the size factor ensures that any cell-specific biases are removed.
Spike-in-specific size factors present are applied to normalize the spike-in
transcripts separately from the endogenous features:

```{r normalise}
sce.norm <- normalize(sce.filtered)
```

<!-- As HTML pages are built independently, objects need to be saved to disk
and imported as needed by the other R mardown files -->

```{r saveRDS, include=FALSE}
saveRDS(sce.norm, "rds/sce.norm.rds")
```
