---
title: "Overview of normalised expression data"
---

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(scater)
library(scran)
library(ggplot2)
library(SummarizedExperiment)
sce.norm <- readRDS("rds/sce.norm.rds")
sce.all <- readRDS("rds/sce.all.rds")
```

# Plotting themes

First of all, let us define theme elements used throughout various figures
in the following sections:

```{r PCAtheme}
PCAtheme <- theme(
    panel.grid = element_blank(),
    legend.position = "bottom", legend.box = "vertical"
  )
```


# Relative importance of experimental factors {#explanatoryVariables .tabset}

Having normalised the expression data for detected features in cells that
passed quality control, the relative importance of experimental factors---both
technical and biological---provides insight into uninteresting technical
factors that may need to be accounted for when explaining variance in the
expression data.

## Explanatory variables

Let us examine the proportion of variance explained by known experimental
factors, as well as the total spike-in count in each cell:

```{r plotExplanatoryVariables, echo=FALSE}
plotExplanatoryVariables(
  sce.norm,
  exprs_values = "norm_exprs",
  variables = c(
    "counts_feature_controls_ERCC",
    "log10_counts_feature_controls_ERCC",
    "Plate", "Lane", "Time", "Status", "Infection"
    )
)
```

The above figure reinforces the effect of `Time` as the primary source of
variance in gene expression, with substantial proportion of variance
explained for a subset of features. Next are `Status` and `Infection`
factors, while other technical factors share explanation of lower amount
of the residual variance in the data set.

**Note:**
Due to technical constraints, the the representation of time points is uneven
across the four plates used for library preparation.
In summary, while 75% of each plate is balanced, the remaining 25% are:

* a block of single cells collected at 2h
* a block of single cells collected at 4h
* a block of single cells collected at 6h
* an even mixture of cells from 2h, 4h, and 6h

As a result, it expected some variance explained by the `Plate` factor,
as it is partially confounded with the powerful `Time` factor.

For reference, the plate design is shown in the second tab.

## Plate design (full)

```{r plateDesign_full, echo=FALSE}
plateDesign <-
  dplyr::mutate(
    pData(sce.all)[,c("Plate","Well","Time","Infection","Status")],
    Row = as.factor(gsub("([[:alpha:]])[[:digit:]]*", "\\1", Well)),
    Column = as.factor(as.numeric(gsub("[[:alpha:]]([[:digit:]]*)", "\\1", Well))),
    Plate = gsub("P", "Plate ", Plate)
  )
ggplot(plateDesign, aes(Column, Row, fill = Time, colour = Infection, alpha = Status)) +
  geom_tile(size = 1) + facet_wrap(~ Plate) + labs(x="",y="") +
  scale_y_discrete(limits = rev(levels(plateDesign$Row))) +
  theme_minimal() + theme(panel.grid = element_blank()) +
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Set1") +
  guides(
    fill = guide_legend(order = 1),
    colour = guide_legend(order = 2),
    shape = guide_legend(order = 3)
  )
```

## Plate design (Cells passed QC)

```{r plateDesign_pass, echo=FALSE}
plateDesign <-
  dplyr::mutate(
    pData(sce.norm)[,c("Plate","Well","Time","Infection","Status")],
    Row = as.factor(gsub("([[:alpha:]])[[:digit:]]*", "\\1", Well)),
    Column = as.factor(as.numeric(gsub("[[:alpha:]]([[:digit:]]*)", "\\1", Well))),
    Plate = gsub("P", "Plate ", Plate)
  )
ggplot(plateDesign, aes(Column, Row, fill = Time, colour = Infection, alpha = Status)) +
  geom_tile(size = 1) + facet_wrap(~ Plate) + labs(x="",y="") +
  scale_y_discrete(limits = rev(levels(plateDesign$Row))) +
  theme_minimal() + theme(panel.grid = element_blank()) +
  scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Set1") +
  scale_alpha_discrete(range = c(0.1, 3/5)) +
  guides(
    fill = guide_legend(order = 1),
    colour = guide_legend(order = 2),
    shape = guide_legend(order = 3)
  )
```


# Clustering

Let us examine the unsupervised clustering of cells using different approaches.

## Principal component analysis (PCA) {#PCA}

### All cells

Let us initially focus our attention on the two principal components that
explain the largest amount of variance in the endogenous features.

**Note 1:** Some important *default* values are:

* `ntop=500`: only the `500` most variable features are used
* `scale_features=TRUE`: standardises expression values so that each feature
  has unit variance; this procedure normalises the relative importance of
  features expressed at high and low levels in the calculation of principal
  components.

**Note 2:** Some important *non-default* values are:

* `exprs_values="norm_exprs"`:
  the normalised log~2~-transformed CPM are used.
* `feature_set=!isSpike(sce.norm)`:
  ERCC spike-in features are not considered
  (see this [earlier section](05_normalisation.html#setSpike) where the
  set of spike-in features is defined).

```{r plotPCASCESet}
norm_fpkm(sce.norm) <- calculateFPKM(sce.norm, fData(sce.norm)[,"width"])
sce.norm <- plotPCASCESet(
  sce.norm, exprs_values = "norm_fpkm",
  colour_by = "Treatment", shape_by = "Time",
  feature_set = !isSpike(sce.norm), return_SCESet = TRUE, draw_plot = FALSE
)
```

#### Principal components 1 and 2 {.tabset}

##### Group

```{r plotPCASCESet_Group, fig.height=6}
plotReducedDim.SCESet(sce.norm, 2, colour_by = "Group", shape_by = "Time") +
  PCAtheme + theme(legend.text = element_text(size = 6))
```

The above figure emphasises the importance of the `Time` factor, significantly
associated with `PC1`; in the figure, cells progress from the earliest time
points at the lowest coordinates on `PC1` to the latest time points at the
highest coordinates of `PC1`.

In addition, `PC2` contributes to the increasingly marked separation of
stimulated cells (*i.e.*, `infected` and `stimulated`,
in orange, green, red, and blue) from `uninfected` cells (in blue);
`uninfected` cells generally appear at lower coordinates than their `infected`
and `exposed` counterparts on both `PC2` and `PC1`, with this trend becoming
increasingly apparent over time.

Overall, the observations above suggest different transcriptional trajectories
taken by cells in response to experimental stimuli.

##### Time

```{r plotPCASCESet_Time, fig.height=5}
plotReducedDim.SCESet(sce.norm, 2, "Time", "Plate") +
  PCAtheme
```

##### Infection

```{r plotPCASCESet_Infection, fig.height=5}
plotReducedDim.SCESet(sce.norm, 2, "Infection", "Time") +
  PCAtheme
```

##### Status

```{r plotPCASCESet_Status, fig.height=5}
plotReducedDim.SCESet(sce.norm, 2, "Status", "Time") +
  PCAtheme
```

##### Plate

```{r plotPCASCESet_Plate, fig.height=5}
plotReducedDim.SCESet(sce.norm, 2, "Plate", "Time") +
  PCAtheme
```

#### Additional principal components {.tabset}

The analysis may be extended to more principal components; however, by
definition those principal components will report decreasing amount of variance
in the data set:

##### Percentage variance explained

Percentage of variance explained by prinicipal components:

```{r PercentVariance_PCA, echo=FALSE}
ggplot(
  data.frame(
    PercentVariance = attr(reducedDimension(sce.norm), "percentVar"),
    PC = factor(seq(1:10))
  )
) + labs(x = "Principal component", y = "Percentage variance explained") +
  geom_col(aes(PC, PercentVariance)) + theme_minimal()
```

##### Treatment

```{r plotPCASCESet_PC1to3_Treatment}
plotReducedDim.SCESet(sce.norm, 3, colour_by = "Treatment", shape_by = "Time") +
  PCAtheme + theme(legend.text = element_text(size = 6))
```

##### Time

```{r plotPCASCESet_PC1to3_Time}
plotReducedDim.SCESet(sce.norm, 3, colour_by = "Time", shape_by = "Plate") +
  PCAtheme
```

##### Status

```{r plotPCASCESet_PC1to3_Status}
plotReducedDim.SCESet(sce.norm, 3, colour_by = "Status", shape_by = "Time") +
  PCAtheme
```

##### Infection

```{r plotPCASCESet_PC1to3_Infection}
plotReducedDim.SCESet(sce.norm, 3, colour_by = "Infection", shape_by = "Time") +
  PCAtheme
```

### PCA analyses within time points {.tabset}

In the above [section](#PCA), the overwhelming influence of the `Time` factor
likely shadows the relatively weaker effect of other biologically relevant
factors `Infection` and `Status`.

Let us perform PCA at each time point:

```{r time.sce}
time.sce <- list()
for (time in levels(sce.norm$Time)){
  tmp.sce <- sce.norm[,sce.norm$Time == time]
  keep_feature <- (rowMeans(counts(tmp.sce)) >= 1)
  tmp.sce <- tmp.sce[keep_feature,]
  tmp.sce <- plotPCASCESet(
    tmp.sce, exprs_values = "norm_fpkm",
    feature_set = !isSpike(tmp.sce), return_SCESet = TRUE, draw_plot = FALSE
  )
  time.sce[[time]] <- tmp.sce
  rm(tmp.sce)
}
rm(time)
```

**Note:**
Below, the tab name indicates the experimental factor used to *colour*
single cells (*i.e.*, `Infection` or `Status`); the other factor is then
used to *shape* each data point.

#### 2h {.tabset}

##### Infection

```{r 2h-Infection, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  time.sce[["2h"]], 3, colour_by = "Infection", shape_by = "Status") + PCAtheme
```

##### Status

```{r 2h-Status, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  time.sce[["2h"]], 3, colour_by = "Status", shape_by = "Infection") + PCAtheme
```

##### Treatment

```{r 2h-Treatment, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(time.sce[["2h"]], 3, colour_by = "Treatment") + PCAtheme
```

#### 4h {.tabset}

##### Infection

```{r 4h-Infection, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  time.sce[["4h"]], 3, colour_by = "Infection", shape_by = "Status") + PCAtheme
```

##### Status

```{r 4h-Status, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  time.sce[["4h"]], 3, colour_by = "Status", shape_by = "Infection") + PCAtheme
```

##### Treatment

```{r 4h-Treatment, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(time.sce[["4h"]], 3, colour_by = "Treatment") + PCAtheme
```

#### 6h {.tabset}

##### Infection

```{r 6h-Infection, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  time.sce[["6h"]], 3, colour_by = "Infection", shape_by = "Status") + PCAtheme
```

##### Status

```{r 6h-Status, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  time.sce[["6h"]], 3, colour_by = "Status", shape_by = "Infection") + PCAtheme
```

##### Treatment

```{r 6h-Treatment, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(time.sce[["6h"]], 3, colour_by = "Treatment") + PCAtheme
```

## t-distributed stochastic neighbour embedding (t-SNE)

### Experimental factors {.tabset}

Let us apply the t-distributed stochastic neighbour embedding (t-SNE) dimensionality reduction technique on the normalised expression of the
**500** most variable endogenous features to check for substructure in the data set.
Importantly, let us set the `perplexity` parameter of the `Rtsne` function to
half the average size of an experimental group of single cells
(*i.e.*, `r round(mean(table(sce.norm$Group)) / 2)`; see this
[earlier section](04_feature_QC.html#selectFeatures) for the count of cells
per group).

**Note**: "In t-SNE, the perplexity may be viewed as a knob that sets the number
of effective nearest neighbors. It is comparable with the number of nearest
neighbors $k$ that is employed in many manifold learners
(see the GitHub website"
[Laurens van der Maaten](https://lvdmaaten.github.io/tsne/)).
As a consequence, an appropriate value of perplexity is necessary to
allow data points (here, single cells) to form clusters with a reasonable
number of nearest neighbours;
instead, excessively large values of perplexity usually result
in large 'balls' of uniformly distributed points, as all data points attempt
to be equidistant from each other:

```{r plotTSNE}
sce.norm <- plotTSNE(
  sce.norm, ncomponents = 2, exprs_values = "norm_fpkm",
  colour_by = "Status", shape_by = "Time", return_SCESet = TRUE,
  feature_set = !isSpike(sce.norm),
  perplexity = round(mean(table(sce.norm$Group)) / 2),
  rand_seed = 1794, draw_plot = FALSE
)
```

```{r saveRDS, include=FALSE}
saveRDS(sce.norm, 'rds/sce.norm.fpkm.tSNE.rds')
```

```{r postertSNE, include=FALSE}
postertSNE <- reducedDimension(sce.norm)
colnames(postertSNE) <- paste("Dimension", 1:2)
posterSamples <- rownames(postertSNE)
postertSNE <- cbind(postertSNE, pData(sce.norm)[
  posterSamples,c("Status","Infection","Time")])
lc <- RColorBrewer::brewer.pal(3, "Set1")[c(1,3,2)] # RGB
posterLabels <- data.frame(
  x = c(-20, -2.5, -5, 10, 15),
  y = c(-12.5, -5, 10, -2.5, 10),
  label = c("2h", "4h", "6h", "4h", "6h"),
  col = c("black", lc[1], lc[1], "black", "black")
)
posterPlot <- ggplot(
  postertSNE,
  aes_string("`Dimension 1`","`Dimension 2`")) +
  geom_point(aes(colour=Status,shape=Infection)) +
  theme_light()
posterPlot
ggsave("20170601_06_fpkm/tSNE_right.pdf", posterPlot, width = 7, height = 6)
ggsave(
  "20170601_06_fpkm/tSNE_bottom.pdf",
  posterPlot + theme(
    legend.position = "bottom", legend.box = "vertical"
  ),
  width = 6, height = 6)
```

#### Time {#tSNEtime}

```{r plotTSNE_time}
plotReducedDim.SCESet(sce.norm,2,colour_by="Time",shape_by="Status") + PCAtheme
```

#### Status {#tSNEstatus}

```{r plotTSNE_status}
plotReducedDim.SCESet(sce.norm,2,colour_by="Status",shape_by="Time") + PCAtheme
```

#### Infection

```{r plotTSNE_infection}
plotReducedDim.SCESet(sce.norm,2,colour_by="Infection",shape_by="Time") + PCAtheme
```

#### Treatment

```{r plotTSNE_Treatment}
plotReducedDim.SCESet(sce.norm,2,colour_by="Treatment",shape_by="Time") + PCAtheme
```

#### Plate (batch)

```{r plotTSNE_Plate}
plotReducedDim.SCESet(sce.norm,2,colour_by="Plate",shape_by="Time") + PCAtheme
```

#### quickCluster (normalisation)

We may also overlay the clusters identify during
[normalisation](05_normalisation.html) onto the t-SNE:

```{r plotTSNE_quickCluster}
plotReducedDim.SCESet(sce.norm,2,colour_by="Cluster",shape_by="Time") + PCAtheme
```

### Genes {.tabset}

Alternatively, we may overlay the expression level of a particular gene onto
the data points as a colour gradient.

For this purpose, let us first set a common colour scale that ranges from
the lowest to the highest expression values observed in the data set:

```{r exprScale}
exprScale <- range(exprs(sce.norm))
colourScale <- rev(RColorBrewer::brewer.pal(9, "YlGnBu"))
```


#### TNF

```{r plotTSNE_TNF, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="TNF"))) + PCAtheme +
  scale_colour_gradientn(name="TNF",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### IFNB1

```{r plotTSNE_IFNB1, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="IFNB1"))) + PCAtheme +
  scale_colour_gradientn(name="IFNB1",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### IL12B

```{r plotTSNE_IL12B, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="IL12B"))) + PCAtheme +
  scale_colour_gradientn(name="IL12B",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### IL36G

```{r plotTSNE_IL36G, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="IL36G"))) + PCAtheme +
  scale_colour_gradientn(name="IL36G",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### IL1B

```{r plotTSNE_IL1B, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="IL1B"))) + PCAtheme +
  scale_colour_gradientn(name="IL1B",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD1A

```{r plotTSNE_CD1A, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="CD1A"))) + PCAtheme +
  scale_colour_gradientn(name="CD1A",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD1B

```{r plotTSNE_CD1B, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm), gene_name=="CD1B"))) + PCAtheme +
  scale_colour_gradientn(name="CD1B",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD1C

```{r plotTSNE_CD1C, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="CD1C"))) + PCAtheme +
  scale_colour_gradientn(name="CD1C",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### RAMP1

```{r plotTSNE_RAMP1, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="RAMP1"))) + PCAtheme +
  scale_colour_gradientn(name="RAMP1",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD40

```{r plotTSNE_CD40, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="CD40"))) + PCAtheme +
  scale_colour_gradientn(name="CD40",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD80

```{r plotTSNE_CD80, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="CD80"))) + PCAtheme +
  scale_colour_gradientn(name="CD80",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD83

```{r plotTSNE_CD83, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="CD83"))) + PCAtheme +
  scale_colour_gradientn(name="CD83",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### ALOX15

```{r plotTSNE_ALOX15, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="ALOX15"))) + PCAtheme +
  scale_colour_gradientn(name="ALOX15",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD209

```{r plotTSNE_CD209, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="CD209"))) + PCAtheme +
  scale_colour_gradientn(name="CD209",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CTSB

```{r plotTSNE_CTSB, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="CTSB"))) + PCAtheme +
  scale_colour_gradientn(name="CTSB",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

#### CD86

```{r tSNE_CD86, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="CD86"))) + PCAtheme +
  scale_colour_gradientn(name="CD86",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

```{r tSNE_CD86_pdf, include=FALSE}
ggsave("20170601_06_fpkm/tSNE_CD86.pdf", width = 5, height = 5)
```

#### MARCH1

```{r tSNE_MARCH1, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="MARCH1"))) + PCAtheme +
  scale_colour_gradientn(name="MARCH1",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

```{r tSNE_MARCH1_pdf, include=FALSE}
ggsave("20170601_06_fpkm/tSNE_MARCH1.pdf", width = 5, height = 5)
```

#### IL10

```{r tSNE_IL10, message=FALSE, echo=FALSE, fig.height=6}
plotReducedDim.SCESet(
  sce.norm, ncomponents = 2, shape_by = "Time",
  colour_by = rownames(subset(fData(sce.norm),gene_name=="IL10"))) + PCAtheme +
  scale_colour_gradientn(name="IL10",limits=exprScale,colours=colourScale) +
  guides(colour = guide_colourbar(order = 1),shape = guide_legend(order = 2))
```

```{r tSNE_IL10_pdf, include=FALSE}
ggsave("20170601_06_fpkm/tSNE_IL10.pdf", width = 5, height = 5)
```

## Diffusion map

Finally, let us also produce a diffusion map plot of two components as a
dimensionality reduction of the endogenous features:

```{r plotDiffusionMapSCESet}
plotDiffusionMapSCESet(
  sce.norm, exprs_values = "norm_fpkm",
  colour_by = "Status", shape_by = "Time",
  feature_set = !isSpike(sce.norm),
  rand_seed = 1794) + PCAtheme
```
