---
title: "Overview of normalised expression data"
---

```{r checkPkgs, child="_checkLibraries.Rmd", include=FALSE}
```

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(scater)
library(scran)
library(ggplot2)
library(SummarizedExperiment)
library(DT)
sce.norm <- readRDS("rds/sce.norm.rds")
```

# Highly variable genes (HVGs)
## Identifying variable genes {#identifyHVGs}

### Using ERCC spike-in features

Let us fit a trend to the variance of spike-in features while accounting
for the minor effect of technical factors described
[earlier](#explanatoryVariables).
We do not block on the key experimental factors `Time`, `Infection`, and
`Status`, as doing so may regress out biological effects of interest.

```{r model.matrix_technical}
dm <- model.matrix(~Plate+Lane, data = pData(sce.norm))
colnames(dm)
```

Having defined the design matrix, we fit the mean-dependent trend to the
gene-specific variances in the data set:

```{r trendVar_ercc}
var.fit.ercc <- trendVar(
  sce.norm, assay = "norm_exprs", trend = "loess", design = dm,
  span = 0.4
)
names(var.fit.ercc)
```

We may then decompose the gene-specific variance into biological and technical
components using the ERCC spike-in trend as an estimate of technical variance:

```{r decomposeVar_ercc}
var.out.ercc <- decomposeVar(sce.norm, var.fit.ercc, assay = "norm_exprs")
names(var.out.ercc)
```

Finally, we may plot the fitted trend, while highlighting ERCC spike-in
features used to obtain it:

```{r plot_var.out.ercc}
plot(
  var.out.ercc$mean, var.out.ercc$total, pch = 16, cex = 0.6,
  xlab = "Mean log-expression", ylab = "Variance of log-expression"
)
points(var.fit.ercc$mean, var.fit.ercc$var, col="red", pch=16)
o <- order(var.out.ercc$mean)
lines(var.out.ercc$mean[o], var.out.ercc$tech[o], col="red", lwd=2)
```

While the above figure suggests that the trend fits accurately spike-in
features at extreme values the detection range, it also indicate substantial
scatter around the trend for feature at intermediate levels of detection,
and likely understimates technical variance in both ERCC spike-in and
endogenous features at moderately high levels of detection
(*i.e.*, mean log-expression in the range `5<X<10`).

### Using endogenous features

Considering the technical variance in ERCC spike-in detection levels observed
in the previous section, an alternate method of estimating technical variation
across the dynamic range of expression uses endogenous features expression
levels under the assumption that most endogenous features exhibit mostly
technical variation, in contrast to little biological variation.

Similarly to the use of ERCC spike-in features, let us use the
[scran](http://bioconductor.org/packages/scran) `trendVar` method
to estimate technical variation under the assumption stated above,
accouting for the same technical factors of the experimental design:

```{r trendVar_endo}
var.fit.endo <- trendVar(
  sce.norm, assay = "norm_exprs", trend = "loess", design = dm,
  use.spikes = FALSE, span = 0.1
)
names(var.fit.endo)
```

Similarly, the estimate variance is decomposed into its technical and
biological components:

```{r decomposeVar_endo}
var.out.endo <- decomposeVar(sce.norm, var.fit.endo)
names(var.out.endo)
```

We assess the suitability of the trend fitted to the endogenous variances by examining whether it is consistent with the
spike-in variances


```{r plot_var.out.endo}
plot(
  var.out.endo$mean, var.out.endo$total, pch=16, cex=0.6,
  xlab="Mean log-expression", ylab="Variance of log-expression"
)
o <- order(var.out.endo$mean)
lines(var.out.endo$mean[o], var.out.endo$tech[o], col="dodgerblue", lwd=2)
spike.fit <- trendVar(sce.norm, use.spikes=TRUE) # To compute spike-in variances.
points(spike.fit$mean, spike.fit$var, col="red", pch=16)
```

\bioccomment{
The help page for `trendVar` suggests fiddling with arguments for smoothing
such as `span`, `degree`, or `family` or using `semiloess` to produce the
trend. I haved tried the `gaussian` family which barely affects the smoothing
of the ERCC spike-in features, and the `semiloess` trend which does not
converge for ERCC spike-in features.
}

## Model selection

The two methods (*i.e.*, ERCC spike-in and endogenous features) yield markedly
distinct trends.

In particular, the limited count of spike-in features robustly detected
across cells (*i.e.*, `44` ERCC spike-in features with an average count greater
than `1` were retained), scarcer at moderate levels of expression, may
explain to some extent the substantial scatter around the trend at moderate
levels of expression (*i.e.*, mean log-expression in the range `2.5<X<7.5`).
Nevertheless, this trend matches closely observations at low detection levels
dense in ERCC spike-in features, and high detection levels that show limited
variability.

In contrast, the use of endogenous features---under the assumption that most
genes exhibit technical variation and little biological variation---yields a
trend that may understimate technical variation at low levels of detection and
overestimate technical variation at high levels of detection, considering
ERCC spike-in for reference. However, at moderate detection levels,
the overall trend provides a more representative average estimate of variance
matching both ERCC spike-in and endogenous features (always under the
assumption that most spike-in and endogenous features only show technical
variation).

Let us adopt a conservative approach and use the latter trend (*i.e.*,
estimated from the endogenous features) as it is consistent with the spike-in
variances, and as such supporting the assumption that most genes have low
levels of biological variability is valid in this data set.

## Top endogenous features

HVGs are defined as genes with biological components that are significantly
greater than zero at a false discovery rate (FDR) of 5%.
These genes are interesting as they are most likely to drive differences in the
expression profiles between cells that belong to different experimental groups,
while more generally explaining a substantial proportion of the
transcriptional variance across all cells in the data set.

In addition to the definition above, let us define as HVGs endogenous features
with an estimated biological component of variance greater than or equal to
`0.5`
For transformed expression values on the log~2~ scale, this means that the
average difference in true expression between any two cells will be at least
`2`-fold. (This reasoning assumes that the true log-expression
values are Normally distributed with variance of 0.5) (**cite paper 67**).

```{r hvg.out}
hvg.out <- var.out.endo[which(
  var.out.endo$FDR <= 0.05 & var.out.endo$bio >= 0.5
  ),]
hvg.out <- hvg.out[order(hvg.out$bio, decreasing=TRUE),]
nrow(hvg.out)
```

Let us annotate features with their associated gene name:

```{r annotated_hvg.out}
hvg.out <- cbind(
  gene_name = fData(sce.norm)[rownames(hvg.out),]$gene_name,
  hvg.out
)
```

We may then examine the HVGs in a table initially ordered by decreasing
biological variance:

```{r pandoc_topBioVariance}
datatable(
  hvg.out,
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

## Expression levels

Let us examine the distribution of expression values for the top HVGs across
cells to ensure that the variance estimate is not being dominated by a small
count of extreme outliers.

We may highlight the relation between `Time` and those HVGs
(irrespective of other key experimental factors), generally indicating
a progressive *up*-regulation of gene expression over time:

```{r top.hvg_Time}
top.hvg <- rownames(hvg.out)[1:10]
plotExpressionSCESet(
  sce.norm, top.hvg, exprs_values = "norm_exprs", colour_by = "Time"
) +
  scale_x_discrete(labels = fData(sce.norm[top.hvg,])$gene_name)
```

We may also relate the biological variance observed in those same HVGs to the
`Infection` factor, further linking the high variance of those genes to
the differences between bacteria-stimulated and mock-infected cells:

```{r top.hvg_Infection}
plotExpressionSCESet(
  sce.norm, top.hvg, exprs_values = "norm_exprs", colour_by = "Infection"
) +
  scale_x_discrete(labels = fData(sce.norm[top.hvg,])$gene_name)
```

Finally, we may also relate the biological variance in the same HVGs to the 
`Status` factor, which reinforces clear-cut differences between cells
stimulated by bacterial infection or exposure and `uninfected` cells:

```{r top.hvg_Status}
plotExpressionSCESet(
  sce.norm, top.hvg, exprs_values = "norm_exprs", colour_by = "Status"
) +
  scale_x_discrete(labels = fData(sce.norm[top.hvg,])$gene_name)
```

# Expression heatmap {.tabset}

We may then visualise normalised and scaled expression values in
a heat map manually organised by phenotype, while allowing genes with similar
expression profiles to cluster.

Let us first subset the normalised data set to retain only endogenous features:

```{r excludeERCC}
sce.endo <- sce.norm[!fData(sce.norm)$is_feature_control_ERCC,]
```

In addition, let us also reorder samples by `Time`, `Status`, and `Infection`,
to facilitate interpretation of the expression data:

```{r orderSamples}
sce.endo <- sce.endo[,with(pData(sce.endo), order(Time, Status, Infection))]
dim(sce.endo)
```

Let us prepare colours to indicate phenotype levels:

```{r palette9}
col9 <- brewer.pal(10, "Set3")[c(1:8, 10)]
```

First let us define a heat map annotation panel that indicates key phenotype
levels for each cell:

```{r h_column}
h_column <- HeatmapAnnotation(
  df = pData(sce.endo)[,c("Time", "Status", "Infection")],
  col = list(
    Time = c("2h" = col9[1], "4h" = col9[2], "6h" = col9[3]),
    Status=c("uninfected"=col9[4],"exposed"=col9[5],"infected"=col9[6]),
    Infection = c("Mock" = col9[7], "LT2" = col9[8], "D23580" = col9[9])
  )
)
```

We may now examine the expression data of the `r nrow(sce.endo)`
endogenous HVGs:

```{r topHVG.norm}
hvg.norm <- assayData(sce.endo)[["norm_exprs"]][
    rownames(hvg.out)[order(hvg.out$bio, decreasing = TRUE)],
  ]
```

## Normalised

```{r ht.norm, fig.height=7}
ht.norm <- Heatmap(
  hvg.norm,
  name = "norm_exprs", column_title = "Normalised expression",
  top_annotation = h_column,
  cluster_rows = TRUE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE,
  show_row_dend = TRUE
)
draw(ht.norm)
```

## Normalised-scaled

```{r ht.scaled, fig.height=7}
ht.scaled <- Heatmap(
  t(scale(t(hvg.norm))),
  name = "scaled_exprs", column_title = "Scaled expression",
  top_annotation = h_column,
  cluster_rows = TRUE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE,
  show_row_dend = TRUE
)
draw(ht.scaled)
```


<!-- As HTML pages are built independently, objects need to be saved to disk
and imported as needed by the other R mardown files -->

```{r saveRDS, include=FALSE}
saveRDS(hvg.out, "rds/hvg.out.rds")
```
