---
title: "Differential expression using scde"
bibliography:
  bibtex.bib
---

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(scater)
library(scran)
library(scde)
library(ensembldb)
library(RColorBrewer)
sce.norm <- readRDS("rds/sce.norm.rds")
EnsDb.Hsapiens.v79 <- EnsDb.Hsapiens.v79::EnsDb.Hsapiens.v79
```

# Plotting themes

First of all, let us define theme elements used throughout various figures
in the following sections:

```{r exprsViolinTheme}
exprsViolinTheme <- theme(
    panel.grid.major.x = element_blank(),
    legend.position = "bottom", legend.box = "vertical")
```

# Prepare data

The analysis starts with a matrix of read counts,
filtered based on gene and cell requirements.
In this case, let us use the raw numbers of reads mapped to each endogenous
feature, recast as `integer` type:

```{r cleanCounts}
cd <- clean.counts(
  counts(sce.norm)[!isSpike(sce.norm),],
  min.lib.size = 1000, min.reads = 1, min.detected = 1
)
cd <- apply(cd, 2, function(x){storage.mode(x) <- 'integer'; x})
dim(cd)
```

\bioccomment{
The documentation of the `clean.counts` function should be improved.
Two genes are discarded above because they are detected in a single cell each.
The thresholds supplied are taken must be strictly met
(i.e., stricly more than N reads, strictly more than N cells).
}

Let us also define a factor that separate cells from each experimental group:

```{r sg}
sg <- as.factor(
  with(pData(sce.norm), paste(Time, Infection, Status, sep = "_"))
)
names(sg) <- colnames(sce.norm)  
```

# Fitting error models

Here, we fit the error models on which all subsequent calculations
will rely. The fitting process relies on a subset of robust genes that are
detected in multiple cross-cell comparisons. Here we supply the groups
defined above to the `groups` argument, so that the error models for each
experimental group of cells are fit independently
(using `r length(levels(sg))` different sets of "robust" genes in this case).
If the groups argument is omitted, the models would be fit using a common set.

**Note:** this step takes a considerable amount of time, even when multiple
cores are used (~ 1 h using 4 cores to fit error models to each of the
`r length(levels(sg))` experimental groups).

<!--
Switch eval to TRUE below and other chunks in this file
when results should be updated.
-->

```{r oifm, eval=FALSE}
o.ifm <- scde.error.models(counts = cd, groups = sg, n.cores = 4)
```

```{r saveRdsOifm, eval=FALSE, include=FALSE}
saveRDS(o.ifm, "rds/o.ifm.rds")
```

```{r readOIfm, include=FALSE}
o.ifm <- readRDS("rds/o.ifm.rds")
```

The `o.ifm` is a `data.frame` with error model coefficients for each cell.

```{r oIfmTrim}
o.ifm.trim <- apply(o.ifm, 2, round, digits = 2)
DT::datatable(
  o.ifm.trim,
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

In the table above:

* `corr.a` and `corr.b` are slope and intercept of the correlated component fit
* `conc.*` refer to the concomitant fit, `corr.theta` is the NB over-dispersion
* `fail.r` is the background Poisson rate (fixed)

Particularly poor cells may result in abnormal fits, most commonly showing
negative `corr.a`, and should be removed.

```{r validCells}
valid.cells <- o.ifm$corr.a > 0
table(valid.cells)
o.ifm <- o.ifm[valid.cells, ]
```

Here, all the fits were valid (most likely owing to the removal of outliers
during earlier [sample QC](03_sampleQC.html)); as a consequence,
the above chunk of code does not actually do anything to the data.

Finally, we need to define an expression magnitude prior for the genes.
Its main function, however, is to define a grid of expression magnitude values
on which the numerical calculations will be carried out.

```{r oPrior}
o.prior <- scde.expression.prior(
  models = o.ifm, counts = cd, length.out = 400, show.plot = TRUE
)
```

Here we used a grid of **400** points, and let the maximum expression magnitude
be determined by the default **0.999** quantile
(default `max.quantile` setting;
use `max.value` parameter to specify the maximum expression magnitude
explicitly --- on log~10~ scale).

# DE relative to uninfected cells {#DEcontrol}
## Testing for DE genes {.tabset}

To test for differential expression:

* we first match the order of single cells in the fitted error models to the
  order of the single cells in the original count matrix;
* we use the above mapping to identify the experimental group and batch of each
  single cell in the order of the fitted error models,
* we define for each comparison a factor in the form of a vector that specifies
  which two groups of single cells are to be compared; the factor elements
  correspond to the rows of the fitted error models, and contain `NA` values
  for single cells that are not included in either group,
* we let [scde](http://bioconductor.org/packages/scde) estimate differential
  expression between the two groups of single cells, after batch correction;
  *Note that results without batch corrections are also available in this way*,
* we sort genes by decreasing absolute (uncorrected) Z-score of expression
  difference; this order highlights the most significantly DE genes at the top
  of the table, regardless of their direction of differential expression,
* we annotate genes with their gene name (at this stage, genes were only
  identified by their Ensembl gene identifier).
* we store in three separate lists the *full* tables of differential expression
  results (*i.e.*, even if not significantly DE):
    + without batch correction
    + with batch correction applied by plate processed
    + estimation of the batch effect

**Note:** this step takes a considerable amount of time, even when multiple
cores are used (~ 1 h using a single core to estimate batch effect and
differential expression between each group of stimulated single cells and
their matched control).

<!--
Switch eval to TRUE below and other chunks in this file
when results should be updated.
-->

```{r orderSCDEphenotypes}
newOrder <- match(rownames(o.ifm), colnames(sce.norm))
sg.all <- sg[newOrder]
batch <- sce.norm$Plate[newOrder]
```

```{r scdeExprDiff, eval=FALSE, warning=FALSE, message=FALSE}
DE.res.raw <- list()
DE.res.adjusted <- list()
DE.res.batch <- list()
for (time in levels(sce.norm$Time)){
  for (infection in levels(sce.norm$Infection)[-1]){
    for (status in levels(sce.norm$Status)[-1]){
      sg.ref <- paste(
        time,
        levels(sce.norm$Infection)[1],
        levels(sce.norm$Status)[1],
        sep = "_"
      )
      sg.tar <- paste(time, infection, status, sep = "_")
      contrast <- sprintf("%s-%s", sg.tar, sg.ref);# message(contrast)
      groups <- rep(NA, length(sg.all))
      names(groups) <- names(sg.all)
      groups[which(sg.all == sg.ref)] <- sg.ref
      groups[which(sg.all == sg.tar)] <- sg.tar
      groups <- factor(groups, c(sg.tar, sg.ref)) # target - reference
      scde.res <- scde.expression.difference(
        o.ifm, cd, o.prior,
        groups = groups, batch = batch,
        n.cores  =  4
      )
      
      de.raw <- scde.res$results
      de.adjusted <- scde.res$batch.adjusted
      de.batch <- scde.res$batch.effect
      
      de.raw <- de.raw[order(abs(de.raw$Z), decreasing = TRUE),]
      de.adjusted <- de.adjusted[order(abs(de.adjusted$Z), decreasing = TRUE),]
      de.batch <- de.batch[order(abs(de.batch$Z), decreasing = TRUE),]
      
      de.raw <- cbind(
        GENENAME=mapIds(EnsDb.Hsapiens.v79,rownames(de.raw),"GENENAME","GENEID"),
        de.raw
      )
      de.adjusted <- cbind(
        GENENAME=mapIds(EnsDb.Hsapiens.v79,rownames(de.adjusted),"GENENAME","GENEID"),
        de.adjusted
      )
      de.batch <- cbind(
        GENENAME=mapIds(EnsDb.Hsapiens.v79,rownames(de.batch),"GENENAME","GENEID"),
        de.batch
      )
      
      DE.res.raw[[contrast]] <- de.raw
      DE.res.adjusted[[contrast]] <- de.adjusted
      DE.res.batch[[contrast]] <- de.batch
    }
  }
}
```

```{r saveRdsDEvsUninf, eval=FALSE, include=FALSE}
saveRDS(DE.res.raw, "rds/DE.res.raw.rds")
saveRDS(DE.res.adjusted, "rds/DE.res.adjusted.rds")
saveRDS(DE.res.batch, "rds/DE.res.batch.rds")
```

```{r readRdsDEvsUninf, include=FALSE}
DE.res.raw <- readRDS("rds/DE.res.raw.rds")
DE.res.adjusted <- readRDS("rds/DE.res.adjusted.rds")
DE.res.batch <- readRDS("rds/DE.res.batch.rds")
```

## Count of DE genes

Let us display the count of differentially expressed genes:

* Without batch correction (`Raw`)
* With batch correction applied by plate processed (`Adjusted`).
  *Note:
  all samples of any of the `4` plates were processed on the same of the `2`
  sequencing lanes; as a result, accounting for a plate batch implicitely
  accounts for a lane batch.*

```{r countDEvsUninf}
contrastsMock <- grep("Mock", names(DE.res.raw), value = TRUE)
DT::datatable(data.frame(
  "Raw" =
    sapply(contrastsMock, function(n){sum(DE.res.raw[[n]]$cZ != 0)}),
  "Adjusted" =
    sapply(contrastsMock, function(n){sum(DE.res.adjusted[[n]]$cZ != 0)}),
  "Batch" =
    sapply(contrastsMock, function(n){sum(DE.res.batch[[n]]$cZ != 0)})
  ),
  options = list(pageLength = 12)
)
```

## Tables of DE genes {.tabset}

Let us display the tables of DE genes for the various contrasts calculated
above.
For convience, let us display only the `1000` most significant DE genes
(*i.e.*, largest Z-score) in cases where larger counts of DE genes were
detected (*i.e.*, corrected Z-score ≠ 0).

```{r topDElimit}
topDElimit <- 1000
DE.adjusted.limit <- list()
for (res.name in names(DE.res.adjusted)){
  de.adjusted <- DE.res.adjusted[[res.name]]
  DE.adjusted.limit[[res.name]] <-
    head(de.adjusted, min(topDElimit,sum(de.adjusted$cZ != 0)))
}
```

### 2h_D23580_exposed

```{r 2h_D23580_exposed}
DT::datatable(
  DE.adjusted.limit[["2h_D23580_exposed-2h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 2h_D23580_infected

```{r 2h_D23580_infected}
DT::datatable(
  DE.adjusted.limit[["2h_D23580_infected-2h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 2h_LT2_exposed

```{r 2h_LT2_exposed}
DT::datatable(
  DE.adjusted.limit[["2h_LT2_exposed-2h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 2h_LT2_infected

```{r 2h_LT2_infected}
DT::datatable(
  DE.adjusted.limit[["2h_LT2_infected-2h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 4h_D23580_exposed

```{r 4h_D23580_exposed}
DT::datatable(
  DE.adjusted.limit[["4h_D23580_exposed-4h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 4h_D23580_infected

```{r 4h_D23580_infected}
DT::datatable(
  DE.adjusted.limit[["4h_D23580_infected-4h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 4h_LT2_exposed

```{r 4h_LT2_exposed}
DT::datatable(
  DE.adjusted.limit[["4h_LT2_exposed-4h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 4h_LT2_infected

```{r 4h_LT2_infected}
DT::datatable(
  DE.adjusted.limit[["4h_LT2_infected-4h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 6h_D23580_exposed

```{r 6h_D23580_exposed}
DT::datatable(
  DE.adjusted.limit[["6h_D23580_exposed-6h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 6h_D23580_infected

```{r 6h_D23580_infected}
DT::datatable(
  DE.adjusted.limit[["6h_D23580_infected-6h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 6h_LT2_exposed

```{r 6h_LT2_exposed}
DT::datatable(
  DE.adjusted.limit[["6h_LT2_exposed-6h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 6h_LT2_infected

```{r 6h_LT2_infected}
DT::datatable(
  DE.adjusted.limit[["6h_LT2_infected-6h_Mock_uninfected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

## Expression profile of DE genes {.tabset}

### Top 1

Let us display the expression level of *ISG15*,
the most significantly DE gene at the `4h` time-point in
the `4h_LT2_infected-4h_Mock_uninfected` contrast (incidentally, also the
most significantly DE gene in all groups of stimulated cells relative to their
matched group of `uninfected` single cells at that time-point):

```{r topDE4hVSuninfected}
geneId <- rownames(DE.res.adjusted[["4h_LT2_infected-4h_Mock_uninfected"]])[1]
geneName <- mapIds(EnsDb.Hsapiens.v79, geneId, "GENENAME", "GENEID")
geneExpr <- norm_exprs(sce.norm)[geneId,]
sampleAes <- pData(sce.norm)[,c("Time","Infection","Status")]
ggData <- cbind(norm_exprs = geneExpr, sampleAes)
ggplot(ggData, aes(Time, norm_exprs)) +
  geom_violin(draw_quantiles = c(0.25,0.5,0.75)) + geom_jitter(width = 0.1) +
  facet_grid(Infection ~ Status) +
  ggtitle(sprintf("%s (symbol: %s)", geneId, geneName))
```

### Top 10

As an alternative, we may also visualise, for the same contrast,
the 10 most significant DE genes, placing all 15 groups of cells side by side,
ordering groups by `Status>Infection>Time`:

```{r top10_LT2vsUninfected, fig.height=10}
geneIds <-
  head(rownames(DE.res.adjusted[["4h_LT2_infected-4h_Mock_uninfected"]]), 10)
genesExpr <- norm_exprs(sce.norm)[geneIds,]
samplesAes <- pData(sce.norm)[,c("Time","Infection","Status","Group","Sample")]
ggData <- cbind(t(genesExpr), samplesAes)
ggLong <- reshape2::melt(
  ggData,
  measure.vars = grep("^ENSG", colnames(ggData)),
  variable.name = "Feature", value.name = "norm_exprs")
label_gene <- function(labels, multi_line = TRUE){
  labels <- lapply(labels, function(x){
      mapIds(EnsDb.Hsapiens.v79, as.character(x), "GENENAME", "GENEID")
  })
}
ggplot(ggLong, aes(Group, norm_exprs)) +
  geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
  geom_jitter(aes(colour = Time), size = 0.2, width = 0.1) +
  facet_grid(Feature ~ ., labeller = "label_gene") +
  theme(
    axis.text.x = element_text(angle=90),
    strip.text.y = element_text(size = 10)) + exprsViolinTheme
```

## Correlation of DE genes {.tabset}

In this section, let us compared the fold-change of expression between
groups of stimulated cells relative to time-matched uninfected cells
(*e.g.*, fold-change of expression in the contrast
`2h_D23580_infected-2h_Mock_uninfected`
compared against the fold-change of expression in the contrast
`2h_LT2_infected-2h_Mock_uninfected`).

### Preprocessing

The code below assembles differential expression results in a format
compatible with the grammar of graphics implemented in `ggplot2`.
A list of three tables is created; one for each time-point.
Each table is subsequently used to produce a figure with four panels;
one for each comparison between two groups of stimulated cells.
Each table is stores the following information:

* `Contrast`: the two groups of stimulated cells being compared
  (relative to `uninfected` cells; *e.g.*,
  `2h_D23580_infected-2h_LT2_infected`)
* `Gene`: the Ensembl gene identifier
* `mle.Y`, `mle.Y`: the maximum likelihood estimate (MLE) ofexpression-fold
  change for each group of stimulated cells relative to time-matched
  uninfected cells
* `DE`: whether the gene is significantly differentially expressed (DE) in
  `Group Y` (the group of stimulated cells on the left-hand side of the
  comparison, represented on the Y axis), `Group X` (the group of stimulated
  cells on the right-hand side of the comparison), `Both` groups, or `None`.
* `Alpha`, `Size`: highlights genes significantly DE in any of the two
  groups of stimulated cells being compared

```{r corTables}
Contrasts <- c(
  "D23580_infected-D23580_exposed", "LT2_infected-LT2_exposed",
  "D23580_infected-LT2_infected","D23580_exposed-LT2_exposed")
corTables <- list("2h"=data.frame(), "4h"=data.frame(), "6h"=data.frame())
for (time in levels(sce.norm$Time)){
  ctrlGroup <- sprintf("%s_Mock_uninfected", time)
  for (Contrast in Contrasts){
    groupNames <- strsplit(Contrast, "-")[[1]]
    GroupY <- sprintf("%s_%s", time, groupNames[1])
    GroupX <- sprintf("%s_%s", time, groupNames[2])
    DE.res.Y <- DE.res.adjusted[[sprintf("%s-%s", GroupY, ctrlGroup)]]
    DE.res.X <- DE.res.adjusted[[sprintf("%s-%s", GroupX, ctrlGroup)]]
    ensGenes <- rownames(DE.res.Y)
    plotData <- data.frame(
      Contrast = gsub("(.*)-(.*)", "\\1 (Y)\n\\2 (X)", Contrast),
      Gene = ensGenes,
      mle.Y = DE.res.Y[ensGenes,"mle"],
      mle.X = DE.res.X[ensGenes,"mle"],
      DE = c("None" ,"Group Y", "Group X", "Both")[
        1+(DE.res.Y[ensGenes,"cZ"]!=0) + 2*(DE.res.X[ensGenes,"cZ"]!=0)],
      Alpha = c(0.5, 0.75)[
        1+((DE.res.Y[ensGenes,"cZ"]!=0)|(DE.res.X[ensGenes,"cZ"]!=0))],
      Size = c(0.4, 0.6)[
        1+((DE.res.Y[ensGenes,"cZ"]!=0)|(DE.res.X[ensGenes,"cZ"]!=0))]
    )
    corTables[[time]] <- rbind(corTables[[time]], plotData)
  }
}
```

Let us manually pick colours to indicate genes differentially expressed
in either, both, or none of the two groups of stimulated cells being
compared:

```{r corDEcolors}
corDEcolors <- brewer.pal(9, "Pastel1")[c(1:3, 9)]
names(corDEcolors) <- c("Both", "Group Y", "Group X", "None")
```

Let us identify the extremum values of MLE, to set identical axis ranges
for all panels of all figures:

```{r mleRange}
mleRange <- range(c(
  corTables[["2h"]][,c("mle.Y","mle.X")],
  corTables[["4h"]][,c("mle.Y","mle.X")],
  corTables[["6h"]][,c("mle.Y","mle.X")]
))
```

Here, let us store shared elements of the figures used across the following
panels:

```{r corMleTheme}
Wrap <- facet_wrap(~Contrast, nrow = 2)
Labs <- labs(
    x="maximum likelihood estimate of expression-fold change",
    y="mle of fold-change")
Scale_color <- scale_color_manual(values = corDEcolors)
Scale_size <- scale_size(range = 0:1)
Scale_x <- scale_x_continuous(limits=mleRange)
Scale_y <- scale_y_continuous(limits=mleRange)
Guides <- guides(alpha = "none", size = "none")
```

### 2h

```{r ggplotCor_2h, fig.height=6}
ggplot(corTables[["2h"]]) +
  geom_point(aes(mle.X, mle.Y, colour = DE, alpha = Alpha, size = Size)) +
  theme_minimal()+Wrap+Labs+Scale_color+Scale_size+Scale_x+Scale_y+Guides
```

### 4h

```{r ggplotCor_4h, fig.height=6}
ggplot(corTables[["4h"]]) +
  geom_point(aes(mle.X, mle.Y, colour = DE, alpha = Alpha, size = Size)) +
  theme_minimal()+Wrap+Labs+Scale_color+Scale_size+Scale_x+Scale_y+Guides
```

### 6h

```{r ggplotCor_6h, fig.height=6}
ggplot(corTables[["6h"]]) +
  geom_point(aes(mle.X, mle.Y, colour = DE, alpha = Alpha, size = Size)) +
  theme_minimal()+Wrap+Labs+Scale_color+Scale_size+Scale_x+Scale_y+Guides
```

### Differences

In the `6h` panel, a number of genes appears to be differentially expressed
in opposite direction between `D23580` and `LT2` infections (relative
to `uninfected` time-mached controls).
In this section, let us identify those genes.

```{r corTables_Direction}
corTables[["6h"]][,"Direction"] <- NA
corTables[["6h"]][
  with(corTables[["6h"]], which(DE != "Both")),"Size"] <- 0.4
corTables[["6h"]][
  with(corTables[["6h"]], which(DE == "Both" & mle.X > 0 & mle.Y > 0)),
  "Direction"] <- "X: up - Y: up"
corTables[["6h"]][
  with(corTables[["6h"]], which(DE == "Both" & mle.X > 0 & mle.Y < 0)),
  "Direction"] <- "X: up - Y: down"
corTables[["6h"]][
  with(corTables[["6h"]], which(DE == "Both" & mle.X < 0 & mle.Y > 0)),
  "Direction"] <- "X: down - Y: up"
corTables[["6h"]][
  with(corTables[["6h"]], which(DE == "Both" & mle.X < 0 & mle.Y < 0)),
  "Direction"] <- "X: down - Y: down"
```

### 6h - Opposite

```{r ggplot_Direction, fig.height=6}
ggplot(corTables[["6h"]]) +
  geom_point(aes(mle.X, mle.Y, colour = Direction, alpha = Alpha, size = Size)) +
  theme_minimal()+Wrap+Labs+Scale_size+Scale_x+Scale_y+Guides
```

```{r opposite_6h_CSVs, include=FALSE}
for (tmpContrast in unique(corTables[["6h"]]$Contrast)){
  csvName <- gsub("^(.*) \\(Y\\)\n(.*) \\(X\\)$", "\\1-\\2", tmpContrast)
  tableOut <- subset(
      corTables[["6h"]],
      Contrast == tmpContrast & ((mle.Y * mle.X) < 0) & DE == "Both"
    )
  tableOut[,"Contrast"] <- gsub("\n", "-", tableOut[,"Contrast"])
  tableOut[,"GENENAME"] <- mapIds(
    EnsDb.Hsapiens.v79, as.character(tableOut[,"Gene"]), "GENENAME", "GENEID")
  write.csv(tableOut, sprintf("SCDE/6h_opposite_%s.csv", csvName)
  )
}
```


# DE between groups of stimulated cells

## Testing for DE genes {.tabset}

Similarly to the above code chunk to estimate differential expression
relative to [uninfected cells](#DEcontrol),
let us compare groups of stimulated cells between themselves:

**Note:** this step takes a considerable amount of time, even when multiple
cores are used (~ 5 min per comparison using a single core to estimate
batch effect and differential expression).
For this purpose, the chunk of code below is not executed during
automated compilation; instead it is manually recomputed as necessary.

<!--
During automated compilation, only load manually pre-computed DE statistics
saved to file.
-->

```{r scdeStimulated, eval=FALSE, warning=FALSE, message=FALSE}
contrasts <- list(
  c("%s_D23580_infected","%s_LT2_infected"),
  c("%s_D23580_infected","%s_D23580_exposed"),
  c("%s_LT2_infected","%s_LT2_exposed"),
  c("%s_D23580_exposed","%s_LT2_exposed")
)
for (time in levels(sce.norm$Time)){
  for (cindex in 1:4){
      sg.ref <- sprintf(contrasts[[cindex]][2], time)
      sg.tar <- sprintf(contrasts[[cindex]][1], time)
      contrast <- sprintf("%s-%s", sg.tar, sg.ref); #message(contrast)
      groups <- rep(NA, length(sg.all))
      names(groups) <- names(sg.all)
      groups[which(sg.all == sg.ref)] <- sg.ref
      groups[which(sg.all == sg.tar)] <- sg.tar
      groups <- factor(groups, c(sg.tar, sg.ref)) # target - reference
      scde.res <- scde.expression.difference(
        o.ifm, cd, o.prior,
        groups = groups,
        batch = batch,
        n.randomizations = 100,
        n.cores  =  1, verbose  =  1
      )
      
      de.raw <- scde.res$results
      de.adjusted <- scde.res$batch.adjusted
      de.batch <- scde.res$batch.effect
      
      de.raw <- de.raw[order(abs(de.raw$Z), decreasing = TRUE),]
      de.adjusted <- de.adjusted[order(abs(de.adjusted$Z), decreasing = TRUE),]
      de.batch <- de.batch[order(abs(de.batch$Z), decreasing = TRUE),]
      
      de.raw <- cbind(
        GENENAME=mapIds(EnsDb.Hsapiens.v79,rownames(de.raw),"GENENAME","GENEID"),
        de.raw
      )
      de.adjusted <- cbind(
        GENENAME=mapIds(EnsDb.Hsapiens.v79,rownames(de.adjusted),"GENENAME","GENEID"),
        de.adjusted
      )
      de.batch <- cbind(
        GENENAME=mapIds(EnsDb.Hsapiens.v79,rownames(de.batch),"GENENAME","GENEID"),
        de.batch
      )
      
      DE.res.raw[[contrast]] <- de.raw
      DE.res.adjusted[[contrast]] <- de.adjusted
      DE.res.batch[[contrast]] <- de.batch
  }
}
```

```{r saveRdsDEbtwStimulated, eval=FALSE, include=FALSE}
saveRDS(DE.res.raw, "rds/DE.res.raw.rds")
saveRDS(DE.res.adjusted, "rds/DE.res.adjusted.rds")
saveRDS(DE.res.batch, "rds/DE.res.batch.rds")
for (groupName in names(DE.res.raw)){
  write.csv(
    DE.res.raw[[groupName]], sprintf("SCDE/raw_%s.csv",groupName))
  write.csv(
    DE.res.adjusted[[groupName]], sprintf("SCDE/adjusted_%s.csv",groupName))
  write.csv(DE.res.batch[[groupName]], sprintf("SCDE/batch_%s.csv",groupName))
}
```

```{r readRdsDEbtwStimulated, include=FALSE}
DE.res.raw <- readRDS("rds/DE.res.raw.rds")
DE.res.adjusted <- readRDS("rds/DE.res.adjusted.rds")
DE.res.batch <- readRDS("rds/DE.res.batch.rds")
```

## Count of DE genes

Let us display the count of differentially expressed genes:

* Without batch correction (`Raw`)
* With batch correction applied by plate processed (`Adjusted`). *Note that
  all samples of any of the `4` plates were processed on the same of the `2`
  sequencing lanes; as a result, accounting for a plate batch implicitely
  accounts for a lane batch.*

```{r countDEbtwStimulated}
contrastsStimulated <- grep("Mock", names(DE.res.raw), invert = TRUE, value = TRUE)
DT::datatable(data.frame(
  "Raw" = sapply(contrastsStimulated, function(n){sum(DE.res.raw[[n]]$cZ != 0)}),
  "Adjusted" =
    sapply(contrastsStimulated, function(n){sum(DE.res.adjusted[[n]]$cZ != 0)}),
  "Batch" =
    sapply(contrastsStimulated, function(n){sum(DE.res.batch[[n]]$cZ != 0)})
  ),
  options = list(pageLength = 12)
)
```

## Tables of DE genes {.tabset}

Let us display the tables of DE genes for the various contrasts calculated
above.

Again, for convience, let us display only the `1000` most significant DE genes
(*i.e.*, largest Z-score) in cases where larger counts of DE genes were
detected (*i.e.*, corrected Z-score ≠ 0).

```{r topDElimitStimulated}
for (res.name in names(DE.res.batch)){
  de.adjusted <- DE.res.adjusted[[res.name]]
  DE.adjusted.limit[[res.name]] <-
    head(de.adjusted, min(topDElimit,sum(de.adjusted$cZ != 0)))
}
```

### 2h_D23_inf-2h_LT2_inf

```{r 2h_D23580_inf-2h_LT2_inf}
DT::datatable(
  DE.adjusted.limit[["2h_D23580_infected-2h_LT2_infected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 2h_D23_inf-2h_D23_exp

```{r 2h_D23580_inf-2h_D23580_exp}
DT::datatable(
  DE.adjusted.limit[["2h_D23580_infected-2h_D23580_exposed"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 2h_LT2_inf-2h_LT2_exp

```{r 2h_LT2_inf-2h_LT2_exp}
DT::datatable(
  DE.adjusted.limit[["2h_LT2_infected-2h_LT2_exposed"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 2h_D23_exp-2h_LT2_exp

```{r 2h_D23580_exp-2h_LT2_exp}
DT::datatable(
  DE.adjusted.limit[["2h_D23580_exposed-2h_LT2_exposed"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 4h_D23_inf-4h_LT2_inf

```{r 4h_D23580_inf-4h_LT2_inf}
DT::datatable(
  DE.adjusted.limit[["4h_D23580_infected-4h_LT2_infected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 4h_D23_inf-4h_D23_exp

```{r 4h_D23580_inf-4h_D23580_exp}
DT::datatable(
  DE.adjusted.limit[["4h_D23580_infected-4h_D23580_exposed"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 4h_LT2_inf-4h_LT2_exp

```{r 4h_LT2_inf-4h_LT2_exp}
DT::datatable(
  DE.adjusted.limit[["4h_LT2_infected-4h_LT2_exposed"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 4h_D23_exp-4h_LT2_exp

```{r 4h_D23580_exp-4h_LT2_exp}
DT::datatable(
  DE.adjusted.limit[["4h_D23580_exposed-4h_LT2_exposed"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 6h_D23_inf-6h_LT2_inf

```{r 6h_D23580_inf-6h_LT2_inf}
DT::datatable(
  DE.adjusted.limit[["6h_D23580_infected-6h_LT2_infected"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 6h_D23_inf-6h_D23_exp

```{r 6h_D23580_inf}
DT::datatable(
  DE.adjusted.limit[["6h_D23580_infected-6h_D23580_exposed"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 6h_LT2_inf-6h_LT2_exp

```{r 6h_LT2_exp}
DT::datatable(
  DE.adjusted.limit[["6h_LT2_infected-6h_LT2_exposed"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

### 6h_D23_exp-6h_LT2_exp

```{r 6h_D23580_exp-6h_LT2_exp}
DT::datatable(
  DE.adjusted.limit[["6h_D23580_exposed-6h_LT2_exposed"]],
  options = list(pageLength = 10, searching = TRUE), filter = "top"
)
```

## Expression profile of DE genes {.tabset}

Let us display the expression level of *CCL7*,
a DE gene significantly more expressed in
`4h_D23580_infected` than `4h_D23580_exposed`:

```{r CCL7violin}
geneName <- "CCL7"
geneId <- mapIds(EnsDb.Hsapiens.v79, geneName, "GENEID", "GENENAME")
geneExpr <- norm_exprs(sce.norm)[geneId,]
sampleAes <- pData(sce.norm)[,c("Time","Infection","Status","Plate")]
ggData <- cbind(norm_exprs = geneExpr, sampleAes)
ggplot(ggData, aes(Status, norm_exprs, colour = Plate)) +
  geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
  geom_jitter(position = position_dodge(width = 0.9)) +
  facet_grid(Infection ~ Time) + exprsViolinTheme +
  ggtitle(sprintf("%s (symbol: %s)", geneId, geneName))
```

Let us compare the posterior of *CCL7* for the four comparisons of interest:

### 6h_D23_inf - 6h_D23_exp

```{r 6h_D23_inf - 6h_D23_exp}
groups <- rep(NA, length(sg.all)); names(groups) <- names(sg.all);
sg.ref <- "6h_D23580_exposed"; sg.tar <- "6h_D23580_infected";
groups[which(sg.all==sg.ref)]<-sg.ref; groups[which(sg.all==sg.tar)]<-sg.tar;
groups <- factor(groups, c(sg.tar, sg.ref)) # target - reference
scde.test.gene.expression.difference(
  geneId, models = o.ifm, counts = cd, prior = o.prior,
  groups = groups, batch = batch)
```

### 6h_LT2_inf - 6h_LT2_exp

```{r 6h_LT2_inf - 6h_LT2_exp}
groups <- rep(NA, length(sg.all)); names(groups) <- names(sg.all)
sg.ref <- "6h_LT2_exposed"; sg.tar <- "6h_LT2_infected"
groups[which(sg.all == sg.ref)] <- sg.ref; groups[which(sg.all == sg.tar)] <- sg.tar
groups <- factor(groups, c(sg.tar, sg.ref)) # target - reference
scde.test.gene.expression.difference(
  geneId, models = o.ifm, counts = cd, prior = o.prior,
  groups = groups, batch = batch)
```

### 6h_D23_inf - 6h_LT2_inf

```{r 6h_D23_inf - 6h_LT2_inf}
groups <- rep(NA, length(sg.all)); names(groups) <- names(sg.all)
sg.ref <- "6h_LT2_infected"; sg.tar <- "6h_D23580_infected"
groups[which(sg.all == sg.ref)] <- sg.ref; groups[which(sg.all == sg.tar)] <- sg.tar
groups <- factor(groups, c(sg.tar, sg.ref)) # target - reference
scde.test.gene.expression.difference(
  geneId, models = o.ifm, counts = cd, prior = o.prior,
  groups = groups, batch = batch)
```

### 6h_D23_exp - 6h_LT2_exp

```{r 6h_D23_exp - 6h_LT2_exp}
groups <- rep(NA, length(sg.all)); names(groups) <- names(sg.all)
sg.ref <- "6h_LT2_exposed"; sg.tar <- "6h_D23580_exposed"
groups[which(sg.all == sg.ref)] <- sg.ref; groups[which(sg.all == sg.tar)] <- sg.tar
groups <- factor(groups, c(sg.tar, sg.ref)) # target - reference
scde.test.gene.expression.difference(
  geneId, models = o.ifm, counts = cd, prior = o.prior,
  groups = groups, batch = batch)
```

<!-- As HTML pages are built independently, objects need to be saved to disk
and imported as needed by the other R mardown files -->
