---
title: "Differential expression using BASiCS"
bibliography:
  bibtex.bib
---

**IMPORTANT**:
This analysis has not been completed successfully yet.
Please ignore what follows.

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(scran)
library(scater)
library(BASiCS)
library(ggplot2)
sce.norm <- readRDS("rds/sce.norm.rds")
mixC <- readRDS("rds/mixC.rds")
EnsDb.Hsapiens.v79 <- EnsDb.Hsapiens.v79::EnsDb.Hsapiens.v79
```

# Data pre-processing

## Input data

For this analysis we use:

* the matrix of expression *counts*
* single cells that passed all quality control filters (see this earlier
  [section](03_sampleQC.html#cellSelection))
* features with an average count equal or higher than $1$ (see this earlier
  [section](04_featureSelection#selectFeatures))

**Note:** this contrasts with the stringent condition of ≥20 RPM on average
across all cells, demonstrated in [@BASiCS_2016_DE]. However, feature
selection is highly dependent on the nature of the experimental design.

As a result, the input data contains `r nrow(sce.norm)` features and
`r ncol(sce.norm)` cells.

# BASiCS analysis

## The input dataset

### Spike-in genes information

In subsequent steps of the analysis, technical spike-in features are identified using the `scater::isSpike` method:

```{r isSpike, include=FALSE}
table(isSpike(sce.norm))
```

This confirms that `r sum(isSpike(sce.norm))` ERCC spike-in features are left in
the filtered dataset.

### Creating the input objects

To use BASiCS, we need to create a `BASiCS_Data` object containing the
expression counts, a vector of logical spike-in gene indicators
and the input number of mRNA molecules for each spike-in gene.

```{r Data.input, results='hide', message=FALSE, warning=FALSE}
Data.input <- list()
ERCCnames <- rownames(sce.norm)[isSpike(sce.norm)]
for (groupName in unique(sce.norm$Group)){
  cellIndex <- which(sce.norm$Group == groupName)
  Data.input[[groupName]] <- newBASiCS_Data(
    Counts = counts(sce.norm)[,cellIndex],
    Tech = isSpike(sce.norm),
    SpikeInfo = mixC[ERCCnames, c("ID","ERCCmolecules")],
    BatchInfo = sce.norm$Plate[cellIndex])
}
```

**Note:**
Due to the nature of the feature selection filters, in the above code chunk
the `newBASiCS_Data` method prints the following message for all 15 groups
of cells:

```{r warning_newData, echo=FALSE}
warning("Some genes have zero counts across all cells. Unless running a differential expression analysis, please remove those genes. Otherwise, the BASiCS_Data object is still a valid object. However, due to the lack of counts, posterior estimates for mu[i] and delta[i] associated to those genes will be driven by the prior. In such case, you must specify `PriorDelta = 'log-normal' in BASiCS_MCMC function.")
```

## Fitting the BASiCS model (for differential expression)

To run the MCMC algorithm for differential expression,
we use the function `BASiCS_MCMC`, following the recommendation prompted
by the `newBASiCS_Data` function:

* `PriorDelta` argument set to `"log-normal"`
* due to the presence, within each experimental group of cells, of genes with
  zero counts across all cells within that group, only use those `BASiCS_Data`
  objects for differential expression analysis, *not*
  differential over-dispersion

```{r BASiCS_MCMC, results='hide', message=FALSE, eval=FALSE}
MCMC_Output_DE <- list()
N <- 20E3; Thin <- 10; Burn <- 10E3; chains.path <- "BASiCS"
for (groupName in names(Data.input)){
  MCMC_Output_DE[[groupName]] <- BASiCS_MCMC(
    Data.input[[groupName]], N = N, Thin = Thin, Burn = Burn,
    PrintProgress = TRUE, StoreChains = TRUE, PriorDelta = 'log-normal',
    StoreDir = chains.path, RunName = paste(groupName, N, sep= "_"))
}
```

```{r saveRDS, eval=FALSE}
saveRDS(MCMC_Output_DE, "rds/MCMC_Output.rds")
```

```{r readRDS_MCMC, include=FALSE}
MCMC_Output_DE <- readRDS("rds/MCMC_Output.rds")
```

Confirm that all groups of cells were run:

```{r deleteMe}
names(MCMC_Output_DE)
```

## Convergence diagnostics

To assess convergence of each chain, some selected parameters may be
viualised in traceplots. For a randomly selected experimental group:

```{r, fig.height=6}
rGroup <- sample(names(MCMC_Output_DE), 1)
par(mar = c(4,6,4,2)); par(mfrow = c(3,2))
plot(
  apply(MCMC_Output_DE[[rGroup]]@mu,1,median), type = "l", cex.lab = 2,
  ylab = expression(paste("Median of ",mu[iJ])), xlab = "Iteration")
plot(
  apply(MCMC_Output_DE[[rGroup]]@delta,1,median), type = "l", cex.lab = 2,
  ylab = expression(paste("Median of ",delta[iJ])), xlab = "Iteration")
plot(
  apply(MCMC_Output_DE[[rGroup]]@phi,1,median), type = "l", cex.lab = 2,
  ylab = expression(paste("Median of ",phi[iJ])), xlab = "Iteration")
plot(
  apply(MCMC_Output_DE[[rGroup]]@s,1,median), type = "l", cex.lab = 2,
  ylab = expression(paste("Median of ",s[j[I]])), xlab = "Iteration")
plot(
  apply(MCMC_Output_DE[[rGroup]]@nu,1,median), type = "l", cex.lab = 2,
  ylab = expression(paste("Median of ",nu[j[I]])), xlab = "Iteration")
plot(
  apply(MCMC_Output_DE[[rGroup]]@theta,1,median), type = "l", cex.lab = 2,
  ylab = expression(paste("Median of ",theta[j[I]])), xlab = "Iteration")

par(mar = c(5, 4, 4, 2) + 0.1); par(mfrow = c(1,1))
```

## Differential expression (mean and over-dispersion)

Combine `BASiCS` data and chains for two groups to compare:

```{r CombineBASiCS_Data}
Data_2h_D23exp_mock <- CombineBASiCS_Data(
  Data.input[["2h_D23580_exposed"]], Data.input[["2h_Mock_uninfected"]])
MCMC_2h_D23exp_mock <- CombineBASiCS_Chain(
  MCMC_Output_DE[["2h_D23580_exposed"]], MCMC_Output_DE[["2h_Mock_uninfected"]])
```

Test with **0%** fold-change tolerance:

```{r Test_D23exp_mock_0, eval=FALSE}
Test_D23exp_mock_0 <- BASiCS_D_TestDE(
  Data_2h_D23exp_mock, MCMC_2h_D23exp_mock,
  GeneNames = Data_2h_D23exp_mock@GeneNames[!Data_2h_D23exp_mock@Tech],
  EpsilonM = 0, EpsilonD = 0,
  EFDR_M = 0.05, EFDR_D = 0.05,
  OrderVariable = "GeneIndex",
  GroupLabelRef = "2h_uninfected", GroupLabelTest = "2h_D23580_exposed",
  OffSet = FALSE)
```

Test with **40%** fold-change tolerance:

```{r Test_D23exp_mock_40, eval=FALSE}
Test_D23exp_mock_40 <- BASiCS_D_TestDE(
  Data_2h_D23exp_mock, MCMC_2h_D23exp_mock,
  GeneNames = Data_2h_D23exp_mock@GeneNames[!Data_2h_D23exp_mock@Tech],
  EpsilonM = 0.4, EpsilonD = 0.4,
  EFDR_M = 0.05, EFDR_D = 0.05,
  OrderVariable = "GeneIndex",
  GroupLabelRef = "2h_uninfected", GroupLabelTest = "2h_D23580_exposed",
  OffSet = FALSE)
```

```{r, eval=FALSE}
tableD23exp_2h <- cbind(
  Symbol = fData(sce.norm)[Test_D23exp_mock_40$Table$GeneNames, "gene_name"],
  Test_D23exp_mock_40$Table)
```

Try again with a more interesting group:

```{r CombineBASiCS_Data_4h, eval=FALSE}
Data_4h_D23exp_mock <- CombineBASiCS_Data(
  Data.input[["4h_D23580_exposed"]], Data.input[["4h_Mock_uninfected"]])
MCMC_4h_D23exp_mock <- CombineBASiCS_Chain(
  MCMC_Output_DE[["4h_D23580_exposed"]], MCMC_Output_DE[["4h_Mock_uninfected"]])
Test_4h_D23exp_mock_40 <- BASiCS_D_TestDE(
  Data_4h_D23exp_mock, MCMC_4h_D23exp_mock,
  GeneNames = Data_4h_D23exp_mock@GeneNames[!Data_4h_D23exp_mock@Tech],
  EpsilonM = 0.4, EpsilonD = 0.4,
  EFDR_M = 0.05, EFDR_D = 0.05,
  OrderVariable = "GeneIndex",
  GroupLabelRef = "4h_uninfected", GroupLabelTest = "4h_D23580_exposed",
  OffSet = FALSE)
tableD23exp_4h <- cbind(
  Symbol = fData(sce.norm)[Test_4h_D23exp_mock_40$Table$GeneNames, "gene_name"],
  Test_4h_D23exp_mock_40$Table)
```


<!-- As HTML pages are built independently, objects need to be saved to disk
and imported as needed by the other R mardown files -->
