---
title: "Differential expression"
bibliography:
  bibtex.bib
---

```{r checkPkgs, child="_checkLibraries.Rmd", include=FALSE}
```

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(scran)
library(scater)
library(BASiCS)
library(data.table) 
library(EnsDb.Hsapiens.v79)
library(ggplot2)
sce.norm <- readRDS("rds/sce.norm.rds")
mixC <- readRDS("rds/mixC.rds")
```

# Data pre-processing

## Input data

For this analysis we use:
* the matrix of expression *counts*,
* single cells that passed all quality control filters (see this earlier
  [section](03_sampleQC.html#cellSelection))
* features with an average count equal or higher than $1$ (see this earlier
  [section](04_featureSelection#selectFeatures))

```{r scePassCounts, results='hide'}
counts(sce.norm)
```

**Note:** this contrasts with the stringent condition of ≥20 RPM on average
across all cells, demonstrated in [@BASiCS2016DE].

As a result, the input data contains `r nrow(sce.norm)` features and
`r ncol(sce.norm)` cells.

# BASiCS analysis

## The input dataset

### Spike-in genes information

In subsequent steps of the analysis, technical spike-in features are identified
as follows:

```{r, results='hide'}
isSpike(sce.norm)
```

```{r}
table(isSpike(sce.norm))
```





<!-- As HTML pages are built independently, objects need to be saved to disk
and imported as needed by the other R mardown files -->
