---
title: "Differential expression"
bibliography:
  bibtex.bib
---

```{r checkPkgs, child="_checkLibraries.Rmd", include=FALSE}
```

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(broom)
library(EnsDb.Hsapiens.v79)
library(ggplot2)
sce.norm <- readRDS("rds/sce.norm.rds")
```

# Prepare data

This analysis uses the expression data set normalised in a separate
[section](05_normalisation.html).

```{r}
dim(sce.norm)
names(assayData(sce.norm))
dim(norm_exprs(sce.norm))
levels(sce.norm$Time)
levels(sce.norm$Infection)
levels(sce.norm$Status)
```

# *t*-test

Let us start with a simple *t*-test to evaluate differential expression
between groups of single cells.

First, let us initialise the list of DE results and the function:

```{r}
tTest.single <- function(x, idx.r, idx.t){
  tidy(t.test(x[idx.target], x[idx.ref]))
}
DE.list <- list()
```

We may then compare each group of stimulated single cells to their matching
control uninfected group of single cells, and store the resulting table of
results as an accoridingly named element of the list initialised above:

```{r}
for (time in levels(sce.norm$Time)){
  for (inf.target in levels(sce.norm$Infection)[-1]){
    for (sta.target in levels(sce.norm$Status)[-1]){
      inf.ref <- levels(sce.norm$Infection)[1]
      sta.ref <- levels(sce.norm$Status)[1]
      contrast <- sprintf(
        "%s_%s_%s-%s_%s_%s",
        time, inf.target, sta.target, time, inf.ref, sta.ref)
      message(contrast)
      idx.ref <- with(
        pData(sce.norm),
        which(Time == time & Infection == inf.ref & Status == sta.ref))
      idx.target <- with(
        pData(sce.norm),
        which(Time == time & Infection == inf.target & Status == sta.target))
      de.res <- do.call(
        "rbind",
        apply(norm_exprs(sce.norm), 1, tTest.single, idx.ref, idx.target))
      de.res <- de.res[order(de.res$p.value),]
      de.res$adj.p.value <- p.adjust(de.res$p.value, "bonferroni")
      de.res$symbol <-
        mapIds(EnsDb.Hsapiens.v79, rownames(de.res), "GENENAME", "GENEID")
      message(sprintf("Bonf < 0.05: %i", sum(de.res$adj.p.value < 0.05)))
      DE.list[[contrast]] <- de.res
    }
  }
}
```

# Display tables of results

TODO

```{r}
DE.list[["6h_LT2_infected-6h_Mock_uninfected"]]
```


# Display expression profile for DE genes

Let us 

```{r}
geneId <- rownames(DE.list[["6h_LT2_infected-6h_Mock_uninfected"]])[5]
geneName <- mapIds(EnsDb.Hsapiens.v79, geneId, "GENENAME", "GENEID")
geneExpr <- norm_exprs(sce.norm)[geneId,]
sampleAes <- pData(sce.norm)[,c("Time","Infection","Status")]
ggData <- cbind(norm_exprs = geneExpr, sampleAes)
ggplot(ggData, aes(Time, norm_exprs)) +
  geom_violin(draw_quantiles = c(0.25,0.5,0.75)) + geom_jitter(width = 0.1) +
  facet_grid(Infection ~ Status) +
  ggtitle(sprintf("%s (symbol: %s)", geneId, geneName))
```


<!-- Let us export the DE tables in CSV files. -->

```{r, include=FALSE}
for (contrast in names(DE.list)){
  message(contrast)
  print(dim(DE.list[[contrast]]))
  write.csv(DE.list[[contrast]], sprintf("t-test/%s.csv", contrast))
}
```

<!-- As HTML pages are built independently, objects need to be saved to disk
and imported as needed by the other R mardown files -->
