---
title: "Sub-group analysis"
---

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(scater)
library(scran)
library(ggplot2)
library(SummarizedExperiment)
library(ComplexHeatmap)
sce.norm <- readRDS("rds/sce.norm.rds")
HVGs.group <- readRDS("rds/HVGsByGroups.rds")
```

# Objective

In previous sections, we have observed that the detection of
[significant differences in gene expression](08_SCDE_v8.html)
between groups of stimulated cells is limited by variability and bi-/multi-modal
distribution of gene expression within each group.

As a result, let us explore cell-level similarities within each experimental
group, using the group-level highly-variables genes (HVGs) identified
[earlier](07_highlyVariableGenes.html).

For each group, let us:

* apply the [scran](http://bioconductor.org/packages/scran) `quickCluster`
  method---using Spearman correlation-based distance---to
  identify potentially distinct behaviours within each experimental condition
* overlay the above cluster membership onto a principal components analysis
  (PCA) based on the HVGs identified for each group
* Overlay the above cluster membership onto a heat map representation of
  normalised gene expression following hierarchical clustering of cells
  within each experimental group

# Preprocessing

In this section, let us search for potential sub-structure within each
experimental group of cells, as follows:

* fetch the group-level HVGs identified in an earlier
  [section](07_highlyVariableGenes.html)
* fetch normalised expression data for the corresponding genes and cells
* apply the [scran](http://bioconductor.org/packages/scran) `quickCluster`
  method
* calculate pairwise distance between cells using Spearman correlation
* perform hierarchical clustering on the above pairwise distances
* perform a PCA
* present in a heat map the normalised gene expression data, with cells
  organised according to the above hierarchical clustering, while leaving
  genes to cluster according to the default Euclidian distance between their
  expression profile across cells

First, let us pick a set of colours, to set a manual colour scale for the
PCA figures, and also later used in the heat map figures:

```{r}
col8 <- RColorBrewer::brewer.pal(8, "Set1")
colGrey <- RColorBrewer::brewer.pal(9, "Set1")[9]
pcaCol <- c(colGrey, col8)
names(pcaCol) <- seq_along(pcaCol) - 1
pcaScaleManual <- scale_color_manual(values = pcaCol)
```

We may then performs the analytical steps listed above on each experimental
group of cells, using the selected colours to indicate cluster membership
in the heat map:

**Note:**
For the `quickCluster` method, `min.size` is here set to **5**, suitable to
detect up to `r trunc(min(table(sce.norm$Group)) / 5)` sub-groups in the
smallest experimental group of `r min(table(sce.norm$Group))` cells.

```{r}
subgroupPlots <- list()
for (groupName in levels(sce.norm$Group)){
  HVGs.sub <- HVGs.group[[groupName]]
  sce.sub <- sce.norm[rownames(HVGs.sub), sce.norm$Group == groupName]
  sce.sub$quickCluster <- quickCluster(sce.sub, min.size = 5)
  dist.sub <- MKmisc::corDist(t(norm_exprs(sce.sub)), method="spearman")
  clust.sub <- hclust(dist.sub)
  pca.sub <- as.data.frame(prcomp(t(norm_exprs(sce.sub)))$x[,1:10]) # top 10 PCs
  pca.sub[,"quickCluster"] <- pData(sce.sub)[rownames(pca.sub),"quickCluster"]
  ht_col <- col9[1:sum(levels(sce.sub$quickCluster) != 0)]
  if (0 %in% levels(sce.sub$quickCluster)) {ht_col <- c(colGrey, ht_col)}
  names(ht_col) <- levels(sce.sub$quickCluster)
  h_column <- HeatmapAnnotation(
    df = pData(sce.sub)[,c("quickCluster"), drop = FALSE],
    col = list(quickCluster = ht_col)
  )
  ht_subset <- Heatmap(
    norm_exprs(sce.sub),
    name = "norm_exprs", column_title = "Normalised expression",
    top_annotation = h_column, cluster_columns = clust.sub,
    show_row_names = FALSE, show_column_names = TRUE
  )
  subgroupPlots[[groupName]] <- list(PCA = pca.sub, Heatmap = ht_subset)
}
```

# PCAs {.tabset}

## 2h_Mock_uninfected

```{r}
ggplot(subgroupPlots[["2h_Mock_uninfected"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster)) + pcaScaleManual
```

## 4h_Mock_uninfected

```{r}
ggplot(subgroupPlots[["4h_Mock_uninfected"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 6h_Mock_uninfected

```{r}
ggplot(subgroupPlots[["6h_Mock_uninfected"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 2h_D23580_exposed

```{r}
ggplot(subgroupPlots[["2h_D23580_exposed"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 4h_D23580_exposed

```{r}
ggplot(subgroupPlots[["4h_D23580_exposed"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 6h_D23580_exposed

```{r}
ggplot(subgroupPlots[["6h_D23580_exposed"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 2h_D23580_infected

```{r}
ggplot(subgroupPlots[["2h_D23580_infected"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 4h_D23580_infected

```{r}
ggplot(subgroupPlots[["4h_D23580_infected"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 6h_D23580_infected

```{r}
ggplot(subgroupPlots[["6h_D23580_infected"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 2h_LT2_exposed

```{r}
ggplot(subgroupPlots[["2h_LT2_exposed"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 4h_LT2_exposed

```{r}
ggplot(subgroupPlots[["4h_LT2_exposed"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 6h_LT2_exposed

```{r}
ggplot(subgroupPlots[["6h_LT2_exposed"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 2h_LT2_infected

```{r}
ggplot(subgroupPlots[["2h_LT2_infected"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 4h_LT2_infected

```{r}
ggplot(subgroupPlots[["4h_LT2_infected"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

## 6h_LT2_infected

```{r}
ggplot(subgroupPlots[["6h_LT2_infected"]][["PCA"]]) +
  geom_point(aes(PC1, PC2, colour = quickCluster))
```

# tSNE

```{r}
# sce.subset <- plotTSNE(
#   sce.subset, ncomponents = 2, exprs_values = "norm_exprs",
#   return_SCESet = TRUE, colour_by = "quickCluster",
#   perplexity = 5, rand_seed = 1794, draw_plot = TRUE
# )
```
