---
  title: "Sub-group analysis"
---

  <!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(scater)
library(scran)
library(ggplot2)
library(scde)
sce.norm <- readRDS("rds/sce.norm.tSNE.rds")
HVGs.group <- readRDS("rds/HVGsByGroups.rds")
```

Subset to only cells at **6h** and challenged with bacteria
(*i.e.*, exclude `Mock`):

```{r sce.6h}
sce.6hStim <- sce.norm[,sce.norm$Time == '6h' & sce.norm$Infection != 'Mock']
dim(sce.6hStim)
```

Identify HVGs in those cells:

```{r HVGs.6hStim}
dm <- model.matrix(~Plate+Lane, data = pData(sce.6hStim))
var.fit <- trendVar(
  sce.6hStim, assay = "norm_exprs", trend = "loess", design = dm,
  use.spikes = FALSE, span = 0.1
)
# sce.6hStim <- normalise(sce.6hStim)
var.out <- decomposeVar(sce.6hStim, var.fit)
var.out <- var.out[with(var.out, order(p.value)),]
var.out <- cbind(
  GENENAME = mapIds(EnsDb.Hsapiens.v79,rownames(var.out),'GENENAME','GENEID'),
  var.out
)
HVGs.6hStim <- (var.out$FDR < 0.05)
table(HVGs.6hStim)
```

Display the `r sum(HVGs.6hStim)` significant endogenous features
(*i.e.*, **FDR < 0.05**):

```{r datatable_HVGs.6hStim}
DT::datatable(var.out[which(HVGs.6hStim),])
```

Identify clusters, given a minimal size of `r min(table(sce.6hStim$Group))`:

```{r quickCluster.6hStim}
sce.6hStim <- sce.6hStim[which(HVGs.6hStim),]
sce.6hStim$Group <- droplevels(sce.6hStim$Group)
clusters <- quickCluster(sce.6hStim, min.size = min(table(sce.6hStim$Group)))
names(clusters) <- sampleNames(sce.6hStim)
sce.6hStim$quickCluster <- clusters
sce.norm$quickCluster.6hStim <- 0
pData(sce.norm)[names(clusters),'quickCluster.6hStim'] <- clusters
table(clusters)
```

Overlay cluster membership and `Treatment` phenotype onto the tSNE coordinate
originally calculated from the complete data set:

```{r plotReducedDim_quickCluster.6hStim}
plotReducedDim(
  sce.norm, colour_by = 'quickCluster.6hStim', shape_by = 'Treatment'
)
```

```{r}
cd.6hStim <- clean.counts(
  counts(sce.6hStim)
)
cd <- apply(cd, 2, function(x){storage.mode(x) <- 'integer'; x})
dim(cd)
```


