---
title: "Mann Whitney differential expression"
---

<!-- As HTML pages are built independently, large objects need to be
reimported from disk, and smaller ones quickly recomputed. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(scater)
library(broom)
library(reshape2)
library(ensembldb)
EnsDb.Hsapiens.v79 <- EnsDb.Hsapiens.v79::EnsDb.Hsapiens.v79
sce.norm <- readRDS("rds/sce.norm.rds")
```

# Plotting themes

First of all, let us identify the minimum and maximum expression values in the
normalised data set:

```{r mleRange}
exprRange <- range(norm_exprs(sce.norm))
```

Let us also define theme elements used throughout various figures
in the following sections:

```{r exprsViolinTheme}
exprsTheme <- theme(
  panel.grid.major.x = element_blank(),
  legend.position = "bottom", legend.box = "vertical",
  axis.text.x = element_text(size = rel(0.75))
)
```

# Pairwise comparison of all stimulated groups {.tabset}

For simplicity, let us remove spike-in features:

```{r}
sce.endo <- sce.norm[!isSpike(sce.norm),]
```

## Functions

To estimate differential expression:

* For each pairwise comparison, let us test only features that are detected
  above **10** CPM in at least **25%** of cells in at least one of the
  two groups

```{r}
genewise.wilcox.test <- function(groupRef, groupTarget){
  stopifnot(all(c(groupRef, groupTarget) %in% sce.norm$Group))
  refIdx<-sce.norm$Group == groupRef;targetIdx<-sce.norm$Group==groupTarget
  cat(sprintf(
    "%s (%i) - %s (%i)", groupTarget, sum(targetIdx), groupRef, sum(refIdx)
  ))
  detectedRef <-
    rowSums(norm_exprs(sce.endo)[,refIdx] > 10) > (0.25 * sum(refIdx))
  detectedTarget <-
    rowSums(norm_exprs(sce.endo)[,targetIdx] > 10) > (0.25 * sum(targetIdx))
  featureIdx <- (detectedRef | detectedTarget)
  cat(sprintf("\n%i endogenous features tested", sum(featureIdx)))
  wt <- do.call("rbind", apply(
    norm_exprs(sce.norm)[featureIdx,], 1, function(x){
      tidy(wilcox.test(x[refIdx], x[targetIdx]))[,c("statistic","p.value")]
      }
  ))
  wt[,"BH"] <- p.adjust(wt[,"p.value"], "BH")
  wt[,"medianDiff"] <- apply(
    norm_exprs(sce.norm)[featureIdx,], 1, function(x){
      median(x[targetIdx]) - median(x[refIdx])
      }
  )
  cat(sprintf(
    " => FDR < 0.05: %i\n", sum(wt[,"BH"] < 0.05, na.rm = TRUE)
  ))
  return(wt)
}
```

To visualise the top **9** genes (by *P*-value / statistic):

```{r}
ggTop9 <- function(cName){
  cNames <- strsplit(cName, "-")[[1]]
  cIndex <- sce.norm$Group %in% cNames
  cGroup <- sce.norm$Group[cIndex]
  names(cGroup) <- sampleNames(sce.norm)[cIndex]
  cTable <- wt.list[[cName]]
  top9geneIds <- head(rownames(cTable[order(cTable[,"p.value"]),]), 9)
  top9geneNames <- mapIds(EnsDb.Hsapiens.v79, top9geneIds, "GENENAME", "GENEID")
  exprWide <- norm_exprs(sce.norm)[top9geneIds, cIndex]
  exprLong <- melt(
    exprWide, value.name = "norm_exprs", varnames = c('GENEID','Sample'))
  exprLong[,"Group"] <- cGroup[as.character(exprLong[,"Sample"])]
  exprLong[,"Feature"] <- factor(
    paste(exprLong[,"GENEID"],top9geneNames[exprLong[,"GENEID"]],sep=" - "),
    paste(top9geneIds, top9geneNames, sep=" - ")
  )
  ggplot(exprLong, aes(Group, norm_exprs)) +
    geom_violin(draw_quantiles = 0.5) + geom_jitter(width = 0.1) +
    facet_wrap(~Feature) + scale_y_continuous(limits = exprRange) + exprsTheme +
    ggtitle(cName)
}
```

## Preprocessing

```{r, warning=FALSE}
wt.list <- list()
contrasts <- list(
  c("%s_D23580_infected","%s_LT2_infected"),
  c("%s_D23580_infected","%s_D23580_exposed"),
  c("%s_LT2_infected","%s_LT2_exposed"),
  c("%s_D23580_exposed","%s_LT2_exposed")
)
for (time in levels(sce.norm$Time)){
  for (cindex in 1:4){
    sg.ref <- sprintf(contrasts[[cindex]][2], time)
    sg.tar <- sprintf(contrasts[[cindex]][1], time)
    contrast <- sprintf("%s-%s", sg.tar, sg.ref);# message(contrast)
    wt.list[[contrast]] <- genewise.wilcox.test(sg.ref, sg.tar)
  }
}
```

```{r, include=FALSE}
for (cName in names(wt.list)){
  cTable <- wt.list[[cName]]
  cTable <- cTable[order(cTable[,"p.value"]),]
  cTable <- cbind(
    GENENAME = mapIds(
      EnsDb.Hsapiens.v79, rownames(cTable), "GENENAME", "GENEID"
    ),
    cTable
  )
  write.csv(
    cTable, sprintf("wilcox.test/%s.csv", cName)
  )
}
```

## DE count

```{r}
DT::datatable(
  data.frame(
  DE = sapply(wt.list, function(x){sum(x[,"BH"] < 0.05, na.rm = TRUE)}),
  row.names = names(wt.list)
)
)
```

## `r names(wt.list)[1]`

```{r}
ggTop9(names(wt.list)[1])
```

## `r names(wt.list)[2]`

```{r}
ggTop9(names(wt.list)[2])
```

## `r names(wt.list)[3]`

```{r}
ggTop9(names(wt.list)[3])
```

## `r names(wt.list)[4]`

```{r}
ggTop9(names(wt.list)[4])
```

## `r names(wt.list)[5]`

```{r}
ggTop9(names(wt.list)[5])
```

## `r names(wt.list)[6]`

```{r}
ggTop9(names(wt.list)[6])
```

## `r names(wt.list)[7]`

```{r}
ggTop9(names(wt.list)[7])
```

## `r names(wt.list)[8]`

```{r}
ggTop9(names(wt.list)[8])
```

## `r names(wt.list)[9]`

```{r}
ggTop9(names(wt.list)[9])
```

## `r names(wt.list)[10]`

```{r}
ggTop9(names(wt.list)[10])
```

## `r names(wt.list)[11]`

```{r}
ggTop9(names(wt.list)[11])
```

## `r names(wt.list)[12]`

```{r}
ggTop9(names(wt.list)[12])
```


