---
title: Preprocessing of alignment-based quantitation data
author:
  - name: Anna Aulicino
    email: anna.aulicino@ndm.ox.ac.uk
    affiliation: MRC,Human Immunology Unit, Weatherall Institute of Molecular
        Medicine, NIHR Biomedical research centre, University of Oxford,
        John Radcliffe Hospital, Oxford, UK.
  - name: Kévin Rue-Albrecht
    email: kevinrue67@gmail.com
    affiliation: Department of Medicine, Imperial College London,
        Hammersmith Campus, Du Cane Road, London, W12 0NN, UK
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
abstract: >
  Import and preprocess quantitation data obtained using an alignment-based
  pipeline made of _Trim Galore!_, _HISAT2_, and _featureCounts_.
vignette: >
  %\VignetteIndexEntry{Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document2:
    toc_newpage: true
---
  
# Prior work

An alignment-based pipeline was implemented using:

* [Trim Galore!](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore)
* [HISAT2](https://ccb.jhu.edu/software/hisat2/index.shtml)
* [featureCounts](http://bioinf.wehi.edu.au/featureCounts)

### Trim Galore!

On this data set, trimming was found to improve alignment rate by ~5%
using _HISAT2_.
All samples were adapter- and quality-trimmed as follows:

```
trim_galore \
    --output_dir $trimgaloreDir \
    --paired \
    $fastqFolder/${cell}_1.fastq.gz $fastqFolder/${cell}_2.fastq.gz
```

\fixme{
Report some trimming statistics here.
}

### HISAT2

#### Composite genome

A composite genome (chromosomes and ERCC spike-in sequences) was assembled
as follows.

First the sequence and annotations for the ERCC92 spike-ins were obtained:

```
wget https://tools.thermofisher.com/content/sfs/manuals/ERCC92.zip
unzip ERCC92.zip
```

A composite genome was assembled by concatenating the ERCC92 sequences to
the GRCh38 build:

```
wget ftp://ftp.ensembl.org/pub/release-86/fasta/homo_sapiens/dna/Homo_sapiens.GRCh38.dna.primary_assembly.fa.gz
gunzip Homo_sapiens.GRCh38.dna.primary_assembly.fa.gz
cat Homo_sapiens.GRCh38.dna.primary_assembly.fa ERCC92.fa > GRCh38.dna.primary_assembly_with_ERCC92.fa
```
A composite genome annotation file was assembled by concatenating the ERCC92 annotations to the the GRCh38 GTF file:

```
wget ftp://ftp.ensembl.org/pub/release-86/gtf/homo_sapiens/Homo_sapiens.GRCh38.86.chr.gtf.gz
gunzip Homo_sapiens.GRCh38.86.chr.gtf.gz
cat Homo_sapiens.GRCh38.86.chr.gtf ERCC92.gtf > GRCh38.86.chr_with_ERCC.gtf
```

#### Genome index

The composite genome was indexed as follows:

```
hisat2-build $fastaFiles $hisatIdxBase
```

#### Alignment

Trimmed read pairs were aligned to the composite genome as follows:

```
hisat2 \
    -x ${hisatIdxBase} \
    -1 ${fastqFolder}/${cell}_1_val_1.fq.gz \
    -2 ${fastqFolder}/${cell}_2_val_2.fq.gz |
    samtools view -hb - > $hisatDir/${cell}.bam
```

\fixme{
Report some alignment statistics here.
}

### featureCounts

Aligned read pairs were assigned to annotated features (including ERCC spike-in
features) as follows:

```
featureCounts \
    -s 1 \
    --primary \
    --ignoreDup \
    -p \
    -a $gtfFile \
    -o $countFolder/$cell \
    $hisatDir/${cell}.bam
```

\fixme{
Report some assignment statistics here.
}

# Settings

Packages required for this vignette:

```{r packages, message=FALSE}
library(scater)
library(edgeR)
library(dplyr)
library(ggplot2)
library(EnsDb.Hsapiens.v79)
```

Let us set the working directory to execute the following code chunks:

```{r workdir}
workdir <- ifelse(interactive(), ".", "..")
```

Let us set a few paths relevant to the analysis:

```{r paths}
folder.expData <- file.path(workdir, "expData")
folder.rds <- file.path(workdir, "rds")
folder.count <- file.path(
  workdir, "trimgalore", "HISAT2", "featureCounts", "trimmed"
)
```

# Import preprocessed data

## Alignment pipeline

Then, let us import experimental information on samples:

```{r expData}
expdata <- read.csv(file.path(folder.expData, "samples.csv"))
rownames(expdata) <- expdata$sample
expdata$Lane <- as.factor(expdata$Lane)
expdata$sample <- as.character(expdata$sample)
summary(expdata)
```

Using the above information, let us add additional columns to define the file
paths and the sample groups:

```{r fixExpData}
expdata$files <- file.path(sprintf(
  "WTCHG_%s_%s", expdata$Lane, rownames(expdata)
))
expdata$group <- as.factor(paste(
  expdata$Time, expdata$Infection, expdata$Status,
  sep = "_"
))
levels(expdata$group)
sort(table(expdata$group))
```

We may now load the data from the _featureCounts_ count files:

```{r readDGE}
dge_all <- readDGE(expdata, folder.count, c(1,7), skip = 1)
```

And then save the serialised object to file for faster loading in subsequent
vignettes:

```{r saveRDS}
saveRDS(dge_all, file.path(folder.rds, "dge_all.rds"))
```

# Overview of alignment pipeline

## Blanks

Let us verify that ERCC spike-in features are the most abundant features
in the two `Blank` samples.

For this purpose, let us first extract the two `Blank` samples from the
`DGEList`:

```{r subsetBlanks}
dge.blanks <- dge_all[,dge_all$samples$Status == "Blank"]
```

Let us then reorder the subsetted `DGEList` by decreasing average
count:

```{r reorderBlanks}
avg.count.blanks <- rowMeans(dge.blanks$counts)
dge.blanks <- dge.blanks[order(avg.count.blanks, decreasing = TRUE),]
```

We may then examine the features with highest counts (not normalised for
feature length) in `Blank` samples:

```{r topFeaturesBlanks, results='asis'}
mappedSymbols <- mapIds(
  EnsDb.Hsapiens.v79, rownames(dge.blanks$counts), "GENENAME", "GENEID"
)
pander::pandoc.table(
  head(data.frame(
    Symbol = mappedSymbols,
    "Average count" = sort(avg.count.blanks, decreasing = TRUE)
  ), 25)
)
rm(dge.blanks, mappedSymbols)
```

# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
